\newpage
\setcounter{page}{1}

# Supplementary information for *Persson (2024). The acoustic characteristics of Central Swedish* {-}
Both the main text and these supplementary information (SI) are derived from the same R markdown document available via OSF at  [https://osf.io/7uvj4/](https://osf.io/7uvj4/).

\setcounter{section}{1}
\setcounter{footnote}{0}
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{equation}{0}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\thefootnote}{S\arabic{footnote}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\theequation}{S\arabic{equation}}

## Required software {#sec:SI-software}

The document was compiled using \texttt{knitr} [@xie2023] in RStudio with R:

```{r}
version
```

The following R packages were used to create this document: `r cite_r("latex-stuff/packages.bib")`

```{r}
#knitr::write_bib(file = 'latex-stuff/packages.bib')
```

## Neutralization of short /e/ and /`r linguisticsdown::cond_cmpl("ɛ")`/ {#sec:SI-hedd-hädd}
Figure \@ref(fig:SI-hedd) demonstrates that the short allophones of /`r linguisticsdown::cond_cmpl("e")`/ and /`r linguisticsdown::cond_cmpl("ɛ")`/ neutralize as [`r linguisticsdown::cond_cmpl("ɛ")`] in Central Swedish.

(ref:SI-hedd) The *hedd* and *hädd* words in the SwehVd vowel data in F1-F2 space. Points show recordings of the *hedd* and *hädd* words ([`r linguisticsdown::cond_cmpl("ɛ")`]) by 48 native talkers in the database, averaged across the five measurement points within each vowel segment. Word labels indicate word means across talkers. Since *hädd* and *hedd* resulted in the same allophone, all *hädd* words are excluded from the subset used in this study. This facilitates comparison of, for example, densities across vowels (see diagonal of Figure \@ref(fig:swe-vowels-all-cues)).

```{r SI-hedd, fig.width=base.width*2, fig.height=base.height*2, out.width='50%', fig.align='center', fig.pos="ht", fig.cap="(ref:SI-hedd)"}
d.SwehVd %>%
  filter(Transcribed_vowel == "targeted", Word %in% c("hedd", "hädd")) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Talker, Word, Token) %>%
  summarise(across(c(F1, F2), ~ geometric.mean(.x))) %>%
  ggplot(
    aes(
      x = F2,
      y = F1,
      color = Word)) +
  geom_point(alpha = .4) +
  stat_ellipse() +
  # Add between-talker mean
  geom_label(
    data =
      ~ .x %>%
      group_by(Word) %>%
      summarise(across(c(F1, F2), ~ geometric.mean(.x))),
    mapping = aes(
      colour = Word,
      label = Word),
    alpha = .8,
    size = 4,
    label.size = NA) +
  scale_x_reverse("F2 (in Hz)", position = "top", breaks = scales::breaks_pretty(6)) +
  scale_y_reverse("F1 (in Hz)", position = "right", breaks = scales::breaks_pretty(6)) +
  guides(color = "none") +
  theme(legend.position = "top")
```

## The effect of different measurement points {#sec:SI-measurement-points}
Figure \@ref(fig:SI-time-points) visualizes the SwehVd vowel space under four different assumptions about measurement points. The figure suggests that the use of different measurement points primarily affects vowels for which formant dynamics are particularly important, e.g., [`r linguisticsdown::cond_cmpl("eː")`] and [`r linguisticsdown::cond_cmpl("oː")`]. 

(ref:SI-time-points) The SwehVd vowel data in F1-F2 space, faceted by the time points used for measuring formants: averaging across the first three time-points (at 20, 35, and 50% into the vowel; **panel A**), averaging across the mid three time-points (at 35, 50, and 65% into the vowel; **panel B**), taking the mid time-point (at 50%; **panel C**), and averaging across the final three mid-points (at 50, 65, and 80% into the vowel; **panel D**). Points show recordings of each of the 21 Central Swedish vowels, vowel labels indicate vowel means across talkers. Long vowels are boldfaced. Vowels that mismatched intended label are excluded (`r signif((d.SwehVd %>% filter(Transcribed_vowel != "targeted") %>% nrow() / d.SwehVd %>% nrow()) * 100)`% of all recordings).

```{r SI-time-points, fig.width=base.width*4.5, fig.height=base.height*4.5+.5, out.width = '100%', fig.pos="ht", fig.align='center', fig.cap="(ref:SI-time-points)"}
p.20_50 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(20,35,50)) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x)))) +
  guides(shape = "none")

p.35_65 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(35,50,65)) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x)))) +
  guides(shape = "none")

p.50 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint == 50)) +
  guides(shape = "none")

p.50_80 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(50,65,80)) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x)))) +
  guides(shape = "none")

(p.20_50 + p.35_65) / (p.50 + p.50_80) +
  plot_annotation(tag_levels = list(c("A", "B", "C", "D"))) +
  plot_layout(
    guides = "collect") & 
  theme(legend.position = "right",
        legend.justification = "right",
        legend.margin = margin(0, 0, 0, 0, unit = "pt"),
        plot.tag = element_text(face = "bold"))
```

## Mean cue values across talkers {#sec:SI-mean-cues}
The overall mean cue values for the long and short vowels are presented in Tables \@ref(tab:SI-cue-values-long) and \@ref(tab:SI-cue-values-short), respectively.

\begin{landscape}
\begin{table}
\centering
\caption{\label{tab:SI-cue-values-long}The spectral and temporal properties of the long vowels (averaged across talkers, by gender\; in Hertz)}
\centering
\fontsize{6}{8}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}}
\toprule
Cue & Gender & [`r linguisticsdown::cond_cmpl("iː")`]  & [`r linguisticsdown::cond_cmpl("yː")`] & [`r linguisticsdown::cond_cmpl("ʉː")`] & [`r linguisticsdown::cond_cmpl("eː")`] & [`r linguisticsdown::cond_cmpl("ɛː")`] & [`r linguisticsdown::cond_cmpl("æː")`] & [`r linguisticsdown::cond_cmpl("øː")`] & [`r linguisticsdown::cond_cmpl("œː")`]& [`r linguisticsdown::cond_cmpl("ɑː")`] & [`r linguisticsdown::cond_cmpl("oː")`] & [`r linguisticsdown::cond_cmpl("uː")`]\\
\midrule
Duration & F & 0.22 (SD=0.05) & 0.24 (SD=0.04) & 0.23 (SD=0.05) & 0.25 (SD=0.05) & 0.25 (SD=0.04) & 0.25 (SD=0.05) & 0.25 (SD=0.04) & 0.25 (SD=0.05) & 0.23 (SD=0.05) & 0.25 (SD=0.05) & 0.23 (SD=0.05)\\
 & M & 0.2 (SD=0.05) & 0.21 (SD=0.05) & 0.22 (SD=0.06) & 0.23 (SD=0.05) & 0.23 (SD=0.05) & 0.25 (SD=0.06) & 0.23 (SD=0.05) & 0.23 (SD=0.06) & 0.23 (SD=0.06) & 0.24 (SD=0.05) & 0.21 (SD=0.05)\\
F0 & F & 202.26 (SD=26.14) & 197.51 (SD=29.77) & 197.48 (SD=25.48) & 195.77 (SD=24.6) & 190.62 (SD=23.72) & 190.29 (SD=24.88) & 190.94 (SD=24.49) & 191.83 (SD=24.97) & 194.12 (SD=25.23) & 196.05 (SD=26.78) & 200.21 (SD=25.91)\\
 & M & 118.12 (SD=21.56) & 113.46 (SD=20.17) & 115.04 (SD=21.02) & 114.87 (SD=19.3) & 108.51 (SD=18.94) & 110.06 (SD=20.16) & 109.15 (SD=21.28) & 113.48 (SD=25.12) & 111.96 (SD=20.08) & 118.94 (SD=21.93) & 115.61 (SD=18.52)\\
F1 & F & 379.73 (SD=40.28) & 365.11 (SD=33.21) & 394.43 (SD=29.37) & 428.32 (SD=43.19) & 789.24 (SD=98.65) & 840.84 (SD=122.92) & 641 (SD=76.26) & 615.78 (SD=74.78) & 652.24 (SD=84.95) & 443.45 (SD=46.98) & 398.69 (SD=37.25)\\
\addlinespace
 & M & 310.53 (SD=36.19) & 298.85 (SD=35.72) & 332.16 (SD=39.26) & 393.81 (SD=41.04) & 683.73 (SD=65.1) & 727.05 (SD=62.18) & 543.16 (SD=44.26) & 548.09 (SD=45.31) & 578.64 (SD=57.42) & 423.07 (SD=28.97) & 350.19 (SD=33.43)\\
F2 & F & 2070.82 (SD=200.88) & 1997.58 (SD=175.05) & 1776.61 (SD=114.1) & 2406.57 (SD=310.74) & 1801.31 (SD=128.8) & 1648.55 (SD=88.26) & 1529.59 (SD=94.52) & 1304.61 (SD=114.89) & 1055.28 (SD=107.03) & 770.11 (SD=93.38) & 786.72 (SD=105.56)\\
 & M & 1937.17 (SD=184.55) & 1864.78 (SD=169.32) & 1602.27 (SD=81.54) & 2106.4 (SD=150.09) & 1573.55 (SD=111.05) & 1469.25 (SD=109.55) & 1335.38 (SD=118.33) & 1136.89 (SD=88.89) & 883.2 (SD=81.81) & 712.57 (SD=78.05) & 768.6 (SD=73.48)\\
F3 & F & 3148.44 (SD=360.71) & 3273.13 (SD=139.43) & 2658.86 (SD=173.38) & 3063.51 (SD=168.32) & 2764.53 (SD=243.69) & 2726.22 (SD=273.58) & 2681.93 (SD=144.12) & 2746.77 (SD=138.83) & 2860.06 (SD=197.94) & 2921.52 (SD=187.61) & 2844.49 (SD=185.93)\\
 & M & 2937.13 (SD=115.16) & 2821.75 (SD=107.5) & 2368.81 (SD=119.22) & 2693.68 (SD=138.95) & 2492.35 (SD=176.49) & 2498.89 (SD=163.73) & 2375.36 (SD=204.2) & 2401.73 (SD=160.65) & 2671.62 (SD=219.92) & 2653.25 (SD=170.04) & 2548.81 (SD=148.75)\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

```{r SI-cue-values-long}
# Keep code for reproducibility
# d.cue.values.forTable <- d.cue.values %>%
#   group_by(Quantity, Gender, category) %>%
#   mutate(
#     F0 = paste0(F0.mean, " (SD=", F0.sd, ")"),
#     F1 = paste0(F1.mean, " (SD=", F1.sd, ")"),
#     F2 = paste0(F2.mean, " (SD=", F2.sd, ")"),
#     F3 = paste0(F3.mean, " (SD=", F3.sd, ")"),
#     Duration = paste0(Duration.mean, " (SD=", Duration.sd, ")")) %>%
#   ungroup() %>%
#   select(-c(contains(c(".mean", ".sd")), Quantity)) %>%
#   arrange(factor(category, levels = levels.vowel.IPA.swe))
# 
# d.cue.values.forTable %<>%
#   pivot_longer(
#     cols = c(starts_with("F"), "Duration"),
#     names_to = "Cue",
#     values_to = "Cue.Value") %>%
#   pivot_wider(
#     names_from = "category",
#     values_from = "Cue.Value") %>%
#   relocate("Gender", .after = Cue) %>%
#   arrange(Cue) %>%
#   mutate(
#     Gender = ifelse(Gender == "female", "F", "M"),
#     Cue = ifelse(row_number() %% 2 == 0, "", Cue))
# 
# kable(d.cue.values.forTable %>%
#         select(Cue, Gender, contains(levels.vowel.IPA.swe.long)),
#       format = "latex",
#       booktabs = TRUE,
#       caption = "The spectral and temporal properties of the long vowels (averaged across talkers, by gender)") %>%
#   kable_styling(font_size = 6) %>%
#   column_spec(1:2, width = ".7cm") %>%
#   column_spec(3:13, width = "1.4cm")
```

\begin{landscape}
\begin{table}
\centering
\caption{\label{tab:SI-cue-values-short}The spectral and temporal properties of the short vowels (averaged across talkers, by gender\; in Hertz)}
\centering
\fontsize{6}{8}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}}
\toprule
Cue & Gender & [`r linguisticsdown::cond_cmpl("ɪ")`] & [`r linguisticsdown::cond_cmpl("ʏ")`] & [`r linguisticsdown::cond_cmpl("ɵ")`] & [`r linguisticsdown::cond_cmpl("ɛ")`] & [`r linguisticsdown::cond_cmpl("æ")`] & [`r linguisticsdown::cond_cmpl("ø")`] & [`r linguisticsdown::cond_cmpl("œ")`] & [`r linguisticsdown::cond_cmpl("a")`] & [`r linguisticsdown::cond_cmpl("ɔ")`] & [`r linguisticsdown::cond_cmpl("ʊ")`]\\
\midrule
Duration & F & 0.12 (SD=0.02) & 0.12 (SD=0.02) & 0.13 (SD=0.02) & 0.12 (SD=0.02) & 0.16 (SD=0.03) & 0.12 (SD=0.03) & 0.17 (SD=0.03) & 0.13 (SD=0.03) & 0.13 (SD=0.02) & 0.13 (SD=0.03)\\
 & M & 0.11 (SD=0.02) & 0.11 (SD=0.03) & 0.13 (SD=0.02) & 0.12 (SD=0.02) & 0.13 (SD=0.03) & 0.12 (SD=0.03) & 0.14 (SD=0.03) & 0.13 (SD=0.02) & 0.12 (SD=0.02) & 0.12 (SD=0.03)\\
F0 & F & 218.55 (SD=29.41) & 221.58 (SD=27.82) & 219.72 (SD=28.7) & 213.5 (SD=26.46) & 203.8 (SD=28.32) & 214.19 (SD=26.7) & 206.31 (SD=26.72) & 208.47 (SD=26.74) & 214.02 (SD=28.06) & 224.89 (SD=28.4)\\
 & M & 128.98 (SD=19.73) & 130.12 (SD=18.44) & 128.58 (SD=19.22) & 119.25 (SD=21.69) & 119.05 (SD=20.05) & 125.29 (SD=17.45) & 123.63 (SD=19.27) & 117.74 (SD=19.19) & 123.68 (SD=19.17) & 132.89 (SD=18.48)\\
F1 & F & 385.31 (SD=38.08) & 387.64 (SD=36.94) & 424.93 (SD=36.49) & 635.03 (SD=69.25) & 799.8 (SD=100.22) & 573.8 (SD=64.24) & 532.76 (SD=55.58) & 788.18 (SD=99.76) & 501.44 (SD=54.57) & 418.49 (SD=37.01)\\
\addlinespace
 & M & 330.61 (SD=30.04) & 334.94 (SD=31.31) & 379.7 (SD=35.54) & 547.46 (SD=41.76) & 703.17 (SD=51.91) & 494.75 (SD=38.76) & 499.02 (SD=37.88) & 687.54 (SD=55.08) & 459.94 (SD=40.86) & 371.24 (SD=28.89)\\
F2 & F & 2511.76 (SD=288.04) & 2376.37 (SD=157.36) & 1329.55 (SD=115.67) & 1946.83 (SD=181.2) & 1755.92 (SD=143.49) & 1666.86 (SD=114.75) & 1405.18 (SD=156.48) & 1372.95 (SD=99.7) & 937.21 (SD=97.85) & 854.39 (SD=109.9)\\
 & M & 2242.23 (SD=124.13) & 2012.84 (SD=105.02) & 1166.35 (SD=87.07) & 1720.93 (SD=112.77) & 1517.06 (SD=113.71) & 1417.46 (SD=136.96) & 1205.86 (SD=90.4) & 1134.2 (SD=86.67) & 793.65 (SD=75.38) & 768.48 (SD=76.95)\\
F3 & F & 3081.21 (SD=171.41) & 2804.95 (SD=166.88) & 2780.03 (SD=197.87) & 2735 (SD=247.64) & 2791.93 (SD=262.46) & 2704.28 (SD=162.68) & 2781.63 (SD=175.79) & 2844.63 (SD=268.99) & 3035.1 (SD=189.32) & 2940.08 (SD=182.4)\\
 & M & 2757.15 (SD=137.69) & 2462.19 (SD=109.26) & 2442.7 (SD=138.25) & 2501.48 (SD=153.16) & 2532.25 (SD=157.51) & 2426.6 (SD=140.27) & 2440.67 (SD=159.42) & 2662.24 (SD=207.5) & 2724.42 (SD=184.96) & 2605.44 (SD=146.73)\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

```{r SI-cue-values-short}
# Keep code for reproducibility
#For each cue, the highest and lowest mean and SD should be boldfaced.
# kable(d.cue.values.forTable %>%
#         select(Cue, Gender, contains(levels.vowel.IPA.swe.short)),
#       format = "latex",
#       booktabs = TRUE,
#       caption = "The spectral and temporal properties of the short vowels (averaged across the female and male talkers)") %>%
#   kable_styling(font_size = 6) %>%
#   column_spec(1:2, width = ".7cm") %>%
#   column_spec(3:12, width = "1.4cm")
```
```{r}
# itsadug::plot_smooth(m.F3.short, view = "timepoint", plot_all = "category")
# itsadug::plot_smooth(m.F3.long, view = "timepoint", plot_all = "category")
#itsadug::plot_diff(m.F3.long, view = "timepoint", comp = list(category = c("[yː]", "[iː]")))
```

## Cue matrix plot in unnormalized Hertz {#sec:SI-raw}
Figure \@ref(fig:SI-swe-vowels-all-cues-hz) displays the pairwise combinations of the five spectral and temporal cues---F0, F1, F2, F3 and duration---in raw Hertz. For F0, there are two overlapping but separated peaks, reflecting the overall difference in vocal tract size for the two groups of female and male talkers in the database.

(ref:SI-swe-vowels-all-cues-hz) The SwehVd vowels in unnormalized Hertz space. Points show recordings of each of the 21 Central Swedish vowels by the talkers in the database in all five cue spaces, averaged across the three midpoints. Vowel labels indicate vowel means across talkers. Ellipses show bivariate Gaussian 95% confidence interval of vowel means. Note that, unlike in Figure \@ref(fig:swe-vowels), axis directions are not reversed. Vowels that mismatched the intended label are excluded (`r signif((d.SwehVd %>% filter(Transcribed_vowel != "targeted") %>% nrow() / d.SwehVd %>% nrow()) * 100)`% of all recordings).

```{r SI-swe-vowels-all-cues-hz, fig.width=base.width*4+.5, fig.height=base.height*4+.5, out.width = '100%', fig.align='center', fig.pos="ht", fig.cap="(ref:SI-swe-vowels-all-cues-hz)"}
p.matrix.cues %+%
  (d.SwehVd.forStudy.long.static %>%
     filter(Normalization.Type == "no normalization (Hz)", F0 < 320))
```

## Talker differences in the [i:] - [y:] contrast
Figure \@ref(fig:SI-swe-merge) visualizes two example talkers that differ in their phonetic realization of [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] when F3 is considered.

(ref:SI-swe-merge) The effect of including F3 for the [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] distinction, illustrated in a reduced vowel space. **Panel A** shows two example talkers with overlapping [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] in F1-F2 space. **Panel B** shows how considering F3 affects the space: for talker SW_028, there is very little effect of F3, thus maintained [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] overlap, while for talker SW_027, separability increases.

```{r SI-swe-merge, fig.width=base.width*3, fig.height=base.height*3, out.width='80%', fig.align='center', fig.cap="(ref:SI-swe-merge)", fig.pos="ht", message=FALSE}
p.merge <- d.SwehVd.forStudy.long.static %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", Quantity == "long", Talker %in% c("SW_028", "SW_027"), category %in% c("[iː]", "[yː]", "[æː]", "[ɑː]", "[oː]", "[uː]")) %>%
  ggplot(
    aes(
      x = F2, 
      y = F1)) +
  geom_point(
    aes(
      colour = category,
      shape = Quantity), alpha = 0.5, size = 1.5) + 
  # Add within-talker mean
  geom_label(
    data =
      . %>%
      group_by(category, Quantity, Normalization.Type, Talker) %>%
      summarise(across(c(F1, F2), ~ mean(.x))),
    mapping = aes(
      colour = category,
      label = category,
      fontface = ifelse(category %in% c("[iː]", "[yː]"), 2, 1)),
    size = 4,
    alpha = .4,
    label.size = NA,
    label.padding = unit(.06, "cm")) +
  scale_colour_manual(name = "category", values = colors.vowel.swe) +
  scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 4)) +
  scale_y_reverse("F1", position = "right", scales::pretty_breaks(n = 4)) +
  guides(alpha = "none", color = "none", shape = "none", size = "none") +
  facet_wrap(~ Talker, ncol=2) 

p.merge.F3 <-  d.SwehVd.forStudy.long.static %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", Quantity == "long", Talker %in% c("SW_028", "SW_027"), category %in% c("[iː]", "[yː]", "[æː]", "[ɑː]", "[oː]", "[uː]")) %>%
  ggplot(
    aes(
      x = F2, 
      y = F3)) +
  geom_point(
    aes(
      colour = category,
      shape = Quantity), alpha = 0.5, size = 1.5) + 
  # Add within-talker mean
  geom_label(
    data =
      . %>%
      group_by(category, Quantity, Normalization.Type, Talker) %>%
      summarise(across(c(F3, F2), ~ mean(.x))),
    mapping = aes(
      colour = category,
      label = category,
      fontface = ifelse(category %in% c("[iː]", "[yː]"), 2, 1)),
    size = 4,
    alpha = .4,
    label.size = NA, 
    label.padding = unit(.06, "cm")) +
  scale_colour_manual(name = "category", values = colors.vowel.swe) +
  scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 4)) +
  scale_y_reverse("F3", position = "right", scales::pretty_breaks(n = 4)) +
  guides(alpha = "none", color = "none", shape = "none", size = "none") +
  facet_wrap(~ Talker, ncol=2)
  
p.merge / p.merge.F3 + theme(axis.title.x = element_blank()) +
  plot_annotation(tag_levels = 'A') &
  theme(legend.margin = margin(0, 0, 0, 0, unit = "pt"),
        plot.tag = element_text(face = "bold"))
#ggsave(file = "../../figures/p.png", width = 9, height = 5, dpi=1200)
```

## Additional information concerning static analyses
### Summary tables of LMMs predicting category separability from cue space {#sec:SI-LMMs-study1}

\begin{table}
\caption{LMMs predicting category separability from cue space}
\centering
\begin{tabular}[t]{lccc}
\addlinespace[0.5em]
\toprule
\multicolumn{4}{l}{\textit{unrounded vs. outrounded}}\\
\midrule
& `r linguisticsdown::cond_cmpl("[iː]")` vs. `r linguisticsdown::cond_cmpl("[yː]")` & `r linguisticsdown::cond_cmpl("[eː]")` vs. `r linguisticsdown::cond_cmpl("[yː]")` & `r linguisticsdown::cond_cmpl("[ɪ]")` vs. `r linguisticsdown::cond_cmpl("[ʏ]")`\\
\midrule \hspace{1em}(Intercept) & 0.091 & 0.319 & 0.145\\
\hspace{1em} & se = 0.008 & se = 0.013 & se = 0.0097\\
\hspace{1em} & t-value = 10.765 & t-value = 25.329 & t-value = 15.0\\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = \vphantom{3} <0.001\\
\hspace{1em} cue space F1F2F3 & 0.032 & 0.012 & 0.045\\
\hspace{1em} & se = 0.01 & se = 0.002 & se = 0.004\\
\hspace{1em} & t-value = 3.191 & t-value = 5.916 & t-value = 11.57\\
\hspace{1em} & p.value = <0.003 & p.value = <0.001 & p.value = \vphantom{2} <0.001\\
\hspace{1em}SD (Intercept Talker) & 0.028 & 0.08 & 0.059\\
\hspace{1em}SD (Observations) & 0.046 & 0.009 & 0.018\\
\addlinespace[0.5em]
\toprule
\multicolumn{4}{l}{\textit{outrounded vs. inrounded}}\\
\midrule
& `r linguisticsdown::cond_cmpl("[yː]")` vs. `r linguisticsdown::cond_cmpl("[ʉː]")` & `r linguisticsdown::cond_cmpl("[oː]")` vs. `r linguisticsdown::cond_cmpl("[uː]")` & `r linguisticsdown::cond_cmpl("[ɔ]")` vs. `r linguisticsdown::cond_cmpl("[ʊ]")`\\
\midrule \hspace{1em}(Intercept) & 0.191 & 0.1999 & 0.244\\
\hspace{1em} & se = 0.013 & se = 0.012 & se = 0.012\\
\hspace{1em} & t-value = 15.2 & t-value = 17.2 & t-value = 19.98\\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = \vphantom{1} <0.001\\
\hspace{1em} cue space F1F2F3 & 0.099 & 0.014 & 0.01\\
\hspace{1em} & se = 0.008 & se = 0.0016 & se = 0.0017\\
\hspace{1em} & t-value = 13.02 & t-value = 8.95 & t-value = 6.29\\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001\\
\hspace{1em}SD (Intercept Talker) & 0.073 & 0.074 & 0.077\\
\hspace{1em}SD (Observations) & 0.034 & 0.007 & 0.0078\\
\bottomrule
\end{tabular}
\end{table}

```{r}
# panels <- list(
#  "long vs. short" = list("A" = m.sep.I, "B" = m.sep.Y, "C" = m.sep.U, "D" = m.sep.E, "E" = m.sep.Ä, "F" = m.sep.ÄR),
#  "long vs. short" = list("G" = m.sep.Ö, "H" = m.sep.ÖR, "I" = m.sep.A, "J" = m.sep.Å, "K" = m.sep.O))
# modelsummary::modelsummary(panels, output = "latex", title = "LMM predicting category separability from cue space for all long-short contrasts", estimate = "{estimate}", statistic = c("se = {std.error}", "t-value = {statistic}", "p.value = {p.value}"), shape = "rbind", gof_map = NA)
rm(m.sep.IIYY, m.sep.EEYY, m.sep.IY, m.sep.YYUU, m.sep.ÅÅOO, m.sep.ÅO)
```

\begin{landscape}
\begin{table}
\tiny
\caption{LMMs predicting category separability from cue space for all long-short contrasts}
\centering
\begin{tabular}[t]{lcccccc}
\toprule
  & `r linguisticsdown::cond_cmpl("[iː]")` vs. `r linguisticsdown::cond_cmpl("[ɪ]")` & `r linguisticsdown::cond_cmpl("[yː]")` vs. `r linguisticsdown::cond_cmpl("[ʏ]")` & `r linguisticsdown::cond_cmpl("[ʉː]")` vs. `r linguisticsdown::cond_cmpl("[ɵ]")` & `r linguisticsdown::cond_cmpl("[eː]")` vs. `r linguisticsdown::cond_cmpl("[ɛ]")` & `r linguisticsdown::cond_cmpl("[ɛː]")` vs. `r linguisticsdown::cond_cmpl("[ɛ]")` & `r linguisticsdown::cond_cmpl("[æː]")` vs. `r linguisticsdown::cond_cmpl("[æ]")`\\
\midrule
\addlinespace[0.5em]
\multicolumn{6}{l}{\textit{long vs. short}}\\
\midrule \hspace{1em}(Intercept) & 0.23 & 0.201 & 0.348 & 0.468 & 0.243 & 0.12 \\
\hspace{1em} & se = 0.012 & se = 0.01 & se = 0.014 & se = 0.018 & se = 0.016 & se = 0.014 \\
\hspace{1em} & t-value = 18.584 & t-value = 20.899 & t-value = 24.974 & t-value = 26.61 & t-value = 15.375 & t-value = 8.853 \\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001\\
\hspace{1em} cue space F1F2F3 & 0.025 & 0.058 & 0.008 & 0.015 & 0.01 & 0.025  \\
\hspace{1em} & se = 0.005 & se = 0.004 & se = 0.002 & se = 0.003 & se = 0.004 & se = 0.007 \\
\hspace{1em} & t-value = 5.444 & t-value = 14.004 & t-value = 4.049 & t-value = 5.381 & t-value = 2.678 & t-value = 3.473 \\
\hspace{1em} & p.value = <0.0001 & p.value = <0.001 & p.value = 0.00012 & p.value = <0.001 & p.value = 0.009 & p.value = 0.0008 \\
\hspace{1em} cue space F1F2Duration & 0.028 & 0.037 & 0.019 & 0.019 & 0.038 & 0.049 \\
\hspace{1em} & se = 0.005 & se = 0.004 & se = 0.002 & se = 0.003 & se = 0.004 & se = 0.007 \vphantom{2} \\
\hspace{1em} & t-value = 6.27 & t-value = 8.956 & t-value = 9.882 & t-value = 7.074 & t-value = 9.806 & t-value = 6.814  \\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001  \\
\hspace{1em}SD (Intercept Talker) & 0.076 & 0.059 & 0.089 & 0.112 & 0.1 & 0.081  \\
\hspace{1em}SD (Observations) & 0.02 & 0.019 & 0.008 & 0.012 & 0.017 & 0.033  \\
\addlinespace[0.5em]
\toprule
  & `r linguisticsdown::cond_cmpl("[øː]")` vs. `r linguisticsdown::cond_cmpl("[ø]")` & `r linguisticsdown::cond_cmpl("[œː]")` vs. `r linguisticsdown::cond_cmpl("[œ]")` & `r linguisticsdown::cond_cmpl("[ɑː]")` vs. `r linguisticsdown::cond_cmpl("[a]")` & `r linguisticsdown::cond_cmpl("[oː]")` vs. `r linguisticsdown::cond_cmpl("[ɔ]")` & `r linguisticsdown::cond_cmpl("[uː]")` vs. `r linguisticsdown::cond_cmpl("[ʊ]")` & \\
\multicolumn{6}{l}{\textit{long vs. short }}\\
\midrule \hspace{1em}(Intercept) & 0.154 & 0.169 & 0.334 & 0.257 & 0.146 & \\
\hspace{1em} & se = 0.011 & se = 0.013 & se = 0.013 & se = 0.016 & se = 0.01 & \\
\hspace{1em} & t-value = 13.765 & t-value = 12.851 & t-value = 25.325 & t-value = 15.713 & t-value = 14.013 & \\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = \vphantom{2} <0.001 & \\
\hspace{1em} cue space F1F2F3 &  0.011 & 0.009 & 0.011 & 0.0085 & 0.014 & \\
\hspace{1em} & se = 0.004 & se = 0.003 & se = 0.003 & se = 0.003 & se = 0.003 & \\
\hspace{1em} & t-value = 2.924 & t-value = 2.636 & t-value = 3.756 & t-value = 3.162 & t-value = 4.183 & \\
\hspace{1em} & p.value = 0.00449 & p.value = 0.0101 & p.value = 0.0003 & p.value = 0.0022 & p.value = <0.001 & \\
\hspace{1em} cue space F1F2Duration &  0.0476 & 0.0313 & 0.011 & 0.033 & 0.037 & \\
\hspace{1em} & se = 0.004 & se = 0.003 & se = 0.003 & se = 0.003 & se = 0.003 & \\
\hspace{1em} & t-value = 12.292 & t-value = 9.074 & t-value = 6.197 & t-value = 12.279 & t-value = 11.177 & \\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & \\
\hspace{1em}SD (Intercept Talker) &  0.069 & 0.083 & 0.083 & 0.104 & 0.064 & \\
\hspace{1em}SD (Observations) &  0.0175 & 0.0156 & 0.013 & 0.012 & 0.015 & \\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

```{r}
rm(m.sep.I, m.sep.Y, m.sep.U, m.sep.E, m.sep.Ä, m.sep.ÄR, m.sep.Ö, m.sep.ÖR, m.sep.A, m.sep.Å, m.sep.O)
```

### Category separability indices for additional cue combinations {#sec:SI-separability}
Mirroring results from Figure \@ref(fig:cat-sep-long-short), Figure \@ref(fig:separability-SI) indicates that the F1-F2-duration combination generates the highest category separability for all vowel pairs besides the [`r linguisticsdown::cond_cmpl("yː")`] - [`r linguisticsdown::cond_cmpl("ʏ")`] contrast, for which F1-F2-F3 achieves higher separability. The second highest separability is overall achieved by F1-F2-F3, besides for the [`r linguisticsdown::cond_cmpl("ɛː")`] - [`r linguisticsdown::cond_cmpl("ɛ")`], [`r linguisticsdown::cond_cmpl("æː")`] - [`r linguisticsdown::cond_cmpl("æ")`], [`r linguisticsdown::cond_cmpl("øː")`] - [`r linguisticsdown::cond_cmpl("ø")`] and [`r linguisticsdown::cond_cmpl("œː")`] - [`r linguisticsdown::cond_cmpl("œ")`] contrasts where F1-duration achieves the second best separability, and for [`r linguisticsdown::cond_cmpl("uː")`] - [`r linguisticsdown::cond_cmpl("ʊ")`], for which F2-duration is second-best. This would seem to suggest duration dependencies for these contrasts. Interestingly, for several contrasts, the category separability combining one spectral cue with duration decreases relative to F1-F2, highlighting the reliance on spectral cues for these quantity contrasts (e.g., [`r linguisticsdown::cond_cmpl("ʉː")`] - [`r linguisticsdown::cond_cmpl("ɵ")`], [`r linguisticsdown::cond_cmpl("eː")`] - [`r linguisticsdown::cond_cmpl("ɛ")`], [`r linguisticsdown::cond_cmpl("oː")`] - [`r linguisticsdown::cond_cmpl("ɔ")`], [`r linguisticsdown::cond_cmpl("ɑː")`] - [`r linguisticsdown::cond_cmpl("a")`]).

For the [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("ɪ")`], [`r linguisticsdown::cond_cmpl("ʉː")`] - [`r linguisticsdown::cond_cmpl("ɵ")`], [`r linguisticsdown::cond_cmpl("oː")`] - [`r linguisticsdown::cond_cmpl("ɔ")`], [`r linguisticsdown::cond_cmpl("uː")`] - [`r linguisticsdown::cond_cmpl("ʊ")`] and [`r linguisticsdown::cond_cmpl("ɑː")`] - [`r linguisticsdown::cond_cmpl("a")`] vowel pairs, the F1-F3 cue combination decreases separability most relative to F1-F2 baseline, which would seem to suggest the informativity of F2 for distinguishing between these pairs. For the allophones to /`r linguisticsdown::cond_cmpl("ɛ")`/ and /`r linguisticsdown::cond_cmpl("ø")`/: [`r linguisticsdown::cond_cmpl("eː")`] - [`r linguisticsdown::cond_cmpl("ɛ")`], [`r linguisticsdown::cond_cmpl("ɛː")`] - [`r linguisticsdown::cond_cmpl("ɛ")`], [`r linguisticsdown::cond_cmpl("æː")`] - [`r linguisticsdown::cond_cmpl("æ")`], [`r linguisticsdown::cond_cmpl("øː")`] - [`r linguisticsdown::cond_cmpl("ø")`] and [`r linguisticsdown::cond_cmpl("œː")`] - [`r linguisticsdown::cond_cmpl("œ")`], the F2-F3 cue combination produces the lowest separability, presumably highlighting the importance of F1.

(ref:separability-SI) The increase in category separability relative to F1-F2-baseline for different cue combinations and the long-short vowel pairs. Pointranges indicate mean and 95% bootstrapped CIs of the category separability summarized across talkers for each cue combination.

```{r separability-SI, message=FALSE, fig.width=base.width*4, fig.height=base.height*3, fig.align='center', fig.cap="(ref:separability-SI)"}
d.SwehVd.cat.separability.for.longshort.SI %<>%
  mutate(
    cue_dimension = factor(cue_dimension, levels = c("F1F2", "F1F3", "F2F3", "F1F2F3", "F1Duration", "F2Duration", "F1F2Duration")))

# Calculate the relative increase of including F3 and duration for the quantity contrasts
d.SwehVd.cat.separability.comparison.longshort.SI <- d.SwehVd.cat.separability.long.short %>%
  group_by(comparison) %>%
  mutate(
    separabilityDifference_F1F3 = (F1F3_separability - F1F2_separability) / F1F2_separability,
    separabilityDifference_F2F3 = (F2F3_separability - F1F2_separability) / F1F2_separability,
    separabilityDifference_F3 = (F1F2F3_separability - F1F2_separability) / F1F2_separability,
    separabilityDifference_F1Duration = (F1Duration_separability - F1F2_separability) / F1F2_separability,
    separabilityDifference_F2Duration = (F2Duration_separability - F1F2_separability) / F1F2_separability,
    separabilityDifference_Duration = (F1F2Duration_separability - F1F2_separability) / F1F2_separability)

p.category.separability.diff.SI <- p.category.separability.diff.unrounded %+%
  (d.SwehVd.cat.separability.comparison.longshort.SI %>%
     select(comparison, separabilityDifference_F1F3, separabilityDifference_F2F3, separabilityDifference_F3, separabilityDifference_F1Duration, separabilityDifference_F2Duration, separabilityDifference_Duration) %>%
     pivot_longer(
       cols = starts_with("separabilityDifference"),
       names_to = "cue_difference") %>%
     mutate(
        cue_difference = factor(cue_difference, levels = c("separabilityDifference_F1F3", "separabilityDifference_F2F3", "separabilityDifference_F3", "separabilityDifference_F1Duration", "separabilityDifference_F2Duration", "separabilityDifference_Duration"), labels = c("F1-F3", "F2-F3", "F1-F2-F3", "F1-duration", "F2-duration", "F1-F2-duration"))) %>%
     rename(separability_diff = value) %>%
     group_by(comparison, cue_difference) %>%
     mutate(
       ci = list(enframe(Hmisc::smean.cl.boot(separability_diff)))) %>%
     unnest(ci) %>%
     spread(name, value) %>%
     group_by(cue_difference, comparison) %>%
     summarise(
       mean_CI_lower = mean(Lower),
       mean_CI_upper = mean(Upper),
       mean = mean(Mean)) %>%
     mutate(comparison = factor(comparison, levels = c("iː - ɪ",  "yː - ʏ", "ʉː - ɵ", "eː - ɛ", "oː - ɔ", "uː - ʊ", "ɛː - ɛ", "æː - æ", "øː - ø", "œː - œ", "ɑː - a")))) +
  scale_y_continuous("Increase in category separability relative to F1-F2", limits = c(-.7, 1)) +
  theme(legend.title = element_blank())
p.category.separability.diff.SI
```

```{r}
contrasts(d.SwehVd.cat.separability.for.longshort.SI$cue_dimension) <- contr.treatment(7)

# Rename contrast labels
d.SwehVd.cat.separability.for.longshort.SI$cue_dimension <- factor(d.SwehVd.cat.separability.for.longshort.SI$cue_dimension, labels = c("F1F2", "F1F3", "F2F3", "F1F2F3", "F1F2Duration", "F1Duration", "F2Duration"))

m.sep.I.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "iː - ɪ"))

m.sep.Y.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "yː - ʏ"))

m.sep.U.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "ʉː - ɵ"))

m.sep.E.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "eː - ɛ"))

m.sep.Ä.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "ɛː - ɛ"))

m.sep.ÄR.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "æː - æ"))

m.sep.Ö.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "øː - ø"))

m.sep.ÖR.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "œː - œ"))

m.sep.A.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "ɑː - a"))

m.sep.Å.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "oː - ɔ"))

m.sep.O.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "uː - ʊ"))
```

```{r}
rm(d.SwehVd.centroids)
```

### Category separability assuming talker-independent representations {#sec:SI-separability-talkerIndep}
The category separability index presented in the main paper assumed talker-specific category representations. Given that the data is normalized using Nearey's uniform scaling, one might instead argue talker-independent representations. Figures \@ref(fig:SI-cat-sep-F3) and \@ref(fig:SI-cat-sep-long-short) visualize the category separability index for both the rounding and the quantity evaluations while assuming talker-independent representations. As indicated in Figures \@ref(fig:SI-cat-sep-F3) and \@ref(fig:SI-cat-sep-long-short), the overall results replicate in both evaluations. Of note, Figure \@ref(fig:SI-cat-sep-F3) suggests that the relative increase in category separability for the [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] contrast when including F3 might possibly be driven by between-talker differences in the realizations of these vowels. There is further a smaller relative benefit of including duration comparing to F3 for the [`r linguisticsdown::cond_cmpl("æː")`] - [`r linguisticsdown::cond_cmpl("æ")`] and [`r linguisticsdown::cond_cmpl("øː")`] - [`r linguisticsdown::cond_cmpl("ø")`] contrasts when talker-independent representations are assumed.

(ref:SI-cat-sep-F3) The effect of including F3 in measures of category separability for the distinction between neighboring unrounded vs. outrounded vowels (**top row**), and outrounded vs. inrounded vowels (**bottom row**), under the assumption of talker-independent representations. **Left panels** plot the category separability for F1-F2 and F1-F2-F3 cue combinations. **Right panels** plot the proportional increase in category separability relative to F1-F2-baseline. Axis ranges are held constant across columns.

```{r}
get_category_separability_talkerIndep <- function(
    data,
    data_means,
    category_A, 
    category_B) 
  {
  data %>% 
    filter(category == category_A) %>%
    left_join(data_means %>% 
                filter(category == category_B),
              by = "Normalization.Type") %>%
    rename(category = category.x, mean_category = category.y) %>%
    mutate(
      distance_F1F2 = sqrt((F1 - F1_mean)^2 + (F2 - F2_mean)^2),
      distance_F1F2F3 = sqrt((F1 - F1_mean)^2 + (F2 - F2_mean)^2 + (F3 - F3_mean)^2),
      distance_F1F2duration = sqrt((F1 - F1_mean)^2 + (F2 - F2_mean)^2 + (Duration - Duration_mean)^2),
      # Calculate SEs for alternative way of assessing CIs
      SE_F1F2 = sd(distance_F1F2) / sqrt(length(distance_F1F2)),
      SE_F1F2F3 = sd(distance_F1F2F3) / sqrt(length(distance_F1F2F3)),
      SE_F1F2duration = sd(distance_F1F2duration) / sqrt(length(distance_F1F2duration)),
      category_separabilityF1F2 = sum(distance_F1F2) / length(distance_F1F2),
      category_separabilityF1F2F3 = sum(distance_F1F2F3) / length(distance_F1F2F3),
      category_separabilityF1F2duration = sum(distance_F1F2duration) / length(distance_F1F2duration))
  }

d.SwehVd.centroids.talkerIndep <- d.SwehVd.for.cat.separability %>%
      group_by(Normalization.Type, category) %>%
      summarise(across(c(starts_with("F"), "Duration"), ~ mean(.x), .names = '{col}_mean'))

d.SwehVd.cat.separability.talkerIndep <- rbind(
  # Get category separability of i: from y:
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[iː]", category_B = "[yː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    # label the same pairwise comparison as there are no differences between the categories for these contrasts
    mutate(comparison = "iː - yː"),
  # and then y: from i:
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[yː]", category_B = "[iː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
  mutate(comparison = "iː - yː"),
  d.SwehVd.for.cat.separability %>%
    get_category_separability_talkerIndep(category_A = "[yː]", category_B = "[ʉː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "yː - ʉː"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ʉː]", category_B = "[yː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "yː - ʉː"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[eː]", category_B = "[yː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "eː - yː"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[yː]", category_B = "[eː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "eː - yː"),
  d.SwehVd.for.cat.separability %>%
    get_category_separability_talkerIndep(category_A = "[iː]", category_B = "[eː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "iː - eː"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[eː]", category_B = "[iː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "iː - eː"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ɪ]", category_B = "[ʏ]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "ɪ - ʏ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ʏ]", category_B = "[ɪ]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "ɪ - ʏ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ɛː]", category_B = "[æː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "ɛː - æː"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[æː]", category_B = "[ɛː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "ɛː - æː"),
  d.SwehVd.for.cat.separability %>%
    get_category_separability_talkerIndep(category_A = "[øː]", category_B = "[œː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "øː - œː"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[œː]", category_B = "[øː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "øː - œː"),
  d.SwehVd.for.cat.separability %>%
    get_category_separability_talkerIndep(category_A = "[ø]", category_B = "[œ]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "ø - œ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[œ]", category_B = "[ø]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "ø - œ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[oː]", category_B = "[uː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "oː - uː"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[uː]", category_B = "[oː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "oː - uː"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ɔ]", category_B = "[ʊ]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "ɔ - ʊ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ʊ]", category_B = "[ɔ]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "ɔ - ʊ")) %>%
  #average the separability across the two distances
  group_by(comparison) %>%
  summarise(
      F1F2_SE = geometric.mean(SE_F1F2),
      F1F2F3_SE = geometric.mean(SE_F1F2F3),
      F1F2_separability = mean(category_separabilityF1F2, na.rm = T),
      F1F2F3_separability = mean(category_separabilityF1F2F3, na.rm = T))
```

```{r SI-cat-sep-F3, fig.align='center', fig.width=base.width*3.5, fig.height=base.height*4, out.width='90%', fig.cap="(ref:SI-cat-sep-F3)", message=FALSE}
colors.comparison <- c("#FC8D62", "#FB9A99", "#1B9E77", "#8DA0CB", "#A6CEE3", "#33A02C")
names(colors.comparison) <- c("iː - yː",  "eː - yː", "ɪ - ʏ", "yː - ʉː", "oː - uː", "ɔ - ʊ")

d.SwehVd.cat.separability.for.rounding.talkerIndep <- d.SwehVd.cat.separability.talkerIndep %>%
  filter(comparison %in% c("iː - yː",  "eː - yː", "ɪ - ʏ", "yː - ʉː", "oː - uː", "ɔ - ʊ")) %>%
  mutate(
    comparison = factor(comparison, levels = c("iː - yː",  "eː - yː", "ɪ - ʏ", "yː - ʉː", "oː - uː", "ɔ - ʊ")),
    rounding = ifelse(comparison %in% c("iː - yː",  "eː - yː", "ɪ - ʏ"), "unrounded vs. outrounded", "outrounded vs. inrounded")) %>%
  pivot_longer(
    cols = starts_with("F"),
    names_to = c("cue_dimension", "measure"),
    names_sep = "_") %>%
  pivot_wider(
    names_from = "measure",
    values_from = "value") %>%
  mutate(
    cue_dimension = factor(cue_dimension, levels = c("F1F2", "F1F2F3")))

# Calculate the relative increase of including F3 for the rounded contrasts
d.SwehVd.cat.separability.comparison.talkerIndep <- d.SwehVd.cat.separability.talkerIndep %>%
  filter(comparison %in% c(c("iː - yː",  "eː - yː", "ɪ - ʏ", "yː - ʉː", "oː - uː", "ɔ - ʊ"))) %>%
  group_by(comparison) %>%
  mutate(
    separabilityDifference_F3 = (F1F2F3_separability - F1F2_separability) / F1F2_separability)
  
p.category.separability.unrounded.talkerIndep <- d.SwehVd.cat.separability.for.rounding.talkerIndep %>%
  filter(rounding == "unrounded vs. outrounded") %>%
  ggplot(
    aes(x = cue_dimension,
        y = separability,
        ymin = (separability - SE),
        ymax = (separability + SE),
        color = comparison)) +
  geom_pointrange(
    position = position_dodge2(width = .3),
    fatten = 2.5,
    size = .8) +
  geom_line(aes(group = comparison), position = position_dodge(width = 0.3), linetype = "dashed") +
  scale_color_manual(name = "comparison", values = colors.comparison) + 
  scale_y_continuous("Category separability index", limits = c(.05, .53)) +
  guides(shape = "none", color = guide_legend(ncol = 2)) +
  myGplot.defaults() +
  theme(axis.text.x = ggtext::element_markdown(angle = 60, vjust = 1, hjust = 1),
        axis.title.x = element_blank(),
        legend.title = element_blank(),
        legend.position = "bottom")

p.category.separability.inrounded.talkerIndep <- p.category.separability.unrounded.talkerIndep %+%
  (d.SwehVd.cat.separability.for.rounding.talkerIndep %>%
  filter(rounding == "outrounded vs. inrounded"))

p.category.separability.diff.unrounded.talkerIndep <- d.SwehVd.cat.separability.comparison.talkerIndep %>%
  select(comparison, separabilityDifference_F3) %>%
  filter(comparison %in% c("iː - yː",  "eː - yː", "ɪ - ʏ")) %>%
  pivot_longer(
    cols = starts_with("separabilityDifference"),
    names_to = "cue_difference") %>%
  rbind(d.SwehVd.cat.separability.comparison.talkerIndep %>% 
          select(comparison, separabilityDifference_F3) %>%
          filter(comparison %in% c("iː - yː",  "eː - yː", "ɪ - ʏ")) %>%
          ungroup() %>%
          summarise(
            separabilityDifference_F3 = mean(separabilityDifference_F3)) %>%
          pivot_longer(
            cols = starts_with("separabilityDifference"),
            names_to = "cue_difference") %>%
          mutate(comparison = "overall")) %>%
  mutate(
    comparison = factor(comparison, levels = c("iː - yː",  "eː - yː", "ɪ - ʏ", "overall")),
    rounding = "unrounded vs. outrounded",
    cue_difference = factor(cue_difference, levels = "separabilityDifference_F3", labels = "F3-inclusion")) %>%
  ggplot(
    aes(x = comparison,
        y = value,
        ymin = value,
        ymax = value,
        color = cue_difference)) +  
  geom_pointrange(
    position = position_dodge2(width = .6),
    fatten = 2.5,
    size = .8) +
  geom_hline(
    aes(
      yintercept = 0),
    color = "gray", 
    linetype = 2) +
  scale_color_brewer(palette = "Dark2") +
  scale_y_continuous("Increase in category separability relative to F1-F2", limits = c(-.05, 1)) +
  guides(shape = "none", color = guide_legend(ncol = 2)) +
  myGplot.defaults() +
  theme(axis.text.x = ggtext::element_markdown(angle = 60, vjust = 1, hjust = 1),
        axis.title.x = element_blank(),
        legend.title = element_blank(),
        legend.position = "bottom")

p.category.separability.diff.inrounded.talkerIndep <- p.category.separability.diff.unrounded.talkerIndep %+%
  (d.SwehVd.cat.separability.comparison.talkerIndep %>%
     select(comparison, separabilityDifference_F3) %>%
     filter(comparison %in% c("yː - ʉː", "oː - uː", "ɔ - ʊ")) %>%
     pivot_longer(
       cols = starts_with("separabilityDifference"),
       names_to = "cue_difference") %>%
     rbind(d.SwehVd.cat.separability.comparison.talkerIndep %>% 
             select(comparison, separabilityDifference_F3) %>%
             filter(comparison %in% c("yː - ʉː", "oː - uː", "ɔ - ʊ")) %>%
             ungroup() %>%
             summarise(
               separabilityDifference_F3 = mean(separabilityDifference_F3)) %>%
             pivot_longer(
               cols = starts_with("separabilityDifference"),
               names_to = "cue_difference") %>%
             mutate(comparison = "overall")) %>%
     mutate(
       comparison = factor(comparison, levels = c("yː - ʉː", "oː - uː", "ɔ - ʊ", "overall")),
       rounding = "outrounded vs. inrounded",
       cue_difference = factor(cue_difference, levels = "separabilityDifference_F3", labels = "F3-inclusion")))

(p.category.separability.unrounded.talkerIndep + facet_wrap(~rounding) + p.category.separability.diff.unrounded.talkerIndep + facet_wrap(~rounding)) /
(p.category.separability.inrounded.talkerIndep + facet_wrap(~rounding) + p.category.separability.diff.inrounded.talkerIndep + facet_wrap(~rounding)) +
  plot_annotation(tag_levels = list(c('A', 'B', '', '')))

```

```{r}
#Calculate separability for all long-short pairs for three-way comparison
d.SwehVd.cat.separability.long.short.talkerIndep <- rbind(
  # Get category separability of i: from y:
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[iː]", category_B = "[ɪ]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    # label the same pairwise comparison as there are no differences between the categories for these contrasts
    mutate(comparison = "iː - ɪ"),
  # and then y: from i:
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ɪ]", category_B = "[iː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "iː - ɪ"),
  d.SwehVd.for.cat.separability %>%
    get_category_separability_talkerIndep(category_A = "[yː]", category_B = "[ʏ]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "yː - ʏ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ʏ]", category_B = "[yː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "yː - ʏ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ʉː]", category_B = "[ɵ]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "ʉː - ɵ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ɵ]", category_B = "[ʉː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "ʉː - ɵ"),
  d.SwehVd.for.cat.separability %>%
    get_category_separability_talkerIndep(category_A = "[eː]", category_B = "[ɛ]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "eː - ɛ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ɛ]", category_B = "[eː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "eː - ɛ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ɛː]", category_B = "[ɛ]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "ɛː - ɛ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ɛ]", category_B = "[ɛː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "ɛː - ɛ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[æː]", category_B = "[æ]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "æː - æ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[æ]", category_B = "[æː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "æː - æ"),
    d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[øː]", category_B = "[ø]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "øː - ø"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ø]", category_B = "[øː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "øː - ø"),
  d.SwehVd.for.cat.separability %>%
    get_category_separability_talkerIndep(category_A = "[œː]", category_B = "[œ]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "œː - œ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[œ]", category_B = "[œː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "œː - œ"),
  d.SwehVd.for.cat.separability %>%
    get_category_separability_talkerIndep(category_A = "[ɑː]", category_B = "[a]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "ɑː - a"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[a]", category_B = "[ɑː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "ɑː - a"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[oː]", category_B = "[ɔ]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "oː - ɔ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ɔ]", category_B = "[oː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "oː - ɔ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[uː]", category_B = "[ʊ]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "uː - ʊ"),
  d.SwehVd.for.cat.separability %>%
  get_category_separability_talkerIndep(category_A = "[ʊ]", category_B = "[uː]", data_means = d.SwehVd.centroids.talkerIndep) %>%
    mutate(comparison = "uː - ʊ")) %>%
  #average the separability across the two distances
  group_by(comparison) %>%
  summarise(
      F1F2_SE = geometric.mean(SE_F1F2),
      F1F2F3_SE = geometric.mean(SE_F1F2F3),
      F1F2Duration_SE = geometric.mean(SE_F1F2duration),
      F1F2_separability = mean(category_separabilityF1F2, na.rm = T),
      F1F2F3_separability = mean(category_separabilityF1F2F3, na.rm = T),
      F1F2Duration_separability = mean(category_separabilityF1F2duration, na.rm = T))

d.SwehVd.cat.separability.for.longshort.talkerIndep <- d.SwehVd.cat.separability.long.short.talkerIndep %>%
  mutate(
    comparison = factor(comparison, levels = c("iː - ɪ",  "yː - ʏ", "ʉː - ɵ", "eː - ɛ", "oː - ɔ", "uː - ʊ", "ɛː - ɛ", "æː - æ", "øː - ø", "œː - œ", "ɑː - a"))) %>%
  pivot_longer(
    cols = starts_with("F"),
    names_to = c("cue_dimension", "measure"),
    names_sep = "_") %>%
  pivot_wider(
    names_from = "measure",
    values_from = "value") %>%
  mutate(
    height = ifelse(comparison %in% c("iː - ɪ",  "yː - ʏ", "ʉː - ɵ", "eː - ɛ", "oː - ɔ", "uː - ʊ"), "high vowels", "lower-mid vowels"))

d.SwehVd.cat.separability.for.longshort.talkerIndep %<>%
  filter(cue_dimension %in% c("F1F2", "F1F2F3", "F1F2Duration")) %>%
  mutate(
    cue_dimension = factor(cue_dimension, levels = c("F1F2", "F1F2F3", "F1F2Duration")))
```

(ref:SI-cat-sep-long-short) The effect of including F3 and duration in measures of category separability for long-short vowel pair distinctions, under the assumption of talker-independent representations. For visualization purposes, the pairs are split into high vowels (**top row**), and lower-mid vowels (**bottom row**). **Left panels** plot the category separability for F1-F2, F1-F2-F3 and F1-F2-duration cue combinations. **Right panels** plot the increase in category separability relative to F1-F2-baseline. Axis ranges are held constant across columns.

```{r SI-cat-sep-long-short, fig.width=base.width*4, fig.height=base.height*4.5, out.width='100%', fig.align='center', fig.cap="(ref:SI-cat-sep-long-short)", message=FALSE}
colors.comparison <- c("#66C2A5", "#FC8D62", "#8DA0CB", "#FB9A99", "#E7298A", "#A6D854", "#FFD92F", "#E5C494", "#B3B3B3", "#A6CEE3", "#33A02C")
names(colors.comparison) <- c("iː - ɪ",  "yː - ʏ", "ʉː - ɵ", "eː - ɛ", "ɛː - ɛ", "æː - æ", "øː - ø", "œː - œ", "ɑː - a",  "oː - ɔ", "uː - ʊ")

p.category.separability.high.talkerIndep <- p.category.separability.unrounded.talkerIndep %+%
  (d.SwehVd.cat.separability.for.longshort.talkerIndep %>%
      filter(height == "high vowels")) + 
  scale_color_manual(name = "comparison", values = colors.comparison)

p.category.separability.lowermid.talkerIndep <- p.category.separability.unrounded.talkerIndep %+%
  (d.SwehVd.cat.separability.for.longshort.talkerIndep %>%
      filter(height == "lower-mid vowels")) + 
  scale_color_manual(name = "comparison", values = colors.comparison)
  
# Calculate the relative increase of including F3 and duration for the quantity contrasts
d.SwehVd.cat.separability.comparison.longshort.talkerIndep <- d.SwehVd.cat.separability.long.short.talkerIndep %>%
  group_by(comparison) %>%
  mutate(
    separabilityDifference_F3 = (F1F2F3_separability - F1F2_separability) / F1F2_separability,
    separabilityDifference_Duration = (F1F2Duration_separability - F1F2_separability) / F1F2_separability,
    height = ifelse(comparison %in% c("iː - ɪ",  "yː - ʏ", "ʉː - ɵ", "eː - ɛ", "oː - ɔ", "uː - ʊ"), "high vowels", "lower-mid vowels"))

p.category.separability.diff.high.talkerIndep <- p.category.separability.diff.unrounded.talkerIndep %+%
  (d.SwehVd.cat.separability.comparison.longshort.talkerIndep %>%
     select(comparison, height, separabilityDifference_F3, separabilityDifference_Duration) %>%
     filter(height == "high vowels") %>%
     pivot_longer(
       cols = starts_with("separabilityDifference"),
       names_to = "cue_difference") %>%
     mutate(
       comparison = factor(comparison, levels = c("iː - ɪ",  "yː - ʏ", "ʉː - ɵ", "eː - ɛ", "oː - ɔ", "uː - ʊ", "overall")),
       height = "high vowels",
       cue_difference = factor(cue_difference, levels = c("separabilityDifference_F3", "separabilityDifference_Duration"), labels = c("F3-inclusion", "Duration-inclusion"))))

p.category.separability.diff.lowermid.talkerIndep <- p.category.separability.diff.high.talkerIndep %+%
  (d.SwehVd.cat.separability.comparison.longshort.talkerIndep %>%
     select(comparison, height, separabilityDifference_F3, separabilityDifference_Duration) %>%
     filter(height == "lower-mid vowels") %>%
     pivot_longer(
       cols = starts_with("separabilityDifference"),
       names_to = "cue_difference") %>%
     mutate(
       comparison = factor(comparison, levels = c("ɛː - ɛ", "æː - æ", "øː - ø", "œː - œ", "ɑː - a", "overall")),
       height = "lower-mid vowels",
       cue_difference = factor(cue_difference, levels = c("separabilityDifference_F3", "separabilityDifference_Duration"), labels = c("F3-inclusion", "Duration-inclusion"))))

(p.category.separability.high.talkerIndep + facet_wrap(~height) + p.category.separability.diff.high.talkerIndep + facet_wrap(~height)) /
(p.category.separability.lowermid.talkerIndep + facet_wrap(~height) + (p.category.separability.diff.lowermid.talkerIndep + facet_wrap(~height) + theme(legend.position = "none"))) +
  plot_annotation(tag_levels = list(c('A', 'B', '', ''))) #+
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom",
        legend.title = element_blank(),
        #legend.margin = margin(0, 0, 0, 0, unit = "pt"),
        plot.tag = element_text(face = "bold"))
```
```{r}
rm(d.SwehVd.cat.separability.comparison.longshort.talkerIndep, d.SwehVd.cat.separability.for.longshort.talkerIndep, d.SwehVd.cat.separability.comparison.talkerIndep, d.SwehVd.cat.separability.for.rounding.talkerIndep, d.SwehVd.centroids.talkerIndep, d.SwehVd.cat.separability.talkerIndep)
```

```{r}
#gamtabs(m.F2.front.long.Y.ord, caption='Model summary, coefficients for parametric and smooth terms', label = "tab:m.F2.front.long.Y.ord")
#gamtabs(m.F3.front.long.Y.ord, caption='Model summary, coefficients for parametric and smooth terms', label = "tab:m.F3.front.long.Y.ord")
```

## Additional information concerning dynamic analyses
### Summary tables of GAMMs {#sec:SI-GAMMs}

\begin{table}[ht]
\caption{\label{tab:m.F1.front.long.Y.ord}Model summary, coefficients for parametric and smooth terms for [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.2505 & 0.0175 & -71.6225 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[yː]")` & -0.0290 & 0.0051 & -5.6460 & $<$ 0.0001 \\
  Gendermale & -0.0793 & 0.0244 & -3.2529 & 0.0012 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.0168 & 3.4940 & 49.7632 & $<$ 0.0001 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 1.8190 & 2.2251 & 1.0603 & 0.4002 \\ 
  s(Talker) & 35.5354 & 39.0000 & 519.3428 & $<$ 0.0001 \\ 
  s(Talker,category) & 37.2811 & 79.0000 & 22.0717 & 0.0987 \\ 
   \hline
\end{tabular}
\end{table}


\begin{table}[ht]
\centering
\caption{\label{tab:m.F2.front.long.Y.ord}Model summary, coefficients for parametric and smooth terms for [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] predicting F2} 
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.4445 & 0.0185 & 24.0769 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[yː]")` & -0.0259 & 0.0038 & -6.7647 & $<$ 0.0001 \\ 
  Gendermale & 0.0539 & 0.0258 & 2.0876 & 0.0369 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.4321 & 2.9246 & 26.6703 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 1.5879 & 1.9288 & 1.0032 & 0.4309 \\ 
  s(Talker) & 37.2737 & 39.0000 & 3559.5780 & $<$ 0.0001 \\ 
  s(Talker,category) & 38.4339 & 79.0000 & 69.9332 & 0.3622 \\ 
   \hline
\end{tabular}
\end{table}


\begin{table}[ht]
\centering
\caption{\label{tab:m.F3.front.long.Y.ord}Model summary, coefficients for parametric and smooth terms for [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] predicting F3} 
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.8997 & 0.0119 & 75.8438 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[yː]")` & -0.0021 & 0.0102 & -0.2023 & 0.8397 \\ 
  Gendermale & 0.0163 & 0.0166 & 0.9830 & 0.3257 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 1.0001 & 1.0002 & 19.2566 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 2.7282 & 3.2476 & 6.8551 & 0.0001 \\ 
  s(Talker) & 9.5569 & 39.0000 & 216.3509 & 0.0605 \\ 
  s(Talker,category) & 68.0168 & 79.0000 & 347.3984 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F1.front.long.U.ord}Model summary, coefficients for parametric and smooth terms for [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("ʉː")`] predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.2127 & 0.0161 & -75.3482 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʉː")`]  & 0.0336 & 0.0081 & 4.1184 & $<$ 0.0001 \\ 
  Gendermale & -0.0663 & 0.0225 & -2.9509 & 0.0032 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.0241 & 3.4728 & 46.6752 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʉː")`]  & 3.0198 & 3.4684 & 21.0872 & $<$ 0.0001 \\ 
  s(Talker) & 28.7465 & 39.0000 & 709.6758 & $<$ 0.0001 \\ 
  s(Talker,category) & 47.3292 & 79.0000 & 121.4009 & 0.0251 \\ 
   \hline
\end{tabular}
\end{table}


\begin{table}[ht]
\centering
\caption{\label{tab:m.F2.front.long.U.ord}Model summary, coefficients for parametric and smooth terms for [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("ʉː")`] predicting F2} 
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.3865 & 0.0131 & 29.4871 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʉː")`]  & -0.1204 & 0.0091 & -13.1598 & $<$ 0.0001 \\ 
  Gendermale & 0.0368 & 0.0183 & 2.0105 & 0.0445 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 1.8214 & 2.2273 & 29.9642 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʉː")`]  & 2.9817 & 3.4626 & 81.3483 & $<$ 0.0001 \\ 
  s(Talker) & 19.5350 & 39.0000 & 2205.8452 & 0.0017 \\ 
  s(Talker,category) & 58.5627 & 79.0000 & 1071.5104 & 0.0019 \\ 
   \hline
\end{tabular}
\end{table}


\begin{table}[ht]
\centering
\caption{\label{tab:m.F3.front.long.U.ord}Model summary, coefficients for parametric and smooth terms for [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("ʉː")`] predicting F3} 
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7963 & 0.0113 & 70.5968 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʉː")`] & -0.1368 & 0.0111 & -12.2835 & $<$ 0.0001 \\ 
  Gendermale & 0.0321 & 0.0158 & 2.0365 & 0.0418 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 1.6940 & 2.0710 & 8.8735 & 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʉː")`] & 2.9234 & 3.4138 & 10.9632 & $<$ 0.0001 \\ 
  s(Talker) & 0.0001 & 39.0000 & 0.0000 & 0.7681 \\ 
  s(Talker,category) & 77.4959 & 79.0000 & 54.7815 & $<$ 0.0001 \\
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F1.front.long.E.ord}Model summary, coefficients for parametric and smooth terms for [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("eː")`] predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.1692 & 0.0139 & -84.2519 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("eː")`] & 0.1307 & 0.0119 & 10.9686 & $<$ 0.0001 \\ 
  Gendermale & -0.0176 & 0.0194 & -0.9067 & 0.3646 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.7489 & 3.2572 & 46.3007 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("eː")`] & 3.8695 & 3.9748 & 790.8174 & $<$ 0.0001 \\ 
  s(Talker) & 9.5453 & 39.0000 & 87.1163 & 0.0580 \\ 
  s(Talker,category) & 67.0920 & 79.0000 & 162.7500 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\end{table}


\begin{table}[ht]
\centering
\caption{\label{tab:m.F2.front.long.E.ord}Model summary, coefficients for parametric and smooth terms for [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("eː")`] predicting F2} 
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.5364 & 0.0145 & 36.9465 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("eː")`] & 0.0800 & 0.0143 & 5.5759 & $<$ 0.0001 \\ 
  Gendermale & 0.0209 & 0.0203 & 1.0278 & 0.3041 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 1.2386 & 1.4340 & 16.9517 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("eː")`] & 3.7960 & 3.9729 & 264.2232 & $<$ 0.0001 \\ 
  s(Talker) & 0.0019 & 39.0000 & 0.0001 & 0.4672 \\ 
  s(Talker,category) & 77.3206 & 79.0000 & 52.8965 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\end{table}


\begin{table}[ht]
\centering
\caption{\label{tab:m.F3.front.long.E.ord}Model summary, coefficients for parametric and smooth terms for [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("eː")`] predicting F3} 
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.8664 & 0.0131 & 66.0824 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("eː")`] & -0.0417 & 0.0096 & -4.3198 & $<$ 0.0001 \\ 
  Gendermale & 0.0265 & 0.0183 & 1.4446 & 0.1487 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 1.0001 & 1.0001 & 14.3112 & 0.0002 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("eː")`] & 3.5533 & 3.8876 & 63.5281 & $<$ 0.0001 \\ 
  s(Talker) & 17.3913 & 39.0000 & 366.7896 & 0.0004 \\ 
  s(Talker,category) & 59.3862 & 79.0000 & 225.6487 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\end{table}

```{r}
```

\begin{table}[ht]
\caption{\label{tab:m.F1.front.short}Model summary, coefficients for parametric and smooth terms for the high front short vowels predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.2153 & 0.0127 & -95.5924 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʏ")`]  & 0.0075 & 0.0067 & 1.1083 & 0.2678 \\ 
  Gendermale & -0.0254 & 0.0178 & -1.4318 & 0.1523 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.3979 & 2.9060 & 3.3187 & 0.0175 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʏ")`] & 1.0000 & 1.0000 & 4.2695 & 0.0389 \\ 
  s(Talker) & 27.7413 & 39.0000 & 468.5362 & $<$ 0.0001 \\ 
  s(Talker,category) & 47.9833 & 79.0000 & 90.8256 & 0.0206 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.front.short}Model summary, coefficients for parametric and smooth terms for the high front short vowels predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.6273 & 0.0128 & 48.8949 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʏ")`]  & -0.0586 & 0.0116 & -5.0568 & $<$ 0.0001 \\ 
  Gendermale & -0.0149 & 0.0179 & -0.8331 & 0.4048 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.3278 & 3.7567 & 13.3047 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʏ")`] & 1.0000 & 1.0001 & 3.1409 & 0.0764 \\ 
  s(Talker) & 6.3833 & 39.0000 & 86.6745 & 0.1901 \\ 
  s(Talker,category) & 71.1834 & 79.0000 & 263.8538 & $<$ 0.0001 \\
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.front.short}Model summary, coefficients for parametric and smooth terms for the high front short vowels predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.8120 & 0.0094 & 86.7831 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʏ")`] & -0.0727 & 0.0050 & -14.5131 & $<$ 0.0001 \\ 
  Gendermale & 0.0044 & 0.0131 & 0.3376 & 0.7357 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.7760 & 3.2549 & 65.5966 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʏ")`] & 3.0148 & 3.4704 & 52.0579 & $<$ 0.0001 \\ 
  s(Talker) & 27.5301 & 39.0000 & 645.2514 & $<$ 0.0001 \\ 
  s(Talker,category) & 48.8674 & 79.0000 & 133.6283 & 0.0486 \\
   \hline
\end{tabular}
\end{table}

```{r}
```

\begin{table}[ht]
\caption{\label{tab:m.F1.back.long}Model summary, coefficients for parametric and smooth terms for the high back long vowels predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.1296 & 0.0124 & -91.4148 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("uː")`] & -0.1061 & 0.0091 & -11.6801 & $<$ 0.0001 \\ 
  Gendermale & 0.0329 & 0.0173 & 1.9054 & 0.0568 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.7637 & 3.9505 & 416.3048 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("uː")`] & 3.6849 & 3.9255 & 369.2896 & $<$ 0.0001 \\ 
  s(Talker) & 17.3835 & 39.0000 & 94.2355 & 0.0032 \\ 
  s(Talker,category) & 57.6310 & 79.0000 & 63.2542 & $<$ 0.0001 \\
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.back.long}Model summary, coefficients for parametric and smooth terms for the high back long vowels predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.5157 & 0.0174 & -29.5961 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("uː")`] & 0.0398 & 0.0103 & 3.8663 & 0.0001 \\ 
  Gendermale & 0.0682 & 0.0243 & 2.8031 & 0.0051 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.8784 & 3.9843 & 884.3931 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("uː")`] & 3.7862 & 3.9658 & 174.8411 & $<$ 0.0001 \\ 
  s(Talker) & 25.0350 & 39.0000 & 134.7690 & $<$ 0.0001 \\ 
  s(Talker,category) & 48.6357 & 79.0000 & 36.1400 & 0.0058 \\
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.back.long}Model summary, coefficients for parametric and smooth terms for the high back long vowels predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7881 & 0.0120 & 65.5786 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("uː")`] & -0.0236 & 0.0051 & -4.6657 & $<$ 0.0001 \\ 
  Gendermale & 0.0267 & 0.0168 & 1.5887 & 0.1122 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.2571 & 3.6883 & 15.2473 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("uː")`] & 1.6521 & 2.0207 & 19.1269 & $<$ 0.0001 \\ 
  s(Talker) & 31.8986 & 39.0000 & 812.7661 & $<$ 0.0001 \\ 
  s(Talker,category) & 43.8638 & 79.0000 & 89.2372 & 0.1221 \\  
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F1.back.short}Model summary, coefficients for parametric and smooth terms for the high back short vowels predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.0462 & 0.0116 & -89.8077 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʊ")`] & -0.1386 & 0.0097 & -14.2818 & $<$ 0.0001 \\ 
  Gendermale & 0.0232 & 0.0162 & 1.4336 & 0.1518 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.0603 & 3.5497 & 75.9391 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʊ")`] & 1.0002 & 1.0003 & 5.4461 & 0.0197 \\ 
  s(Talker) & 10.9301 & 39.0000 & 104.7983 & 0.0852 \\ 
  s(Talker,category) & 64.8352 & 78.0000 & 148.9877 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.back.short}Model summary, coefficients for parametric and smooth terms for the high back short vowels predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.3769 & 0.0156 & -24.2012 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʊ")`] & -0.0424 & 0.0088 & -4.8025 & $<$ 0.0001 \\ 
  Gendermale & -0.0112 & 0.0217 & -0.5154 & 0.6063 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.5562 & 3.8467 & 76.9066 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʊ")`] & 2.7071 & 3.1932 & 71.2364 & $<$ 0.0001 \\ 
  s(Talker) & 26.0859 & 39.0000 & 272.0183 & $<$ 0.0001 \\ 
  s(Talker,category) & 48.0521 & 78.0000 & 66.7823 & 0.0096 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.back.short}Model summary, coefficients for parametric and smooth terms for the high back short vowels predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.8255 & 0.0113 & 73.0739 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʊ")`] & -0.0292 & 0.0056 & -5.2185 & $<$ 0.0001 \\ 
  Gendermale & 0.0124 & 0.0157 & 0.7887 & 0.4303 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 1.0001 & 1.0002 & 95.7868 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʊ")`] & 1.9991 & 2.4488 & 6.2454 & 0.0015 \\ 
  s(Talker) & 29.1275 & 39.0000 & 2403.7161 & $<$ 0.0001 \\ 
  s(Talker,category) & 47.3171 & 78.0000 & 374.0028 & 0.0998 \\ 
   \hline
\end{tabular}
\end{table}

```{r}
```

\begin{table}[ht]
\caption{\label{tab:m.F1.center.long}Model summary, coefficients for parametric and smooth terms for the lower-mid front long vowels predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.4664 & 0.0205 & -22.7936 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æː")`] & 0.0460 & 0.0084 & 5.4424 & $<$ 0.0001 \\ 
  Gendermale & -0.0209 & 0.0286 & -0.7300 & 0.4655 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.3985 & 2.8801 & 5.5492 & 0.0009 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æː")`] & 2.9271 & 3.4029 & 42.2298 & $<$ 0.0001 \\ 
  s(Talker) & 32.1863 & 39.0000 & 852.9541 & $<$ 0.0001 \\ 
  s(Talker,category) & 43.1275 & 79.0000 & 89.3690 & 0.0311 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.center.long}Model summary, coefficients for parametric and smooth terms for the lower-mid front long vowels predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.2778 & 0.0099 & 28.0182 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æː")`] & -0.0559 & 0.0065 & -8.6094 & $<$ 0.0001 \\ 
  Gendermale & -0.0062 & 0.0139 & -0.4452 & 0.6562 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.4254 & 3.7885 & 244.3314 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æː")`] & 3.6101 & 3.8806 & 242.5392 & $<$ 0.0001 \\ 
  s(Talker) & 21.8765 & 39.0000 & 1052.7781 & $<$ 0.0001 \\ 
  s(Talker,category) & 55.3558 & 79.0000 & 410.4479 & 0.0004 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.center.long}Model summary, coefficients for parametric and smooth terms for the lower-mid front long vowels
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7418 & 0.0160 & 46.4515 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æː")`] & -0.0063 & 0.0071 & -0.8957 & 0.3705 \\ 
  Gendermale & 0.0264 & 0.0223 & 1.1837 & 0.2366 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.5570 & 3.0749 & 18.5927 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æː")`] & 1.0000 & 1.0001 & 13.8709 & 0.0002 \\ 
  s(Talker) & 31.1397 & 39.0000 & 914.4816 & $<$ 0.0001 \\ 
  s(Talker,category) & 44.5499 & 79.0000 & 99.2827 & 0.0933 \\ 
   \hline
\end{tabular} 
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F1.center.short}Model summary, coefficients for parametric and smooth terms for the lower-mid front short vowels
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.6015 & 0.0130 & -46.3345 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æ")`] & 0.1726 & 0.0104 & 16.5493 & $<$ 0.0001 \\ 
  Gendermale & -0.0163 & 0.0181 & -0.8991 & 0.3687 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.4917 & 3.8332 & 51.7731 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æ")`] & 1.9327 & 2.3680 & 45.2830 & $<$ 0.0001 \\ 
  s(Talker) & 13.1542 & 39.0000 & 160.7058 & 0.0705 \\ 
  s(Talker,category) & 63.7164 & 79.0000 & 177.0030 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.center.short}Model summary, coefficients for parametric and smooth terms for the lower-mid front short vowels
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.3499 & 0.0113 & 31.0690 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æ")`] & -0.0797 & 0.0082 & -9.6873 & $<$ 0.0001 \\ 
  Gendermale & -0.0123 & 0.0157 & -0.7810 & 0.4349 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 1.1719 & 1.3066 & 365.5786 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æ")`] & 2.3852 & 2.8844 & 208.8811 & $<$ 0.0001 \\ 
  s(Talker) & 17.6245 & 39.0000 & 429.5987 & 0.0028 \\ 
  s(Talker,category) & 59.5197 & 79.0000 & 269.1815 & $<$ 0.0001 \\  
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.center.short}Model summary, coefficients for parametric and smooth terms for the lower-mid front short vowels
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7520 & 0.0128 & 58.8018 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æ")`] & 0.0099 & 0.0055 & 1.7964 & 0.0725 \\ 
  Gendermale & 0.0297 & 0.0179 & 1.6637 & 0.0963 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.7661 & 3.2897 & 4.4730 & 0.0025 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æ")`] & 1.0001 & 1.0001 & 0.8789 & 0.3486 \\ 
  s(Talker) & 31.6397 & 39.0000 & 296.4814 & $<$ 0.0001 \\ 
  s(Talker,category) & 41.9894 & 79.0000 & 30.9853 & 0.0725 \\
   \hline
\end{tabular}
\end{table}

```{r}
#gamtabs(m.F3.mid.short, caption='Model summary, coefficients for parametric and smooth terms', label = "tab:m.F3.mid.short")
```

\begin{table}[ht]
\caption{\label{tab:m.F1.mid.long}Model summary, coefficients for parametric and smooth terms for the mid center long vowels predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.7265 & 0.0169 & -42.9651 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œː")`] & -0.0098 & 0.0073 & -1.3412 & 0.1799 \\ 
  Gendermale & -0.0176 & 0.0236 & -0.7457 & 0.4559 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.9469 & 3.4536 & 4.4850 & 0.0033 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œː")`] & 1.0001 & 1.0001 & 39.1344 & $<$ 0.0001 \\ 
  s(Talker) & 31.5576 & 39.0000 & 1258.5044 & $<$ 0.0001 \\ 
  s(Talker,category) & 44.6301 & 79.0000 & 145.9365 & 0.0505 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.mid.long}Model summary, coefficients for parametric and smooth terms for the mid center long vowels predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.0744 & 0.0117 & 6.3573 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œː")`] & -0.1171 & 0.0114 & -10.2512 & $<$ 0.0001 \\ 
  Gendermale & -0.0104 & 0.0163 & -0.6372 & 0.5241 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.2941 & 3.7089 & 193.6028 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œː")`] & 3.7297 & 3.9282 & 381.5898 & $<$ 0.0001 \\ 
  s(Talker) & 0.8813 & 39.0000 & 6.9354 & 0.4171 \\ 
  s(Talker,category) & 77.3538 & 79.0000 & 254.5904 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.mid.long}Model summary, coefficients for parametric and smooth terms for the mid center long vowels predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7323 & 0.0108 & 67.6613 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œː")`] & 0.0104 & 0.0052 & 1.9782 & 0.0480 \\ 
  Gendermale & -0.0025 & 0.0151 & -0.1681 & 0.8665 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.3374 & 3.7227 & 59.1292 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œː")`] & 3.4294 & 3.7772 & 67.9323 & $<$ 0.0001 \\ 
  s(Talker) & 29.6535 & 39.0000 & 1744.4730 & $<$ 0.0001 \\ 
  s(Talker,category) & 47.2299 & 79.0000 & 260.6346 & 0.0391 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F1.mid.short}Model summary, coefficients for parametric and smooth terms for the mid center short vowels
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.8545 & 0.0123 & -69.4714 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œ")`] & -0.0158 & 0.0097 & -1.6344 & 0.1023 \\ 
  Gendermale & 0.0125 & 0.0172 & 0.7262 & 0.4678 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.5796 & 3.8588 & 173.3591 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œ")`] & 2.9340 & 3.4113 & 16.0463 & $<$ 0.0001 \\ 
  s(Talker) & 14.1702 & 39.0000 & 302.4206 & 0.0380 \\ 
  s(Talker,category) & 63.1586 & 79.0000 & 279.1835 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.mid.short}Model summary, coefficients for parametric and smooth terms for the mid center short vowels
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.1595 & 0.0141 & 11.3009 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œ")`] & -0.1191 & 0.0114 & -10.4280 & $<$ 0.0001 \\ 
  Gendermale & -0.0344 & 0.0197 & -1.7436 & 0.0813 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.2604 & 2.7351 & 34.5559 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œ")`] & 3.4647 & 3.8071 & 349.8467 & $<$ 0.0001 \\ 
  s(Talker) & 12.8319 & 39.0000 & 2243.9486 & 0.0298 \\ 
  s(Talker,category) & 65.5976 & 79.0000 & 2258.5568 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.mid.short}Model summary, coefficients for parametric and smooth terms for the mid center short vowels
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7432 & 0.0098 & 75.6290 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œ")`] & 0.0114 & 0.0056 & 2.0340 & 0.0420 \\ 
  Gendermale & 0.0062 & 0.0137 & 0.4523 & 0.6511 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.2841 & 3.6832 & 149.3740 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œ")`] & 3.3298 & 3.7122 & 106.1946 & $<$ 0.0001 \\ 
  s(Talker) & 25.9119 & 39.0000 & 1422.1973 & $<$ 0.0001 \\ 
  s(Talker,category) & 51.4251 & 79.0000 & 345.2444 & 0.0184 \\ 
   \hline
\end{tabular}
\end{table}

```{r}
```


\begin{table}[ht]
\caption{\label{tab:m.F1.I}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("i")`/
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.2427 & 0.0157 & -78.9328 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɪ]")` & 0.0371 & 0.0137 & 2.7134 & 0.0067 \\ 
  Gendermale & -0.0545 & 0.0198 & -2.7534 & 0.0059 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.9563 & 3.4642 & 55.3871 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɪ]")` & 1.0000 & 1.0001 & 68.4212 & $<$ 0.0001 \\ 
  s(Talker) & 20.3749 & 39.0000 & 505.9699 & 0.0001 \\ 
  s(Talker,category) & 56.5165 & 79.0000 & 231.1143 & 0.0002 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.I}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("i")`/
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.4737 & 0.0174 & 27.1554 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɪ]")` & 0.1704 & 0.0196 & 8.6881 & $<$ 0.0001 \\ 
  Gendermale & 0.0327 & 0.0202 & 1.6233 & 0.1046 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 1.3830 & 1.6530 & 18.0138 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɪ]")` & 3.2796 & 3.7133 & 6.7933 & 0.0001 \\ 
  s(Talker) & 2.0415 & 39.0000 & 11.9241 & 0.4174 \\ 
  s(Talker,category) & 75.7153 & 79.0000 & 171.1447 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.I}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("i")`/
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.8922 & 0.0147 & 60.5368 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɪ]")` & -0.0441 & 0.0130 & -3.3795 & 0.0007 \\ 
  Gendermale & 0.0337 & 0.0185 & 1.8258 & 0.0680 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 1.0001 & 1.0003 & 15.5226 & 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɪ]")` & 2.3498 & 2.8529 & 35.7660 & $<$ 0.0001 \\ 
  s(Talker) & 19.5361 & 39.0000 & 568.5014 & 0.0003 \\ 
  s(Talker,category) & 57.6016 & 79.0000 & 289.8411 & 0.0001 \\ 
   \hline
\end{tabular}
\end{table}


\begin{table}[ht]
\caption{\label{tab:m.F1.Y}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("y")`/
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.2862 & 0.0149 & -86.2412 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʏ]")` & 0.0889 & 0.0120 & 7.4271 & $<$ 0.0001 \\ 
  Gendermale & -0.0502 & 0.0191 & -2.6311 & 0.0085 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.1766 & 3.6029 & 65.3943 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʏ]")` & 2.5556 & 3.0388 & 19.2641 & $<$ 0.0001 \\ 
  s(Talker) & 23.6319 & 39.0000 & 469.8048 & $<$ 0.0001 \\ 
  s(Talker,category) & 52.8112 & 79.0000 & 152.3845 & 0.0026 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.Y}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("y")`/
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.4507 & 0.0144 & 31.2079 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʏ]")` & 0.1242 & 0.0159 & 7.8100 & $<$ 0.0001 \\ 
  Gendermale & 0.0062 & 0.0168 & 0.3656 & 0.7147 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.9464 & 3.4178 & 47.4311 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʏ]")` & 2.3870 & 2.8642 & 3.4414 & 0.0162 \\ 
  s(Talker) & 4.2101 & 39.0000 & 365.6413 & 0.3015 \\ 
  s(Talker,category) & 74.3386 & 79.0000 & 1645.4089 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.Y}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("y")`/
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.9134 & 0.0086 & 106.0065 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʏ]")` & -0.1439 & 0.0066 & -21.8142 & $<$ 0.0001 \\ 
  Gendermale & -0.0130 & 0.0111 & -1.1686 & 0.2427 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.2360 & 3.6465 & 17.5977 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʏ]")` & 3.3754 & 3.7403 & 16.9235 & $<$ 0.0001 \\ 
  s(Talker) & 25.2775 & 39.0000 & 949.5166 & $<$ 0.0001 \\ 
  s(Talker,category) & 51.8369 & 79.0000 & 249.7645 & 0.0314 \\ 
   \hline
\end{tabular}
\end{table}


\begin{table}[ht]
\caption{\label{tab:m.F1.U}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("ʉ")`/
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.2142 & 0.0147 & -82.7945 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɵ]")` & 0.1134 & 0.0118 & 9.6084 & $<$ 0.0001 \\ 
  Gendermale & -0.0200 & 0.0188 & -1.0662 & 0.2864 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.5016 & 3.8160 & 173.2009 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɵ]")` & 3.0360 & 3.4928 & 43.2258 & $<$ 0.0001 \\ 
  s(Talker) & 23.5404 & 39.0000 & 442.8618 & $<$ 0.0001 \\ 
  s(Talker,category) & 52.8094 & 79.0000 & 145.9149 & 0.0019 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.U}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("ʉ")`/
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.3157 & 0.0122 & 25.9013 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɵ]")` & -0.3064 & 0.0116 & -26.3315 & $<$ 0.0001 \\ 
  Gendermale & 0.0068 & 0.0150 & 0.4535 & 0.6502 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.9953 & 3.4585 & 298.3962 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɵ]")` & 3.4809 & 3.8035 & 584.4777 & $<$ 0.0001 \\ 
  s(Talker) & 15.3919 & 39.0000 & 773.9038 & 0.0152 \\ 
  s(Talker,category) & 62.4428 & 79.0000 & 579.9508 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.U}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("ʉ")`/
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7145 & 0.0106 & 67.3415 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɵ]")` & 0.0352 & 0.0079 & 4.4349 & $<$ 0.0001 \\ 
  Gendermale & 0.0033 & 0.0137 & 0.2431 & 0.8079 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.4854 & 3.8067 & 27.2527 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɵ]")` & 3.0080 & 3.4685 & 59.1125 & $<$ 0.0001 \\ 
  s(Talker) & 25.9963 & 39.0000 & 1478.7326 & $<$ 0.0001 \\ 
  s(Talker,category) & 51.3885 & 79.0000 & 377.1923 & 0.0153 \\ 
   \hline
\end{tabular}
\end{table}


\begin{table}[ht]
\caption{\label{tab:m.F1.E}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("e")`/
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.0969 & 0.0134 & -81.6369 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɛ]")` & 0.3621 & 0.0141 & 25.6152 & $<$ 0.0001 \\ 
  Gendermale & 0.0082 & 0.0160 & 0.5169 & 0.6053 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.9225 & 3.9937 & 1197.4189 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛ]")` & 3.8937 & 3.9906 & 921.5105 & $<$ 0.0001 \\ 
  s(Talker) & 8.3683 & 39.0000 & 39.6280 & 0.1113 \\ 
  s(Talker,category) & 67.6617 & 79.0000 & 89.3655 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.E}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("e")`/
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.6114 & 0.0163 & 37.4577 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɛ]")` & -0.2070 & 0.0138 & -14.9844 & $<$ 0.0001 \\ 
  Gendermale & -0.0066 & 0.0207 & -0.3172 & 0.7511 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.8022 & 3.9591 & 425.7994 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛ]")` & 3.6306 & 3.9048 & 123.7554 & $<$ 0.0001 \\ 
  s(Talker) & 21.5500 & 39.0000 & 365.0006 & $<$ 0.0001 \\ 
  s(Talker,category) & 54.5449 & 79.0000 & 145.6635 & 0.0001 \\
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.E}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("e")`/
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.8436 & 0.0116 & 72.5287 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɛ]")` & -0.0916 & 0.0088 & -10.3843 & $<$ 0.0001 \\ 
  Gendermale & 0.0157 & 0.0150 & 1.0448 & 0.2962 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.5769 & 3.8592 & 82.3482 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛ]")` & 3.4562 & 3.7990 & 61.6135 & $<$ 0.0001 \\ 
  s(Talker) & 25.5586 & 39.0000 & 250.4581 & $<$ 0.0001 \\ 
  s(Talker,category) & 49.3150 & 79.0000 & 61.7838 & 0.0098 \\ 
   \hline
\end{tabular}
\end{table}


\begin{table}[ht]
\caption{\label{tab:m.F1.Ä}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("ɛ")`/
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.4979 & 0.0168 & -29.5536 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɛ]")` & -0.2217 & 0.0144 & -15.4139 & $<$ 0.0001 \\ 
  Gendermale & -0.0221 & 0.0213 & -1.0374 & 0.2996 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.8687 & 3.3403 & 8.5123 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛ]")` & 3.0057 & 3.4615 & 14.7926 & $<$ 0.0001 \\ 
  s(Talker) & 21.1842 & 39.0000 & 753.9500 & $<$ 0.0001 \\ 
  s(Talker,category) & 55.9935 & 79.0000 & 305.7499 & 0.0017 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.Ä}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("ɛ")`/
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.3201 & 0.0122 & 26.1441 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɛ]")` & 0.0857 & 0.0099 & 8.6797 & $<$ 0.0001 \\ 
  Gendermale & -0.0090 & 0.0157 & -0.5780 & 0.5633 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.3495 & 3.7247 & 157.0077 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛ]")` & 2.6391 & 3.1272 & 3.9379 & 0.0076 \\ 
  s(Talker) & 23.4744 & 39.0000 & 728.9755 & $<$ 0.0001 \\ 
  s(Talker,category) & 53.3979 & 79.0000 & 252.7058 & 0.0014 \\
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.Ä}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("ɛ")`/
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7461 & 0.0139 & 53.8644 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɛ]")` & 0.0003 & 0.0076 & 0.0447 & 0.9643 \\ 
  Gendermale & 0.0267 & 0.0186 & 1.4333 & 0.1519 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.7718 & 3.2960 & 17.0974 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛ]")` & 1.0000 & 1.0001 & 8.1519 & 0.0043 \\ 
  s(Talker) & 32.4840 & 39.0000 & 382.3264 & $<$ 0.0001 \\ 
  s(Talker,category) & 41.0810 & 79.0000 & 32.9008 & 0.0589 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F1.ÄR}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("æ")`/
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.4389 & 0.0196 & -22.3632 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æ]")` & -0.0430 & 0.0136 & -3.1549 & 0.0016 \\ 
  Gendermale & -0.0151 & 0.0257 & -0.5872 & 0.5571 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.3641 & 3.7547 & 108.3002 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æ]")` & 1.8893 & 2.3113 & 1.9999 & 0.1150 \\ 
  s(Talker) & 28.0353 & 39.0000 & 586.8502 & $<$ 0.0001 \\ 
  s(Talker,category) & 47.7162 & 79.0000 & 115.2095 & 0.0079 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.ÄR}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("æ")`/
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.2405 & 0.0119 & 20.2467 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æ]")` & 0.0521 & 0.0091 & 5.7258 & $<$ 0.0001 \\ 
  Gendermale & -0.0093 & 0.0153 & -0.6040 & 0.5459 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.3987 & 3.7515 & 43.0948 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æ]")` & 2.7770 & 3.2571 & 8.0074 & $<$ 0.0001 \\ 
  s(Talker) & 25.2241 & 39.0000 & 1125.4241 & $<$ 0.0001 \\ 
  s(Talker,category) & 51.7897 & 79.0000 & 314.6290 & 0.0025 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.ÄR}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("æ")`/
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7357 & 0.0163 & 45.2652 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æ]")` & 0.0233 & 0.0106 & 2.2046 & 0.0276 \\ 
  Gendermale & 0.0297 & 0.0215 & 1.3819 & 0.1671 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.3374 & 2.8366 & 3.2304 & 0.0391 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æ]")` & 1.0001 & 1.0001 & 2.8277 & 0.0927 \\ 
  s(Talker) & 29.5293 & 39.0000 & 739.5767 & $<$ 0.0001 \\ 
  s(Talker,category) & 46.3596 & 79.0000 & 104.4224 & 0.0879 \\ 
   \hline
\end{tabular}
\end{table}

```{r}
```


\begin{table}[ht]
\caption{\label{tab:m.F1.Ö}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("ø")`/
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.7103 & 0.0156 & -45.5354 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ø]")` & -0.1058 & 0.0117 & -9.0252 & $<$ 0.0001 \\ 
  Gendermale & -0.0351 & 0.0202 & -1.7397 & 0.0820 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.3099 & 2.7827 & 1.6642 & 0.1254 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ø]")` & 3.2258 & 3.6473 & 75.5103 & $<$ 0.0001 \\ 
  s(Talker) & 25.8634 & 39.0000 & 911.2704 & $<$ 0.0001 \\ 
  s(Talker,category) & 51.0366 & 79.0000 & 221.2125 & 0.0176 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.Ö}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("ø")`/
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.1670 & 0.0152 & 10.9938 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ø]")` & 0.0734 & 0.0117 & 6.2872 & $<$ 0.0001 \\ 
  Gendermale & -0.0267 & 0.0196 & -1.3656 & 0.1721 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.2560 & 3.6754 & 207.5789 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ø]")` & 2.1014 & 2.5558 & 64.4651 & $<$ 0.0001 \\ 
  s(Talker) & 25.1208 & 39.0000 & 4944.0262 & $<$ 0.0001 \\ 
  s(Talker,category) & 53.0234 & 79.0000 & 1358.2399 & 0.0109 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.Ö}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("ø")`/
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7172 & 0.0109 & 65.8742 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ø]")` & 0.0141 & 0.0062 & 2.2891 & 0.0221 \\ 
  Gendermale & 0.0105 & 0.0146 & 0.7168 & 0.4736 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.4723 & 3.8446 & 75.4970 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ø]")` & 1.0001 & 1.0002 & 41.4930 & $<$ 0.0001 \\ 
  s(Talker) & 32.0273 & 39.0000 & 1485.3644 & $<$ 0.0001 \\ 
  s(Talker,category) & 44.4773 & 79.0000 & 160.7187 & 0.0545 \\ 
   \hline
\end{tabular}
\end{table}


\begin{table}[ht]
\caption{\label{tab:m.F1.ÖR}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("œ")`/
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.7590 & 0.0157 & -48.2066 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[œ]")` & -0.1173 & 0.0122 & -9.6135 & $<$ 0.0001 \\ 
  Gendermale & 0.0300 & 0.0203 & 1.4771 & 0.1397 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.9969 & 3.4959 & 37.6112 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œ]")` & 1.0000 & 1.0001 & 20.8779 & $<$ 0.0001 \\ 
  s(Talker) & 24.8639 & 39.0000 & 927.1593 & $<$ 0.0001 \\ 
  s(Talker,category) & 52.0699 & 79.0000 & 266.9870 & 0.0088 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.ÖR}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("œ")`/
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.0023 & 0.0171 & -0.1343 & 0.8931 \\ 
  category `r linguisticsdown::cond_cmpl("[œ]")` & 0.0721 & 0.0139 & 5.2018 & $<$ 0.0001 \\ 
  Gendermale & -0.0181 & 0.0218 & -0.8276 & 0.4079 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.8523 & 3.9696 & 266.9332 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œ]")` & 3.4298 & 3.8031 & 44.3554 & $<$ 0.0001 \\ 
  s(Talker) & 23.2700 & 39.0000 & 5567.4541 & $<$ 0.0001 \\ 
  s(Talker,category) & 54.9619 & 79.0000 & 1835.2586 & 0.0081 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.ÖR}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("œ")`/
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7415 & 0.0117 & 63.5705 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[œ]")` & 0.0160 & 0.0059 & 2.7304 & 0.0064 \\ 
  Gendermale & -0.0065 & 0.0158 & -0.4126 & 0.6799 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.9308 & 3.4371 & 19.7617 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œ]")` & 1.0000 & 1.0001 & 4.4916 & 0.0341 \\ 
  s(Talker) & 33.5974 & 39.0000 & 1500.4023 & $<$ 0.0001 \\ 
  s(Talker,category) & 42.3813 & 79.0000 & 109.4314 & 0.1543 \\ 
   \hline
\end{tabular}
\end{table}


\begin{table}[ht]
\caption{\label{tab:m.F1.A}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("ɑ")`/
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.6845 & 0.0176 & -38.8058 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[a]")` & 0.1755 & 0.0114 & 15.4202 & $<$ 0.0001 \\ 
  Gendermale & -0.0061 & 0.0233 & -0.2621 & 0.7933 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.9673 & 3.4347 & 41.8548 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[a]")` & 2.2977 & 2.7647 & 22.1895 & $<$ 0.0001 \\ 
  s(Talker) & 29.7089 & 39.0000 & 488.5858 & $<$ 0.0001 \\ 
  s(Talker,category) & 45.6542 & 79.0000 & 70.3904 & 0.0658 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.A}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("ɑ")`/
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.2110 & 0.0134 & -15.7262 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[a]")` & 0.2581 & 0.0110 & 23.3829 & $<$ 0.0001 \\ 
  Gendermale & -0.0598 & 0.0171 & -3.4988 & 0.0005 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.7570 & 3.9380 & 61.3738 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[a]")` & 3.3958 & 3.7743 & 56.2100 & $<$ 0.0001 \\ 
  s(Talker) & 22.7061 & 39.0000 & 1721.6117 & $<$ 0.0001 \\ 
  s(Talker,category) & 54.9854 & 79.0000 & 604.0724 & 0.0016 \\
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.A}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("ɑ")`/
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7854 & 0.0152 & 51.7589 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[a]")` & -0.0043 & 0.0101 & -0.4226 & 0.6726 \\ 
  Gendermale & 0.0569 & 0.0200 & 2.8485 & 0.0044 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 1.0000 & 1.0001 & 14.9221 & 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[a]")` & 2.0927 & 2.5609 & 1.6685 & 0.1931 \\ 
  s(Talker) & 29.0459 & 39.0000 & 743.3069 & $<$ 0.0001 \\ 
  s(Talker,category) & 47.0456 & 79.0000 & 122.1371 & 0.0246 \\
   \hline
\end{tabular}
\end{table}


\begin{table}[ht]
\caption{\label{tab:m.F1.Å}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("o")`/
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.0655 & 0.0137 & -77.5512 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɔ]")` & 0.1012 & 0.0127 & 7.9773 & $<$ 0.0001 \\ 
  Gendermale & 0.0554 & 0.0170 & 3.2524 & 0.0012 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.8116 & 3.9665 & 520.6087 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɔ]")` & 3.7159 & 3.9400 & 428.5582 & $<$ 0.0001 \\ 
  s(Talker) & 17.3746 & 39.0000 & 126.5860 & 0.0095 \\ 
  s(Talker,category) & 58.3434 & 79.0000 & 81.3096 & 0.0001 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.Å}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("o")`/
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.5081 & 0.0180 & -28.2486 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɔ]")` & 0.1549 & 0.0129 & 11.9984 & $<$ 0.0001 \\ 
  Gendermale & 0.0006 & 0.0235 & 0.0273 & 0.9782 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.9347 & 3.9940 & 1679.1649 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɔ]")` & 3.8070 & 3.9740 & 512.5442 & $<$ 0.0001 \\ 
  s(Talker) & 27.1884 & 39.0000 & 408.4524 & $<$ 0.0001 \\ 
  s(Talker,category) & 48.3598 & 79.0000 & 84.9479 & 0.0194 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.Å}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("o")`/
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.8060 & 0.0127 & 63.3878 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɔ]")` & 0.0339 & 0.0067 & 5.0600 & $<$ 0.0001 \\ 
  Gendermale & 0.0244 & 0.0171 & 1.4220 & 0.1551 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.3570 & 3.7273 & 19.6656 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɔ]")` & 2.9268 & 3.3935 & 26.7933 & $<$ 0.0001 \\ 
  s(Talker) & 33.0410 & 39.0000 & 1704.5743 & $<$ 0.0001 \\ 
  s(Talker,category) & 43.4334 & 79.0000 & 132.6541 & 0.2583 \\ 
   \hline
\end{tabular}
\end{table}


\begin{table}[ht]
\caption{\label{tab:m.F1.O}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("u")`/
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.1881 & 0.0139 & -85.1839 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʊ]")` & 0.0563 & 0.0089 & 6.3124 & $<$ 0.0001 \\ 
  Gendermale & 0.0000 & 0.0185 & 0.0017 & 0.9986 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.5083 & 3.0028 & 85.9692 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʊ]")` & 1.6891 & 2.0617 & 7.9543 & 0.0004 \\ 
  s(Talker) & 29.9150 & 39.0000 & 253.4452 & $<$ 0.0001 \\ 
  s(Talker,category) & 42.8163 & 78.0000 & 35.4706 & 0.0246 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.O}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("u")`/
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.4820 & 0.0187 & -25.8430 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʊ]")` & 0.0378 & 0.0123 & 3.0855 & 0.0020 \\ 
  Gendermale & 0.0581 & 0.0247 & 2.3574 & 0.0185 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 3.7956 & 3.9618 & 140.2768 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʊ]")` & 1.9794 & 2.4184 & 10.1612 & $<$ 0.0001 \\ 
  s(Talker) & 29.3475 & 39.0000 & 119.5479 & $<$ 0.0001 \\ 
  s(Talker,category) & 41.3514 & 78.0000 & 17.6264 & 0.0200 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.O}Model summary, coefficients for parametric and smooth terms for /`r linguisticsdown::cond_cmpl("u")`/
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7765 & 0.0125 & 62.3403 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʊ]")` & 0.0239 & 0.0058 & 4.1451 & $<$ 0.0001 \\ 
  Gendermale & 0.0169 & 0.0169 & 0.9999 & 0.3175 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint) & 2.8207 & 3.3106 & 19.9309 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʊ]")` & 2.0381 & 2.4777 & 2.5524 & 0.0505 \\ 
  s(Talker) & 34.4885 & 39.0000 & 721.2592 & $<$ 0.0001 \\ 
  s(Talker,category) & 39.3159 & 78.0000 & 45.0897 & 0.2425 \\ 
   \hline
\end{tabular}
\end{table}


```{r}
```

### Visualizations of GAMMs {#sec:SI-GAMMs-figs}
Figures \@ref(fig:SI-m-I-Y), \@ref(fig:SI-m-U-E), \@ref(fig:SI-m-ER), \@ref(fig:SI-m-OER), \@ref(fig:SI-m-OA), and \@ref(fig:SI-m-O) visualize the GAMMs fit to the long-short vowel pairs.

(ref:SI-m-I-Y) Fitted smooths of GAMM for predicting F1-F2-F3 and 95% confidence intervals for the [`r linguisticsdown::cond_cmpl("iː")`]-[`r linguisticsdown::cond_cmpl("ɪ")`]  (**left**) and [`r linguisticsdown::cond_cmpl("yː")`]-[`r linguisticsdown::cond_cmpl("ʏ")`]  (**right**) vowel pairs.

```{r SI-m-I-Y, message=FALSE, fig.width=base.width*4.5, fig.height=base.height*4.5, fig.align='center', fig.cap="(ref:SI-m-I-Y)"}
par(mfrow=c(3,2))
plot_diff(m.F1.I, view = "timepoint", comp = list(categoryO = c("[iː]", "[ɪ]")), print.summary = F, col = "#66C2A5", hide.label = T, main = "A", col.diff = "black")
plot_diff(m.F1.Y, view = "timepoint", comp = list(categoryO = c("[yː]", "[ʏ]")), print.summary = F, col = "#FC8D62", hide.label = T, main = "B", col.diff = "black")
plot_diff(m.F2.I, view = "timepoint", comp = list(categoryO = c("[iː]", "[ɪ]")), print.summary = F, col = "#66C2A5",hide.label = T, main = "C", col.diff = "black")
plot_diff(m.F2.Y, view = "timepoint", comp = list(categoryO = c("[yː]", "[ʏ]")), print.summary = F, col = "#FC8D62", hide.label = T, main = "D", col.diff = "black")
plot_diff(m.F3.I, view = "timepoint", comp = list(categoryO = c("[iː]", "[ɪ]")), print.summary = F, col = "#66C2A5", hide.label = T, main = "E", col.diff = "black")
plot_diff(m.F3.Y, view = "timepoint", comp = list(categoryO = c("[yː]", "[ʏ]")), print.summary = F, col = "#FC8D62", hide.label = T, main = "F", col.diff = "black")
```

(ref:SI-m-U-E) Fitted smooths of GAMM for predicting F1-F2-F3 and 95% confidence intervals for the [`r linguisticsdown::cond_cmpl("ʉː")`]-[`r linguisticsdown::cond_cmpl("ɵ")`]  (**left**) and [`r linguisticsdown::cond_cmpl("eː")`]-[`r linguisticsdown::cond_cmpl("ɛ")`]  (**right**) vowel pairs.

```{r SI-m-U-E, message=FALSE, fig.width=base.width*4.5, fig.height=base.height*4.5, fig.align='center', fig.cap="(ref:SI-m-U-E)"}
par(mfrow=c(3,2))
plot_diff(m.F1.U, view = "timepoint", comp = list(categoryO = c("[ʉː]", "[ɵ]")), print.summary = F, col = "#8DA0CB", hide.label = T, main = "A", col.diff = "black")
plot_diff(m.F1.E, view = "timepoint", comp = list(categoryO = c("[eː]", "[ɛ]")), print.summary = F, col = "#FB9A99", hide.label = T, main = "B", col.diff = "black")
plot_diff(m.F2.U, view = "timepoint", comp = list(categoryO = c("[ʉː]", "[ɵ]")), print.summary = F, col = "#8DA0CB",hide.label = T, main = "C", col.diff = "black")
plot_diff(m.F2.E, view = "timepoint", comp = list(categoryO = c("[eː]", "[ɛ]")), print.summary = F, col = "#FB9A99", hide.label = T, main = "D", col.diff = "black")
plot_diff(m.F3.U, view = "timepoint", comp = list(categoryO = c("[ʉː]", "[ɵ]")), print.summary = F, col = "#8DA0CB", hide.label = T, main = "E", col.diff = "black")
plot_diff(m.F3.E, view = "timepoint", comp = list(categoryO = c("[eː]", "[ɛ]")), print.summary = F, col = "#FB9A99", hide.label = T, main = "F", col.diff = "black")
```

(ref:SI-m-ER) Fitted smooths of GAMM for predicting F1-F2-F3 and 95% confidence intervals for the [`r linguisticsdown::cond_cmpl("ɛː")`]-[`r linguisticsdown::cond_cmpl("ɛ")`] (**left**) and [`r linguisticsdown::cond_cmpl("æː")`]-[`r linguisticsdown::cond_cmpl("æ")`] (**right**) vowel pairs.

```{r SI-m-ER, message=FALSE, fig.width=base.width*4.5, fig.height=base.height*4.5, fig.align='center', fig.cap="(ref:SI-m-ER)"}
par(mfrow=c(3,2))
plot_diff(m.F1.Ä, view = "timepoint", comp = list(categoryO = c("[ɛː]", "[ɛ]")), print.summary = F, col = "#E7298A", hide.label = T, main = "A", col.diff = "black")
plot_diff(m.F1.ÄR, view = "timepoint", comp = list(categoryO = c("[æː]", "[æ]")), print.summary = F, col = "#A6D854", hide.label = T, main = "B", col.diff = "black")
plot_diff(m.F2.Ä, view = "timepoint", comp = list(categoryO = c("[ɛː]", "[ɛ]")), print.summary = F, col = "#E7298A",hide.label = T, main = "C", col.diff = "black")
plot_diff(m.F2.ÄR, view = "timepoint", comp = list(categoryO = c("[æː]", "[æ]")), print.summary = F, col = "#A6D854", hide.label = T, main = "D", col.diff = "black")
plot_diff(m.F3.Ä, view = "timepoint", comp = list(categoryO = c("[ɛː]", "[ɛ]")), print.summary = F, col = "#E7298A", hide.label = T, main = "E", col.diff = "black")
plot_diff(m.F3.ÄR, view = "timepoint", comp = list(categoryO = c("[æː]", "[æ]")), print.summary = F, col = "#A6D854", hide.label = T, main = "F", col.diff = "black")
```

(ref:SI-m-OER) Fitted smooths of GAMM for predicting F1-F2-F3 and 95% confidence intervals for the [`r linguisticsdown::cond_cmpl("øː")`]-[`r linguisticsdown::cond_cmpl("ø")`] (**left**) and [`r linguisticsdown::cond_cmpl("œː")`]-[`r linguisticsdown::cond_cmpl("œ")`] (**right**) vowel pairs.

```{r SI-m-OER, message=FALSE, fig.width=base.width*4.5, fig.height=base.height*4.5, fig.align='center', fig.cap="(ref:SI-m-OER)"}
par(mfrow=c(3,2))
plot_diff(m.F1.Ö, view = "timepoint", comp = list(categoryO = c("[øː]", "[ø]")), print.summary = F, col = "#FFD92F", hide.label = T, main = "A", col.diff = "black")
plot_diff(m.F1.ÖR, view = "timepoint", comp = list(categoryO = c("[œː]", "[œ]")), print.summary = F, col = "#E5C494", hide.label = T, main = "B", col.diff = "black")
plot_diff(m.F2.Ö, view = "timepoint", comp = list(categoryO = c("[øː]", "[ø]")), print.summary = F, col = "#FFD92F",hide.label = T, main = "C", col.diff = "black")
plot_diff(m.F2.ÖR, view = "timepoint", comp = list(categoryO = c("[œː]", "[œ]")), print.summary = F, col = "#E5C494", hide.label = T, main = "D", col.diff = "black")
plot_diff(m.F3.Ö, view = "timepoint", comp = list(categoryO = c("[øː]", "[ø]")), print.summary = F, col = "#FFD92F", hide.label = T, main = "E", col.diff = "black")
plot_diff(m.F3.ÖR, view = "timepoint", comp = list(categoryO = c("[œː]", "[œ]")), print.summary = F, col = "#E5C494", hide.label = T, main = "F", col.diff = "black")
```

(ref:SI-m-OA) Fitted smooths of GAMM for predicting F1-F2-F3 and 95% confidence intervals for the [`r linguisticsdown::cond_cmpl("ɑː")`]-[`r linguisticsdown::cond_cmpl("a")`] (**left**) and [`r linguisticsdown::cond_cmpl("oː")`]-[`r linguisticsdown::cond_cmpl("œ")`] (**right**) vowel pairs.

```{r SI-m-OA, message=FALSE, fig.width=base.width*4.5, fig.height=base.height*4.5, fig.align='center', fig.cap="(ref:SI-m-OA)"}
par(mfrow=c(3,2))
plot_diff(m.F1.A, view = "timepoint", comp = list(categoryO = c("[ɑː]", "[a]")), print.summary = F, col = "#B3B3B3", hide.label = T, main = "A", col.diff = "black")
plot_diff(m.F1.Å, view = "timepoint", comp = list(categoryO = c("[oː]", "[ɔ]")), print.summary = F, col = "#A6CEE3", hide.label = T, main = "B", col.diff = "black")
plot_diff(m.F2.A, view = "timepoint", comp = list(categoryO = c("[ɑː]", "[a]")), print.summary = F, col = "#B3B3B3",hide.label = T, main = "C", col.diff = "black")
plot_diff(m.F2.Å, view = "timepoint", comp = list(categoryO = c("[oː]", "[ɔ]")), print.summary = F, col = "#A6CEE3", hide.label = T, main = "D", col.diff = "black")
plot_diff(m.F3.A, view = "timepoint", comp = list(categoryO = c("[ɑː]", "[a]")), print.summary = F, col = "#B3B3B3", hide.label = T, main = "E", col.diff = "black")
plot_diff(m.F3.Å, view = "timepoint", comp = list(categoryO = c("[oː]", "[ɔ]")), print.summary = F, col = "#A6CEE3", hide.label = T, main = "F", col.diff = "black")
```

(ref:SI-m-O) Fitted smooths of GAMM for predicting F1-F2-F3 and 95% confidence intervals for the [`r linguisticsdown::cond_cmpl("uː")`]-[`r linguisticsdown::cond_cmpl("ʊ")`] vowel pair.

```{r SI-m-O, message=FALSE, fig.width=base.width*1.5, fig.height=base.height*2, out.width = '60%', fig.align='center', fig.cap="(ref:SI-m-O)"}
par(mfrow=c(3,1))
plot_diff(m.F1.O, view = "timepoint", comp = list(categoryO = c("[uː]", "[ʊ]")), print.summary = F, col = "#B2DF8A", hide.label = T, main = "A", col.diff = "black")
plot_diff(m.F2.O, view = "timepoint", comp = list(categoryO = c("[uː]", "[ʊ]")), print.summary = F, col = "#B2DF8A", hide.label = T, main = "B", col.diff = "black")
plot_diff(m.F3.O, view = "timepoint", comp = list(categoryO = c("[uː]", "[ʊ]")), print.summary = F, col = "#B2DF8A",hide.label = T, main = "C", col.diff = "black")
```
```{r}
# Code kept for reproducibility
# rm(m.F1.I, m.F2.I, m.F3.I, m.F1.Y, m.F2.Y, m.F3.Y, m.F1.U, m.F2.U, m.F3.U, m.F1.E, m.F2.E, m.F3.E, m.F1.Ä, m.F2.Ä, m.F3.Ä, m.F1.ÄR, m.F2.ÄR, m.F3.ÄR, m.F1.Ö, m.F2.Ö, m.F3.Ö, m.F1.ÖR, m.F2.ÖR, m.F3.ÖR, m.F1.A, m.F2.A, m.F3.A)
```
