\newpage
\setcounter{page}{1}

# Supplementary information for *Persson (2024). The acoustic characteristics of Central Swedish* {-}
Both the main text and these supplementary information (SI) are derived from the same R markdown document available via OSF at  [https://osf.io/7uvj4/](https://osf.io/7uvj4/).

\setcounter{section}{1}
\setcounter{footnote}{0}
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{equation}{0}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\thefootnote}{S\arabic{footnote}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\theequation}{S\arabic{equation}}

## Required software {#sec:SI-software}

The document was compiled using \texttt{knitr} [@xie2023] in RStudio with R:

```{r}
version
```

The following R packages were used to create this document: `r cite_r("latex-stuff/packages.bib")`

```{r}
#knitr::write_bib(file = 'latex-stuff/packages.bib')
```

## Additional information on Study 1

### Neutralization of short /e/ and /`r linguisticsdown::cond_cmpl("ɛ")`/ {#sec:SI-hedd-hädd}
Figure \@ref(fig:SI-hedd) demonstrates that the short allophones of /`r linguisticsdown::cond_cmpl("e")`/ and /`r linguisticsdown::cond_cmpl("ɛ")`/ neutralize as [`r linguisticsdown::cond_cmpl("ɛ")`] in Central Swedish.

(ref:SI-hedd) The *hedd* and *hädd* words in the SwehVd vowel data in F1-F2 space. Points show recordings of the *hedd* and *hädd* words ([`r linguisticsdown::cond_cmpl("ɛ")`]) by 44 native talkers in the database, averaged across the five measurement points within each vowel segment. Word labels indicate word means across talkers. Since *hädd* and *hedd* resulted in the same allophone, all *hädd* words are excluded from the subset used in this study. This facilitates comparison of, for example, densities across vowels (see diagonal of Figure \@ref(fig:swe-vowels-all-cues)).

```{r SI-hedd, fig.width=base.width*2, fig.height=base.height*2, out.width='50%', fig.align='center', fig.pos="ht", fig.cap="(ref:SI-hedd)"}
d.SwehVd %>%
  filter(Transcribed_vowel == "targeted", Word %in% c("hedd", "hädd")) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Talker, Word, Token) %>%
  summarise(across(c(F1, F2), ~ geometric.mean(.x))) %>%
  ggplot(
    aes(
      x = F2,
      y = F1,
      color = Word)) +
  geom_point(alpha = .4) +
  stat_ellipse() +
  # Add between-talker mean
  geom_label(
    data =
      ~ .x %>%
      group_by(Word) %>%
      summarise(across(c(F1, F2), ~ geometric.mean(.x))),
    mapping = aes(
      colour = Word,
      label = Word),
    alpha = .8,
    size = 4,
    label.size = NA) +
  scale_x_reverse("F2 (in Hz)", position = "top", breaks = scales::breaks_pretty(6)) +
  scale_y_reverse("F1 (in Hz)", position = "right", breaks = scales::breaks_pretty(6)) +
  guides(color = "none") +
  theme(legend.position = "top")
```

### The effect of different measurement points {#sec:SI-measurement-points}
Figure \@ref(fig:SI-time-points) visualizes the SwehVd vowel space under four different assumptions about measurement points. The figure suggests that the use of different measurement points primarily affects vowels for which formant dynamics are particularly important, e.g., [`r linguisticsdown::cond_cmpl("eː")`] and [`r linguisticsdown::cond_cmpl("oː")`]. 

(ref:SI-time-points) The SwehVd vowel data in F1-F2 space, faceted by the time points used for measuring formants: averaging across the first three time-points (at 20, 35, and 50% into the vowel; **panel A**), averaging across the mid three time-points (at 35, 50, and 65% into the vowel; **panel B**), taking the mid time-point (at 50%; **panel C**), and averaging across the final three mid-points (at 50, 65, and 80% into the vowel; **panel D**). Points show recordings of each of the 21 Central Swedish vowels by the 44 L1 talkers in the database. Vowel labels indicate vowel means across talkers. Long vowels are boldfaced. Vowels that mismatched intended label are excluded (`r signif((d.SwehVd %>% filter(Transcribed_vowel != "targeted") %>% nrow() / d.SwehVd %>% nrow()) * 100)`% of all recordings).

```{r SI-time-points, fig.width=base.width*4.5, fig.height=base.height*4.5+.5, out.width = '100%', fig.pos="ht", fig.align='center', fig.cap="(ref:SI-time-points)"}
p.20_50 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(20,35,50)) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x)))) +
  guides(shape = "none")

p.35_65 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(35,50,65)) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x)))) +
  guides(shape = "none")

p.50 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint == 50)) +
  guides(shape = "none")

p.50_80 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(50,65,80)) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x)))) +
  guides(shape = "none")

(p.20_50 + p.35_65) / (p.50 + p.50_80) +
  plot_annotation(tag_levels = list(c("A", "B", "C", "D"))) +
  plot_layout(
    guides = "collect") & 
  theme(legend.position = "right",
        legend.justification = "right",
        legend.margin = margin(0, 0, 0, 0, unit = "pt"),
        plot.tag = element_text(face = "bold"))
```

### Mean cue values across talkers {#sec:SI-mean-cues}
The overall mean cue values for the long and short vowels are presented in Tables \@ref(tab:SI-cue-values-long) and \@ref(tab:SI-cue-values-short), respectively.

\begin{landscape}

\begin{table}

\caption{\label{tab:SI-cue-values-long}The spectral and temporal properties of the long vowels (averaged across talkers, by gender)}
\centering
\fontsize{6}{8}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}}
\toprule
Cue & Gender & `r linguisticsdown::cond_cmpl("[iː]")`& `r linguisticsdown::cond_cmpl("[yː]")`& `r linguisticsdown::cond_cmpl("[ʉː]")` & `r linguisticsdown::cond_cmpl("[eː]")` & `r linguisticsdown::cond_cmpl("[ɛː]")` & `r linguisticsdown::cond_cmpl("[æː]")` & `r linguisticsdown::cond_cmpl("[øː]")` & `r linguisticsdown::cond_cmpl("[œː]")` & `r linguisticsdown::cond_cmpl("[ɑː]")` & `r linguisticsdown::cond_cmpl("[oː]")` & `r linguisticsdown::cond_cmpl("[uː]")`\\
\midrule
Duration & F & 0.23 (SD=0.05) & 0.24 (SD=0.05) & 0.23 (SD=0.05) & 0.26 (SD=0.05) & 0.25 (SD=0.04) & 0.25 (SD=0.05) & 0.25 (SD=0.04) & 0.26 (SD=0.05) & 0.23 (SD=0.05) & 0.25 (SD=0.05) & 0.23 (SD=0.05)\\
 & M & 0.21 (SD=0.05) & 0.22 (SD=0.06) & 0.22 (SD=0.06) & 0.23 (SD=0.06) & 0.24 (SD=0.06) & 0.24 (SD=0.06) & 0.23 (SD=0.05) & 0.23 (SD=0.06) & 0.23 (SD=0.06) & 0.24 (SD=0.05) & 0.22 (SD=0.05)\\
F0 & F & 200.53 (SD=27.45) & 196.29 (SD=30.09) & 196.1 (SD=26.89) & 195.04 (SD=25.38) & 189.86 (SD=24.03) & 187.74 (SD=25.95) & 189.77 (SD=25) & 190.65 (SD=27.04) & 192.84 (SD=25.61) & 195.51 (SD=29.33) & 199.89 (SD=26.49)\\
 & M & 118.09 (SD=22.59) & 114.94 (SD=21.79) & 116.85 (SD=22.18) & 116.12 (SD=19.12) & 109.89 (SD=22.61) & 111.56 (SD=23.07) & 112.17 (SD=28.07) & 114.12 (SD=29.46) & 113.37 (SD=NA) & 122.23 (SD=26.64) & 116.84 (SD=20.36)\\
F1 & F & 374.2 (SD=40.95) & 361.09 (SD=34.22) & 391.34 (SD=30.39) & 424.37 (SD=44.27) & 791.24 (SD=101.72) & 848.7 (SD=115.25) & 638.55 (SD=76.26) & 614.75 (SD=75.62) & 649.15 (SD=83.44) & 440.1 (SD=46.87) & 396.75 (SD=37.13)\\
\addlinespace
 & M & 315.64 (SD=37.31) & 302.43 (SD=36.81) & 333.44 (SD=41.18) & 393.6 (SD=41.58) & 682.61 (SD=69.57) & 726.06 (SD=68.57) & 553.09 (SD=41.36) & 548.04 (SD=42.07) & 582.84 (SD=54.5) & 424.82 (SD=30.21) & 354.55 (SD=32.95)\\
F2 & F & 2093.73 (SD=217.16) & 2009 (SD=186.87) & 1783.62 (SD=113.99) & 2430.25 (SD=300.79) & 1794.7 (SD=129.12) & 1648.51 (SD=86.13) & 1525.93 (SD=99.29) & 1299.7 (SD=110.97) & 1050.33 (SD=105.69) & 760.92 (SD=94.01) & 789.15 (SD=108.23)\\
 & M & 1887.17 (SD=171.28) & 1843.46 (SD=155.66) & 1590.83 (SD=88.26) & 2132.97 (SD=174.9) & 1587.19 (SD=112.9) & 1463.63 (SD=114.66) & 1358.95 (SD=106.46) & 1134.96 (SD=92.03) & 893.01 (SD=88.48) & 703.9 (SD=80.61) & 762.44 (SD=76.79)\\
F3 & F & 3173.35 (SD=351.41) & 3270.07 (SD=136.69) & 2655.81 (SD=165.76) & 3077.56 (SD=168.89) & 2785.39 (SD=244.12) & 2754.6 (SD=265.82) & 2690.74 (SD=142.9) & 2743.12 (SD=140.38) & 2864.53 (SD=195.42) & 2910.92 (SD=180.43) & 2835.14 (SD=181.03)\\
 & M & 2946.54 (SD=152.43) & 2837.89 (SD=137.97) & 2364.8 (SD=131.58) & 2697.06 (SD=166.57) & 2491.89 (SD=185.15) & 2495.84 (SD=172.42) & 2370.48 (SD=172.29) & 2417.6 (SD=164.1) & 2638.12 (SD=215.24) & 2645.54 (SD=166.4) & 2567.77 (SD=151.27)\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}


```{r SI-cue-values-long}
# Keep code for reproducibility
# d.cue.values.forTable <- d.cue.values %>%
#   group_by(Quantity, Gender, category) %>%
#   mutate(
#     F0 = paste0(F0.mean, " (SD=", F0.sd, ")"),
#     F1 = paste0(F1.mean, " (SD=", F1.sd, ")"),
#     F2 = paste0(F2.mean, " (SD=", F2.sd, ")"),
#     F3 = paste0(F3.mean, " (SD=", F3.sd, ")"),
#     Duration = paste0(Duration.mean, " (SD=", Duration.sd, ")")) %>%
#   ungroup() %>%
#   select(-c(contains(c(".mean", ".sd")), Quantity)) %>%
#   arrange(factor(category, levels = levels.vowel.IPA.swe))
# 
# d.cue.values.forTable %<>%
#   pivot_longer(
#     cols = c(starts_with("F"), "Duration"),
#     names_to = "Cue",
#     values_to = "Cue.Value") %>%
#   pivot_wider(
#     names_from = "category",
#     values_from = "Cue.Value") %>%
#   relocate("Gender", .after = Cue) %>%
#   arrange(Cue) %>%
#   mutate(
#     Gender = ifelse(Gender == "female", "F", "M"),
#     Cue = ifelse(row_number() %% 2 == 0, "", Cue))
# 
# kable(d.cue.values.forTable %>%
#         select(Cue, Gender, contains(levels.vowel.IPA.swe.long)),
#       format = "latex",
#       booktabs = TRUE,
#       caption = "The spectral and temporal properties of the long vowels (averaged across talkers, by gender)") %>%
#   kable_styling(font_size = 6) %>%
#   column_spec(1:2, width = ".7cm") %>%
#   column_spec(3:13, width = "1.4cm")
```

\begin{landscape}

\begin{table}

\caption{\label{tab:SI-cue-values-short}The spectral and temporal properties of the short vowels (averaged across the female and male talkers)}
\centering
\fontsize{6}{8}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}}
\toprule
Cue & Gender & `r linguisticsdown::cond_cmpl("[ɪ]")` & `r linguisticsdown::cond_cmpl("[ʏ]")` & `r linguisticsdown::cond_cmpl("[ɵ]")` & `r linguisticsdown::cond_cmpl("[ɛ]")` & `r linguisticsdown::cond_cmpl("[æ]")` & `r linguisticsdown::cond_cmpl("[ø]")` & `r linguisticsdown::cond_cmpl("[œ]")` & `r linguisticsdown::cond_cmpl("[a]")` & `r linguisticsdown::cond_cmpl("[ɔ]")` & `r linguisticsdown::cond_cmpl("[ʊ]")`\\
\midrule
Duration & F & 0.12 (SD=0.02) & 0.12 (SD=0.02) & 0.13 (SD=0.03) & 0.12 (SD=0.02) & 0.16 (SD=0.03) & 0.13 (SD=0.03) & 0.17 (SD=0.03) & 0.13 (SD=0.03) & 0.13 (SD=0.03) & 0.13 (SD=0.03)\\
 & M & 0.11 (SD=0.02) & 0.11 (SD=0.03) & 0.13 (SD=0.02) & 0.12 (SD=0.02) & 0.14 (SD=0.03) & 0.12 (SD=0.03) & 0.14 (SD=0.03) & 0.13 (SD=0.03) & 0.12 (SD=0.03) & 0.12 (SD=0.03)\\
F0 & F & 216.61 (SD=NA) & 219.19 (SD=29.02) & 216.44 (SD=31.66) & 210.79 (SD=27.89) & 202.46 (SD=28.44) & 211.95 (SD=28.48) & 204.68 (SD=29.17) & 205.66 (SD=28.99) & 211.46 (SD=30.03) & 224.54 (SD=28.62)\\
 & M & 131.12 (SD=20.59) & 132.02 (SD=22.6) & 131.06 (SD=20.07) & 121.65 (SD=NA) & 122.6 (SD=37.4) & 128.09 (SD=33.82) & 126.12 (SD=29.65) & 117.64 (SD=NA) & 126.47 (SD=NA) & 131.03 (SD=19.12)\\
F1 & F & 380.91 (SD=38.71) & 383.73 (SD=37.34) & 422.14 (SD=37.68) & 627.89 (SD=70.76) & 801.77 (SD=97.71) & 569.1 (SD=63.26) & 533.26 (SD=55.52) & 796.35 (SD=101.31) & 500.55 (SD=53.01) & 418.77 (SD=37.08)\\
\addlinespace
 & M & 332.8 (SD=28.49) & 337.95 (SD=29.57) & 387.37 (SD=31.06) & 555.36 (SD=47.99) & 722.32 (SD=59.41) & 504.93 (SD=39.4) & 505.32 (SD=38.18) & 690.47 (SD=54.28) & 471.87 (SD=31.94) & 375.94 (SD=26.94)\\
F2 & F & 2526.8 (SD=283.27) & 2372.74 (SD=152.56) & 1322.19 (SD=112.94) & 1961.78 (SD=179.47) & 1762.92 (SD=141.77) & 1666.89 (SD=110.52) & 1399.73 (SD=151.46) & 1367.24 (SD=98.14) & 932.73 (SD=95.5) & 854.43 (SD=109.58)\\
 & M & 2257.6 (SD=149.07) & 2030.22 (SD=124.31) & 1162.53 (SD=95.09) & 1714 (SD=123.35) & 1510.17 (SD=116.33) & 1432.82 (SD=144.94) & 1190.35 (SD=82.51) & 1139.67 (SD=92.26) & 782.81 (SD=80.04) & 774.96 (SD=81.75)\\
F3 & F & 3085.16 (SD=168.28) & 2798.4 (SD=160.74) & 2781.55 (SD=188.02) & 2760.26 (SD=248.05) & 2810.87 (SD=260.86) & 2706.12 (SD=155.43) & 2781.02 (SD=171.5) & 2850.9 (SD=259.87) & 3022.98 (SD=190.97) & 2939.17 (SD=182.25)\\
 & M & 2772.45 (SD=180.21) & 2479.21 (SD=128.84) & 2459.57 (SD=148.01) & 2522.15 (SD=154.8) & 2545.13 (SD=165.44) & 2424.64 (SD=134.98) & 2470.4 (SD=165.12) & 2659.06 (SD=204.29) & 2716.22 (SD=182.89) & 2596.2 (SD=158)\\
\bottomrule
\end{tabular}
\end{table}

\end{landscape}


```{r SI-cue-values-short}
# Keep code for reproducibility
#For each cue, the highest and lowest mean and SD should be boldfaced.
# kable(d.cue.values.forTable %>%
#         select(Cue, Gender, contains(levels.vowel.IPA.swe.short)),
#       format = "latex",
#       booktabs = TRUE,
#       caption = "The spectral and temporal properties of the short vowels (averaged across the female and male talkers)") %>%
#   kable_styling(font_size = 6) %>%
#   column_spec(1:2, width = ".7cm") %>%
#   column_spec(3:12, width = "1.4cm")
```
```{r}
# itsadug::plot_smooth(m.F3.short, view = "timepoint", plot_all = "category")
# itsadug::plot_smooth(m.F3.long, view = "timepoint", plot_all = "category")
#itsadug::plot_diff(m.F3.long, view = "timepoint", comp = list(category = c("[yː]", "[iː]")))
```

### Cue matrix plot in unnormalized Hertz {#sec:SI-raw}
Figure \@ref(fig:SI-swe-vowels-all-cues-hz) displays the pairwise combinations of the five spectral and temporal cues---F0, F1, F2, F3 and duration---in raw Hertz. For F0, there are two overlapping but separated peaks, reflecting the overall difference in vocal tract size for the two groups of female and male talkers in the database.

(ref:SI-swe-vowels-all-cues-hz) The SwehVd vowels in unnormalized Hertz space. Points show recordings of each of the 21 Central Swedish vowels by the talkers in the database in all five cue spaces, averaged across the three midpoints. Vowel labels indicate vowel means across talkers. Ellipses show bivariate Gaussian 95% confidence interval of vowel means. Note that, unlike in Figure \@ref(fig:swe-vowels), axis directions are not reversed. Vowels that mismatched the intended label are excluded (`r signif((d.SwehVd %>% filter(Transcribed_vowel != "targeted") %>% nrow() / d.SwehVd %>% nrow()) * 100)`% of all recordings).

```{r SI-swe-vowels-all-cues-hz, fig.width=base.width*4+.5, fig.height=base.height*4+.5, out.width = '100%', fig.align='center', fig.pos="ht", fig.cap="(ref:SI-swe-vowels-all-cues-hz)"}
p.matrix.cues %+%
  (d.SwehVd.forStudy.long.static %>%
     filter(Normalization.Type == "no normalization (Hz)", F0 < 320))
```

### Talker differences in the [i:] - [y:] contrast
(ref:SI-swe-merge) The effect of including F3 for the [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] distinction, illustrated in a reduced vowel space. **Panel A** shows two example talkers with overlapping [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] in F1-F2 space. **Panel B** shows how considering F3 affects the space: for talker SW_028, there is very little effect of F3, thus maintained [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] overlap, while for talker SW_027, separability increases.

```{r SI-swe-merge, fig.width=base.width*3, fig.height=base.height*3, out.width='80%', fig.align='center', fig.cap="(ref:SI-swe-merge)", message=FALSE}
p.merge <- d.SwehVd.forStudy.long.static %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", Quantity == "long", Talker %in% c("SW_028", "SW_027"), category %in% c("[iː]", "[yː]", "[æː]", "[ɑː]", "[oː]", "[uː]")) %>%
  ggplot(
    aes(
      x = F2, 
      y = F1)) +
  geom_point(
    aes(
      colour = category,
      shape = Quantity), alpha = 0.5, size = 1.5) + 
  # Add within-talker mean
  geom_label(
    data =
      . %>%
      group_by(category, Quantity, Normalization.Type, Talker) %>%
      summarise(across(c(F1, F2), ~ mean(.x))),
    mapping = aes(
      colour = category,
      label = category,
      fontface = ifelse(category %in% c("[iː]", "[yː]"), 2, 1)),
    size = 4,
    alpha = .4,
    label.size = NA,
    label.padding = unit(.06, "cm")) +
  scale_colour_manual(name = "category", values = colors.vowel.swe) +
  scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 4)) +
  scale_y_reverse("F1", position = "right", scales::pretty_breaks(n = 4)) +
  guides(alpha = "none", color = "none", shape = "none", size = "none") +
  facet_wrap(~ Talker, ncol=2) 

p.merge.F3 <-  d.SwehVd.forStudy.long.static %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", Quantity == "long", Talker %in% c("SW_028", "SW_027"), category %in% c("[iː]", "[yː]", "[æː]", "[ɑː]", "[oː]", "[uː]")) %>%
  ggplot(
    aes(
      x = F2, 
      y = F3)) +
  geom_point(
    aes(
      colour = category,
      shape = Quantity), alpha = 0.5, size = 1.5) + 
  # Add within-talker mean
  geom_label(
    data =
      . %>%
      group_by(category, Quantity, Normalization.Type, Talker) %>%
      summarise(across(c(F3, F2), ~ mean(.x))),
    mapping = aes(
      colour = category,
      label = category,
      fontface = ifelse(category %in% c("[iː]", "[yː]"), 2, 1)),
    size = 4,
    alpha = .4,
    label.size = NA, 
    label.padding = unit(.06, "cm")) +
  scale_colour_manual(name = "category", values = colors.vowel.swe) +
  scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 4)) +
  scale_y_reverse("F3", position = "right", scales::pretty_breaks(n = 4)) +
  guides(alpha = "none", color = "none", shape = "none", size = "none") +
  facet_wrap(~ Talker, ncol=2)
  
p.merge / p.merge.F3 + theme(axis.title.x = element_blank()) +
  plot_annotation(tag_levels = 'A') &
  theme(legend.margin = margin(0, 0, 0, 0, unit = "pt"),
        plot.tag = element_text(face = "bold"))
#ggsave(file = "../../figures/p.png", width = 9, height = 5, dpi=1200)
```

### Summary tables of LMMs predicting category separability from cue space {#sec:SI-LMMs-study1}
#### Investigating separability in rounded vs. unrounded neighboring categories

\begin{table}
\caption{LMMs predicting category separability from cue space}
\centering
\begin{tabular}[t]{lccc}
\addlinespace[0.5em]
\toprule
\multicolumn{4}{l}{\textit{unrounded vs. outrounded}}\\
\midrule
& `r linguisticsdown::cond_cmpl("[iː]")` vs. `r linguisticsdown::cond_cmpl("[yː]")` & `r linguisticsdown::cond_cmpl("[eː]")` vs. `r linguisticsdown::cond_cmpl("[yː]")` & `r linguisticsdown::cond_cmpl("[ɪ]")` vs. `r linguisticsdown::cond_cmpl("[ʏ]")`\\
\midrule \hspace{1em}(Intercept) & 0.089 & 0.322 & 0.146\\
\hspace{1em} & se = 0.006 & se = 0.014 & se = 0.011\\
\hspace{1em} & t-value = 14.891 & t-value = 23.742 & t-value = 13.400\\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = \vphantom{3} <0.001\\
\hspace{1em} cue space F1F2F3 & 0.013 & 0.007 & 0.015\\
\hspace{1em} & se = 0.003 & se = 0.001 & se = 0.001\\
\hspace{1em} & t-value = 3.765 & t-value = 11.120 & t-value = 16.460\\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = \vphantom{2} <0.001\\
\hspace{1em}SD (Intercept Talker) & 0.033 & 0.082 & 0.066\\
\hspace{1em}SD (Observations) & 0.021 & 0.004 & 0.005\\
\addlinespace[0.5em]
\toprule
\multicolumn{4}{l}{\textit{outrounded vs. inrounded}}\\
\midrule
& `r linguisticsdown::cond_cmpl("[yː]")` vs. `r linguisticsdown::cond_cmpl("[ʉː]")` & `r linguisticsdown::cond_cmpl("[oː]")` vs. `r linguisticsdown::cond_cmpl("[uː]")` & `r linguisticsdown::cond_cmpl("[ɔ]")` vs. `r linguisticsdown::cond_cmpl("[ʊ]")`\\
\midrule \hspace{1em}(Intercept) & 0.187 & 0.195 & 0.245\\
\hspace{1em} & se = 0.014 & se = 0.012 & se = 0.013\\
\hspace{1em} & t-value = 12.916 & t-value = 16.089 & t-value = 18.506\\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = \vphantom{1} <0.001\\
\hspace{1em} cue space F1F2F3 & 0.046 & 0.004 & 0.005\\
\hspace{1em} & se = 0.002 & se = 0.000 & se = 0.001\\
\hspace{1em} & t-value = 18.642 & t-value = 16.049 & t-value = 8.546\\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001\\
\hspace{1em}SD (Intercept Talker) & 0.087 & 0.074 & 0.079\\
\hspace{1em}SD (Observations) & 0.015 & 0.002 & 0.004\\
\bottomrule
\end{tabular}
\end{table}

```{r}
#panels <- list(
#  "long vs. short" = list("A" = m.sep.I, "B" = m.sep.Y, "C" = m.sep.U, "D" = m.sep.E, "E" = m.sep.Ä, "F" = m.sep.ÄR),
#  "long vs. short" = list("G" = m.sep.Ö, "H" = m.sep.ÖR, "I" = m.sep.A, "J" = m.sep.Å, "K" = m.sep.O))
#modelsummary::modelsummary(panels, output = "latex", title = "LMM predicting category separability from cue space for all long-short contrasts", estimate = "{estimate}", statistic = c("se = {std.error}", "t-value = {statistic}", "p.value = {p.value}"), shape = "rbind", gof_map = NA)
rm(m.sep.IIYY, m.sep.EEYY, m.sep.IY, m.sep.YYUU, m.sep.ÅÅOO, m.sep.ÅO)
```

#### Investigating separability in long vs. short categories

\begin{landscape}
\begin{table}
\tiny
\caption{LMMs predicting category separability from cue space for all long-short contrasts}
\centering
\begin{tabular}[t]{lcccccc}
\toprule
  & `r linguisticsdown::cond_cmpl("[iː]")` vs. `r linguisticsdown::cond_cmpl("[ɪ]")` & `r linguisticsdown::cond_cmpl("[yː]")` vs. `r linguisticsdown::cond_cmpl("[ʏ]")` & `r linguisticsdown::cond_cmpl("[ʉː]")` vs. `r linguisticsdown::cond_cmpl("[ɵ]")` & `r linguisticsdown::cond_cmpl("[eː]")` vs. `r linguisticsdown::cond_cmpl("[ɛ]")` & `r linguisticsdown::cond_cmpl("[ɛː]")` vs. `r linguisticsdown::cond_cmpl("[ɛ]")` & `r linguisticsdown::cond_cmpl("[æː]")` vs. `r linguisticsdown::cond_cmpl("[æ]")`\\
\midrule
\addlinespace[0.5em]
\multicolumn{6}{l}{\textit{long vs. short}}\\
\midrule \hspace{1em}(Intercept) & 0.234 & 0.201 & 0.347 & 0.467 & 0.239 & 0.112 \\
\hspace{1em} & se = 0.012 & se = 0.010 & se = 0.013 & se = 0.018 & se = 0.016 & se = 0.013 \\
\hspace{1em} & t-value = 19.250 & t-value = 20.798 & t-value = 25.758 & t-value = 26.667 & t-value = 15.358 & t-value = 8.398 \\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001\\
\hspace{1em} cue space F1F3 & -0.102 & -0.007 & -0.195 & -0.055 & -0.010 & 0.002 \\
\hspace{1em} & se = 0.008 & se = 0.006 & se = 0.005 & se = 0.007 & se = 0.005 & se = 0.006 \vphantom{2} \\
\hspace{1em} & t-value = -13.259 & t-value = -1.132 & t-value = -37.105 & t-value = -7.617 & t-value = -2.011 & t-value = 0.410  \\
\hspace{1em} & p.value = <0.001 & p.value = 0.258 & p.value = <0.001 & p.value = <0.001 & p.value = 0.045 & p.value = 0.682  \\
\hspace{1em} cue space F2F3 & 0.000 & 0.023 & -0.024 & -0.202 & -0.129 & -0.014  \\
\hspace{1em} & se = 0.008 & se = 0.006 & se = 0.005 & se = 0.007 & se = 0.005 & se = 0.006 \vphantom{1}  \\
\hspace{1em} & t-value = 0.039 & t-value = 3.720 & t-value = -4.488 & t-value = -27.827 & t-value = -25.714 & t-value = -2.560 \\
\hspace{1em} & p.value = 0.969 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = 0.011  \\
\hspace{1em} cue space F1F2F3 & 0.012 & 0.025 & 0.006 & 0.015 & 0.006 & 0.010  \\
\hspace{1em} & se = 0.008 & se = 0.006 & se = 0.005 & se = 0.007 & se = 0.005 & se = 0.006 \\
\hspace{1em} & t-value = 1.619 & t-value = 4.122 & t-value = 1.062 & t-value = 2.052 & t-value = 1.114 & t-value = 1.874 \\
\hspace{1em} & p.value = 0.107 & p.value = <0.001 & p.value = 0.289 & p.value = 0.041 & p.value = 0.266 & p.value = 0.062 \\
\hspace{1em}SD (Intercept Talker) & 0.066 & 0.052 & 0.079 & 0.102 & 0.092 & 0.078  \\
\hspace{1em}SD (Observations) & 0.047 & 0.038 & 0.032 & 0.044 & 0.031 & 0.033  \\
\addlinespace[0.5em]
\toprule
  & `r linguisticsdown::cond_cmpl("[øː]")` vs. `r linguisticsdown::cond_cmpl("[ø]")` & `r linguisticsdown::cond_cmpl("[œː]")` vs. `r linguisticsdown::cond_cmpl("[œ]")` & `r linguisticsdown::cond_cmpl("[ɑː]")` vs. `r linguisticsdown::cond_cmpl("[a]")` & `r linguisticsdown::cond_cmpl("[oː]")` vs. `r linguisticsdown::cond_cmpl("[ɔ]")` & `r linguisticsdown::cond_cmpl("[uː]")` vs. `r linguisticsdown::cond_cmpl("[ʊ]")` & \\
\multicolumn{6}{l}{\textit{long vs. short }}\\
\midrule \hspace{1em}(Intercept) & 0.151 & 0.168 & 0.328 & 0.261 & 0.143 & \\
\hspace{1em} & se = 0.011 & se = 0.012 & se = 0.012 & se = 0.015 & se = 0.009 & \\
\hspace{1em} & t-value = 13.918 & t-value = 13.942 & t-value = 26.696 & t-value = 17.341 & t-value = 15.995 & \\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = \vphantom{2} <0.001 & \\
\hspace{1em} cue space F1F3 &  -0.021 & -0.029 & -0.134 & -0.097 & -0.046 & \\
\hspace{1em} & se = 0.004 & se = 0.007 & se = 0.007 & se = 0.006 & se = \vphantom{2} 0.005 & \\
\hspace{1em} & t-value = -4.841 & t-value = -4.055 & t-value = -20.347 & t-value = -15.011 & t-value = -9.606 & \\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = \vphantom{1} <0.001 & \\
\hspace{1em} cue space F2F3 & -0.051 & -0.064 & -0.046 & -0.048 & -0.019 & \\
\hspace{1em} &  se = 0.004 & se = 0.007 & se = 0.007 & se = 0.006 & se = \vphantom{1} 0.005 & \\
\hspace{1em} &  t-value = -12.042 & t-value = -9.055 & t-value = -6.968 & t-value = -7.447 & t-value = -3.907 & \\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & \\
\hspace{1em} cue space F1F2F3 &  0.003 & 0.003 & 0.008 & 0.004 & 0.004 & \\
\hspace{1em} & se = 0.004 & se = 0.007 & se = 0.007 & se = 0.006 & se = 0.005 & \\
\hspace{1em} & t-value = 0.673 & t-value = 0.401 & t-value = 1.233 & t-value = 0.639 & t-value = 0.747 & \\
\hspace{1em} & p.value = 0.501 & p.value = 0.689 & p.value = 0.219 & p.value = 0.524 & p.value = 0.456 & \\
\hspace{1em}SD (Intercept Talker) &  0.063 & 0.067 & 0.069 & 0.087 & 0.049 & \\
\hspace{1em}SD (Observations) &  0.026 & 0.043 & 0.040 & 0.039 & 0.029 & \\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

```{r}
rm(m.sep.I, m.sep.Y, m.sep.U, m.sep.E, m.sep.Ä, m.sep.ÄR, m.sep.Ö, m.sep.ÖR, m.sep.A, m.sep.Å, m.sep.O)
```

### Summary tables of GAMMs {#sec:SI-GAMMs}

\begin{table}[ht]
\caption{\label{tab:m-F1-long}Model summary, coefficients for parametric and smooth terms for the high front long vowels predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.2468 & 0.0163 & -76.3272 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[yː]")` & -0.0410 & 0.0136 & -3.0195 & 0.0025 \\ 
  category `r linguisticsdown::cond_cmpl("[ʉː]")` & 0.0460 & 0.0136 & 3.3884 & 0.0007 \\ 
  category `r linguisticsdown::cond_cmpl("[eː]")` & 0.1773 & 0.0136 & 13.0179 & $<$ 0.0001 \\ 
  Gendermale & -0.0387 & 0.0207 & -1.8640 & 0.0624 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[iː]")` & 2.9311 & 3.4400 & 57.0277 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 3.0869 & 3.5748 & 70.2809 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʉː]")` & 3.4486 & 3.8299 & 179.1962 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[eː]")` & 3.9101 & 3.9950 & 1004.9035 & $<$ 0.0001 \\ 
  s(Talker) & 27.4456 & 35.0000 & 1492.1121 & $<$ 0.0001 \\ 
  s(Talker,category) & 110.0590 & 143.0000 & 131.7453 & 0.0012 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.front.long}
\end{table}

\begin{table}[ht]
\centering
\caption{\label{tab:m.F1.front.short}Model summary, coefficients for parametric and smooth terms for the high front short vowels predicting F1} 
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.2205 & 0.0131 & -92.8715 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʏ]")` & 0.0103 & 0.0104 & 0.9899 & 0.3223 \\ 
  Gendermale & -0.0260 & 0.0178 & -1.4605 & 0.1442 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɪ]")` & 2.4343 & 2.9458 & 3.4971 & 0.0140 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʏ]")` & 1.4602 & 1.7789 & 10.4211 & 0.0001 \\ 
  s(Talker) & 22.8931 & 35.0000 & 452.9953 & $<$ 0.0001 \\ 
  s(Talker,category) & 45.4715 & 71.0000 & 115.8641 & 0.0108 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.front.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.front.long}Model summary, coefficients for parametric and smooth terms for the high front long vowels predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.4714 & 0.0158 & 29.7433 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[yː]")` & -0.0337 & 0.0169 & -1.9920 & 0.0464 \\ 
  category `r linguisticsdown::cond_cmpl("[ʉː]")` & -0.1640 & 0.0169 & -9.6862 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[eː]")` & 0.1220 & 0.0170 & 7.1905 & $<$ 0.0001 \\ 
  Gendermale & 0.0226 & 0.0177 & 1.2800 & 0.2006 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[iː]")` & 1.2063 & 1.3816 & 26.3039 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 1.8420 & 2.2681 & 26.5236 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʉː]")` & 2.7349 & 3.2517 & 179.2476 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[eː]")` & 3.8488 & 3.9863 & 660.7542 & $<$ 0.0001 \\ 
  s(Talker) & 18.8071 & 35.0000 & 1958.0198 & 0.0003 \\ 
  s(Talker,category) & 121.8467 & 143.0000 & 461.3724 & 0.0002 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.front.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.front.short}Model summary, coefficients for parametric and smooth terms for the high front short vowels predicting F2} 
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.6673 & 0.0161 & 41.4614 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʏ]")` & -0.0799 & 0.0181 & -4.4158 & $<$ 0.0001 \\ 
  Gendermale & -0.0130 & 0.0196 & -0.6607 & 0.5088 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɪ]")` & 3.3567 & 3.7755 & 9.2388 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʏ]")` & 2.5556 & 3.0757 & 11.1705 & $<$ 0.0001 \\ 
  s(Talker) & 5.0689 & 35.0000 & 71.2973 & 0.2234 \\ 
  s(Talker,category) & 64.6715 & 71.0000 & 256.0953 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.front.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.front.long}Model summary, coefficients for parametric and smooth terms for the high front long vowels predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.9027 & 0.0120 & 75.2840 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("yː")`] & 0.0007 & 0.0135 & 0.0552 & 0.9560 \\ 
  category [`r linguisticsdown::cond_cmpl("ʉː")`] & -0.1923 & 0.0135 & -14.2802 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("eː")`] & -0.0584 & 0.0135 & -4.3279 & $<$ 0.0001 \\ 
  Gendermale & 0.0082 & 0.0128 & 0.6409 & 0.5216 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("iː")`] & 2.4433 & 2.9498 & 10.5927 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("yː")`] & 2.7424 & 3.2613 & 6.8240 & 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʉː")`] & 3.1406 & 3.6172 & 15.3775 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("eː")`] & 3.6758 & 3.9392 & 100.9240 & $<$ 0.0001 \\ 
  s(Talker) & 15.5843 & 35.0000 & 691.3561 & 0.0056 \\ 
  s(Talker,category) & 124.3566 & 143.0000 & 222.1766 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F3.front.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.front.short}Model summary, coefficients for parametric and smooth terms for the high front short vowels predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.8632 & 0.0101 & 85.5852 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʏ")`] & -0.1024 & 0.0077 & -13.3324 & $<$ 0.0001 \\ 
  Gendermale & 0.0054 & 0.0138 & 0.3956 & 0.6925 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɪ")`] & 2.8288 & 3.3486 & 62.2052 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʏ")`] & 2.3569 & 2.8620 & 6.1364 & 0.0005 \\ 
  s(Talker) & 24.0220 & 35.0000 & 717.3496 & $<$ 0.0001 \\ 
  s(Talker,category) & 44.8619 & 71.0000 & 162.8064 & 0.0522 \\ 
   \hline
\end{tabular}
\label{tab:m.F3.front.short}
\end{table}

```{r}
```

\begin{table}[ht]
\caption{\label{tab:m.F1.back.long}Model summary, coefficients for parametric and smooth terms for the high back long vowels predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.0587 & 0.0141 & -75.1949 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("uː")`] & -0.1418 & 0.0128 & -11.0521 & $<$ 0.0001 \\ 
  Gendermale & 0.0425 & 0.0185 & 2.3003 & 0.0215 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("oː")`] & 3.7652 & 3.9665 & 396.9714 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("uː")`] & 2.5930 & 3.1042 & 87.5653 & $<$ 0.0001 \\ 
  s(Talker) & 18.0364 & 35.0000 & 136.5104 & 0.0006 \\ 
  s(Talker,category) & 49.4368 & 71.0000 & 67.1671 & 0.0005 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.back.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.back.long}Model summary, coefficients for parametric and smooth terms for the high back long vowels predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.5435 & 0.0187 & -29.0043 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("uː")`]  & 0.0552 & 0.0161 & 3.4218 & 0.0006 \\ 
  Gendermale & 0.0697 & 0.0250 & 2.7914 & 0.0053 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("oː")`] & 3.8644 & 3.9886 & 808.6381 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("uː")`]  & 3.7776 & 3.9699 & 117.8905 & $<$ 0.0001 \\ 
  s(Talker) & 20.2977 & 35.0000 & 106.8326 & $<$ 0.0001 \\ 
  s(Talker,category) & 45.9985 & 71.0000 & 38.2194 & 0.0023 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.back.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.back.long}Model summary, coefficients for parametric and smooth terms for the high back long vowels predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.8020 & 0.0130 & 61.6839 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("uː")`] & -0.0279 & 0.0072 & -3.8880 & 0.0001 \\ 
  Gendermale & 0.0252 & 0.0184 & 1.3686 & 0.1712 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("oː")`] & 3.0516 & 3.5419 & 15.4137 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("uː")`] & 2.7295 & 3.2436 & 16.7865 & $<$ 0.0001 \\ 
  s(Talker) & 29.6774 & 35.0000 & 848.4999 & $<$ 0.0001 \\ 
  s(Talker,category) & 38.1508 & 71.0000 & 74.8089 & 0.1462 \\ 
   \hline
\end{tabular}
\label{tab:m.F3.back.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F1.back.short}Model summary, coefficients for parametric and smooth terms for the high back short vowels predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.9472 & 0.0128 & -74.0368 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʊ")`] & -0.1975 & 0.0145 & -13.6429 & $<$ 0.0001 \\ 
  Gendermale & 0.0378 & 0.0157 & 2.4136 & 0.0159 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɔ")`] & 3.1438 & 3.6173 & 88.6017 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʊ")`] & 2.1220 & 2.5898 & 66.7215 & $<$ 0.0001 \\ 
  s(Talker) & 4.9611 & 35.0000 & 31.5664 & 0.3458 \\ 
  s(Talker,category) & 63.2273 & 70.0000 & 123.1536 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.back.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.back.short}Model summary, coefficients for parametric and smooth terms for the high back short vowels predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.3485 & 0.0176 & -19.8130 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʊ")`] & -0.0574 & 0.0139 & -4.1379 & $<$ 0.0001 \\ 
  Gendermale & -0.0121 & 0.0239 & -0.5046 & 0.6139 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɔ")`] & 3.3590 & 3.7732 & 58.3532 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʊ")`] & 3.5984 & 3.9059 & 269.8991 & $<$ 0.0001 \\ 
  s(Talker) & 23.1772 & 35.0000 & 270.7720 & $<$ 0.0001 \\ 
  s(Talker,category) & 43.3625 & 70.0000 & 68.6349 & 0.0106 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.back.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.back.short}Model summary, coefficients for parametric and smooth terms for the high back short vowels predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.8453 & 0.0125 & 67.7240 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʊ")`]  & -0.0396 & 0.0087 & -4.5752 & $<$ 0.0001 \\ 
  Gendermale & 0.0084 & 0.0173 & 0.4861 & 0.6269 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɔ")`] & 1.0000 & 1.0000 & 79.4844 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʊ")`]  & 1.8262 & 2.2471 & 11.1195 & $<$ 0.0001 \\ 
  s(Talker) & 26.2191 & 35.0000 & 2923.8268 & $<$ 0.0001 \\ 
  s(Talker,category) & 42.4906 & 70.0000 & 443.8224 & 0.1140 \\ 
   \hline
\end{tabular}
\label{tab:m.F3.back.short}
\end{table}

```{r}
```

\begin{table}[ht]
\caption{\label{tab:m.F1.center.long}Model summary, coefficients for parametric and smooth terms for the lower-mid front long vowels predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.4989 & 0.0219 & -22.8167 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æː")`] & 0.0658 & 0.0130 & 5.0547 & $<$ 0.0001 \\ 
  Gendermale & -0.0346 & 0.0308 & -1.1223 & 0.2618 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɛː")`] & 1.4851 & 1.8142 & 7.0119 & 0.0038 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æː")`] & 3.2083 & 3.6706 & 111.7106 & $<$ 0.0001 \\ 
  s(Talker) & 28.7148 & 35.0000 & 805.9948 & $<$ 0.0001 \\ 
  s(Talker,category) & 38.9026 & 71.0000 & 86.8982 & 0.0405 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.center.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.center.long}Model summary, coefficients for parametric and smooth terms for the lower-mid front long vowels predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.3202 & 0.0113 & 28.2110 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æː")`] & -0.0846 & 0.0096 & -8.7790 & $<$ 0.0001 \\ 
  Gendermale & -0.0038 & 0.0152 & -0.2518 & 0.8012 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɛː")`]  & 3.3457 & 3.7693 & 239.3935 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æː")`] & 3.3336 & 3.7592 & 44.0107 & $<$ 0.0001 \\ 
  s(Talker) & 20.7802 & 35.0000 & 1200.4655 & $<$ 0.0001 \\ 
  s(Talker,category) & 48.5959 & 71.0000 & 406.9209 & 0.0016 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.center.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.center.long}Model summary, coefficients for parametric and smooth terms for the lower-mid front long vowels
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7477 & 0.0171 & 43.8376 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æː")`] & -0.0119 & 0.0105 & -1.1309 & 0.2582 \\ 
  Gendermale & 0.0290 & 0.0239 & 1.2118 & 0.2257 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɛː")`] & 2.5579 & 3.0789 & 16.9912 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æː")`] & 1.9849 & 2.4346 & 3.4538 & 0.0236 \\ 
  s(Talker) & 28.1870 & 35.0000 & 982.4866 & $<$ 0.0001 \\ 
  s(Talker,category) & 39.8401 & 71.0000 & 101.3499 & 0.1088 \\ 
   \hline
\end{tabular} 
\label{tab:m.F3.center.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F1.center.short}Model summary, coefficients for parametric and smooth terms for the lower-mid front short vowels
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.7269 & 0.0151 & -48.0611 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æ")`] & 0.2512 & 0.0136 & 18.4943 & $<$ 0.0001 \\ 
  Gendermale & -0.0162 & 0.0199 & -0.8137 & 0.4159 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɛ")`] & 3.4131 & 3.8118 & 50.0923 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æ")`] & 3.0619 & 3.5569 & 209.1458 & $<$ 0.0001 \\ 
  s(Talker) & 18.6400 & 35.0000 & 368.7870 & 0.0008 \\ 
  s(Talker,category) & 50.2350 & 71.0000 & 167.6331 & 0.0024 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.center.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.center.short}Model summary, coefficients for parametric and smooth terms for the lower-mid front short vowels
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.4089 & 0.0127 & 32.1002 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æ")`] & -0.1181 & 0.0124 & -9.5505 & $<$ 0.0001 \\ 
  Gendermale & -0.0201 & 0.0164 & -1.2217 & 0.2219 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɛ")`] & 1.3303 & 1.5854 & 290.5422 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æ")`] & 2.4991 & 3.0151 & 57.1634 & $<$ 0.0001 \\ 
  s(Talker) & 15.0381 & 35.0000 & 352.1814 & 0.0102 \\ 
  s(Talker,category) & 54.2360 & 71.0000 & 243.0783 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.center.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.center.short}Model summary, coefficients for parametric and smooth terms for the lower-mid front short vowels
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7458 & 0.0136 & 55.0097 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æ")`] & 0.0123 & 0.0085 & 1.4409 & 0.1497 \\ 
  Gendermale & 0.0315 & 0.0190 & 1.6568 & 0.0977 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɛ")`] & 2.2157 & 2.7079 & 4.4190 & 0.0084 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æ")`] & 2.5770 & 3.0972 & 6.6918 & 0.0001 \\ 
  s(Talker) & 27.8932 & 35.0000 & 288.5886 & $<$ 0.0001 \\ 
  s(Talker,category) & 38.4020 & 71.0000 & 32.8884 & 0.0793 \\ 
   \hline
\end{tabular}
\label{tab:m.F3.center.short}
\end{table}

```{r}
#gamtabs(m.F3.mid.short, caption='Model summary, coefficients for parametric and smooth terms', label = "tab:m.F3.mid.short")
```

\begin{table}[ht]
\caption{\label{tab:m.F1.mid.long}Model summary, coefficients for parametric and smooth terms for the mid center long vowels predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.7178 & 0.0177 & -40.5263 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œː")`] & -0.0171 & 0.0111 & -1.5411 & 0.1234 \\ 
  Gendermale & -0.0175 & 0.0248 & -0.7054 & 0.4806 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("øː")`] & 2.0230 & 2.4847 & 2.0968 & 0.0968 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œː")`] & 2.6059 & 3.1188 & 56.5090 & $<$ 0.0001 \\ 
  s(Talker) & 27.9922 & 35.0000 & 1639.3603 & $<$ 0.0001 \\ 
  s(Talker,category) & 40.8306 & 71.0000 & 206.6339 & 0.0475 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.mid.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.mid.long}Model summary, coefficients for parametric and smooth terms for the mid center long vowels predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.1620 & 0.0143 & 11.3386 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œː")`] & -0.1755 & 0.0167 & -10.4903 & $<$ 0.0001 \\ 
  Gendermale & -0.0040 & 0.0171 & -0.2360 & 0.8135 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("øː")`] & 3.3494 & 3.7708 & 206.3966 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œː")`] & 3.8405 & 3.9842 & 264.4796 & $<$ 0.0001 \\ 
  s(Talker) & 1.2539 & 35.0000 & 17.5588 & 0.4091 \\ 
  s(Talker,category) & 69.0578 & 71.0000 & 335.5738 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.mid.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.mid.long}Model summary, coefficients for parametric and smooth terms for the mid center long vowels predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7238 & 0.0110 & 65.6956 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œː")`] & 0.0171 & 0.0076 & 2.2335 & 0.0256 \\ 
  Gendermale & -0.0094 & 0.0153 & -0.6174 & 0.5370 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("øː")`] & 3.3171 & 3.7495 & 53.2943 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œː")`] & 2.7352 & 3.2497 & 16.0149 & $<$ 0.0001 \\ 
  s(Talker) & 26.1642 & 35.0000 & 1328.1128 & $<$ 0.0001 \\ 
  s(Talker,category) & 42.7042 & 71.0000 & 212.7082 & 0.0380 \\ 
   \hline
\end{tabular} 
\label{tab:m.F3.mid.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F1.mid.short}Model summary, coefficients for parametric and smooth terms for the mid center short vowels
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.8401 & 0.0143 & -58.6160 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œ")`] & -0.0280 & 0.0139 & -2.0198 & 0.0435 \\ 
  Gendermale & 0.0146 & 0.0185 & 0.7916 & 0.4287 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ø")`] & 3.5157 & 3.8680 & 170.5141 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œ")`] & 2.5538 & 3.0669 & 111.0694 & $<$ 0.0001 \\ 
  s(Talker) & 15.2210 & 35.0000 & 539.6243 & 0.0133 \\ 
  s(Talker,category) & 54.4128 & 71.0000 & 357.8582 & 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.mid.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.mid.short}Model summary, coefficients for parametric and smooth terms for the mid center short vowels
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.2497 & 0.0162 & 15.4595 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œ")`] & -0.1809 & 0.0165 & -10.9577 & $<$ 0.0001 \\ 
  Gendermale & -0.0352 & 0.0205 & -1.7185 & 0.0858 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ø")`] & 2.3312 & 2.8344 & 41.2673 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œ")`] & 3.6083 & 3.9108 & 518.6392 & $<$ 0.0001 \\ 
  s(Talker) & 12.1043 & 35.0000 & 2648.4818 & 0.0306 \\ 
  s(Talker,category) & 58.3994 & 71.0000 & 2464.2257 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.mid.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.mid.short}Model summary, coefficients for parametric and smooth terms for the mid center short vowels
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7335 & 0.0109 & 67.1090 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œ")`] & 0.0192 & 0.0082 & 2.3509 & 0.0188 \\ 
  Gendermale & 0.0059 & 0.0150 & 0.3955 & 0.6925 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ø")`] & 3.2158 & 3.6784 & 128.8414 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œ")`] & 2.5288 & 3.0405 & 7.8057 & $<$ 0.0001 \\ 
  s(Talker) & 24.5124 & 35.0000 & 1752.1939 & $<$ 0.0001 \\ 
  s(Talker,category) & 45.0178 & 71.0000 & 354.4428 & 0.0421 \\ 
   \hline
\end{tabular}
\label{tab:m.F3.mid.short}
\end{table}

```{r}
```
 
\begin{table}[ht]
\small
\caption{\label{tab:m-F1-paired}Model summary, coefficients for parametric and smooth terms for all 21 vowels predicting F1} 
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.2618 & 0.0142 & -89.1177 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɪ]")` & 0.0324 & 0.0157 & 2.0593 & 0.0395 \\ 
  category `r linguisticsdown::cond_cmpl("[yː]")` & -0.0410 & 0.0157 & -2.6085 & 0.0091 \\ 
  category `r linguisticsdown::cond_cmpl("[ʏ]")` & 0.0426 & 0.0157 & 2.7044 & 0.0068 \\ 
  category `r linguisticsdown::cond_cmpl("[ʉː]")` & 0.0456 & 0.0157 & 2.8987 & 0.0037 \\ 
  category `r linguisticsdown::cond_cmpl("[ɵ]")` & 0.1600 & 0.0157 & 10.1678 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[eː]")` & 0.1794 & 0.0158 & 11.3701 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɛ]")` & 0.5294 & 0.0157 & 33.7688 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɛː]")` & 0.7494 & 0.0158 & 47.5815 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æː]")` & 0.8138 & 0.0158 & 51.6221 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æ]")` & 0.7805 & 0.0157 & 49.5770 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[øː]")` & 0.5390 & 0.0157 & 34.2376 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ø]")` & 0.4322 & 0.0157 & 27.4658 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[œː]")` & 0.5222 & 0.0158 & 33.1307 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[œ]")` & 0.4039 & 0.0157 & 25.6634 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɑː]")` & 0.5853 & 0.0157 & 37.1926 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[a]")` & 0.7541 & 0.0157 & 47.9239 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[oː]")` & 0.2189 & 0.0157 & 13.9132 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɔ]")` & 0.3373 & 0.0157 & 21.4318 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[uː]")` & 0.0861 & 0.0157 & 5.4716 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʊ]")` & 0.1406 & 0.0159 & 8.8655 & $<$ 0.0001 \\ 
  Gendermale & -0.0063 & 0.0134 & -0.4692 & 0.6389 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[iː]")` & 2.9699 & 3.4749 & 60.8984 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɪ]")` & 2.3196 & 2.8202 & 2.7657 & 0.0380 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 3.1211 & 3.6026 & 75.2795 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʏ]")` & 1.2836 & 1.5107 & 9.6941 & 0.0003 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʉː]")` & 3.4823 & 3.8488 & 192.2389 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɵ]")` & 2.5131 & 3.0207 & 32.6885 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[eː]")` & 3.9160 & 3.9957 & 1083.5453 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛ]")` & 3.6352 & 3.9234 & 92.8670 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛː]")` & 2.0567 & 2.5243 & 7.0863 & 0.0003 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æː]")` & 3.3827 & 3.7909 & 164.8723 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æ]")` & 3.0379 & 3.5369 & 199.9285 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[øː]")` & 1.8388 & 2.2686 & 1.5376 & 0.1954 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ø]")` & 3.3342 & 3.7613 & 107.8693 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œː]")` & 2.3894 & 2.8882 & 43.6960 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œ]")` & 2.2966 & 2.7874 & 75.0532 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɑː]")` & 2.6751 & 3.1893 & 69.7824 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[a]")` & 3.3890 & 3.7948 & 182.4588 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[oː]")` & 3.8321 & 3.9826 & 575.2500 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɔ]")` & 2.9739 & 3.4744 & 66.6260 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[uː]")` & 2.9624 & 3.4643 & 114.3566 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʊ]")` & 1.9404 & 2.3813 & 52.5447 & $<$ 0.0001 \\ 
  s(Talker) & 30.3648 & 35.0000 & 7129.7891 & $<$ 0.0001 \\ 
  s(Talker,category) & 700.8765 & 754.0000 & 82.5811 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.paired}
\end{table}

\begin{table}[ht]
\small
\caption{\label{tab:m-F2-paired}Model summary, coefficients for parametric and smooth terms for all 21 vowels predicting F2} 
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.4831 & 0.0135 & 35.8530 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɪ]")` & 0.1793 & 0.0170 & 10.5309 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[yː]")` & -0.0338 & 0.0170 & -1.9831 & 0.0474 \\ 
  category `r linguisticsdown::cond_cmpl("[ʏ]")` & 0.0995 & 0.0170 & 5.8417 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʉː]")` & -0.1643 & 0.0170 & -9.6510 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɵ]")` & -0.4735 & 0.0170 & -27.8011 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[eː]")` & 0.1209 & 0.0171 & 7.0847 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[ɛ]")` & -0.0824 & 0.0170 & -4.8503 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɛː]")` & -0.1641 & 0.0170 & -9.6292 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æː]")` & -0.2473 & 0.0171 & -14.5034 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æ]")` & -0.1997 & 0.0170 & -11.7242 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[øː]")` & -0.3210 & 0.0170 & -18.8458 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ø]")` & -0.2480 & 0.0170 & -14.5610 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[œː]")` & -0.4977 & 0.0171 & -29.1877 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[œ]")` & -0.4302 & 0.0170 & -25.2542 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[ɑː]")` & -0.7165 & 0.0170 & -42.0672 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[a]")` & -0.4597 & 0.0170 & -26.9935 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[oː]")` & -1.0077 & 0.0170 & -59.1781 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɔ]")` & -0.8391 & 0.0170 & -49.2613 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[uː]")` & -0.9441 & 0.0170 & -55.4250 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[ʊ]")` & -0.9004 & 0.0172 & -52.4751 & $<$ 0.0001 \\
  Gendermale & -0.0030 & 0.0097 & -0.3103 & 0.7563 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[iː]")` & 1.0015 & 1.0029 & 25.5650 & $<$ 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɪ]")` & 3.3221 & 3.7528 & 8.6397 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 1.6674 & 2.0543 & 21.7906 & $<$ 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʏ]")` & 2.5259 & 3.0445 & 10.5300 & $<$ 0.0001 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʉː]")` & 2.4654 & 2.9713 & 145.9897 & $<$ 0.0001 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɵ]")` & 3.3562 & 3.7716 & 135.6853 & $<$ 0.0001 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[eː]")` & 3.8051 & 3.9774 & 494.3527 & $<$ 0.0001 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛ]")` & 1.7680 & 2.1855 & 278.9595 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛː]")` & 2.9022 & 3.4184 & 98.6021 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æː]")` & 2.8558 & 3.3707 & 17.9822 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æ]")` & 2.2613 & 2.7546 & 42.0383 & $<$ 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[øː]")` & 2.5127 & 3.0298 & 91.1535 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ø]")` & 1.1352 & 1.2570 & 29.5323 & $<$ 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œː]")` & 3.6238 & 3.9165 & 95.8035 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œ]")` & 3.2293 & 3.6851 & 184.8442 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɑː]")` & 3.4899 & 3.8515 & 25.4819 & $<$ 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[a]")` & 1.7816 & 2.1970 & 42.6025 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[oː]")` & 3.9594 & 3.9990 & 2902.3723 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɔ]")` & 3.5890 & 3.9014 & 111.6825 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[uː]")` & 3.9363 & 3.9974 & 420.6330 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʊ]")` & 3.7607 & 3.9653 & 524.5672 & $<$ 0.0001 \\ 
  s(Talker) & 24.6716 & 35.0000 & 3703.0200 & $<$ 0.0001 \\ 
  s(Talker,category) & 711.8453 & 754.0000 & 111.1971 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.paired}
\end{table}

\begin{table}[ht]
\small
\caption{\label{tab:m-F3-paired}Model summary, coefficients for parametric and smooth terms for all 21 vowels predicting F3} 
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.8994 & 0.0116 & 77.7339 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɪ]")` & -0.0413 & 0.0119 & -3.4809 & 0.0005 \\
  category `r linguisticsdown::cond_cmpl("[yː]")` & 0.0007 & 0.0118 & 0.0576 & 0.9541 \\ 
  category `r linguisticsdown::cond_cmpl("[ʏ]")` & -0.1432 & 0.0119 & -12.0801 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʉː]")` & -0.1924 & 0.0118 & -16.2359 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɵ]")` & -0.1555 & 0.0119 & -13.1199 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[eː]")` & -0.0588 & 0.0119 & -4.9460 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[ɛ]")` & -0.1459 & 0.0118 & -12.3596 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɛː]")` & -0.1450 & 0.0119 & -12.2203 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[æː]")` & -0.1572 & 0.0119 & -13.2361 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æ]")` & -0.1340 & 0.0119 & -11.2963 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[øː]")` & -0.1875 & 0.0119 & -15.8153 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[ø]")` & -0.1709 & 0.0119 & -14.4191 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[œː]")` & -0.1698 & 0.0119 & -14.3056 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[œ]")` & -0.1511 & 0.0119 & -12.7489 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[ɑː]")` & -0.0996 & 0.0119 & -8.4030 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[a]")` & -0.1003 & 0.0119 & -8.4656 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[oː]")` & -0.0921 & 0.0119 & -7.7723 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɔ]")` & -0.0580 & 0.0119 & -4.8932 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[uː]")` & -0.1210 & 0.0119 & -10.2044 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[ʊ]")` & -0.0974 & 0.0119 & -8.1535 & $<$ 0.0001 \\
  Gendermale & 0.0155 & 0.0121 & 1.2866 & 0.1982 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[iː]")` & 2.3877 & 2.8890 & 10.2860 & $<$ 0.0001 \\   
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɪ]")`  & 2.5706 & 3.0908 & 42.4087 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 2.7213 & 3.2404 & 6.5528 & 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʏ]")` & 2.0793 & 2.5501 & 4.1808 & 0.0088 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʉː]")` & 3.1212 & 3.6017 & 14.7398 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɵ]")` & 1.0007 & 1.0013 & 124.4234 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[eː]")` & 3.6630 & 3.9345 & 96.5489 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛ]")` & 3.0119 & 3.5168 & 12.8402 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛː]")` & 2.8123 & 3.3344 & 24.2192 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æː]")` & 2.2083 & 2.6896 & 5.0740 & 0.0028 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æ]")` & 3.0922 & 3.5819 & 10.8873 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[øː]")` & 3.0677 & 3.5614 & 32.2446 & $<$ 0.0001 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ø]")` & 2.8947 & 3.4110 & 68.1325 & $<$ 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œː]")` & 2.4586 & 2.9632 & 9.8438 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œ]")` & 2.1449 & 2.6169 & 4.3514 & 0.0095 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɑː]")` & 1.0002 & 1.0003 & 17.7578 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[a]")` & 2.4811 & 2.9927 & 7.9447 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[oː]")` & 3.0203 & 3.5151 & 14.4622 & $<$ 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɔ]")` & 1.0001 & 1.0003 & 38.5056 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[uː]")` & 2.6925 & 3.2065 & 15.8756 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʊ]")` & 1.2141 & 1.3948 & 7.2875 & 0.0023 \\ 
  s(Talker) & 31.7596 & 35.0000 & 11782.6492 & $<$ 0.0001 \\ 
  s(Talker,category) & 699.8265 & 754.0000 & 88.8884 & 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F3.paired}
\end{table}

\begin{table}[ht]
\small
\caption{\label{tab:m-duration-paired}Model summary, coefficients for parametric and smooth terms for all 21 vowels predicting duration} 
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.2247 & 0.0075 & 29.7842 & $<$ 0.0001 \\ 
    category `r linguisticsdown::cond_cmpl("[ɪ]")` & -0.1035 & 0.0051 & -20.3505 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[yː]")` & 0.0147 & 0.0051 & 2.8873 & 0.0039 \\
  category `r linguisticsdown::cond_cmpl("[ʏ]")` & -0.1047 & 0.0051 & -20.5962 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʉː]")` & 0.0124 & 0.0051 & 2.4408 & 0.0147 \\  
  category `r linguisticsdown::cond_cmpl("[ɵ]")` & -0.0904 & 0.0051 & -17.7653 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[eː]")` & 0.0271 & 0.0051 & 5.3302 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[ɛ]")` & -0.0970 & 0.0051 & -19.1206 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[ɛː]")` & 0.0249 & 0.0051 & 4.8883 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[æː]")` & 0.0309 & 0.0051 & 6.0607 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æ]")` & -0.0673 & 0.0051 & -13.2361 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[øː]")` & 0.0220 & 0.0051 & 4.3213 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ø]")` & -0.0954 & 0.0051 & -18.7537 & $<$ 0.0001 \\   
  category `r linguisticsdown::cond_cmpl("[œː]")` & 0.0254 & 0.0051 & 4.9813 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[œ]")` & -0.0630 & 0.0051 & -12.3867 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɑː]")` & 0.0137 & 0.0051 & 2.6987 & 0.0070 \\  
  category `r linguisticsdown::cond_cmpl("[a]")` & -0.0876 & 0.0051 & -17.2311 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[oː]")` & 0.0278 & 0.0051 & 5.4724 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[ɔ]")` & -0.0907 & 0.0051 & -17.8292 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[uː]")` & 0.0072 & 0.0051 & 1.4083 & 0.1590 \\ 
  category `r linguisticsdown::cond_cmpl("[ʊ]")` & -0.0898 & 0.0051 & -17.5237 & $<$ 0.0001 \\ 
  Gendermale & -0.0158 & 0.0099 & -1.5985 & 0.1099 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[iː]")` & 1.0003 & 1.0005 & 0.4070 & 0.5235 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɪ]")`  & 1.0001 & 1.0002 & 0.0001 & 0.9953 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 1.0002 & 1.0003 & 0.3187 & 0.5725 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʏ]")` & 1.0001 & 1.0002 & 0.0123 & 0.9121 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʉː]")` & 1.0001 & 1.0002 & 2.3733 & 0.1234 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɵ]")` & 1.0001 & 1.0003 & 0.0859 & 0.7696 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[eː]")` & 1.0002 & 1.0004 & 0.0092 & 0.9244 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛ]")` & 1.0002 & 1.0005 & 0.5153 & 0.4730 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛː]")` & 1.0002 & 1.0004 & 1.0377 & 0.3084 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æː]")` & 1.0001 & 1.0003 & 0.1511 & 0.6976 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æ]")` & 1.0002 & 1.0003 & 0.0611 & 0.8052 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[øː]")` & 1.0001 & 1.0002 & 0.1833 & 0.6686 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ø]")` & 1.0002 & 1.0004 & 1.1941 & 0.2746 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œː]")` & 1.0001 & 1.0003 & 0.2322 & 0.6300 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œ]")` & 1.0001 & 1.0002 & 0.1012 & 0.7504 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɑː]")` & 1.0002 & 1.0004 & 0.4794 & 0.4888 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[a]")` & 1.0001 & 1.0002 & 0.1999 & 0.6549 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[oː]")` & 1.0001 & 1.0002 & 0.0503 & 0.8228 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɔ]")` & 1.0001 & 1.0002 & 0.0442 & 0.8337 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[uː]")` & 1.0004 & 1.0008 & 0.8059 & 0.3692 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʊ]")` & 1.0001 & 1.0002 & 0.0110 & 0.9169 \\ 
  s(Talker) & 34.1047 & 35.0000 & 127803.3103 & $<$ 0.0001 \\ 
  s(Talker,category) & 705.9825 & 754.0000 & 147.2948 & 0.0251 \\ 
   \hline
\end{tabular} 
\label{tab:m.duration.paired}
\end{table}

```{r}
```

#### Visualizations of GAMMs {#sec:SI-GAMMs-figs}

(ref:SI-m-F1-paired) Fitted smooths of GAMM for predicting F1 and 95% confidence intervals for all vowel pairs. For visualization purposes, the pairs are split into high vowels (**left panel**), and lower-mid vowels (**right panel**).

```{r SI-m-F1-paired, message=FALSE, fig.width=base.width*4, fig.height=base.height*1.5, fig.align='center', fig.pos="ht", fig.cap="(ref:SI-m-F1-paired)"}
par(mfrow=c(1,2))
itsadug::plot_smooth(m.F1.paired, view = "timepoint", plot_all = "category", cond = list(category = c("[iː]", "[ɪ]", "[yː]", "[ʏ]", "[ʉː]", "[ɵ]", "[eː]", "[ɛ]", "[oː]", "[ɔ]", "[uː]", "[ʊ]")), col = c("#66C2A5", "#1B9E77", "#FC8D62", "#D95F02", "#8DA0CB", "#7570B3", "#FB9A99", "#E78AC3", "#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C"), print.summary = FALSE, hide.label = T)
itsadug::plot_smooth(m.F1.paired, view = "timepoint", plot_all = "category", cond = list(category = c("[ɛː]", "[æː]", "[æ]", "[øː]", "[ø]", "[œː]", "[œ]", "[ɑː]", "[a]")), col = c("#E7298A", "#A6D854", "#66A61E", "#FFD92F", "#E6AB02", "#E5C494", "#A6761D", "#B3B3B3", "#666666"), print.summary = FALSE, hide.label = T)
```

(ref:SI-m-F2-paired) Fitted smooths of GAMM for predicting F2 and 95% confidence intervals for all vowel pairs. For visualization purposes, the pairs are split into high vowels (**left panel**), and lower-mid vowels (**right panel**).

```{r SI-m-F2-paired, message=FALSE, fig.width=base.width*4, fig.height=base.height*1.5, fig.align='center', fig.pos="ht", fig.cap="(ref:SI-m-F2-paired)"}
par(mfrow=c(1,2))
itsadug::plot_smooth(m.F2.paired, view = "timepoint", plot_all = "category", cond = list(category = c("[iː]", "[ɪ]", "[yː]", "[ʏ]", "[ʉː]", "[ɵ]", "[eː]", "[ɛ]", "[oː]", "[ɔ]", "[uː]", "[ʊ]")), col = c("#66C2A5", "#1B9E77", "#FC8D62", "#D95F02", "#8DA0CB", "#7570B3", "#FB9A99", "#E78AC3", "#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C"), print.summary = FALSE, hide.label = T)
itsadug::plot_smooth(m.F2.paired, view = "timepoint", plot_all = "category", cond = list(category = c("[ɛː]", "[æː]", "[æ]", "[øː]", "[ø]", "[œː]", "[œ]", "[ɑː]", "[a]")), col = c("#E7298A", "#A6D854", "#66A61E", "#FFD92F", "#E6AB02", "#E5C494", "#A6761D", "#B3B3B3", "#666666"), print.summary = FALSE, hide.label = T)
```

(ref:SI-m-F3-paired) Fitted smooths of GAMM for predicting F3 and 95% confidence intervals for all vowel pairs. For visualization purposes, the pairs are split into high vowels (**left panel**), and lower-mid vowels (**right panel**).

```{r SI-m-F3-paired, message=FALSE, fig.width=base.width*4, fig.height=base.height*1.5, fig.align='center', fig.pos="ht", fig.cap="(ref:m-F3-paired)"}
par(mfrow=c(1,2))
itsadug::plot_smooth(m.F3.paired, view = "timepoint", plot_all = "category", cond = list(category = c("[iː]", "[ɪ]", "[yː]", "[ʏ]", "[ʉː]", "[ɵ]", "[eː]", "[ɛ]", "[oː]", "[ɔ]", "[uː]", "[ʊ]")), col = c("#66C2A5", "#1B9E77", "#FC8D62", "#D95F02", "#8DA0CB", "#7570B3", "#FB9A99", "#E78AC3", "#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C"), print.summary = FALSE, hide.label = T)
itsadug::plot_smooth(m.F3.paired, view = "timepoint", plot_all = "category", cond = list(category = c("[ɛː]", "[æː]", "[æ]", "[øː]", "[ø]", "[œː]", "[œ]", "[ɑː]", "[a]")), col = c("#E7298A", "#A6D854", "#66A61E", "#FFD92F", "#E6AB02", "#E5C494", "#A6761D", "#B3B3B3", "#666666"), print.summary = FALSE, hide.label = T)
```
```{r}
# Code kept for reproducibility
#gamtabs(m.duration.paired, caption='Model summary, coefficients for parametric and smooth terms', label = "tab:m.duration.paired")
rm(m.F1.paired, m.F2.paired, m.F3.paired, m.duration.paired)
```

## Additional information on Study 2 {#sec:SI-SweDia}

### Wordlist used for recording of SweDia {#sec:SI-SweDia-wordlist}

\begin{table}
\caption{\label{tab:word-list}Words recorded by talkers of Central Swedish for the SweDia database}
\centering
\begin{tabular}[t]{c|c|c}
\hline
Vowel IPA&Target words&Comment\\
\hline
`r linguisticsdown::cond_cmpl("[æː]")`&lär&\\
\hline
`r linguisticsdown::cond_cmpl("[øː]")`&blöt&\\
\hline
&söt&\\
\hline
&lös&\\
\hline
`r linguisticsdown::cond_cmpl("[ɛː]")`&läs&missing for ym4 and yw4\\
\hline
&nät&\\
\hline
&väg&recorded by ym4 only\\
\hline
`r linguisticsdown::cond_cmpl("[uː]")`&sot&\\
\hline
`r linguisticsdown::cond_cmpl("[ʉː]")`&lus&\\
\hline
`r linguisticsdown::cond_cmpl("[yː]")`&typ&\\
\hline
`r linguisticsdown::cond_cmpl("[oː]")`&lås&\\
\hline
&låt&recorded by ym4 only\\
\hline
`r linguisticsdown::cond_cmpl("[eː]")`&leta&\\
\hline
&gles&recorded by ym4 and yw4 only\\
\hline
`r linguisticsdown::cond_cmpl("[œː]")`&dör&\\
\hline
`r linguisticsdown::cond_cmpl("[iː]")`&dis&\\
\hline
`r linguisticsdown::cond_cmpl("[ɑː]")`&lat&\\
\hline
&tak&recorded by ym4 and yw4 only\\
\hline
`r linguisticsdown::cond_cmpl("[ɪ]")`&sitt&\\
\hline
&disk&recorded by ym4 and yw4 only\\
\hline
&fisk&\\
\hline
`r linguisticsdown::cond_cmpl("[ɔ]")`&lott&\\
\hline
&såll&\\
\hline
`r linguisticsdown::cond_cmpl("[ʏ]")`&flytta&\\
\hline
&tysk&\\
\hline
&sytt&\\
\hline
`r linguisticsdown::cond_cmpl("[a]")`&lass&\\
\hline
&tack&recorded by ym4 only\\
\hline
`r linguisticsdown::cond_cmpl("[ɛ]")`&lätt&\\
\hline
&lett&\\
\hline
&vägg&recorded by ym4 only\\
\hline
`r linguisticsdown::cond_cmpl("[æ]")`&särk&recorded by ym4 and yw4 only\\
\hline
`r linguisticsdown::cond_cmpl("[ɵ]")`&ludd&\\
\hline
`r linguisticsdown::cond_cmpl("[œ]")`&dörr&\\
\hline
`r linguisticsdown::cond_cmpl("[ø]")`&blött&\\
\hline
`r linguisticsdown::cond_cmpl("[ʊ]")`&snott&\\
\hline
\end{tabular}
\end{table}

### Additional visualizations of the SweDia materials {#sec:SI-raw-Swedia}
(ref:SI-raw-Swedia) The SweDia reference talkers' vowel data along F1-F2, in unnormalized Hertz (*left*) and Nearey's uniform scaling space (*right*). Points show the productions of each of the 21 Central Swedish vowels by the 8 (4 female) L1 reference talkers in the database. Vowel labels indicate vowel means across talkers. Vowels that mismatched intended label are excluded (`r signif((d.SwehVd %>% filter(Transcribed_vowel != "targeted") %>% nrow() / d.SwehVd %>% nrow()) * 100)`% of all recordings).

```{r SI-raw-Swedia, fig.width=base.width*4.5, fig.height=base.height*3.5, out.width='90%', fig.align='center', fig.pos="ht", fig.cap="(ref:SI-raw-Swedia)"}
p %+%
  (d.SweDia2000.forStudy.long %>%
     filter(Normalization.Type %in% c("no normalization (Hz)", "Nearey's uniform scaling (log)"), F1 < 1000, timepoint %in% c(35, 50, 65)) %>%
     # Get the geometric mean F1-F2 across the five time points
     group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
     summarise(across(c(F1, F2, F3, Duration), ~ mean(.x))))
```

(ref:SI-Swedia-normalized) The SweDia reference talkers' vowel data in F1-F2 Nearey's uniform scaling space, separated by quantity. Ellipses show bivariate Gaussian 95% confidence interval of vowel means. Vowel labels indicate vowel means across female and male talkers.

```{r SI-Swedia-normalized, fig.width=base.width*4.5, fig.height=base.height*2.5, fig.align='center', fig.pos="ht", fig.cap="(ref:SI-Swedia-normalized)"}
p.normalized %+%
  (d.SweDia2000.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(35, 50, 65)) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F1, F2, F3, Duration), ~ mean(.x))))
```

(ref:SI-swe-vowels-all-cues-nearey-swedia) The SweDia vowels in Nearey's uniform scaling space. Points show recordings of each of the 21 Central Swedish vowels by the talkers in the database in all five cue spaces, averaged across the three midpoints. Vowel labels indicate vowel means across talkers. Ellipses show bivariate Gaussian 95% confidence interval of vowel means. Note that, unlike in Figure \@ref(fig:swe-vowels), axis directions are not reversed. Vowels that mismatched the intended label are excluded (`r signif((d.SwehVd %>% filter(Transcribed_vowel != "targeted") %>% nrow() / d.SwehVd %>% nrow()) * 100)`% of all recordings).

```{r SI-swe-vowels-all-cues-nearey-swedia, fig.width=base.width*4+.5, fig.height=base.height*4+.5, out.width = '100%', fig.align='center', fig.pos="ht", fig.cap="(ref:SI-swe-vowels-all-cues-nearey-swedia)"}
set.seed(63720472)

p.matrix.cues %+%
  (d.SweDia2000.forStudy.long %>%
     filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(35, 50, 65)) %>%
     # Get the geometric mean F1-F2 across the five time points
     group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
     summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x))) %>%
     # For plotting reasons, randomly select 3 tokens of each category
     group_by(Talker, category) %>%
     slice_sample(n = 3))
```

(ref:trajectories-swedia) The trajectory of all vowels in the SweDia reference materials across the five time points, along F1-F2. The arrow indicates the direction of the trajectory and ends at 80% into the vowel. The vowel label is placed at the third time-point, at vowel mid-point (50%).

```{r trajectories-swedia, fig.width=base.width*3.5, fig.height=base.height*2, out.width='90%', fig.align='center', fig.pos="ht", fig.cap="(ref:trajectories-swedia)"}
d.SweDia2000.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)") %>%
  mutate(timepoint = as.character(timepoint)) %>%
  group_by(Talker, category, timepoint, Quantity) %>%
  summarise(across(c(F2, F1), ~mean(.x))) %>%
  group_by(category, timepoint, Quantity) %>%
  summarise(across(c(F2, F1), ~mean(.x))) %>%
  ggplot(
    aes(
      x = F2,
      y = F1,
      color = category,
      group = NULL)) +
  geom_path(
    alpha = .5,
    arrow = arrow(ends = "last", type = "closed", angle = "22.5", length = unit(.1, "cm"))) +
  geom_label(
    data = . %>%
      filter(timepoint == "50"),
    mapping = aes(
      colour = category,
      label = category,
      fontface = ifelse(category %in% levels.vowel.IPA.swe.long, 2, 1)),
    alpha = .6,
    size = 4,
    label.size = NA,
    label.padding = unit(.01, "cm")) +
  # geom_point(
  #   data = d.F2d %>%
  #     #get the mean F1 F2 at 10% into d-closure across talkers
  #     summarise(
  #       F1 = mean(F1_closure),
  #       F2 = mean(F2_closure)),
  #   color = "black",
  #   alpha = .6) +
  scale_color_manual(name = "category", values = colors.vowel.swe) +
  scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 6)) +
  scale_y_reverse("F1", position = "right", scales::pretty_breaks(n = 6)) +
  guides(color = "none") +
  facet_wrap( ~ Quantity, labeller = label_wrap_gen(multi_line=FALSE))
```

### Visualization of the orthogonal projection ratio
(ref:SI-op) Illustration of orthogonal projection ratio. Examples of target categories $V_{vowel}$ (in cyan), and their respective anchor categories, $U_{anchor}$ and $W_{anchor}$ (in red), from the SwehVd database. The $op$ measures the relative distance of $V_{vowel}$ to its anchors, as represented by the lines.

```{r SI-op, fig.width=base.width*2, fig.height=base.height*2, out.width='50%', fig.align='center', fig.pos='ht', fig.cap="(ref:SI-op)"}
d.SwehVd.forStudy.long.static %>%
  # Filter down to test folds
  filter(Normalization.Type == "Nearey's uniform scaling (log)", category %in% c("[ʉː]", "[oː]", "[ɪ]", "[eː]", "[æː]", "[ɑː]", "[ɛː]", "[øː]")) %>%
  mutate(anchor = ifelse(category %in% c("[ʉː]", "[ɛː]", "[øː]"), "vowel", "anchor"),
         op_pair = case_when(category %in% c("[ʉː]", "[oː]", "[ɪ]") ~ "A",
                             category %in% c("[eː]", "[ɑː]", "[øː]") ~ "B",
                             category %in% c("[eː]", "[æː]", "[ɛː]") ~ "C")) %>%
  ggplot(
    aes(
      x = F2, 
      y = F1)) +
  stat_ellipse(
    mapping = aes(
      colour = anchor,
      fill = anchor,
      group = interaction(anchor, category)),
    geom = "polygon",
    alpha = .4) +
  geom_path(data =
              . %>%
              filter(category %in% c("[ɪ]", "[ʉː]", "[oː]")) %>%
              mutate(
                category = factor(category, levels = c("[ɪ]", "[ʉː]", "[oː]")),
                group = "A") %>%
              group_by(category, group, Normalization.Type) %>%
              summarise(across(c(F1, F2), ~ mean(.x))),
            linetype = "dotted") +
  geom_path(data =
              . %>%
              filter(category %in% c("[eː]", "[øː]", "[ɑː]")) %>%
              mutate(
                category = factor(category, levels = c("[eː]", "[øː]", "[ɑː]")),
                group = "B") %>%
              group_by(category, group, Normalization.Type) %>%
              summarise(across(c(F1, F2), ~ mean(.x))),
            linetype = "dashed") +
  geom_path(data =
              . %>%
              filter(category %in% c("[eː]", "[ɛː]", "[æː]")) %>%
              mutate(
                category = factor(category, levels = c("[eː]", "[ɛː]", "[æː]")),
                group = "C") %>%
              group_by(category, group, Normalization.Type) %>%
              summarise(across(c(F1, F2), ~ mean(.x)))) +
  # Add between-talker mean
  geom_label(
    data =
      . %>%  
      mutate(
        anchor = ifelse(category %in% c("[ʉː]", "[ɛː]", "[øː]"), "vowel", "anchor")) %>%
      group_by(category, anchor, Normalization.Type) %>%
      summarise(across(c(F1, F2), ~ mean(.x))),
    mapping = aes(
      colour = anchor,
      label = category),
    alpha = .8,
    size = 5,
    label.size = NA,
    label.padding = unit(.06, "cm")) +
  scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 4)) +
  scale_y_reverse("F1", position = "right", scales::pretty_breaks(n = 4)) +
  guides(alpha = "none", color = "none", shape = "none", fill = "none")
#ggsave(file = "../../figures/p.png", width = 9, height = 5, dpi=1200)
```

<!--### Sampling-with-replacement approach to op
The two datasets were of unequal sample sizes: N = `r nrow(d.SwehVd %>% distinct(Talker, category, Token))` datapoints for SwehVd, and N = `r nrow(d.SweDia2000.forStudy %>% distinct(Talker, category, Token))` datapoints for SweDia. In order to address this imbalance, the SwehVd materials were down-sampled to approximate the size of the SweDia for the calculation of the $op$. The sampling-with-replacement procedure included 1000 bootstrapping rounds. At each round, a sample of 840 tokens from the SwehVd was extracted by first randomly selecting 8 talkers (4 female) from the database, and then randomly selecting 5 tokens per vowel from each talker, with replacement. The $op$ was calculated at each round, and finally averaged across the 1000 rounds for a comparison against the $op$ calculated for the SweDia materials.-->

```{r}
# Set seed for reproducibility
set.seed(15497221)

# Number of bootstrap rounds
n_bootstrap_rounds <- 1000

# Number of talkers to select in each round
n_talkers_per_gender <- 4

# Number of tokens to select in each round per category
n_tokens_per_category <- 5

# Create a function to perform downsampling and store values
perform_bootstrap_round <- function(data) {
  # Randomly select 4 female and 4 male talkers
  selected_talkers <- data %>%
    group_by(Gender) %>%
    slice_sample(n = n_talkers_per_gender, replace = TRUE) %>%
    ungroup()

  # Randomly select 5 tokens per category
  selected_tokens <- selected_talkers %>%
    group_by(category) %>%
    slice_sample(n = n_tokens_per_category, replace = TRUE) %>%
    ungroup()

 # Extract and store the F1 and F2 values, category, and bootstrap round
  values <- selected_tokens %>%
    select(Talker, category, Word, F1, F2) %>%
    mutate(bootstrap_round = cur_group_id()) %>%
    as.data.frame()

  return(values)
}

d.SwehVd.forStudy.opr <- d.SwehVd.forStudy.long.static %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)")

# Perform downsampling and bootstrap
bootstrap_results <- map(
  1:n_bootstrap_rounds,
  ~ perform_bootstrap_round(d.SwehVd.forStudy.opr)
)

# Create a dataframe with bootstrap results in long format
bootstrap_SwehVd <- bind_rows(bootstrap_results, .id = "bootstrap_round")
```
```{r}
get_centroids <- . %>%
  group_by(Talker, category) %>% 
  # compute mean F1 and F2
  summarise(cenF1 = mean(F1),
            cenF2 = mean(F2)) %>%
  # rearrange in columns
  pivot_wider(names_from = category, 
              values_from = c(cenF1, cenF2)) %>%
  ungroup() %>%
  left_join(bootstrap_SwehVd, by = "Talker") %>%
  group_by(category) %>%
  # for each row i.e. for each observation
  rowwise() %>%
  # calculate orth.proj.ratio
  mutate(orth_proj = case_when(category == "[ʏ]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɑː]`, `cenF2_[ɑː]`), c(`cenF1_[ɪ]`, `cenF2_[ɪ]`)),
                               category == "[ʉː]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɪ]`, `cenF2_[ɪ]`), c(`cenF1_[oː]`, `cenF2_[oː]`)),
                               category == "[ɛː]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[eː]`, `cenF2_[eː]`), c(`cenF1_[æː]`, `cenF2_[æː]`)),
                               category == "[øː]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[eː]`, `cenF2_[eː]`), c(`cenF1_[ɑː]`, `cenF2_[ɑː]`)),
                               category == "[œ]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɵ]`, `cenF2_[ɵ]`), c(`cenF1_[œː]`, `cenF2_[œː]`)),
                               category == "[ɔ]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɑː]`, `cenF2_[ɑː]`), c(`cenF1_[oː]`, `cenF2_[oː]`)),
                               TRUE ~ NA))

d.SwehVd.opr <-
  bootstrap_SwehVd %>% 
  # choose the anchor vowels
  filter(category %in% c("[ɪ]", "[ʏ]", "[yː]", "[ɵ]", "[eː]", "[ɛː]", "[æː]", "[œː]", "[œ]", "[ɑː]", "[oː]", "[øː]", "[ʉː]", "[ɔ]")) %>% 
  group_by(bootstrap_round) %>%
  get_centroids() %>%  
  # Then summarise across all bootstrap rounds
  group_by(category, F1, F2, Talker) %>%
  summarise(orth_proj = mean(orth_proj, na.rm = T))
```

### Summary tables for LMMs fit to orthogonal projection ratio {#sec:SI-LMMs}

\begin{table}

\caption{LMM predicting orthogonal projection ratio from database}
\centering
\begin{tabular}[t]{lcccc}
\toprule
  & `r linguisticsdown::cond_cmpl("[ʏ]")` & `r linguisticsdown::cond_cmpl("[ʉː]")` & `r linguisticsdown::cond_cmpl("[ɛː]")` & `r linguisticsdown::cond_cmpl("[øː]")`\\
\addlinespace[0.5em]
\multicolumn{5}{l}{\textit{}}\\
\midrule \hspace{1em}(Intercept) & 0.670 & -0.555 & 0.440 & 0.017\\
\hspace{1em} & se = 0.068 & se = 0.056 & se = 0.078 & se = 0.057\\
\hspace{1em} & t-value = 9.881 & t-value = -9.896 & t-value = 5.651 & t-value = 0.294\\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = 0.769\\
\hspace{1em} SwehVd & 0.189 & 0.115 & 0.283 & 0.204\\
\hspace{1em} & se = 0.075 & se = 0.062 & se = 0.086 & se = 0.063\\
\hspace{1em} & t-value = 2.524 & t-value = 1.856 & t-value = 3.307 & t-value = 3.212\\
\hspace{1em} & p.value = 0.012 & p.value = 0.064 & p.value = 0.001 & p.value = 0.001\\
\hspace{1em}SD (Intercept Talker) & 0.189 & 0.156 & 0.212 & 0.160\\
\hspace{1em}SD (Observations) & 0.091 & 0.046 & 0.139 & 0.071\\
\addlinespace[0.5em]
\multicolumn{5}{l}{\textit{}}\\
\toprule
  & `r linguisticsdown::cond_cmpl("[ø]")` & `r linguisticsdown::cond_cmpl("[œː]")` & `r linguisticsdown::cond_cmpl("[ɔ]")` & \\
\midrule \hspace{1em}(Intercept) & -0.276 & -0.470 & -0.376 & \\
\hspace{1em} & se = 0.072 & se = 0.163 & se = 0.100 & \\
\hspace{1em} & t-value = -3.830 & t-value = -2.889 & t-value = -3.773 & \\
\hspace{1em} & p.value = <0.001 & p.value = 0.004 & p.value = <0.001 & \\
\hspace{1em} SwehVd & 0.249 & 0.756 & 0.457 & \\
\hspace{1em} & se = 0.079 & se = 0.178 & se = 0.110 & \\
\hspace{1em} & t-value = 3.163 & t-value = 4.249 & t-value = 4.161 & \\
\hspace{1em} & p.value = 0.002 & p.value = <0.001 & p.value = <0.001 & \\
\hspace{1em}SD (Intercept Talker) & 0.189 & 0.426 & 0.271 & \\
\hspace{1em}SD (Observations) & 0.132 & 0.304 & 0.228 & \\
\bottomrule
\end{tabular}
\end{table}


```{r}
#panels <- list(
# "first line" = list("A" = m.Y, "B" = m.U, "C" = m.Ä, "D" = m.ÖÖ),
# "second line" = list("A" = m.Ö, "B" = m.ÖR, "C" = m.Å))
#modelsummary::modelsummary(panels, output = "latex", title = "LMM predicting orthogonal projection ratio from database", estimate = "{estimate}", statistic = c("se = {std.error}", "t-value = {statistic}", "p.value = {p.value}"), shape = "rbind", gof_map = NA)
rm(m.Y, m.Ä, m.ÖR, m.U, m.ÖÖ, m.Ö, m.Å)
```
```{r subset-data-for-opr}
#Alternative downsampling approach
# set.seed(15497221)
# 
# d.SwehVd.forStudy.opr <- d.SwehVd.forStudy.long %>%
#   filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint == 50)
# 
# d.SweDia2000.forStudy.opr <- d.SweDia2000.forStudy.long %>%
#   filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint == 50)
# 
# selected_talkers <- d.SwehVd.forStudy.opr %>%
#   select(Gender, Talker) %>%
#   distinct() %>%
#   group_by(Gender) %>%
#   slice_sample(n = 4, replace = F) %>%
#   pull(Talker)
# 
# d.SwehVd.forStudy.opr %<>%
#   filter(Talker %in% selected_talkers) %>%
#   group_by(category, Talker) %>%
#   slice_sample(n = 5, replace = F)
# 
# d.SwehVd.SweDia2000.opr <- rbind(d.SwehVd.forStudy.opr, d.SweDia2000.forStudy.opr)
# 
# # Instead sample 36 tokens per category from the entire database of all talkers
# d.SwehVd.SweDia2000.opr_by_category <- d.SwehVd.forStudy.long %>%
#   filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint == 50) %>%
#   group_by(category, Gender) %>%
#   slice_sample(n = 18, replace = F) %>%
#   rbind(d.SweDia2000.forStudy.opr)
# 

# d.SwehVd.SweDia2000.opr.noBootstrap <- rbind(d.SwehVd.forStudy.long %>%
#                                                filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint == 50), 
#                                              d.SweDia2000.forStudy.long %>%
#                                                filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint == 50))
# 
# centroids.noBootstrap <- d.SwehVd.SweDia2000.opr.noBootstrap %>%
#   # choose the anchor vowels
#   filter(category %in% c("[ɪ]", "[ʏ]", "[yː]", "[ɵ]", "[eː]", "[ɛː]", "[æː]", "[œː]", "[œ]", "[ɑː]", "[a]", "[oː]", "[øː]", "[ø]", "[ʉː]", "[ɔ]")) %>%
#   # for each speaker and vowel
#   group_by(Talker, category) %>%
#   # compute mean F1 and F2
#   summarise(cenF1 = mean(F1),
#             cenF2 = mean(F2)) %>%
#   # rearrange in columns
#   pivot_wider(names_from = category,
#               values_from = c(cenF1, cenF2)) %>%
#   ungroup()
# 
# d.SwehVd.SweDia2000.opr.noBootstrap %>%
#   # add the centroids %>%
#   left_join(centroids.noBootstrap, by = "Talker")
# 
# d.orth.subset.noBootstrap <- d.SwehVd.SweDia2000.opr.noBootstrap %>%
#   group_by(category, Database) %>%
#  # for each row i.e. for each observation
#   rowwise() %>%
#   # calculate orth.proj.ratio
#   mutate(orth_proj = case_when(category == "[ʏ]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɑː]`, `cenF2_[ɑː]`), c(`cenF1_[ɪ]`, `cenF2_[ɪ]`)),
#                                category == "[ʉː]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɪ]`, `cenF2_[ɪ]`), c(`cenF1_[oː]`, `cenF2_[oː]`)),
#                                category == "[ɛː]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[eː]`, `cenF2_[eː]`), c(`cenF1_[æː]`, `cenF2_[æː]`)),
#                                category == "[øː]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[eː]`, `cenF2_[eː]`), c(`cenF1_[ɑː]`, `cenF2_[ɑː]`)),
#                                category == "[ø]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[yː]`, `cenF2_[yː]`), c(`cenF1_[a]`, `cenF2_[a]`)),
#                                category == "[œ]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɵ]`, `cenF2_[ɵ]`), c(`cenF1_[œː]`, `cenF2_[œː]`)),
#                                category == "[ɔ]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɑː]`, `cenF2_[ɑː]`), c(`cenF1_[oː]`, `cenF2_[oː]`)),
#                                TRUE ~ NA)) %>%
#   ungroup() %>%
#   filter(category %in% c("[ʏ]", "[ɛː]", "[œ]", "[ʉː]", "[øː]", "[ø]", "[ɔ]"))
# 
# p.orth %+%
#   (d.orth.subset.noBootstrap %>%
#      group_by(Database, category) %>%
#      mutate(
#        # category = factor(category, levels = c("[ʏ]", "[ɛː]", "[œ]", "[ɔ]", "[ʊ]", "[ɵ]"), labels =  c("[ʏ] [ɑː]-[ɪ]", "[ɛː] [eː]-[æː]", "[œ] [ɵ]-[œː]", "[ɔ] [ɑː]-[oː]", "[ʊ] [œː]-[oː]", "[ɵ] [ɛ]-[uː]")),
#        #median = median(orth_proj),
#        ci = list(enframe(Hmisc::smean.cl.boot(orth_proj)))) %>%
#      unnest(ci) %>%
#      spread(name, value) %>%
#      group_by(Database, category) %>%
#      summarise(
#        mean_CI_lower = mean(Lower),
#        mean_CI_upper = mean(Upper),
#        mean = mean(Mean)))
```
```{r}
# d.orth.subset.noBootstrap %>%
#   ggplot +
#   aes(y = orth_proj, x = Database) +
#   geom_boxplot() +
#   facet_wrap(~category, scales = "free")
```

```{r}
# Add code to write csv.file for harrington
# write.csv(
#   d.Swedish,
#   file = "../../../data/phonetic vowel statistics/Swedish/Published databases/Swedish_vowels_Persson2023_Kuronen2000.csv",
#   row.names = FALSE, quote = FALSE)
```
```{r}
# d.Swedish <- d.Kuronen2000 %>%
#   rbind(
#     d.SwehVd %>%
#       # get only mid-point for comparability to kuronen
#       filter(timepoint == 50, Transcribed_vowel == "targeted") %>%
#       select(c(Talker, Age, Gender, category, Quantity, Word, Token, F1, F2)) %>%
#       mutate(
#         Database = "Persson2021"),
#     d.SweDia2000 %>%
#       # get only mid-point for comparability to kuronen
#       filter(timepoint == 50) %>%
#       select(c(Database, Talker, Age, Gender, category, Quantity, Word, Token, F1, F2))) %>%
#     select(c(Database, Talker, Age, Gender, category, Quantity, Word, Token, F1, F2))
# 
# p.db.comparison <- 
#   d.Swedish %>%
#   filter(Gender == "male", Database != "SweDia") %>%
#   ggplot(
#     aes(
#       x = F2,
#       y = F1, 
#       color = category,
#   #shape = Gender,
#   linetype = Database)) +
#   stat_ellipse() +
#   # Add between-talker mean
#   geom_label(
#     data =
#       ~ .x %>%
#       group_by(category, Quantity, Database) %>%
#       summarise(across(c(F1, F2), ~ mean(.x))),
#     mapping = aes(
#       colour = category,
#       label = category),
#       #fontface = ifelse(Database == "SweDia", 2, 1)),
#     alpha = .8,
#     size = 4,
#     label.size = NA,
#     label.padding = unit(.06, "cm")) +
#   #geom_point(alpha = .4) +
#   scale_color_manual(name = "category", values = colors.vowel.swe) +
#   scale_x_reverse() +
#   scale_y_reverse() +
#   guides(color = "none") +
#   facet_grid(Database ~ Quantity)
# p.db.comparison
```

```{r}
# p.vowel.movements %+%
#   (d.Swedish %>%
#      filter(Gender == "male", Database != "SweDia") %>%
#      group_by(Database, Talker, category, Quantity) %>%
#      summarise(across(c(F2, F1), ~mean(.x))) %>%
#      group_by(Database, category, Quantity) %>%
#      summarise(across(c(F2, F1), ~mean(.x)))) +
#   geom_label(
#     data =
#       ~ .x %>%
#       filter(Database == "Kuronen2000") %>%
#       group_by(category, Quantity, Database) %>%
#       summarise(across(c(F1, F2), ~ mean(.x))),
#     mapping = aes(
#       colour = category,
#       label = category),
#     fontface = 2,
#     alpha = .4,
#     size = 4,
#     label.size = NA,
#     label.padding = unit(.06, "cm"))
```