\newpage
\setcounter{page}{1}

# Supplementary information for *Persson (2024). The acoustic characteristics of Central Swedish* {-}
Both the main text and these supplementary information (SI) are derived from the same R markdown document available via OSF at  [https://osf.io/7uvj4/](https://osf.io/7uvj4/).

\setcounter{section}{1}
\setcounter{footnote}{0}
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{equation}{0}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\thefootnote}{S\arabic{footnote}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\theequation}{S\arabic{equation}}

## Required software {#sec:SI-software}

The document was compiled using \texttt{knitr} [@xie2023] in RStudio with R:

```{r}
version
```

The following R packages were used to create this document: `r cite_r("latex-stuff/packages.bib")`

```{r}
#knitr::write_bib(file = 'latex-stuff/packages.bib')
```

## Additional information on Study 1

### Neutralization of short /e/ and /`r linguisticsdown::cond_cmpl("ɛ")`/ {#sec:SI-hedd-hädd}
Figure \@ref(fig:SI-hedd) demonstrates that the short allophones of /`r linguisticsdown::cond_cmpl("e")`/ and /`r linguisticsdown::cond_cmpl("ɛ")`/ neutralize as [`r linguisticsdown::cond_cmpl("ɛ")`] in Central Swedish.

(ref:SI-hedd) The *hedd* and *hädd* words in the SwehVd vowel data in F1-F2 space. Points show recordings of the *hedd* and *hädd* words ([`r linguisticsdown::cond_cmpl("ɛ")`]) by 44 native talkers in the database, averaged across the five measurement points within each vowel segment. Word labels indicate word means across talkers. Since *hädd* and *hedd* resulted in the same allophone, all *hädd* words are excluded from the subset used in this study. This facilitates comparison of, for example, densities across vowels (see diagonal of Figure \@ref(fig:swe-vowels-all-cues)).

```{r SI-hedd, fig.width=base.width*2, fig.height=base.height*2, out.width='50%', fig.align='center', fig.pos="ht", fig.cap="(ref:SI-hedd)"}
d.SwehVd %>%
  filter(Transcribed_vowel == "targeted", Word %in% c("hedd", "hädd")) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Talker, Word, Token) %>%
  summarise(across(c(F1, F2), ~ geometric.mean(.x))) %>%
  ggplot(
    aes(
      x = F2,
      y = F1,
      color = Word)) +
  geom_point(alpha = .4) +
  stat_ellipse() +
  # Add between-talker mean
  geom_label(
    data =
      ~ .x %>%
      group_by(Word) %>%
      summarise(across(c(F1, F2), ~ geometric.mean(.x))),
    mapping = aes(
      colour = Word,
      label = Word),
    alpha = .8,
    size = 4,
    label.size = NA) +
  scale_x_reverse("F2 (in Hz)", position = "top", breaks = scales::breaks_pretty(6)) +
  scale_y_reverse("F1 (in Hz)", position = "right", breaks = scales::breaks_pretty(6)) +
  guides(color = "none") +
  theme(legend.position = "top")
```

### The effect of different measurement points {#sec:SI-measurement-points}
Figure \@ref(fig:SI-time-points) visualizes the SwehVd vowel space under four different assumptions about measurement points. The figure suggests that the use of different measurement points primarily affects vowels for which formant dynamics are particularly important, e.g., [`r linguisticsdown::cond_cmpl("eː")`] and [`r linguisticsdown::cond_cmpl("oː")`]. 

(ref:SI-time-points) The SwehVd vowel data in F1-F2 space, faceted by the time points used for measuring formants: averaging across the first three time-points (at 20, 35, and 50% into the vowel; **panel A**), averaging across the mid three time-points (at 35, 50, and 65% into the vowel; **panel B**), taking the mid time-point (at 50%; **panel C**), and averaging across the final three mid-points (at 50, 65, and 80% into the vowel; **panel D**). Points show recordings of each of the 21 Central Swedish vowels by the 44 L1 talkers in the database. Vowel labels indicate vowel means across talkers. Long vowels are boldfaced. Vowels that mismatched intended label are excluded (`r signif((d.SwehVd %>% filter(Transcribed_vowel != "targeted") %>% nrow() / d.SwehVd %>% nrow()) * 100)`% of all recordings).

```{r SI-time-points, fig.width=base.width*4.5, fig.height=base.height*4.5+.5, out.width = '100%', fig.pos="ht", fig.align='center', fig.cap="(ref:SI-time-points)"}
p.20_50 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(20,35,50)) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x)))) +
  guides(shape = "none")

p.35_65 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(35,50,65)) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x)))) +
  guides(shape = "none")

p.50 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint == 50)) +
  guides(shape = "none")

p.50_80 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(50,65,80)) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x)))) +
  guides(shape = "none")

(p.20_50 + p.35_65) / (p.50 + p.50_80) +
  plot_annotation(tag_levels = list(c("A", "B", "C", "D"))) +
  plot_layout(
    guides = "collect") & 
  theme(legend.position = "right",
        legend.justification = "right",
        legend.margin = margin(0, 0, 0, 0, unit = "pt"),
        plot.tag = element_text(face = "bold"))
```

### Mean cue values across talkers {#sec:SI-mean-cues}
The overall mean cue values for the long and short vowels are presented in Tables \@ref(tab:SI-cue-values-long) and \@ref(tab:SI-cue-values-short), respectively.

\begin{landscape}

\begin{table}

\caption{\label{tab:SI-cue-values-long}The spectral and temporal properties of the long vowels (averaged across talkers, by gender)}
\centering
\fontsize{6}{8}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}}
\toprule
Cue & Gender & `r linguisticsdown::cond_cmpl("[iː]")`& `r linguisticsdown::cond_cmpl("[yː]")`& `r linguisticsdown::cond_cmpl("[ʉː]")` & `r linguisticsdown::cond_cmpl("[eː]")` & `r linguisticsdown::cond_cmpl("[ɛː]")` & `r linguisticsdown::cond_cmpl("[æː]")` & `r linguisticsdown::cond_cmpl("[øː]")` & `r linguisticsdown::cond_cmpl("[œː]")` & `r linguisticsdown::cond_cmpl("[ɑː]")` & `r linguisticsdown::cond_cmpl("[oː]")` & `r linguisticsdown::cond_cmpl("[uː]")`\\
\midrule
Duration & F & 0.23 (SD=0.05) & 0.24 (SD=0.05) & 0.23 (SD=0.05) & 0.26 (SD=0.05) & 0.25 (SD=0.04) & 0.25 (SD=0.05) & 0.25 (SD=0.04) & 0.26 (SD=0.05) & 0.23 (SD=0.05) & 0.25 (SD=0.05) & 0.23 (SD=0.05)\\
 & M & 0.21 (SD=0.05) & 0.22 (SD=0.06) & 0.22 (SD=0.06) & 0.23 (SD=0.06) & 0.24 (SD=0.06) & 0.24 (SD=0.06) & 0.23 (SD=0.05) & 0.23 (SD=0.06) & 0.23 (SD=0.06) & 0.24 (SD=0.05) & 0.22 (SD=0.05)\\
F0 & F & 200.53 (SD=27.45) & 196.29 (SD=30.09) & 196.1 (SD=26.89) & 195.04 (SD=25.38) & 189.86 (SD=24.03) & 187.74 (SD=25.95) & 189.77 (SD=25) & 190.65 (SD=27.04) & 192.84 (SD=25.61) & 195.51 (SD=29.33) & 199.89 (SD=26.49)\\
 & M & 118.09 (SD=22.59) & 114.94 (SD=21.79) & 116.85 (SD=22.18) & 116.12 (SD=19.12) & 109.89 (SD=22.61) & 111.56 (SD=23.07) & 112.17 (SD=28.07) & 114.12 (SD=29.46) & 113.37 (SD=NA) & 122.23 (SD=26.64) & 116.84 (SD=20.36)\\
F1 & F & 374.2 (SD=40.95) & 361.09 (SD=34.22) & 391.34 (SD=30.39) & 424.37 (SD=44.27) & 791.24 (SD=101.72) & 848.7 (SD=115.25) & 638.55 (SD=76.26) & 614.75 (SD=75.62) & 649.15 (SD=83.44) & 440.1 (SD=46.87) & 396.75 (SD=37.13)\\
\addlinespace
 & M & 315.64 (SD=37.31) & 302.43 (SD=36.81) & 333.44 (SD=41.18) & 393.6 (SD=41.58) & 682.61 (SD=69.57) & 726.06 (SD=68.57) & 553.09 (SD=41.36) & 548.04 (SD=42.07) & 582.84 (SD=54.5) & 424.82 (SD=30.21) & 354.55 (SD=32.95)\\
F2 & F & 2093.73 (SD=217.16) & 2009 (SD=186.87) & 1783.62 (SD=113.99) & 2430.25 (SD=300.79) & 1794.7 (SD=129.12) & 1648.51 (SD=86.13) & 1525.93 (SD=99.29) & 1299.7 (SD=110.97) & 1050.33 (SD=105.69) & 760.92 (SD=94.01) & 789.15 (SD=108.23)\\
 & M & 1887.17 (SD=171.28) & 1843.46 (SD=155.66) & 1590.83 (SD=88.26) & 2132.97 (SD=174.9) & 1587.19 (SD=112.9) & 1463.63 (SD=114.66) & 1358.95 (SD=106.46) & 1134.96 (SD=92.03) & 893.01 (SD=88.48) & 703.9 (SD=80.61) & 762.44 (SD=76.79)\\
F3 & F & 3173.35 (SD=351.41) & 3270.07 (SD=136.69) & 2655.81 (SD=165.76) & 3077.56 (SD=168.89) & 2785.39 (SD=244.12) & 2754.6 (SD=265.82) & 2690.74 (SD=142.9) & 2743.12 (SD=140.38) & 2864.53 (SD=195.42) & 2910.92 (SD=180.43) & 2835.14 (SD=181.03)\\
 & M & 2946.54 (SD=152.43) & 2837.89 (SD=137.97) & 2364.8 (SD=131.58) & 2697.06 (SD=166.57) & 2491.89 (SD=185.15) & 2495.84 (SD=172.42) & 2370.48 (SD=172.29) & 2417.6 (SD=164.1) & 2638.12 (SD=215.24) & 2645.54 (SD=166.4) & 2567.77 (SD=151.27)\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}


```{r SI-cue-values-long}
# Keep code for reproducibility
# d.cue.values.forTable <- d.cue.values %>%
#   group_by(Quantity, Gender, category) %>%
#   mutate(
#     F0 = paste0(F0.mean, " (SD=", F0.sd, ")"),
#     F1 = paste0(F1.mean, " (SD=", F1.sd, ")"),
#     F2 = paste0(F2.mean, " (SD=", F2.sd, ")"),
#     F3 = paste0(F3.mean, " (SD=", F3.sd, ")"),
#     Duration = paste0(Duration.mean, " (SD=", Duration.sd, ")")) %>%
#   ungroup() %>%
#   select(-c(contains(c(".mean", ".sd")), Quantity)) %>%
#   arrange(factor(category, levels = levels.vowel.IPA.swe))
# 
# d.cue.values.forTable %<>%
#   pivot_longer(
#     cols = c(starts_with("F"), "Duration"),
#     names_to = "Cue",
#     values_to = "Cue.Value") %>%
#   pivot_wider(
#     names_from = "category",
#     values_from = "Cue.Value") %>%
#   relocate("Gender", .after = Cue) %>%
#   arrange(Cue) %>%
#   mutate(
#     Gender = ifelse(Gender == "female", "F", "M"),
#     Cue = ifelse(row_number() %% 2 == 0, "", Cue))
# 
# kable(d.cue.values.forTable %>%
#         select(Cue, Gender, contains(levels.vowel.IPA.swe.long)),
#       format = "latex",
#       booktabs = TRUE,
#       caption = "The spectral and temporal properties of the long vowels (averaged across talkers, by gender)") %>%
#   kable_styling(font_size = 6) %>%
#   column_spec(1:2, width = ".7cm") %>%
#   column_spec(3:13, width = "1.4cm")
```

\begin{landscape}

\begin{table}

\caption{\label{tab:SI-cue-values-short}The spectral and temporal properties of the short vowels (averaged across the female and male talkers)}
\centering
\fontsize{6}{8}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}}
\toprule
Cue & Gender & `r linguisticsdown::cond_cmpl("[ɪ]")` & `r linguisticsdown::cond_cmpl("[ʏ]")` & `r linguisticsdown::cond_cmpl("[ɵ]")` & `r linguisticsdown::cond_cmpl("[ɛ]")` & `r linguisticsdown::cond_cmpl("[æ]")` & `r linguisticsdown::cond_cmpl("[ø]")` & `r linguisticsdown::cond_cmpl("[œ]")` & `r linguisticsdown::cond_cmpl("[a]")` & `r linguisticsdown::cond_cmpl("[ɔ]")` & `r linguisticsdown::cond_cmpl("[ʊ]")`\\
\midrule
Duration & F & 0.12 (SD=0.02) & 0.12 (SD=0.02) & 0.13 (SD=0.03) & 0.12 (SD=0.02) & 0.16 (SD=0.03) & 0.13 (SD=0.03) & 0.17 (SD=0.03) & 0.13 (SD=0.03) & 0.13 (SD=0.03) & 0.13 (SD=0.03)\\
 & M & 0.11 (SD=0.02) & 0.11 (SD=0.03) & 0.13 (SD=0.02) & 0.12 (SD=0.02) & 0.14 (SD=0.03) & 0.12 (SD=0.03) & 0.14 (SD=0.03) & 0.13 (SD=0.03) & 0.12 (SD=0.03) & 0.12 (SD=0.03)\\
F0 & F & 216.61 (SD=NA) & 219.19 (SD=29.02) & 216.44 (SD=31.66) & 210.79 (SD=27.89) & 202.46 (SD=28.44) & 211.95 (SD=28.48) & 204.68 (SD=29.17) & 205.66 (SD=28.99) & 211.46 (SD=30.03) & 224.54 (SD=28.62)\\
 & M & 131.12 (SD=20.59) & 132.02 (SD=22.6) & 131.06 (SD=20.07) & 121.65 (SD=NA) & 122.6 (SD=37.4) & 128.09 (SD=33.82) & 126.12 (SD=29.65) & 117.64 (SD=NA) & 126.47 (SD=NA) & 131.03 (SD=19.12)\\
F1 & F & 380.91 (SD=38.71) & 383.73 (SD=37.34) & 422.14 (SD=37.68) & 627.89 (SD=70.76) & 801.77 (SD=97.71) & 569.1 (SD=63.26) & 533.26 (SD=55.52) & 796.35 (SD=101.31) & 500.55 (SD=53.01) & 418.77 (SD=37.08)\\
\addlinespace
 & M & 332.8 (SD=28.49) & 337.95 (SD=29.57) & 387.37 (SD=31.06) & 555.36 (SD=47.99) & 722.32 (SD=59.41) & 504.93 (SD=39.4) & 505.32 (SD=38.18) & 690.47 (SD=54.28) & 471.87 (SD=31.94) & 375.94 (SD=26.94)\\
F2 & F & 2526.8 (SD=283.27) & 2372.74 (SD=152.56) & 1322.19 (SD=112.94) & 1961.78 (SD=179.47) & 1762.92 (SD=141.77) & 1666.89 (SD=110.52) & 1399.73 (SD=151.46) & 1367.24 (SD=98.14) & 932.73 (SD=95.5) & 854.43 (SD=109.58)\\
 & M & 2257.6 (SD=149.07) & 2030.22 (SD=124.31) & 1162.53 (SD=95.09) & 1714 (SD=123.35) & 1510.17 (SD=116.33) & 1432.82 (SD=144.94) & 1190.35 (SD=82.51) & 1139.67 (SD=92.26) & 782.81 (SD=80.04) & 774.96 (SD=81.75)\\
F3 & F & 3085.16 (SD=168.28) & 2798.4 (SD=160.74) & 2781.55 (SD=188.02) & 2760.26 (SD=248.05) & 2810.87 (SD=260.86) & 2706.12 (SD=155.43) & 2781.02 (SD=171.5) & 2850.9 (SD=259.87) & 3022.98 (SD=190.97) & 2939.17 (SD=182.25)\\
 & M & 2772.45 (SD=180.21) & 2479.21 (SD=128.84) & 2459.57 (SD=148.01) & 2522.15 (SD=154.8) & 2545.13 (SD=165.44) & 2424.64 (SD=134.98) & 2470.4 (SD=165.12) & 2659.06 (SD=204.29) & 2716.22 (SD=182.89) & 2596.2 (SD=158)\\
\bottomrule
\end{tabular}
\end{table}

\end{landscape}


```{r SI-cue-values-short}
# Keep code for reproducibility
#For each cue, the highest and lowest mean and SD should be boldfaced.
# kable(d.cue.values.forTable %>%
#         select(Cue, Gender, contains(levels.vowel.IPA.swe.short)),
#       format = "latex",
#       booktabs = TRUE,
#       caption = "The spectral and temporal properties of the short vowels (averaged across the female and male talkers)") %>%
#   kable_styling(font_size = 6) %>%
#   column_spec(1:2, width = ".7cm") %>%
#   column_spec(3:12, width = "1.4cm")
```
```{r}
# itsadug::plot_smooth(m.F3.short, view = "timepoint", plot_all = "category")
# itsadug::plot_smooth(m.F3.long, view = "timepoint", plot_all = "category")
#itsadug::plot_diff(m.F3.long, view = "timepoint", comp = list(category = c("[yː]", "[iː]")))
```

### Cue matrix plot in unnormalized Hertz {#sec:SI-raw}
Figure \@ref(fig:SI-swe-vowels-all-cues-hz) displays the pairwise combinations of the five spectral and temporal cues---F0, F1, F2, F3 and duration---in raw Hertz. For F0, there are two overlapping but separated peaks, reflecting the overall difference in vocal tract size for the two groups of female and male talkers in the database.

(ref:SI-swe-vowels-all-cues-hz) The SwehVd vowels in unnormalized Hertz space. Points show recordings of each of the 21 Central Swedish vowels by the talkers in the database in all five cue spaces, averaged across the three midpoints. Vowel labels indicate vowel means across talkers. Ellipses show bivariate Gaussian 95% confidence interval of vowel means. Note that, unlike in Figure \@ref(fig:swe-vowels), axis directions are not reversed. Vowels that mismatched the intended label are excluded (`r signif((d.SwehVd %>% filter(Transcribed_vowel != "targeted") %>% nrow() / d.SwehVd %>% nrow()) * 100)`% of all recordings).

```{r SI-swe-vowels-all-cues-hz, fig.width=base.width*4+.5, fig.height=base.height*4+.5, out.width = '100%', fig.align='center', fig.pos="ht", fig.cap="(ref:SI-swe-vowels-all-cues-hz)"}
p.matrix.cues %+%
  (d.SwehVd.forStudy.long.static %>%
     filter(Normalization.Type == "no normalization (Hz)", F0 < 320))
```

### Talker differences in the [i:] - [y:] contrast
Figure \@ref(fig:SI-swe-merge) visualizes two example talkers that differ in their phonetic realization of [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] when F3 is considered.

(ref:SI-swe-merge) The effect of including F3 for the [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] distinction, illustrated in a reduced vowel space. **Panel A** shows two example talkers with overlapping [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] in F1-F2 space. **Panel B** shows how considering F3 affects the space: for talker SW_028, there is very little effect of F3, thus maintained [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] overlap, while for talker SW_027, separability increases.

```{r SI-swe-merge, fig.width=base.width*3, fig.height=base.height*3, out.width='80%', fig.align='center', fig.cap="(ref:SI-swe-merge)", message=FALSE}
p.merge <- d.SwehVd.forStudy.long.static %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", Quantity == "long", Talker %in% c("SW_028", "SW_027"), category %in% c("[iː]", "[yː]", "[æː]", "[ɑː]", "[oː]", "[uː]")) %>%
  ggplot(
    aes(
      x = F2, 
      y = F1)) +
  geom_point(
    aes(
      colour = category,
      shape = Quantity), alpha = 0.5, size = 1.5) + 
  # Add within-talker mean
  geom_label(
    data =
      . %>%
      group_by(category, Quantity, Normalization.Type, Talker) %>%
      summarise(across(c(F1, F2), ~ mean(.x))),
    mapping = aes(
      colour = category,
      label = category,
      fontface = ifelse(category %in% c("[iː]", "[yː]"), 2, 1)),
    size = 4,
    alpha = .4,
    label.size = NA,
    label.padding = unit(.06, "cm")) +
  scale_colour_manual(name = "category", values = colors.vowel.swe) +
  scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 4)) +
  scale_y_reverse("F1", position = "right", scales::pretty_breaks(n = 4)) +
  guides(alpha = "none", color = "none", shape = "none", size = "none") +
  facet_wrap(~ Talker, ncol=2) 

p.merge.F3 <-  d.SwehVd.forStudy.long.static %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", Quantity == "long", Talker %in% c("SW_028", "SW_027"), category %in% c("[iː]", "[yː]", "[æː]", "[ɑː]", "[oː]", "[uː]")) %>%
  ggplot(
    aes(
      x = F2, 
      y = F3)) +
  geom_point(
    aes(
      colour = category,
      shape = Quantity), alpha = 0.5, size = 1.5) + 
  # Add within-talker mean
  geom_label(
    data =
      . %>%
      group_by(category, Quantity, Normalization.Type, Talker) %>%
      summarise(across(c(F3, F2), ~ mean(.x))),
    mapping = aes(
      colour = category,
      label = category,
      fontface = ifelse(category %in% c("[iː]", "[yː]"), 2, 1)),
    size = 4,
    alpha = .4,
    label.size = NA, 
    label.padding = unit(.06, "cm")) +
  scale_colour_manual(name = "category", values = colors.vowel.swe) +
  scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 4)) +
  scale_y_reverse("F3", position = "right", scales::pretty_breaks(n = 4)) +
  guides(alpha = "none", color = "none", shape = "none", size = "none") +
  facet_wrap(~ Talker, ncol=2)
  
p.merge / p.merge.F3 + theme(axis.title.x = element_blank()) +
  plot_annotation(tag_levels = 'A') &
  theme(legend.margin = margin(0, 0, 0, 0, unit = "pt"),
        plot.tag = element_text(face = "bold"))
#ggsave(file = "../../figures/p.png", width = 9, height = 5, dpi=1200)
```

### Summary tables of LMMs predicting category separability from cue space {#sec:SI-LMMs-study1}
#### Investigating separability in rounded vs. unrounded neighboring categories

\begin{table}
\caption{LMMs predicting category separability from cue space}
\centering
\begin{tabular}[t]{lccc}
\addlinespace[0.5em]
\toprule
\multicolumn{4}{l}{\textit{unrounded vs. outrounded}}\\
\midrule
& `r linguisticsdown::cond_cmpl("[iː]")` vs. `r linguisticsdown::cond_cmpl("[yː]")` & `r linguisticsdown::cond_cmpl("[eː]")` vs. `r linguisticsdown::cond_cmpl("[yː]")` & `r linguisticsdown::cond_cmpl("[ɪ]")` vs. `r linguisticsdown::cond_cmpl("[ʏ]")`\\
\midrule \hspace{1em}(Intercept) & 0.089 & 0.322 & 0.146\\
\hspace{1em} & se = 0.006 & se = 0.014 & se = 0.011\\
\hspace{1em} & t-value = 14.891 & t-value = 23.742 & t-value = 13.400\\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = \vphantom{3} <0.001\\
\hspace{1em} cue space F1F2F3 & 0.013 & 0.007 & 0.015\\
\hspace{1em} & se = 0.003 & se = 0.001 & se = 0.001\\
\hspace{1em} & t-value = 3.765 & t-value = 11.120 & t-value = 16.460\\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = \vphantom{2} <0.001\\
\hspace{1em}SD (Intercept Talker) & 0.033 & 0.082 & 0.066\\
\hspace{1em}SD (Observations) & 0.021 & 0.004 & 0.005\\
\addlinespace[0.5em]
\toprule
\multicolumn{4}{l}{\textit{outrounded vs. inrounded}}\\
\midrule
& `r linguisticsdown::cond_cmpl("[yː]")` vs. `r linguisticsdown::cond_cmpl("[ʉː]")` & `r linguisticsdown::cond_cmpl("[oː]")` vs. `r linguisticsdown::cond_cmpl("[uː]")` & `r linguisticsdown::cond_cmpl("[ɔ]")` vs. `r linguisticsdown::cond_cmpl("[ʊ]")`\\
\midrule \hspace{1em}(Intercept) & 0.187 & 0.195 & 0.245\\
\hspace{1em} & se = 0.014 & se = 0.012 & se = 0.013\\
\hspace{1em} & t-value = 12.916 & t-value = 16.089 & t-value = 18.506\\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = \vphantom{1} <0.001\\
\hspace{1em} cue space F1F2F3 & 0.046 & 0.004 & 0.005\\
\hspace{1em} & se = 0.002 & se = 0.000 & se = 0.001\\
\hspace{1em} & t-value = 18.642 & t-value = 16.049 & t-value = 8.546\\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001\\
\hspace{1em}SD (Intercept Talker) & 0.087 & 0.074 & 0.079\\
\hspace{1em}SD (Observations) & 0.015 & 0.002 & 0.004\\
\bottomrule
\end{tabular}
\end{table}

```{r}
#panels <- list(
#  "long vs. short" = list("A" = m.sep.I, "B" = m.sep.Y, "C" = m.sep.U, "D" = m.sep.E, "E" = m.sep.Ä, "F" = m.sep.ÄR),
#  "long vs. short" = list("G" = m.sep.Ö, "H" = m.sep.ÖR, "I" = m.sep.A, "J" = m.sep.Å, "K" = m.sep.O))
#modelsummary::modelsummary(panels, output = "latex", title = "LMM predicting category separability from cue space for all long-short contrasts", estimate = "{estimate}", statistic = c("se = {std.error}", "t-value = {statistic}", "p.value = {p.value}"), shape = "rbind", gof_map = NA)
rm(m.sep.IIYY, m.sep.EEYY, m.sep.IY, m.sep.YYUU, m.sep.ÅÅOO, m.sep.ÅO)
```

#### Investigating separability in long vs. short categories

\begin{landscape}
\begin{table}
\tiny
\caption{LMMs predicting category separability from cue space for all long-short contrasts}
\centering
\begin{tabular}[t]{lcccccc}
\toprule
  & `r linguisticsdown::cond_cmpl("[iː]")` vs. `r linguisticsdown::cond_cmpl("[ɪ]")` & `r linguisticsdown::cond_cmpl("[yː]")` vs. `r linguisticsdown::cond_cmpl("[ʏ]")` & `r linguisticsdown::cond_cmpl("[ʉː]")` vs. `r linguisticsdown::cond_cmpl("[ɵ]")` & `r linguisticsdown::cond_cmpl("[eː]")` vs. `r linguisticsdown::cond_cmpl("[ɛ]")` & `r linguisticsdown::cond_cmpl("[ɛː]")` vs. `r linguisticsdown::cond_cmpl("[ɛ]")` & `r linguisticsdown::cond_cmpl("[æː]")` vs. `r linguisticsdown::cond_cmpl("[æ]")`\\
\midrule
\addlinespace[0.5em]
\multicolumn{6}{l}{\textit{long vs. short}}\\
\midrule \hspace{1em}(Intercept) & 0.234 & 0.201 & 0.347 & 0.467 & 0.239 & 0.112 \\
\hspace{1em} & se = 0.012 & se = 0.010 & se = 0.013 & se = 0.018 & se = 0.016 & se = 0.013 \\
\hspace{1em} & t-value = 19.250 & t-value = 20.798 & t-value = 25.758 & t-value = 26.667 & t-value = 15.358 & t-value = 8.398 \\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001\\
\hspace{1em} cue space F1F3 & -0.102 & -0.007 & -0.195 & -0.055 & -0.010 & 0.002 \\
\hspace{1em} & se = 0.008 & se = 0.006 & se = 0.005 & se = 0.007 & se = 0.005 & se = 0.006 \vphantom{2} \\
\hspace{1em} & t-value = -13.259 & t-value = -1.132 & t-value = -37.105 & t-value = -7.617 & t-value = -2.011 & t-value = 0.410  \\
\hspace{1em} & p.value = <0.001 & p.value = 0.258 & p.value = <0.001 & p.value = <0.001 & p.value = 0.045 & p.value = 0.682  \\
\hspace{1em} cue space F2F3 & 0.000 & 0.023 & -0.024 & -0.202 & -0.129 & -0.014  \\
\hspace{1em} & se = 0.008 & se = 0.006 & se = 0.005 & se = 0.007 & se = 0.005 & se = 0.006 \vphantom{1}  \\
\hspace{1em} & t-value = 0.039 & t-value = 3.720 & t-value = -4.488 & t-value = -27.827 & t-value = -25.714 & t-value = -2.560 \\
\hspace{1em} & p.value = 0.969 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = 0.011  \\
\hspace{1em} cue space F1F2F3 & 0.012 & 0.025 & 0.006 & 0.015 & 0.006 & 0.010  \\
\hspace{1em} & se = 0.008 & se = 0.006 & se = 0.005 & se = 0.007 & se = 0.005 & se = 0.006 \\
\hspace{1em} & t-value = 1.619 & t-value = 4.122 & t-value = 1.062 & t-value = 2.052 & t-value = 1.114 & t-value = 1.874 \\
\hspace{1em} & p.value = 0.107 & p.value = <0.001 & p.value = 0.289 & p.value = 0.041 & p.value = 0.266 & p.value = 0.062 \\
\hspace{1em}SD (Intercept Talker) & 0.066 & 0.052 & 0.079 & 0.102 & 0.092 & 0.078  \\
\hspace{1em}SD (Observations) & 0.047 & 0.038 & 0.032 & 0.044 & 0.031 & 0.033  \\
\addlinespace[0.5em]
\toprule
  & `r linguisticsdown::cond_cmpl("[øː]")` vs. `r linguisticsdown::cond_cmpl("[ø]")` & `r linguisticsdown::cond_cmpl("[œː]")` vs. `r linguisticsdown::cond_cmpl("[œ]")` & `r linguisticsdown::cond_cmpl("[ɑː]")` vs. `r linguisticsdown::cond_cmpl("[a]")` & `r linguisticsdown::cond_cmpl("[oː]")` vs. `r linguisticsdown::cond_cmpl("[ɔ]")` & `r linguisticsdown::cond_cmpl("[uː]")` vs. `r linguisticsdown::cond_cmpl("[ʊ]")` & \\
\multicolumn{6}{l}{\textit{long vs. short }}\\
\midrule \hspace{1em}(Intercept) & 0.151 & 0.168 & 0.328 & 0.261 & 0.143 & \\
\hspace{1em} & se = 0.011 & se = 0.012 & se = 0.012 & se = 0.015 & se = 0.009 & \\
\hspace{1em} & t-value = 13.918 & t-value = 13.942 & t-value = 26.696 & t-value = 17.341 & t-value = 15.995 & \\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = \vphantom{2} <0.001 & \\
\hspace{1em} cue space F1F3 &  -0.021 & -0.029 & -0.134 & -0.097 & -0.046 & \\
\hspace{1em} & se = 0.004 & se = 0.007 & se = 0.007 & se = 0.006 & se = \vphantom{2} 0.005 & \\
\hspace{1em} & t-value = -4.841 & t-value = -4.055 & t-value = -20.347 & t-value = -15.011 & t-value = -9.606 & \\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = \vphantom{1} <0.001 & \\
\hspace{1em} cue space F2F3 & -0.051 & -0.064 & -0.046 & -0.048 & -0.019 & \\
\hspace{1em} &  se = 0.004 & se = 0.007 & se = 0.007 & se = 0.006 & se = \vphantom{1} 0.005 & \\
\hspace{1em} &  t-value = -12.042 & t-value = -9.055 & t-value = -6.968 & t-value = -7.447 & t-value = -3.907 & \\
\hspace{1em} & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & p.value = <0.001 & \\
\hspace{1em} cue space F1F2F3 &  0.003 & 0.003 & 0.008 & 0.004 & 0.004 & \\
\hspace{1em} & se = 0.004 & se = 0.007 & se = 0.007 & se = 0.006 & se = 0.005 & \\
\hspace{1em} & t-value = 0.673 & t-value = 0.401 & t-value = 1.233 & t-value = 0.639 & t-value = 0.747 & \\
\hspace{1em} & p.value = 0.501 & p.value = 0.689 & p.value = 0.219 & p.value = 0.524 & p.value = 0.456 & \\
\hspace{1em}SD (Intercept Talker) &  0.063 & 0.067 & 0.069 & 0.087 & 0.049 & \\
\hspace{1em}SD (Observations) &  0.026 & 0.043 & 0.040 & 0.039 & 0.029 & \\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

```{r}
rm(m.sep.I, m.sep.Y, m.sep.U, m.sep.E, m.sep.Ä, m.sep.ÄR, m.sep.Ö, m.sep.ÖR, m.sep.A, m.sep.Å, m.sep.O)
```

### Category separability indices for additional cue combinations {#sec:SI-separability}
For the F1-F3 and F2-F3 cue spaces, there seems to be two main groups for which either cue combination generates the lowest separability index. For the [`r linguisticsdown::cond_cmpl("ʉː")`] - [`r linguisticsdown::cond_cmpl("ɵ")`], [`r linguisticsdown::cond_cmpl("ɑː")`] - [`r linguisticsdown::cond_cmpl("a")`], [`r linguisticsdown::cond_cmpl("oː")`] - [`r linguisticsdown::cond_cmpl("ɔ")`], [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("ɪ")`], and [`r linguisticsdown::cond_cmpl("uː")`] - [`r linguisticsdown::cond_cmpl("ʊ")`] vowel pairs, the F1-F3 cue combination generates the lowest separability, which would seem to suggest the informativity of F2 for distinguishing between these pairs. For the second group, mainly consisting of allophones to /`r linguisticsdown::cond_cmpl("ɛ")`/ and /`r linguisticsdown::cond_cmpl("ø")`/: [`r linguisticsdown::cond_cmpl("eː")`] - [`r linguisticsdown::cond_cmpl("ɛ")`], [`r linguisticsdown::cond_cmpl("ɛː")`] - [`r linguisticsdown::cond_cmpl("ɛ")`], [`r linguisticsdown::cond_cmpl("æː")`] - [`r linguisticsdown::cond_cmpl("æ")`], [`r linguisticsdown::cond_cmpl("øː")`] - [`r linguisticsdown::cond_cmpl("ø")`] and [`r linguisticsdown::cond_cmpl("œː")`] - [`r linguisticsdown::cond_cmpl("œ")`], the F2-F3 cue combination produces the lowest separability, presumably highlighting the importance of F1. Interestingly, the [`r linguisticsdown::cond_cmpl("yː")`] - [`r linguisticsdown::cond_cmpl("ʏ")`] pair generates almost identical separability indices in F1-F2 and F1-F3 spaces, and increased separability for both F2-F3 and F1-F2-F3, which seems to signal the importance of all three cues.

For the other comparisons, F1-F2 vs. F1-F3, and F1-F2 vs. F2-F3, the LMMs indicated significant negative changes for both comparisons for all but three vowel pairs: for the [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("ɪ")`], there was a significant effect of cue combination in the F1-F2 vs. F1-F3 comparison only ($\hat\beta = -1.02, SE = .007, p < .0001$), for the [`r linguisticsdown::cond_cmpl("yː")`] - [`r linguisticsdown::cond_cmpl("ʏ")`] and [`r linguisticsdown::cond_cmpl("æː")`] - [`r linguisticsdown::cond_cmpl("æ")`] vowel pairs, there was a significant effect in the F1-F2 vs. F2-F3 comparison only (for [`r linguisticsdown::cond_cmpl("yː")`] - [`r linguisticsdown::cond_cmpl("ʏ")`], $\hat\beta = .023, SE = .006, p < .0002$; for [`r linguisticsdown::cond_cmpl("æː")`] - [`r linguisticsdown::cond_cmpl("æ")`], $\hat\beta = -.014, SE = .005, p < .011$).<!--check this-->

```{r}
d.SwehVd.cat.separability.for.longshort.SI %<>%
  mutate(
    cue_dimension = factor(cue_dimension, levels = c("F1F2", "F1F3", "F2F3", "F1F2F3", "F1Duration", "F2Duration", "F1F2Duration")))

# Calculate the relative increase of including F3 and duration for the quantity contrasts
d.SwehVd.cat.separability.comparison.longshort.SI <- d.SwehVd.cat.separability.long.short %>%
  group_by(comparison) %>%
  mutate(
    separabilityDifference_F1F3 = (F1F3_separability - F1F2_separability) / F1F2_separability,
    separabilityDifference_F2F3 = (F2F3_separability - F1F2_separability) / F1F2_separability,
    separabilityDifference_F3 = (F1F2F3_separability - F1F2_separability) / F1F2_separability,
    separabilityDifference_F1Duration = (F1Duration_separability - F1F2_separability) / F1F2_separability,
    separabilityDifference_F2Duration = (F2Duration_separability - F1F2_separability) / F1F2_separability,
    separabilityDifference_Duration = (F1F2Duration_separability - F1F2_separability) / F1F2_separability)

p.category.separability.diff.SI <- p.category.separability.diff.unrounded %+%
  (d.SwehVd.cat.separability.comparison.longshort.SI %>%
     select(comparison, separabilityDifference_F1F3, separabilityDifference_F2F3, separabilityDifference_F3, separabilityDifference_F1Duration, separabilityDifference_F2Duration, separabilityDifference_Duration) %>%
     pivot_longer(
       cols = starts_with("separabilityDifference"),
       names_to = "cue_difference") %>%
     mutate(
        cue_difference = factor(cue_difference, levels = c("separabilityDifference_F1F3", "separabilityDifference_F2F3", "separabilityDifference_F3", "separabilityDifference_F1Duration", "separabilityDifference_F2Duration", "separabilityDifference_Duration"), labels = c("F1-F3", "F2-F3", "F1-F2-F3", "F1-duration", "F2-duration", "F1-F2-duration"))) %>%
     rename(separability_diff = value) %>%
     group_by(comparison, cue_difference) %>%
     mutate(
       ci = list(enframe(Hmisc::smean.cl.boot(separability_diff)))) %>%
     unnest(ci) %>%
     spread(name, value) %>%
     group_by(cue_difference, comparison) %>%
     summarise(
       mean_CI_lower = mean(Lower),
       mean_CI_upper = mean(Upper),
       mean = mean(Mean)) %>%
     mutate(comparison = factor(comparison, levels = c("iː - ɪ",  "yː - ʏ", "ʉː - ɵ", "eː - ɛ", "oː - ɔ", "uː - ʊ", "ɛː - ɛ", "æː - æ", "øː - ø", "œː - œ", "ɑː - a")))) +
  scale_y_continuous("Increase in category separability relative to F1-F2", limits = c(-.7, 1)) +
  theme(legend.title = element_blank())
p.category.separability.diff.SI
```

```{r}
contrasts(d.SwehVd.cat.separability.for.longshort.SI$cue_dimension) <- contr.treatment(7)

# Rename contrast labels
d.SwehVd.cat.separability.for.longshort.SI$cue_dimension <- factor(d.SwehVd.cat.separability.for.longshort.SI$cue_dimension, labels = c("F1F2", "F1F3", "F2F3", "F1F2F3", "F1F2Duration", "F1Duration", "F2Duration"))

m.sep.I.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "iː - ɪ"))

m.sep.Y.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "yː - ʏ"))

m.sep.U.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "ʉː - ɵ"))

m.sep.E.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "eː - ɛ"))

m.sep.Ä.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "ɛː - ɛ"))

m.sep.ÄR.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "æː - æ"))

m.sep.Ö.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "øː - ø"))

m.sep.ÖR.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "œː - œ"))

m.sep.A.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "ɑː - a"))

m.sep.Å.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "oː - ɔ"))

m.sep.O.SI <- lmerTest::lmer(separability ~ cue_dimension + (1|Talker),
                      data = d.SwehVd.cat.separability.for.longshort.SI %>% filter(comparison == "uː - ʊ"))
```

```{r}
rm(d.SwehVd.centroids)
```

### GAMMs evaluating non-linear patterns
Figures XXX to XXX visualize GAMMs evaluating non-linear patterns in the data. Here, separate smooths were fit to each category, to assess whether all investigated categories displayed non-linear patterns.

```{r}
# Difference code the vowel variable
contrasts(d.model.front.long$category) <- contr.treatment(4)
contrasts(d.model.front.short$category) <- contr.treatment(2)
contrasts(d.model.back.long$category) <- contr.treatment(2)
contrasts(d.model.back.long$category) <- contr.treatment(2)
contrasts(d.model.center.short$category) <- contr.treatment(2)
contrasts(d.model.center.short$category) <- contr.treatment(2)
contrasts(d.model.mid.long$category) <- contr.treatment(2)
contrasts(d.model.mid.short$category) <- contr.treatment(2)

d.model.back.long$categoryO <- as.ordered(d.model.back.long$category)
contrasts(d.model.back.long$categoryO) <- contr.treatment(2)

# Rename contrast labels
d.model.front.long$category <- factor(d.model.front.long$category, labels = c("[iː]", "[yː]", "[ʉː]", "[eː]"))
d.model.front.short$category <- factor(d.model.front.short$category, labels = c("[ɪ]", "[ʏ]"))
d.model.back.long$category <- factor(d.model.back.long$category, labels = c("[oː]", "[uː]"))
d.model.back.short$category <- factor(d.model.back.short$category, labels = c("[ɔ]", "[ʊ]"))
d.model.center.long$category <- factor(d.model.center.long$category, labels = c("[ɛː]", "[æː]"))
d.model.center.short$category <- factor(d.model.center.short$category, labels = c("[ɛ]", "[æ]"))
d.model.mid.long$category <- factor(d.model.mid.long$category, labels = c("[øː]", "[œː]"))
d.model.mid.short$category <- factor(d.model.mid.short$category, labels = c("[ø]", "[œ]"))

# Run all gamms for front vowels

m.F1.front.long <- mgcv::bam(F1 ~ category + Gender + s(timepoint, by = category, k = 5) 
                       + s(Talker, bs = "re")  
                       + s(Talker, category, bs = "re"),
                       # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                       data = d.model.front.long)

m.F2.front.long <- mgcv::bam(F2 ~ category + Gender + s(timepoint, by = category, k = 5) 
                       + s(Talker, bs = "re")  
                       + s(Talker, category, bs = "re"),
                       # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                       data = d.model.front.long)

m.F3.front.long <- mgcv::bam(F3 ~ category + Gender + s(timepoint, by = category, k = 5) 
                       + s(Talker, bs = "re")  
                       + s(Talker, category, bs = "re"),
                       # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                       data = d.model.front.long)

m.F1.front.short <- mgcv::bam(F1 ~ category + Gender + s(timepoint, by = category, k = 5) 
                        + s(Talker, bs = "re")  
                        + s(Talker, category, bs = "re"),
                        # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                        data = d.model.front.short)

m.F2.front.short <- mgcv::bam(F2 ~ category + Gender + s(timepoint, by = category, k = 5) 
                        + s(Talker, bs = "re")  
                        + s(Talker, category, bs = "re"),
                        # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                        data = d.model.front.short)

m.F3.front.short <- mgcv::bam(F3 ~ category + Gender + s(timepoint, by = category, k = 5) 
                        + s(Talker, bs = "re")  
                        + s(Talker, category, bs = "re"),
                        # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                        data = d.model.front.short)
```
```{r}
# Run all gamms for back vowels
m.F1.back.long <- mgcv::bam(F1 ~ category + Gender + s(timepoint, by = category, k = 5) 
                       + s(Talker, bs = "re")  
                       + s(Talker, category, bs = "re"),
                       # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                       data = d.model.back.long)

m.F2.back.long <- mgcv::bam(F2 ~ category + Gender + s(timepoint, by = category, k = 5) 
                       + s(Talker, bs = "re")  
                       + s(Talker, category, bs = "re"),
                       # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                       data = d.model.back.long)

m.F3.back.long <- mgcv::bam(F3 ~ category + Gender + s(timepoint, by = category, k = 5) 
                       + s(Talker, bs = "re")  
                       + s(Talker, category, bs = "re"),
                       # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                       data = d.model.back.long)

m.F1.back.short <- mgcv::bam(F1 ~ category + Gender + s(timepoint, by = category, k = 5) 
                        + s(Talker, bs = "re")  
                        + s(Talker, category, bs = "re"),
                        # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                        data = d.model.back.short)

m.F2.back.short <- mgcv::bam(F2 ~ category + Gender + s(timepoint, by = category, k = 5) 
                        + s(Talker, bs = "re")  
                        + s(Talker, category, bs = "re"),
                        # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                        data = d.model.back.short)

m.F3.back.short <- mgcv::bam(F3 ~ category + Gender + s(timepoint, by = category, k = 5) 
                        + s(Talker, bs = "re")  
                        + s(Talker, category, bs = "re"),
                        # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                        data = d.model.back.short)
```
```{r}
# Run all gamms for center vowels
m.F1.center.long <- mgcv::bam(F1 ~ category + Gender + s(timepoint, by = category, k = 5) 
                       + s(Talker, bs = "re")  
                       + s(Talker, category, bs = "re"),
                       # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                       data = d.model.center.long)

m.F2.center.long <- mgcv::bam(F2 ~ category + Gender + s(timepoint, by = category, k = 5) 
                       + s(Talker, bs = "re")  
                       + s(Talker, category, bs = "re"),
                       # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                       data = d.model.center.long)

m.F3.center.long <- mgcv::bam(F3 ~ category + Gender + s(timepoint, by = category, k = 5) 
                       + s(Talker, bs = "re")  
                       + s(Talker, category, bs = "re"),
                       # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                       data = d.model.center.long)

m.F1.center.short <- mgcv::bam(F1 ~ category + Gender + s(timepoint, by = category, k = 5) 
                        + s(Talker, bs = "re")  
                        + s(Talker, category, bs = "re"),
                        # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                        data = d.model.center.short)

m.F2.center.short <- mgcv::bam(F2 ~ category + Gender + s(timepoint, by = category, k = 5) 
                        + s(Talker, bs = "re")  
                        + s(Talker, category, bs = "re"),
                        # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                        data = d.model.center.short)

m.F3.center.short <- mgcv::bam(F3 ~ category + Gender + s(timepoint, by = category, k = 5) 
                        + s(Talker, bs = "re")  
                        + s(Talker, category, bs = "re"),
                        # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                        data = d.model.center.short)
```
```{r}
# Run all gamms for center vowels
m.F1.mid.long <- mgcv::bam(F1 ~ category + Gender + s(timepoint, by = category, k = 5) 
                       + s(Talker, bs = "re")  
                       + s(Talker, category, bs = "re"),
                       # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                       data = d.model.mid.long)

m.F2.mid.long <- mgcv::bam(F2 ~ category + Gender + s(timepoint, by = category, k = 5) 
                       + s(Talker, bs = "re")  
                       + s(Talker, category, bs = "re"),
                       # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                       data = d.model.mid.long)

m.F3.mid.long <- mgcv::bam(F3 ~ category + Gender + s(timepoint, by = category, k = 5) 
                       + s(Talker, bs = "re")  
                       + s(Talker, category, bs = "re"),
                       # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                       data = d.model.mid.long)

m.F1.mid.short <- mgcv::bam(F1 ~ category + Gender + s(timepoint, by = category, k = 5) 
                        + s(Talker, bs = "re")  
                        + s(Talker, category, bs = "re"),
                        # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                        data = d.model.mid.short)

m.F2.mid.short <- mgcv::bam(F2 ~ category + Gender + s(timepoint, by = category, k = 5) 
                        + s(Talker, bs = "re")  
                        + s(Talker, category, bs = "re"),
                        # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                        data = d.model.mid.short)

m.F3.mid.short <- mgcv::bam(F3 ~ category + Gender + s(timepoint, by = category, k = 5) 
                        + s(Talker, bs = "re")  
                        + s(Talker, category, bs = "re"),
                        # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                        data = d.model.mid.short)
```

(ref:m-high-front) Fitted smooths of GAMM for predicting F1 (**upper row**), F2 (**mid row**), F3 (**bottom row**) and 95% confidence intervals for the long vowels [`r linguisticsdown::cond_cmpl("iː")`], [`r linguisticsdown::cond_cmpl("yː")`], [`r linguisticsdown::cond_cmpl("ʉː")`], and [`r linguisticsdown::cond_cmpl("eː")`] (**left**), and the short vowels [`r linguisticsdown::cond_cmpl("ɪ")`] and [`r linguisticsdown::cond_cmpl("ʏ")`] (**right**).

```{r m-high-front, message=FALSE, fig.width=base.width*4.5, fig.height=base.height*4.5, fig.align='center', fig.cap="(ref:m-high-front)"}
par(mfrow=c(3,2))
itsadug::plot_smooth(m.F1.front.long, view = "timepoint", plot_all = "category", col = c("#66C2A5", "#FC8D62", "#8DA0CB", "#FB9A99"), print.summary = FALSE, ylim = c(-1.4, -.75), hide.label = T, main = "A")
itsadug::plot_smooth(m.F1.front.short, view = "timepoint", plot_all = "category", col = c("#1B9E77", "#D95F02"), print.summary = FALSE, ylim = c(-1.4, -.75), hide.label = T, main = "B")
itsadug::plot_smooth(m.F2.front.long, view = "timepoint", plot_all = "category", col = c("#66C2A5", "#FC8D62", "#8DA0CB", "#FB9A99"), print.summary = FALSE, ylim = c(.2, .75), hide.label = T, main = "C")
itsadug::plot_smooth(m.F2.front.short, view = "timepoint", plot_all = "category", col = c("#1B9E77", "#D95F02"), print.summary = FALSE, ylim = c(.2, .75), hide.label = T, main = "D")
itsadug::plot_smooth(m.F3.front.long, view = "timepoint", plot_all = "category", col = c("#66C2A5", "#FC8D62", "#8DA0CB", "#FB9A99"), print.summary = FALSE, ylim = c(.5, 1), hide.label = T, main = "E")
itsadug::plot_smooth(m.F3.front.short, view = "timepoint", plot_all = "category", col = c("#1B9E77", "#D95F02"), print.summary = FALSE, ylim = c(.5, 1), hide.label = T, main = "F")
#plot_diff(m.F1.front.long, view = "timepoint", comp = list(category = c("[iː]", "[yː]")))
#ggsave(file = "../../figures/p.png", width = 9, height = 5, dpi=1200)
```

(ref:m-high-back) Fitted smooths of GAMM for predicting F1 (**upper row**), F2 (**mid row**), F3 (**bottom row**) and 95% confidence intervals for the long vowels [`r linguisticsdown::cond_cmpl("oː")`] - [`r linguisticsdown::cond_cmpl("uː")`] (**left**), and the short vowels [`r linguisticsdown::cond_cmpl("ɔ")`] - [`r linguisticsdown::cond_cmpl("ʊ")`] (**right**).

```{r m-high-back, message=FALSE, fig.width=base.width*4.5, fig.height=base.height*4.5, fig.align='center', fig.cap="(ref:m-high-back)"}
par(mfrow=c(3,2))
itsadug::plot_smooth(m.F1.back.long, view = "timepoint", plot_all = "category", col = c("#A6CEE3", "#B2DF8A"), print.summary = FALSE, ylim = c(-1.4, -.8), hide.label = T, main = "A")
itsadug::plot_smooth(m.F1.back.short, view = "timepoint", plot_all = "category", col = c("#1F78B4", "#33A02C"), print.summary = FALSE, ylim = c(-1.4, -.8), hide.label = T, main = "B")
itsadug::plot_smooth(m.F2.back.long, view = "timepoint", plot_all = "category", col = c("#A6CEE3", "#B2DF8A"), print.summary = FALSE, ylim = c(-.75 -.3), hide.label = T, main = "C")
itsadug::plot_smooth(m.F2.back.short, view = "timepoint", plot_all = "category", col = c("#1F78B4", "#33A02C"), print.summary = FALSE, ylim = c(-.75 -.3), hide.label = T, main = "D")
itsadug::plot_smooth(m.F3.back.long, view = "timepoint", plot_all = "category", col = c("#A6CEE3", "#B2DF8A"), print.summary = FALSE, ylim = c(.7, 1), hide.label = T, main = "E")
itsadug::plot_smooth(m.F3.back.short, view = "timepoint", plot_all = "category", col = c("#1F78B4", "#33A02C"), print.summary = FALSE, ylim = c(.7, 1), hide.label = T, main = "F")
#ggsave(file = "../../figures/p.png", width = 9, height = 5, dpi=1200)
```

(ref:m-center) Fitted smooths of GAMM for predicting F1 (**upper row**), F2 (**mid row**), F3 (**bottom row**) and 95% confidence intervals for the long vowels [`r linguisticsdown::cond_cmpl("ɛː")`] - [`r linguisticsdown::cond_cmpl("æː")`] (**left**), and the short vowels [`r linguisticsdown::cond_cmpl("ɛ")`] - [`r linguisticsdown::cond_cmpl("æ")`] (**right**).

```{r m-center, message=FALSE, fig.width=base.width*4.5, fig.height=base.height*4.5, fig.align='center', fig.cap="(ref:m-center)"}
par(mfrow=c(3,2))
itsadug::plot_smooth(m.F1.center.long, view = "timepoint", plot_all = "category", col = c("#E78AC3", "#A6D854"), print.summary = FALSE, ylim = c(-.95, -.3), hide.label = T, main = "A")
itsadug::plot_smooth(m.F1.center.short, view = "timepoint", plot_all = "category", col = c("#E7298A", "#66A61E"), print.summary = FALSE, ylim = c(-.95, -.3), hide.label = T, main = "B")
itsadug::plot_smooth(m.F2.center.long, view = "timepoint", plot_all = "category", col = c("#E78AC3", "#A6D854"), print.summary = FALSE, ylim = c(.2, .5), hide.label = T, main = "C")
itsadug::plot_smooth(m.F2.center.short, view = "timepoint", plot_all = "category", col = c("#E7298A", "#66A61E"), print.summary = FALSE, ylim = c(.2, .5), hide.label = T, main = "D")
itsadug::plot_smooth(m.F3.center.long, view = "timepoint", plot_all = "category", col = c("#E78AC3", "#A6D854"), print.summary = FALSE, ylim = c(.65, .85), hide.label = T, main = "E")
itsadug::plot_smooth(m.F3.center.short, view = "timepoint", plot_all = "category", col = c("#E7298A", "#66A61E"), print.summary = FALSE, ylim = c(.65, .85), hide.label = T, main = "F")
```
(ref:m-mid) Fitted smooths of GAMM for predicting F1 (**upper row**), F2 (**mid row**), F3 (**bottom row**) and 95% confidence intervals for the long vowels [`r linguisticsdown::cond_cmpl("øː")`] - [`r linguisticsdown::cond_cmpl("œː")`] (**left**), and the short vowels [`r linguisticsdown::cond_cmpl("ø")`] - [`r linguisticsdown::cond_cmpl("œ")`] (**right**).

```{r m-mid, message=FALSE, fig.width=base.width*4.5, fig.height=base.height*4.5, fig.align='center', fig.cap="(ref:m-mid)"}
par(mfrow=c(3,2))
itsadug::plot_smooth(m.F1.mid.long, view = "timepoint", plot_all = "category", col = c("#FFD92F", "#E5C494"), print.summary = FALSE, ylim = c(-1, -.6), hide.label = T, main = "A")
itsadug::plot_smooth(m.F1.mid.short, view = "timepoint", plot_all = "category", col = c("#E6AB02", "#A6761D"), print.summary = FALSE, ylim = c(-1, -.6), hide.label = T, main = "B")
itsadug::plot_smooth(m.F2.mid.long, view = "timepoint", plot_all = "category", col = c("#FFD92F", "#E5C494"), print.summary = FALSE, ylim = c(-.1, .3), hide.label = T, main = "C")
itsadug::plot_smooth(m.F2.mid.short, view = "timepoint", plot_all = "category", col = c("#E6AB02", "#A6761D"), print.summary = FALSE, ylim = c(-.1, .3), hide.label = T, main = "D")
itsadug::plot_smooth(m.F3.mid.long, view = "timepoint", plot_all = "category", col = c("#FFD92F", "#E5C494"), print.summary = FALSE, ylim = c(.65, .8), hide.label = T, main = "E")
itsadug::plot_smooth(m.F3.mid.short, view = "timepoint", plot_all = "category", col = c("#E6AB02", "#A6761D"), print.summary = FALSE, ylim = c(.65, .8), hide.label = T, main = "F")
#ggsave(file = "../../figures/p.png", width = 9, height = 5, dpi=1200)
```
```{r}
# Dataframe for running GAMMs on all 21 vowels grouped by pair
d.model.paired <- d.SwehVd.vowel_pairs %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", vowel_pair != "ɛ - eː, ɛː") %>%
  mutate(
    #Reorder vowel_pairs to match the levels in category
    vowel_pair = factor(vowel_pair, 
                        levels = c("ɪ - iː", "ʏ - yː", "ɵ - ʉː", "ɛ - eː", "ɛ - ɛː", "æ - æː", "ø - øː", "œ - œː", "a - ɑː", "ɔ - oː", "ʊ - uː")),
    category = factor(category, levels = levels.vowel.IPA.swe)) %>%
  mutate(across(c(Talker, Quantity, Gender), ~ factor(.x)))

# Difference code the vowel pair variable
contrasts(d.model.paired$category) <- MASS::contr.sdif(21)

# Rename contrast labels
d.model.paired$category <- factor(d.model.paired$category, labels = c("[iː]", "[ɪ]", "[yː]", "[ʏ]", "[ʉː]", "[ɵ]", "[eː]", "[ɛ]", "[ɛː]", "[æː]", "[æ]", "[øː]", "[ø]", "[œː]", "[œ]", "[ɑː]", "[a]", "[oː]", "[ɔ]", "[uː]", "[ʊ]"))

m.F1.paired <- mgcv::bam(F1 ~ category + Gender + 
                           s(timepoint, by = category, k = 5) + 
                           s(Talker, bs = "re") +
                           s(Talker, category, bs = "re"),
                         data = d.model.paired)

# m.1 <- brms::bf(category ~ t2(F1, F2, F3, timepoint) + (1 + F1 | Talker),
#                   family = multinomial,
#                   data = d.model.paired)
# 
# m1 <- glm(category ~ F1 + F2 + F3, family = "binomial",
#           data = d.model.paired %>% filter(category %in% c("[iː]", "[ɪ]", "[yː]", "[ʏ]", "[ʉː]", "[ɵ]")))

m.F2.paired <- mgcv::bam(F2 ~ category + Gender + 
                           s(timepoint, by = category, k = 5) + 
                           s(Talker, bs = "re") +
                           s(Talker, category, bs = "re"),
                         data = d.model.paired)

m.F3.paired <- mgcv::bam(F3 ~ category + Gender + 
                           s(timepoint, by = category, k = 5) + 
                           s(Talker, bs = "re") +
                           s(Talker, category, bs = "re"),
                         data = d.model.paired)
```

### Summary tables of GAMMs {#sec:SI-GAMMs}
Tables \@ref(tab:m-F1-long), \@ref(tab:m.F1.front.short), \@ref(tab:m.F2.front.long), \@ref(tab:m.F2.front.short), \@ref(tab:m.F3.front.long), \@ref(tab:m.F3.front.short) summarize the results of the GAMMs fit to neighboring contrasts.

\begin{table}[ht]
\caption{\label{tab:m-F1-long}Model summary, coefficients for parametric and smooth terms for the high front long vowels predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.2468 & 0.0163 & -76.3272 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[yː]")` & -0.0410 & 0.0136 & -3.0195 & 0.0025 \\ 
  category `r linguisticsdown::cond_cmpl("[ʉː]")` & 0.0460 & 0.0136 & 3.3884 & 0.0007 \\ 
  category `r linguisticsdown::cond_cmpl("[eː]")` & 0.1773 & 0.0136 & 13.0179 & $<$ 0.0001 \\ 
  Gendermale & -0.0387 & 0.0207 & -1.8640 & 0.0624 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[iː]")` & 2.9311 & 3.4400 & 57.0277 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 3.0869 & 3.5748 & 70.2809 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʉː]")` & 3.4486 & 3.8299 & 179.1962 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[eː]")` & 3.9101 & 3.9950 & 1004.9035 & $<$ 0.0001 \\ 
  s(Talker) & 27.4456 & 35.0000 & 1492.1121 & $<$ 0.0001 \\ 
  s(Talker,category) & 110.0590 & 143.0000 & 131.7453 & 0.0012 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.front.long}
\end{table}

\begin{table}[ht]
\centering
\caption{\label{tab:m.F1.front.short}Model summary, coefficients for parametric and smooth terms for the high front short vowels predicting F1} 
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.2205 & 0.0131 & -92.8715 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʏ]")` & 0.0103 & 0.0104 & 0.9899 & 0.3223 \\ 
  Gendermale & -0.0260 & 0.0178 & -1.4605 & 0.1442 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɪ]")` & 2.4343 & 2.9458 & 3.4971 & 0.0140 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʏ]")` & 1.4602 & 1.7789 & 10.4211 & 0.0001 \\ 
  s(Talker) & 22.8931 & 35.0000 & 452.9953 & $<$ 0.0001 \\ 
  s(Talker,category) & 45.4715 & 71.0000 & 115.8641 & 0.0108 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.front.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.front.long}Model summary, coefficients for parametric and smooth terms for the high front long vowels predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.4714 & 0.0158 & 29.7433 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[yː]")` & -0.0337 & 0.0169 & -1.9920 & 0.0464 \\ 
  category `r linguisticsdown::cond_cmpl("[ʉː]")` & -0.1640 & 0.0169 & -9.6862 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[eː]")` & 0.1220 & 0.0170 & 7.1905 & $<$ 0.0001 \\ 
  Gendermale & 0.0226 & 0.0177 & 1.2800 & 0.2006 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[iː]")` & 1.2063 & 1.3816 & 26.3039 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 1.8420 & 2.2681 & 26.5236 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʉː]")` & 2.7349 & 3.2517 & 179.2476 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[eː]")` & 3.8488 & 3.9863 & 660.7542 & $<$ 0.0001 \\ 
  s(Talker) & 18.8071 & 35.0000 & 1958.0198 & 0.0003 \\ 
  s(Talker,category) & 121.8467 & 143.0000 & 461.3724 & 0.0002 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.front.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.front.short}Model summary, coefficients for parametric and smooth terms for the high front short vowels predicting F2} 
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.6673 & 0.0161 & 41.4614 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʏ]")` & -0.0799 & 0.0181 & -4.4158 & $<$ 0.0001 \\ 
  Gendermale & -0.0130 & 0.0196 & -0.6607 & 0.5088 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɪ]")` & 3.3567 & 3.7755 & 9.2388 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʏ]")` & 2.5556 & 3.0757 & 11.1705 & $<$ 0.0001 \\ 
  s(Talker) & 5.0689 & 35.0000 & 71.2973 & 0.2234 \\ 
  s(Talker,category) & 64.6715 & 71.0000 & 256.0953 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.front.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.front.long}Model summary, coefficients for parametric and smooth terms for the high front long vowels predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.9027 & 0.0120 & 75.2840 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("yː")`] & 0.0007 & 0.0135 & 0.0552 & 0.9560 \\ 
  category [`r linguisticsdown::cond_cmpl("ʉː")`] & -0.1923 & 0.0135 & -14.2802 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("eː")`] & -0.0584 & 0.0135 & -4.3279 & $<$ 0.0001 \\ 
  Gendermale & 0.0082 & 0.0128 & 0.6409 & 0.5216 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("iː")`] & 2.4433 & 2.9498 & 10.5927 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("yː")`] & 2.7424 & 3.2613 & 6.8240 & 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʉː")`] & 3.1406 & 3.6172 & 15.3775 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("eː")`] & 3.6758 & 3.9392 & 100.9240 & $<$ 0.0001 \\ 
  s(Talker) & 15.5843 & 35.0000 & 691.3561 & 0.0056 \\ 
  s(Talker,category) & 124.3566 & 143.0000 & 222.1766 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F3.front.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.front.short}Model summary, coefficients for parametric and smooth terms for the high front short vowels predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.8632 & 0.0101 & 85.5852 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʏ")`] & -0.1024 & 0.0077 & -13.3324 & $<$ 0.0001 \\ 
  Gendermale & 0.0054 & 0.0138 & 0.3956 & 0.6925 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɪ")`] & 2.8288 & 3.3486 & 62.2052 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʏ")`] & 2.3569 & 2.8620 & 6.1364 & 0.0005 \\ 
  s(Talker) & 24.0220 & 35.0000 & 717.3496 & $<$ 0.0001 \\ 
  s(Talker,category) & 44.8619 & 71.0000 & 162.8064 & 0.0522 \\ 
   \hline
\end{tabular}
\label{tab:m.F3.front.short}
\end{table}

```{r}
```

Tables \@ref(tab:m.F1.back.long), \@ref(tab:m.F1.back.short), \@ref(tab:m.F2.back.long), \@ref(tab:m.F2.back.short), \@ref(tab:m.F3.back.long), \@ref(tab:m.F3.back.short)    summarize the results of the GAMMs fit to neighboring high back contrasts.

\begin{table}[ht]
\caption{\label{tab:m.F1.back.long}Model summary, coefficients for parametric and smooth terms for the high back long vowels predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.0587 & 0.0141 & -75.1949 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("uː")`] & -0.1418 & 0.0128 & -11.0521 & $<$ 0.0001 \\ 
  Gendermale & 0.0425 & 0.0185 & 2.3003 & 0.0215 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("oː")`] & 3.7652 & 3.9665 & 396.9714 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("uː")`] & 2.5930 & 3.1042 & 87.5653 & $<$ 0.0001 \\ 
  s(Talker) & 18.0364 & 35.0000 & 136.5104 & 0.0006 \\ 
  s(Talker,category) & 49.4368 & 71.0000 & 67.1671 & 0.0005 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.back.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.back.long}Model summary, coefficients for parametric and smooth terms for the high back long vowels predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.5435 & 0.0187 & -29.0043 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("uː")`]  & 0.0552 & 0.0161 & 3.4218 & 0.0006 \\ 
  Gendermale & 0.0697 & 0.0250 & 2.7914 & 0.0053 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("oː")`] & 3.8644 & 3.9886 & 808.6381 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("uː")`]  & 3.7776 & 3.9699 & 117.8905 & $<$ 0.0001 \\ 
  s(Talker) & 20.2977 & 35.0000 & 106.8326 & $<$ 0.0001 \\ 
  s(Talker,category) & 45.9985 & 71.0000 & 38.2194 & 0.0023 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.back.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.back.long}Model summary, coefficients for parametric and smooth terms for the high back long vowels predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.8020 & 0.0130 & 61.6839 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("uː")`] & -0.0279 & 0.0072 & -3.8880 & 0.0001 \\ 
  Gendermale & 0.0252 & 0.0184 & 1.3686 & 0.1712 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("oː")`] & 3.0516 & 3.5419 & 15.4137 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("uː")`] & 2.7295 & 3.2436 & 16.7865 & $<$ 0.0001 \\ 
  s(Talker) & 29.6774 & 35.0000 & 848.4999 & $<$ 0.0001 \\ 
  s(Talker,category) & 38.1508 & 71.0000 & 74.8089 & 0.1462 \\ 
   \hline
\end{tabular}
\label{tab:m.F3.back.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F1.back.short}Model summary, coefficients for parametric and smooth terms for the high back short vowels predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.9472 & 0.0128 & -74.0368 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʊ")`] & -0.1975 & 0.0145 & -13.6429 & $<$ 0.0001 \\ 
  Gendermale & 0.0378 & 0.0157 & 2.4136 & 0.0159 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɔ")`] & 3.1438 & 3.6173 & 88.6017 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʊ")`] & 2.1220 & 2.5898 & 66.7215 & $<$ 0.0001 \\ 
  s(Talker) & 4.9611 & 35.0000 & 31.5664 & 0.3458 \\ 
  s(Talker,category) & 63.2273 & 70.0000 & 123.1536 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.back.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.back.short}Model summary, coefficients for parametric and smooth terms for the high back short vowels predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.3485 & 0.0176 & -19.8130 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʊ")`] & -0.0574 & 0.0139 & -4.1379 & $<$ 0.0001 \\ 
  Gendermale & -0.0121 & 0.0239 & -0.5046 & 0.6139 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɔ")`] & 3.3590 & 3.7732 & 58.3532 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʊ")`] & 3.5984 & 3.9059 & 269.8991 & $<$ 0.0001 \\ 
  s(Talker) & 23.1772 & 35.0000 & 270.7720 & $<$ 0.0001 \\ 
  s(Talker,category) & 43.3625 & 70.0000 & 68.6349 & 0.0106 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.back.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.back.short}Model summary, coefficients for parametric and smooth terms for the high back short vowels predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.8453 & 0.0125 & 67.7240 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("ʊ")`]  & -0.0396 & 0.0087 & -4.5752 & $<$ 0.0001 \\ 
  Gendermale & 0.0084 & 0.0173 & 0.4861 & 0.6269 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɔ")`] & 1.0000 & 1.0000 & 79.4844 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ʊ")`]  & 1.8262 & 2.2471 & 11.1195 & $<$ 0.0001 \\ 
  s(Talker) & 26.2191 & 35.0000 & 2923.8268 & $<$ 0.0001 \\ 
  s(Talker,category) & 42.4906 & 70.0000 & 443.8224 & 0.1140 \\ 
   \hline
\end{tabular}
\label{tab:m.F3.back.short}
\end{table}

```{r}
```

Tables \@ref(tab:m.F1.center.long), \@ref(tab:m.F1.center.short), \@ref(tab:m.F2.center.long), \@ref(tab:m.F2.center.short), \@ref(tab:m.F3.center.long), \@ref(tab:m.F3.center.short) summarize the results of the GAMMs fit to neighboring lower-mid front contrasts.

\begin{table}[ht]
\caption{\label{tab:m.F1.center.long}Model summary, coefficients for parametric and smooth terms for the lower-mid front long vowels predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.4989 & 0.0219 & -22.8167 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æː")`] & 0.0658 & 0.0130 & 5.0547 & $<$ 0.0001 \\ 
  Gendermale & -0.0346 & 0.0308 & -1.1223 & 0.2618 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɛː")`] & 1.4851 & 1.8142 & 7.0119 & 0.0038 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æː")`] & 3.2083 & 3.6706 & 111.7106 & $<$ 0.0001 \\ 
  s(Talker) & 28.7148 & 35.0000 & 805.9948 & $<$ 0.0001 \\ 
  s(Talker,category) & 38.9026 & 71.0000 & 86.8982 & 0.0405 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.center.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.center.long}Model summary, coefficients for parametric and smooth terms for the lower-mid front long vowels predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.3202 & 0.0113 & 28.2110 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æː")`] & -0.0846 & 0.0096 & -8.7790 & $<$ 0.0001 \\ 
  Gendermale & -0.0038 & 0.0152 & -0.2518 & 0.8012 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɛː")`]  & 3.3457 & 3.7693 & 239.3935 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æː")`] & 3.3336 & 3.7592 & 44.0107 & $<$ 0.0001 \\ 
  s(Talker) & 20.7802 & 35.0000 & 1200.4655 & $<$ 0.0001 \\ 
  s(Talker,category) & 48.5959 & 71.0000 & 406.9209 & 0.0016 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.center.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.center.long}Model summary, coefficients for parametric and smooth terms for the lower-mid front long vowels
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7477 & 0.0171 & 43.8376 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æː")`] & -0.0119 & 0.0105 & -1.1309 & 0.2582 \\ 
  Gendermale & 0.0290 & 0.0239 & 1.2118 & 0.2257 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɛː")`] & 2.5579 & 3.0789 & 16.9912 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æː")`] & 1.9849 & 2.4346 & 3.4538 & 0.0236 \\ 
  s(Talker) & 28.1870 & 35.0000 & 982.4866 & $<$ 0.0001 \\ 
  s(Talker,category) & 39.8401 & 71.0000 & 101.3499 & 0.1088 \\ 
   \hline
\end{tabular} 
\label{tab:m.F3.center.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F1.center.short}Model summary, coefficients for parametric and smooth terms for the lower-mid front short vowels
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.7269 & 0.0151 & -48.0611 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æ")`] & 0.2512 & 0.0136 & 18.4943 & $<$ 0.0001 \\ 
  Gendermale & -0.0162 & 0.0199 & -0.8137 & 0.4159 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɛ")`] & 3.4131 & 3.8118 & 50.0923 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æ")`] & 3.0619 & 3.5569 & 209.1458 & $<$ 0.0001 \\ 
  s(Talker) & 18.6400 & 35.0000 & 368.7870 & 0.0008 \\ 
  s(Talker,category) & 50.2350 & 71.0000 & 167.6331 & 0.0024 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.center.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.center.short}Model summary, coefficients for parametric and smooth terms for the lower-mid front short vowels
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.4089 & 0.0127 & 32.1002 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æ")`] & -0.1181 & 0.0124 & -9.5505 & $<$ 0.0001 \\ 
  Gendermale & -0.0201 & 0.0164 & -1.2217 & 0.2219 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɛ")`] & 1.3303 & 1.5854 & 290.5422 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æ")`] & 2.4991 & 3.0151 & 57.1634 & $<$ 0.0001 \\ 
  s(Talker) & 15.0381 & 35.0000 & 352.1814 & 0.0102 \\ 
  s(Talker,category) & 54.2360 & 71.0000 & 243.0783 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.center.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.center.short}Model summary, coefficients for parametric and smooth terms for the lower-mid front short vowels
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7458 & 0.0136 & 55.0097 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("æ")`] & 0.0123 & 0.0085 & 1.4409 & 0.1497 \\ 
  Gendermale & 0.0315 & 0.0190 & 1.6568 & 0.0977 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ɛ")`] & 2.2157 & 2.7079 & 4.4190 & 0.0084 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("æ")`] & 2.5770 & 3.0972 & 6.6918 & 0.0001 \\ 
  s(Talker) & 27.8932 & 35.0000 & 288.5886 & $<$ 0.0001 \\ 
  s(Talker,category) & 38.4020 & 71.0000 & 32.8884 & 0.0793 \\ 
   \hline
\end{tabular}
\label{tab:m.F3.center.short}
\end{table}

```{r}
#gamtabs(m.F3.mid.short, caption='Model summary, coefficients for parametric and smooth terms', label = "tab:m.F3.mid.short")
```
Tables \@ref(tab:m.F1.mid.long), \@ref(tab:m.F1.mid.short), \@ref(tab:m.F2.mid.long), \@ref(tab:m.F2.mid.short), \@ref(tab:m.F3.mid.long), \@ref(tab:m.F3.mid.short) summarize the results of the GAMMs fit to mid center neighboring contrasts.

\begin{table}[ht]
\caption{\label{tab:m.F1.mid.long}Model summary, coefficients for parametric and smooth terms for the mid center long vowels predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.7178 & 0.0177 & -40.5263 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œː")`] & -0.0171 & 0.0111 & -1.5411 & 0.1234 \\ 
  Gendermale & -0.0175 & 0.0248 & -0.7054 & 0.4806 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("øː")`] & 2.0230 & 2.4847 & 2.0968 & 0.0968 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œː")`] & 2.6059 & 3.1188 & 56.5090 & $<$ 0.0001 \\ 
  s(Talker) & 27.9922 & 35.0000 & 1639.3603 & $<$ 0.0001 \\ 
  s(Talker,category) & 40.8306 & 71.0000 & 206.6339 & 0.0475 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.mid.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.mid.long}Model summary, coefficients for parametric and smooth terms for the mid center long vowels predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.1620 & 0.0143 & 11.3386 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œː")`] & -0.1755 & 0.0167 & -10.4903 & $<$ 0.0001 \\ 
  Gendermale & -0.0040 & 0.0171 & -0.2360 & 0.8135 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("øː")`] & 3.3494 & 3.7708 & 206.3966 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œː")`] & 3.8405 & 3.9842 & 264.4796 & $<$ 0.0001 \\ 
  s(Talker) & 1.2539 & 35.0000 & 17.5588 & 0.4091 \\ 
  s(Talker,category) & 69.0578 & 71.0000 & 335.5738 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.mid.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.mid.long}Model summary, coefficients for parametric and smooth terms for the mid center long vowels predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7238 & 0.0110 & 65.6956 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œː")`] & 0.0171 & 0.0076 & 2.2335 & 0.0256 \\ 
  Gendermale & -0.0094 & 0.0153 & -0.6174 & 0.5370 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("øː")`] & 3.3171 & 3.7495 & 53.2943 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œː")`] & 2.7352 & 3.2497 & 16.0149 & $<$ 0.0001 \\ 
  s(Talker) & 26.1642 & 35.0000 & 1328.1128 & $<$ 0.0001 \\ 
  s(Talker,category) & 42.7042 & 71.0000 & 212.7082 & 0.0380 \\ 
   \hline
\end{tabular} 
\label{tab:m.F3.mid.long}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F1.mid.short}Model summary, coefficients for parametric and smooth terms for the mid center short vowels
predicting F1}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -0.8401 & 0.0143 & -58.6160 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œ")`] & -0.0280 & 0.0139 & -2.0198 & 0.0435 \\ 
  Gendermale & 0.0146 & 0.0185 & 0.7916 & 0.4287 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ø")`] & 3.5157 & 3.8680 & 170.5141 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œ")`] & 2.5538 & 3.0669 & 111.0694 & $<$ 0.0001 \\ 
  s(Talker) & 15.2210 & 35.0000 & 539.6243 & 0.0133 \\ 
  s(Talker,category) & 54.4128 & 71.0000 & 357.8582 & 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.mid.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F2.mid.short}Model summary, coefficients for parametric and smooth terms for the mid center short vowels
predicting F2}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.2497 & 0.0162 & 15.4595 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œ")`] & -0.1809 & 0.0165 & -10.9577 & $<$ 0.0001 \\ 
  Gendermale & -0.0352 & 0.0205 & -1.7185 & 0.0858 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ø")`] & 2.3312 & 2.8344 & 41.2673 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œ")`] & 3.6083 & 3.9108 & 518.6392 & $<$ 0.0001 \\ 
  s(Talker) & 12.1043 & 35.0000 & 2648.4818 & 0.0306 \\ 
  s(Talker,category) & 58.3994 & 71.0000 & 2464.2257 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.mid.short}
\end{table}

\begin{table}[ht]
\caption{\label{tab:m.F3.mid.short}Model summary, coefficients for parametric and smooth terms for the mid center short vowels
predicting F3}
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.7335 & 0.0109 & 67.1090 & $<$ 0.0001 \\ 
  category [`r linguisticsdown::cond_cmpl("œ")`] & 0.0192 & 0.0082 & 2.3509 & 0.0188 \\ 
  Gendermale & 0.0059 & 0.0150 & 0.3955 & 0.6925 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("ø")`] & 3.2158 & 3.6784 & 128.8414 & $<$ 0.0001 \\ 
  s(timepoint):category [`r linguisticsdown::cond_cmpl("œ")`] & 2.5288 & 3.0405 & 7.8057 & $<$ 0.0001 \\ 
  s(Talker) & 24.5124 & 35.0000 & 1752.1939 & $<$ 0.0001 \\ 
  s(Talker,category) & 45.0178 & 71.0000 & 354.4428 & 0.0421 \\ 
   \hline
\end{tabular}
\label{tab:m.F3.mid.short}
\end{table}

```{r}
```
Tables \@ref(tab:m-F1-paired), \@ref(tab:m-F2-paired), \@ref(tab:m-F3-paired), \@ref(tab:m-duration-paired summarize the results of the GAMMs fit to long-short vowel pairs.

\begin{table}[ht]
\small
\caption{\label{tab:m-F1-paired}Model summary, coefficients for parametric and smooth terms for all 21 vowels predicting F1} 
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.2618 & 0.0142 & -89.1177 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɪ]")` & 0.0324 & 0.0157 & 2.0593 & 0.0395 \\ 
  category `r linguisticsdown::cond_cmpl("[yː]")` & -0.0410 & 0.0157 & -2.6085 & 0.0091 \\ 
  category `r linguisticsdown::cond_cmpl("[ʏ]")` & 0.0426 & 0.0157 & 2.7044 & 0.0068 \\ 
  category `r linguisticsdown::cond_cmpl("[ʉː]")` & 0.0456 & 0.0157 & 2.8987 & 0.0037 \\ 
  category `r linguisticsdown::cond_cmpl("[ɵ]")` & 0.1600 & 0.0157 & 10.1678 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[eː]")` & 0.1794 & 0.0158 & 11.3701 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɛ]")` & 0.5294 & 0.0157 & 33.7688 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɛː]")` & 0.7494 & 0.0158 & 47.5815 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æː]")` & 0.8138 & 0.0158 & 51.6221 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æ]")` & 0.7805 & 0.0157 & 49.5770 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[øː]")` & 0.5390 & 0.0157 & 34.2376 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ø]")` & 0.4322 & 0.0157 & 27.4658 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[œː]")` & 0.5222 & 0.0158 & 33.1307 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[œ]")` & 0.4039 & 0.0157 & 25.6634 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɑː]")` & 0.5853 & 0.0157 & 37.1926 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[a]")` & 0.7541 & 0.0157 & 47.9239 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[oː]")` & 0.2189 & 0.0157 & 13.9132 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɔ]")` & 0.3373 & 0.0157 & 21.4318 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[uː]")` & 0.0861 & 0.0157 & 5.4716 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʊ]")` & 0.1406 & 0.0159 & 8.8655 & $<$ 0.0001 \\ 
  Gendermale & -0.0063 & 0.0134 & -0.4692 & 0.6389 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[iː]")` & 2.9699 & 3.4749 & 60.8984 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɪ]")` & 2.3196 & 2.8202 & 2.7657 & 0.0380 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 3.1211 & 3.6026 & 75.2795 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʏ]")` & 1.2836 & 1.5107 & 9.6941 & 0.0003 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʉː]")` & 3.4823 & 3.8488 & 192.2389 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɵ]")` & 2.5131 & 3.0207 & 32.6885 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[eː]")` & 3.9160 & 3.9957 & 1083.5453 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛ]")` & 3.6352 & 3.9234 & 92.8670 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛː]")` & 2.0567 & 2.5243 & 7.0863 & 0.0003 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æː]")` & 3.3827 & 3.7909 & 164.8723 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æ]")` & 3.0379 & 3.5369 & 199.9285 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[øː]")` & 1.8388 & 2.2686 & 1.5376 & 0.1954 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ø]")` & 3.3342 & 3.7613 & 107.8693 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œː]")` & 2.3894 & 2.8882 & 43.6960 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œ]")` & 2.2966 & 2.7874 & 75.0532 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɑː]")` & 2.6751 & 3.1893 & 69.7824 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[a]")` & 3.3890 & 3.7948 & 182.4588 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[oː]")` & 3.8321 & 3.9826 & 575.2500 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɔ]")` & 2.9739 & 3.4744 & 66.6260 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[uː]")` & 2.9624 & 3.4643 & 114.3566 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʊ]")` & 1.9404 & 2.3813 & 52.5447 & $<$ 0.0001 \\ 
  s(Talker) & 30.3648 & 35.0000 & 7129.7891 & $<$ 0.0001 \\ 
  s(Talker,category) & 700.8765 & 754.0000 & 82.5811 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F1.paired}
\end{table}

\begin{table}[ht]
\small
\caption{\label{tab:m-F2-paired}Model summary, coefficients for parametric and smooth terms for all 21 vowels predicting F2} 
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.4831 & 0.0135 & 35.8530 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɪ]")` & 0.1793 & 0.0170 & 10.5309 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[yː]")` & -0.0338 & 0.0170 & -1.9831 & 0.0474 \\ 
  category `r linguisticsdown::cond_cmpl("[ʏ]")` & 0.0995 & 0.0170 & 5.8417 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʉː]")` & -0.1643 & 0.0170 & -9.6510 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɵ]")` & -0.4735 & 0.0170 & -27.8011 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[eː]")` & 0.1209 & 0.0171 & 7.0847 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[ɛ]")` & -0.0824 & 0.0170 & -4.8503 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɛː]")` & -0.1641 & 0.0170 & -9.6292 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æː]")` & -0.2473 & 0.0171 & -14.5034 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æ]")` & -0.1997 & 0.0170 & -11.7242 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[øː]")` & -0.3210 & 0.0170 & -18.8458 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ø]")` & -0.2480 & 0.0170 & -14.5610 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[œː]")` & -0.4977 & 0.0171 & -29.1877 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[œ]")` & -0.4302 & 0.0170 & -25.2542 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[ɑː]")` & -0.7165 & 0.0170 & -42.0672 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[a]")` & -0.4597 & 0.0170 & -26.9935 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[oː]")` & -1.0077 & 0.0170 & -59.1781 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɔ]")` & -0.8391 & 0.0170 & -49.2613 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[uː]")` & -0.9441 & 0.0170 & -55.4250 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[ʊ]")` & -0.9004 & 0.0172 & -52.4751 & $<$ 0.0001 \\
  Gendermale & -0.0030 & 0.0097 & -0.3103 & 0.7563 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[iː]")` & 1.0015 & 1.0029 & 25.5650 & $<$ 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɪ]")` & 3.3221 & 3.7528 & 8.6397 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 1.6674 & 2.0543 & 21.7906 & $<$ 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʏ]")` & 2.5259 & 3.0445 & 10.5300 & $<$ 0.0001 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʉː]")` & 2.4654 & 2.9713 & 145.9897 & $<$ 0.0001 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɵ]")` & 3.3562 & 3.7716 & 135.6853 & $<$ 0.0001 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[eː]")` & 3.8051 & 3.9774 & 494.3527 & $<$ 0.0001 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛ]")` & 1.7680 & 2.1855 & 278.9595 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛː]")` & 2.9022 & 3.4184 & 98.6021 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æː]")` & 2.8558 & 3.3707 & 17.9822 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æ]")` & 2.2613 & 2.7546 & 42.0383 & $<$ 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[øː]")` & 2.5127 & 3.0298 & 91.1535 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ø]")` & 1.1352 & 1.2570 & 29.5323 & $<$ 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œː]")` & 3.6238 & 3.9165 & 95.8035 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œ]")` & 3.2293 & 3.6851 & 184.8442 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɑː]")` & 3.4899 & 3.8515 & 25.4819 & $<$ 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[a]")` & 1.7816 & 2.1970 & 42.6025 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[oː]")` & 3.9594 & 3.9990 & 2902.3723 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɔ]")` & 3.5890 & 3.9014 & 111.6825 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[uː]")` & 3.9363 & 3.9974 & 420.6330 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʊ]")` & 3.7607 & 3.9653 & 524.5672 & $<$ 0.0001 \\ 
  s(Talker) & 24.6716 & 35.0000 & 3703.0200 & $<$ 0.0001 \\ 
  s(Talker,category) & 711.8453 & 754.0000 & 111.1971 & $<$ 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F2.paired}
\end{table}

\begin{table}[ht]
\small
\caption{\label{tab:m-F3-paired}Model summary, coefficients for parametric and smooth terms for all 21 vowels predicting F3} 
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.8994 & 0.0116 & 77.7339 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɪ]")` & -0.0413 & 0.0119 & -3.4809 & 0.0005 \\
  category `r linguisticsdown::cond_cmpl("[yː]")` & 0.0007 & 0.0118 & 0.0576 & 0.9541 \\ 
  category `r linguisticsdown::cond_cmpl("[ʏ]")` & -0.1432 & 0.0119 & -12.0801 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʉː]")` & -0.1924 & 0.0118 & -16.2359 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɵ]")` & -0.1555 & 0.0119 & -13.1199 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[eː]")` & -0.0588 & 0.0119 & -4.9460 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[ɛ]")` & -0.1459 & 0.0118 & -12.3596 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɛː]")` & -0.1450 & 0.0119 & -12.2203 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[æː]")` & -0.1572 & 0.0119 & -13.2361 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æ]")` & -0.1340 & 0.0119 & -11.2963 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[øː]")` & -0.1875 & 0.0119 & -15.8153 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[ø]")` & -0.1709 & 0.0119 & -14.4191 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[œː]")` & -0.1698 & 0.0119 & -14.3056 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[œ]")` & -0.1511 & 0.0119 & -12.7489 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[ɑː]")` & -0.0996 & 0.0119 & -8.4030 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[a]")` & -0.1003 & 0.0119 & -8.4656 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[oː]")` & -0.0921 & 0.0119 & -7.7723 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɔ]")` & -0.0580 & 0.0119 & -4.8932 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[uː]")` & -0.1210 & 0.0119 & -10.2044 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[ʊ]")` & -0.0974 & 0.0119 & -8.1535 & $<$ 0.0001 \\
  Gendermale & 0.0155 & 0.0121 & 1.2866 & 0.1982 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[iː]")` & 2.3877 & 2.8890 & 10.2860 & $<$ 0.0001 \\   
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɪ]")`  & 2.5706 & 3.0908 & 42.4087 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 2.7213 & 3.2404 & 6.5528 & 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʏ]")` & 2.0793 & 2.5501 & 4.1808 & 0.0088 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʉː]")` & 3.1212 & 3.6017 & 14.7398 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɵ]")` & 1.0007 & 1.0013 & 124.4234 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[eː]")` & 3.6630 & 3.9345 & 96.5489 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛ]")` & 3.0119 & 3.5168 & 12.8402 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛː]")` & 2.8123 & 3.3344 & 24.2192 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æː]")` & 2.2083 & 2.6896 & 5.0740 & 0.0028 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æ]")` & 3.0922 & 3.5819 & 10.8873 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[øː]")` & 3.0677 & 3.5614 & 32.2446 & $<$ 0.0001 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ø]")` & 2.8947 & 3.4110 & 68.1325 & $<$ 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œː]")` & 2.4586 & 2.9632 & 9.8438 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œ]")` & 2.1449 & 2.6169 & 4.3514 & 0.0095 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɑː]")` & 1.0002 & 1.0003 & 17.7578 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[a]")` & 2.4811 & 2.9927 & 7.9447 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[oː]")` & 3.0203 & 3.5151 & 14.4622 & $<$ 0.0001 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɔ]")` & 1.0001 & 1.0003 & 38.5056 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[uː]")` & 2.6925 & 3.2065 & 15.8756 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʊ]")` & 1.2141 & 1.3948 & 7.2875 & 0.0023 \\ 
  s(Talker) & 31.7596 & 35.0000 & 11782.6492 & $<$ 0.0001 \\ 
  s(Talker,category) & 699.8265 & 754.0000 & 88.8884 & 0.0001 \\ 
   \hline
\end{tabular}
\label{tab:m.F3.paired}
\end{table}

\begin{table}[ht]
\small
\caption{\label{tab:m-duration-paired}Model summary, coefficients for parametric and smooth terms for all 21 vowels predicting duration} 
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & 0.2247 & 0.0075 & 29.7842 & $<$ 0.0001 \\ 
    category `r linguisticsdown::cond_cmpl("[ɪ]")` & -0.1035 & 0.0051 & -20.3505 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[yː]")` & 0.0147 & 0.0051 & 2.8873 & 0.0039 \\
  category `r linguisticsdown::cond_cmpl("[ʏ]")` & -0.1047 & 0.0051 & -20.5962 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ʉː]")` & 0.0124 & 0.0051 & 2.4408 & 0.0147 \\  
  category `r linguisticsdown::cond_cmpl("[ɵ]")` & -0.0904 & 0.0051 & -17.7653 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[eː]")` & 0.0271 & 0.0051 & 5.3302 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[ɛ]")` & -0.0970 & 0.0051 & -19.1206 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[ɛː]")` & 0.0249 & 0.0051 & 4.8883 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[æː]")` & 0.0309 & 0.0051 & 6.0607 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æ]")` & -0.0673 & 0.0051 & -13.2361 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[øː]")` & 0.0220 & 0.0051 & 4.3213 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ø]")` & -0.0954 & 0.0051 & -18.7537 & $<$ 0.0001 \\   
  category `r linguisticsdown::cond_cmpl("[œː]")` & 0.0254 & 0.0051 & 4.9813 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[œ]")` & -0.0630 & 0.0051 & -12.3867 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɑː]")` & 0.0137 & 0.0051 & 2.6987 & 0.0070 \\  
  category `r linguisticsdown::cond_cmpl("[a]")` & -0.0876 & 0.0051 & -17.2311 & $<$ 0.0001 \\
  category `r linguisticsdown::cond_cmpl("[oː]")` & 0.0278 & 0.0051 & 5.4724 & $<$ 0.0001 \\  
  category `r linguisticsdown::cond_cmpl("[ɔ]")` & -0.0907 & 0.0051 & -17.8292 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[uː]")` & 0.0072 & 0.0051 & 1.4083 & 0.1590 \\ 
  category `r linguisticsdown::cond_cmpl("[ʊ]")` & -0.0898 & 0.0051 & -17.5237 & $<$ 0.0001 \\ 
  Gendermale & -0.0158 & 0.0099 & -1.5985 & 0.1099 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[iː]")` & 1.0003 & 1.0005 & 0.4070 & 0.5235 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɪ]")`  & 1.0001 & 1.0002 & 0.0001 & 0.9953 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 1.0002 & 1.0003 & 0.3187 & 0.5725 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʏ]")` & 1.0001 & 1.0002 & 0.0123 & 0.9121 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʉː]")` & 1.0001 & 1.0002 & 2.3733 & 0.1234 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɵ]")` & 1.0001 & 1.0003 & 0.0859 & 0.7696 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[eː]")` & 1.0002 & 1.0004 & 0.0092 & 0.9244 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛ]")` & 1.0002 & 1.0005 & 0.5153 & 0.4730 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛː]")` & 1.0002 & 1.0004 & 1.0377 & 0.3084 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æː]")` & 1.0001 & 1.0003 & 0.1511 & 0.6976 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æ]")` & 1.0002 & 1.0003 & 0.0611 & 0.8052 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[øː]")` & 1.0001 & 1.0002 & 0.1833 & 0.6686 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ø]")` & 1.0002 & 1.0004 & 1.1941 & 0.2746 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œː]")` & 1.0001 & 1.0003 & 0.2322 & 0.6300 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œ]")` & 1.0001 & 1.0002 & 0.1012 & 0.7504 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɑː]")` & 1.0002 & 1.0004 & 0.4794 & 0.4888 \\
  s(timepoint):category `r linguisticsdown::cond_cmpl("[a]")` & 1.0001 & 1.0002 & 0.1999 & 0.6549 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[oː]")` & 1.0001 & 1.0002 & 0.0503 & 0.8228 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɔ]")` & 1.0001 & 1.0002 & 0.0442 & 0.8337 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[uː]")` & 1.0004 & 1.0008 & 0.8059 & 0.3692 \\  
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʊ]")` & 1.0001 & 1.0002 & 0.0110 & 0.9169 \\ 
  s(Talker) & 34.1047 & 35.0000 & 127803.3103 & $<$ 0.0001 \\ 
  s(Talker,category) & 705.9825 & 754.0000 & 147.2948 & 0.0251 \\ 
   \hline
\end{tabular} 
\label{tab:m.duration.paired}
\end{table}

```{r}
```

#### Visualizations of GAMMs {#sec:SI-GAMMs-figs}
Figures \@ref(fig:SI-m-F1-paired), \@ref(fig:SI-m-F2-paired), \@ref(fig:SI-m-F3-paired) visualize the GAMMs fit to the long-short vowel pairs.

(ref:SI-m-F1-paired) Fitted smooths of GAMM for predicting F1 and 95% confidence intervals for all vowel pairs. For visualization purposes, the pairs are split into high vowels (**left panel**), and lower-mid vowels (**right panel**).

```{r SI-m-F1-paired, message=FALSE, fig.width=base.width*4, fig.height=base.height*1.5, fig.align='center', fig.pos="ht", fig.cap="(ref:SI-m-F1-paired)"}
par(mfrow=c(1,2))
itsadug::plot_smooth(m.F1.paired, view = "timepoint", plot_all = "category", cond = list(category = c("[iː]", "[ɪ]", "[yː]", "[ʏ]", "[ʉː]", "[ɵ]", "[eː]", "[ɛ]", "[oː]", "[ɔ]", "[uː]", "[ʊ]")), col = c("#66C2A5", "#1B9E77", "#FC8D62", "#D95F02", "#8DA0CB", "#7570B3", "#FB9A99", "#E78AC3", "#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C"), print.summary = FALSE, hide.label = T)
itsadug::plot_smooth(m.F1.paired, view = "timepoint", plot_all = "category", cond = list(category = c("[ɛː]", "[æː]", "[æ]", "[øː]", "[ø]", "[œː]", "[œ]", "[ɑː]", "[a]")), col = c("#E7298A", "#A6D854", "#66A61E", "#FFD92F", "#E6AB02", "#E5C494", "#A6761D", "#B3B3B3", "#666666"), print.summary = FALSE, hide.label = T)
```

(ref:SI-m-F2-paired) Fitted smooths of GAMM for predicting F2 and 95% confidence intervals for all vowel pairs. For visualization purposes, the pairs are split into high vowels (**left panel**), and lower-mid vowels (**right panel**).

```{r SI-m-F2-paired, message=FALSE, fig.width=base.width*4, fig.height=base.height*1.5, fig.align='center', fig.pos="ht", fig.cap="(ref:SI-m-F2-paired)"}
par(mfrow=c(1,2))
itsadug::plot_smooth(m.F2.paired, view = "timepoint", plot_all = "category", cond = list(category = c("[iː]", "[ɪ]", "[yː]", "[ʏ]", "[ʉː]", "[ɵ]", "[eː]", "[ɛ]", "[oː]", "[ɔ]", "[uː]", "[ʊ]")), col = c("#66C2A5", "#1B9E77", "#FC8D62", "#D95F02", "#8DA0CB", "#7570B3", "#FB9A99", "#E78AC3", "#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C"), print.summary = FALSE, hide.label = T)
itsadug::plot_smooth(m.F2.paired, view = "timepoint", plot_all = "category", cond = list(category = c("[ɛː]", "[æː]", "[æ]", "[øː]", "[ø]", "[œː]", "[œ]", "[ɑː]", "[a]")), col = c("#E7298A", "#A6D854", "#66A61E", "#FFD92F", "#E6AB02", "#E5C494", "#A6761D", "#B3B3B3", "#666666"), print.summary = FALSE, hide.label = T)
```

(ref:SI-m-F3-paired) Fitted smooths of GAMM for predicting F3 and 95% confidence intervals for all vowel pairs. For visualization purposes, the pairs are split into high vowels (**left panel**), and lower-mid vowels (**right panel**).

```{r SI-m-F3-paired, message=FALSE, fig.width=base.width*4, fig.height=base.height*1.5, fig.align='center', fig.pos="ht", fig.cap="(ref:m-F3-paired)"}
par(mfrow=c(1,2))
itsadug::plot_smooth(m.F3.paired, view = "timepoint", plot_all = "category", cond = list(category = c("[iː]", "[ɪ]", "[yː]", "[ʏ]", "[ʉː]", "[ɵ]", "[eː]", "[ɛ]", "[oː]", "[ɔ]", "[uː]", "[ʊ]")), col = c("#66C2A5", "#1B9E77", "#FC8D62", "#D95F02", "#8DA0CB", "#7570B3", "#FB9A99", "#E78AC3", "#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C"), print.summary = FALSE, hide.label = T)
itsadug::plot_smooth(m.F3.paired, view = "timepoint", plot_all = "category", cond = list(category = c("[ɛː]", "[æː]", "[æ]", "[øː]", "[ø]", "[œː]", "[œ]", "[ɑː]", "[a]")), col = c("#E7298A", "#A6D854", "#66A61E", "#FFD92F", "#E6AB02", "#E5C494", "#A6761D", "#B3B3B3", "#666666"), print.summary = FALSE, hide.label = T)
```
```{r}
# Code kept for reproducibility
#gamtabs(m.duration.paired, caption='Model summary, coefficients for parametric and smooth terms', label = "tab:m.duration.paired")
rm(m.F1.paired, m.F2.paired, m.F3.paired, m.duration.paired)
```
