---
title             : "The acoustic characteristics of Central Swedish vowels"
date              : "`r format(Sys.time(), '%B %d, %Y')`"

author:
  - name: Anna Persson
    affiliation: '1'
    corresponding: yes
    address: 'Department of Swedish Language and Multilingualism, Stockholm University, SE-106 91 Stockholm, Sweden'
    email: 'anna.persson@su.se'

affiliation:
  - id: '1'
    institution: Department of Swedish Language and Multilingualism, Stockholm University

wordcount         : "X"

bibliography      : ["latex-stuff/librarySweAcoust.bib", "latex-stuff/r-referencesSweAcoust.bib"]
link-citations    : yes
csl               : latex-stuff/apa-6th-edition.csl

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
numbersections    : false
mask              : no
draft             : no

header-includes:
 - \usepackage{animate}
 - \usepackage{amsmath}
 - \usepackage{tikz}
 - \usetikzlibrary{bayesnet}
 - \usepackage{booktabs}
 - \usepackage{siunitx}
 - \usepackage{soul}
 - \usepackage{tabto}
 - \usepackage{xcolor}
 - \usepackage{colortbl}
 - \usepackage{placeins}
 - \setstcolor{red}
 - \usepackage{sectsty}
 - \sectionfont{\color{black}}
 - \subsectionfont{\color{black}}
 - \subsubsectionfont{\color{black}}
 - \usepackage{setspace}\doublespacing
 - \usepackage{subfig}
 - \usepackage{lscape}
 - \usepackage{pdflscape}

documentclass     : "apa6"
classoption       : "man"
fontsize          : 11pt
output:
  papaja::apa6_pdf:
    latex_engine: xelatex
    extra_dependencies: "subfig"
    includes:
      in_header: latex-stuff/header.tex
always_allow_html: true
---

```{r libraries, include=FALSE, message=FALSE}
library(curl)             # Check availability of internet for install of remote libraries

# Install remotes package if necessary
if(!requireNamespace("remotes", quietly = TRUE)) if (has_internet()) install.packages("remotes")
# Install the stable development version from GitHub
#if (has_internet()) remotes::install_github("crsh/papaja")
library(papaja)           # APA formatted ms

library(tidyverse)        # data wrangling and plotting
library(magrittr)         # pipes, my friend, we need pipes
#library(assertthat)       # for error checking
library(rlang)            # quosures and unquoting
library(ggforce)          # correlation matrices
library(plotly)           # for interactive HTML plots
library(patchwork)       # for plotting many plots
library(mvtnorm)          # for identifying outliers
library(mgcv)             # for running generalized additive mixed models
library(gtsummary)        # for converting gamms to tables
library(itsadug)          # for plotting generalized additive mixed models
library(lme4)             # for running linear mixed-effects models

library(linguisticsdown)  # for inserting IPA symbols
library(kableExtra)       # for styling html format
library(modelr)
```

```{r setup, include=FALSE}
library(knitr)

# Set knit defaults for code chunks
opts_chunk$set(
  dev='jpeg', # default format of figures
  fig.ext="jpeg",
  comment="",
  echo=FALSE, warning=FALSE, message=FALSE,
  cache=TRUE)

# some useful formatting functions for output of knitting
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

color_block = function(color) {
  function(x, options) sprintf('\\color{%s}\\begin{verbatim}%s\\end{verbatim}\\color{black}', color, x)
}
knitr::knit_hooks$set(error = color_block('red'))
knitr::knit_hooks$set(warning = color_block('orange'))
```

```{r functions-and-pipes, include=FALSE}
source("constants_functions.R")

myGplot.defaults("paper")

options(knitr.table.format = "latex")
color.transform <- "#C9C0BB"
color.centering <- "#ABCDEF"
color.scaling <- "#DDADAF"

base.width = 2.5
base.height = 2.5
base_size = 10
```

# Abstract
The Swedish vowel space is a relatively densely populated space with 21 categories that differ in quality and quantity [@riad2014]. Existing descriptions of the entire space rest on recordings made in the late 1990s or earlier, while recent work in general have focused on sub-parts of the space [e.g., @engstrand1999; @fant1969; @kuronen2000; @pelzer-boersma2019; @riad2014; @schotz2011; @wenner2010]. The present paper reports on two studies. The first study presents static and dynamic acoustic analyses of the entire vowel space using a recently released database of h-VOWEL-d words [SwehVd; @persson-jaeger2023]. The second study compares the acoustic characteristics of SwehVd against formant data from 8 talkers of the same variety recorded in 1999 [@eriksson2004], in order to investigate the scope and spread of previously reported shifts in the phonetic realization of some categories [e.g., @bjorsten-engstrand1999; @pelzer-boersma2019; @riad2014]. This includes, for instance, the lowering and retraction of [`r linguisticsdown::cond_cmpl("ɛː")`] and the centralization of [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`]. The results highlight the importance of F1, F2, duration and formant dynamics for Swedish vowel category distinction, and indicate possibly ongoing shifts in the high front part of the space.

\newpage
\setcounter{page}{1}

# Introduction {#sec:intro}
The Swedish vowel inventory consists of 21 categories that differ in spectral cues (formant frequencies) and temporal cues (duration). It forms a typologically rather complex space, characterized by a systematic quantity distinction resulting in long and short vowel pairs (i.e., allophony by structural position) [@riad2014], 3 different levels of lip-rounding, and contextually conditioned allophones to /`r linguisticsdown::cond_cmpl("ɛ")`/ and /`r linguisticsdown::cond_cmpl("ø")`/ in position before /r/ or retroflex segments. Given the crowdedness of the space and resulting category overlap for some vowels, previous work has stressed the importance of additional cues besides the two primary determinants to vowel identity cross-linguistically---the first two formants, F1 and F2 [e.g., @joos1948; @Ladefoged1957; @nearey-assmann1986; @peterson1961]. For instance, the third formant (F3) for rounded vs. unrounded categories, and formant dynamics for several neighboring categories in the front part of the space [@kuronen2000; @pelzer-boersma2019]. This paper investigates the acoustic characteristics of modern-day Swedish vowels in two studies that aim to contribute to our understanding of language-specific and langue-general patterns of vowel acoustics. The first study presents a comprehensive description of the primary acoustic cues to vowel identity, using a recently released database of *h*-VOWEL-*d* (short: hVd) words recorded by 44 male and female talkers of Swedish, the SwehVd database [@persson-jaeger2023]. The second study assesses whether the Swedish vowel space have been submitted to changes over the last generation, by comparing the acoustics of SwehVd against that of 8 talkers of the same variety recorded in 1999 [@eriksson2004]. The variety investigated is Central Swedish, the regional standard variety of Swedish spoken in an area around and beyond Stockholm (eastern Svealand), including Mälardalssvenska, Sveamål, Uppsvenska, Mellansvenska [see e.g., @bruce2009; @elert1994; @riad2014].^[For the reader unfamiliar with Central Swedish, Section \@ref(sec:sweVowels) provides an overview of the acoustics of Central Swedish vowel space.] 

Existing descriptions of the entire space of 21 vowels rest on recordings made in the mid- or late $20^{th}$ century [e.g., @engstrand1999; @kuronen2000; @riad2014]. @leinonen2010's comprehensive study on dialectal (and social) variation across more than a 1000 rural locations did include reference talkers of Standard Swedish for comparison, however, XX categories were excluded from the analysis, making @kuronen2000's study the most recent mapping of the entire space. While the materials in @leinonen2010 were recorded around 1999, @kuronen2000 used recordings from 1982 [@leinonen1982]. Over the last two decades, the majority of studies on Central Swedish vowels have instead focused on sub-parts of the space, such as the long vowels for studies on diphthongization [@pelzer-boersma2019], two categories for merger studies [e.g., [`r linguisticsdown::cond_cmpl("ɵ")`] - [`r linguisticsdown::cond_cmpl("œ")`] in @wenner2010], or a single category, such the damped [`r linguisticsdown::cond_cmpl("ɨː")`] [@schotz2011]. A summary of a selection of previous work on Central Swedish vowels is presented in Table \@ref(tab:SI-previous-work) in the Supplementary Information (SI). These studies all provide detailed mappings of different parts of the space, and contribute with important insights into ongoing processes and shifts. However, given their focus on subsets of the space, acoustic mappings of the modern-day Central Swedish vowel space *in its entirety* are largely lacking. In addition, most previous studies differ in the materials used, in terms of the size of the database (e.g., number of talkers, number of repetitions per vowel), the demographics of talkers (e.g., male/female talkers, region of origin), and phonological contexts used for recording. For instance, the majority of previous studies have either not held the phonetic context constant across vowels, or have investigated vowel production out of context or in different CVC contexts (Table \@ref(tab:SI-previous-work)). This diversity restricts comparison across studies on Swedish, as well as cross-linguistically.

The choice of materials and methodological approach adopted in the current paper aims at expanding on and complementing previous work for a comprehensive picture of modern-day Central Swedish vowels. The first study provides an up-to-date acoustic description of the entire Central Swedish vowel space of 21 categories, using the SwehVd database [@persson-jaeger2023]. The use of an hVd database minimizes coarticulatory effects from the surrounding phonetic context, and increase cross-linguistic comparability [e.g., @hillenbrand1995; @Peterson1952]. The main spectral and temporal cues to vowel identity are reported in static analyses [following e.g., @engstrand1999; @fant1969], as well as dynamic analyses, given the well documented importance of formant dynamics on vowel production and perception [e.g., @assmann-katz2005; @eklund1997; @hillenbrand-nearey1999; @kuronen2000; @nearey-assmann1986]. In contrast with previous work investigating the dynamics of Central Swedish vowels, the present study includes analyses of both long and short vowels. General additive mixed models (GAMMs) are implemented to model formant dynamics. GAMMs are increasingly used in phonetic research, due to their suitability in modelling complex phonetic patterns without having to rely on simplifying assumptions of linearity or that vowels can be reduced to a single F1-F2 point estimate [@soskuthy2021]. The use of GAMMs to model formant dynamics and assess the relative weight of different acoustic cues for vowel identity, complements previous work on Central Swedish that have primarily used visual inspection, formant measurements and linear regression (Table \@ref(tab:SI-previous-work)).

Figure \@ref(fig:swe-vowels) shows the vowel space as represented by the talkers in SwehVd. Compared to many established descriptions of Central Swedish [e.g., @eklund1997; @engstrand1999; @fant1969; @kuronen2000; @riad2014], Figure \@ref(fig:swe-vowels) indicates changes in the phonetic realization of the Central Swedish vowels, similar to those found by @pelzer-boersma2019: for instance, [`r linguisticsdown::cond_cmpl("eː")`] is produced substantially higher, [`r linguisticsdown::cond_cmpl("ɛː")`] has lowered, while [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] have centralized<!--, presumably because they were produced with a buzzing off-glide sound, as the 'damped' [`r linguisticsdown::cond_cmpl("ɨː")`]-->. The aim of the second study is to assess the scope and spread of these changes over the last generation, by comparing the acoustic characteristics of SwehVd against vowel data from a database of 8 (female = 4) talkers of Central Swedish recorded in 1999 as part of the SweDia dialect project [e.g., @eriksson2004]. Potential vowel shifts are assessed qualitatively by visualizations, and quantitatively using a distance metric, the orthogonal projection ratio [@stevens-schiel2019].

The paper is organized as follows. A background to the acoustics of Central Swedish vowels is provided by a review of previous work, the general methodology of the paper is then presented, including a description of the materials used. This is followed by the two studies' study-specific methods and results, and finally, a discussion of the results and its consequences for the Central Swedish vowel system. 
All analyses and visualization code for this study can be found in an online repository (<https://osf.io/>).<!--fill in-->

```{r load-SwehVd}
# Load native Swedish vowel data, manually corrected, but with distributional outliers remaining (generated by Persson-2021-outlierCorrected)----NOW REPLACED WITH THE CORRECTED DATA IN SWEHVD-PAPER
d.SwehVd <-
  read_csv('../../data/phonetic vowel statistics/Swedish/Published databases/Persson-2021/outlier-corrected-vowel-data/Persson_2021_L1_vowels_wDistrOutliers.csv') %>%
  mutate(
    Database = "SwehVd")

# What proportion of the most extreme values should be considered outliers? 
# (if set to e.g., .05 that means that points with cumulative densities below
# .025 or above .975 are considered outliers)
outlier_probability_cutoff = .05

# To be conservative, and while waiting to complete measurement correction, remove all outliers outside of 0.5th to the 99.5th quantile for each category and talker
d.SwehVd %<>%
  obtain_densities() %>%
  filter(!is_outlier(cumulative_probability_F1F2)) %>%
  select(-c(cumulative_probability_F1F2, mean_F1F2, cov_F1F2))
```
```{r}
# Make temp exclusions file for identifying no of talkers to exclude
d.SwehVd.exclusions <-
  d.SwehVd %>%
  select(c(Talker, Gender, category, Transcribed_vowel, Token)) %>%
  distinct() %>%
  group_by(Talker, Gender, category) %>%
  summarise(
    N_tokens_per_category = n_distinct(Token),
    Transcribed_vowel = Transcribed_vowel) %>%
  filter(Transcribed_vowel != "targeted") %>%
  group_by(Talker, Gender, category, N_tokens_per_category) %>%
  summarise(
    N_unidentifiable_vowels = length(Transcribed_vowel)) %>%
  group_by(Talker, category, Gender) %>%
  mutate(
    exclude_Talker = case_when(N_tokens_per_category < 6 & N_unidentifiable_vowels > 2 ~ "TRUE",
                               N_tokens_per_category >= 10 & N_unidentifiable_vowels > 7 ~ "TRUE",
                               TRUE ~ "FALSE")) %>%
  filter(exclude_Talker == TRUE)
```

```{r filter-data-for-study}
# Exclude talkers that failed to produce *hodd* altogether + the mispronounced words by remaining talkers. This approach results in exclusion of an additional two talkers, given that they only did five recordings for each vowel. Also exclude measurements at 20 and 80% into the vowel, and all /hädd/ words since they are a duplicate of "[ɛ]" - the short allophone to /e/ neutralizes with the short allophone to /ɛ/. Also exclude all NAs in F0s.
d.SwehVd.forStudy <-
  d.SwehVd %>%
  filter(Transcribed_vowel == "targeted", Word != "hädd", !is.na(F0), Unreliable_measurement == FALSE) %>%
  select(-Transcribed_vowel) %>%
  rename(timepoint = Location) %>%
  group_by(Talker, Age, Gender, category, Word, Token, Trial, timepoint, Quantity, Database) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ geometric.mean(.x))) %>%
  ungroup()
```

```{r normalize-data}
# Normalize all data based on the respective training of each cross-validation group
d.SwehVd.forStudy %<>%
  #Add fold_type in order for norm function to work
  mutate(fold_type = "training") %>%
  group_map(
    .f = ~ apply_all_transformations_and_normalization(
      data = .x,
      normalize_based_on_fold_types = "training"),
    .keep = T) %>%
  reduce(bind_rows) %>%
  ungroup()

# Pivot data to long format for plotting and calculation of separability index
d.SwehVd.forStudy.long <-
  d.SwehVd.forStudy %>%
  select(-starts_with("formants")) %>%
  pivot_longer(
    cols = starts_with("F", ignore.case = FALSE),
    names_to = c("Cue", "Normalization.Scale", "Normalization.Type"),
    names_sep = "_",
    values_to = "Cue.Value") %>%
  pivot_wider(
    names_from = "Cue",
    values_from = "Cue.Value") %>%
  # Mutate both
  mutate(
    Normalization.Type = paste(Normalization.Type, Normalization.Scale, sep = "_")) %>%
  mutate(
    Normalization.Type = factor(plyr::mapvalues(Normalization.Type, levels.normalization, labels.normalization), levels = labels.normalization, ordered = T),
    Duration = ifelse(Normalization.Type %in% c("C-CuRE (Hz)", "C-CuRE (Mel)", "C-CuRE (ERB)", "C-CuRE (Bark)", "C-CuRE (semitones)"), Duration_CCuRE, Duration)) %>%
   select(-c(Duration_CCuRE, Normalization.Scale)) %>%
   relocate(Talker, Age, Gender, Normalization.Type, fold_type, category, Quantity, Token)

d.SwehVd.forStudy.long.static <- d.SwehVd.forStudy.long %>%
  filter(Normalization.Type %in% c("no normalization (Hz)", "Nearey's uniform scaling (log)"), timepoint %in% c(35, 50, 65)) %>%
  # Get the  mean F1-F2 across the five time points
  group_by(Normalization.Type, Talker, Gender, category, Token, Word, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x))) %>%
  ungroup()

d.SwehVd.forStudy.long.dynamic <- d.SwehVd.forStudy.long %>%
  filter(Normalization.Type %in% c("no normalization (Hz)", "Nearey's uniform scaling (log)"))
```

(ref:swe-vowels) The SwehVd vowel data in unnormalized Hertz (*left*) and Nearey's uniform scaling space (*right*), along the first two formants, F1 and F2. Points show recordings of each of the 21 Central Swedish vowels by 44 (24 female) L1 talkers in the database, averaged across the three middle time-points (at 35, 50, 65% into the vowel). Vowel labels are placed at the category mean across talkers. Long vowels are boldfaced. Vowels that mismatched intended label are excluded (`r signif((d.SwehVd %>% filter(Transcribed_vowel != "targeted") %>% nrow() / d.SwehVd %>% nrow()) * 100)`% of all recordings).

```{r swe-vowels, fig.width=base.width*4.5, fig.height=base.height*3.5, out.width='90%', fig.align='center', fig.cap="(ref:swe-vowels)"}
p <- d.SwehVd.forStudy.long.static %>%
  ggplot(
    aes(
      x = F2,
      y = F1)) +
  geom_point(
    aes(
      colour = category,
      shape = Gender),
    alpha = .4) +
  # Add between-talker mean
  geom_label(
    data =
      ~ .x %>%
      group_by(Normalization.Type, category, Quantity) %>%
      summarise(across(c(F1, F2), ~ mean(.x))),
    mapping = aes(
      colour = category,
      label = category,
      fontface = ifelse(category %in% levels.vowel.IPA.swe.long, 2, 1)),
    alpha = .8,
    size = 4,
    label.size = NA,
    label.padding = unit(.06, "cm")) +
  scale_colour_manual(name = "category", values = colors.vowel.swe) +
  scale_x_reverse("F2", position = "top", breaks = scales::breaks_pretty(6)) +
  scale_y_reverse("F1", position = "right", breaks = scales::breaks_pretty(6)) +
  guides(color = "none") +
  facet_wrap(~ Normalization.Type, scales = "free") +
  myGplot.defaults(base_size+2)  +
  theme(legend.position = "top")
p

```

## The acoustics of Central Swedish vowels {#sec:sweVowels}
The 9 vowel phonemes of Central Swedish are most often described as /`r linguisticsdown::cond_cmpl("i")`/, /`r linguisticsdown::cond_cmpl("y")`/, /`r linguisticsdown::cond_cmpl("ʉ")`/, /`r linguisticsdown::cond_cmpl("e")`/, /`r linguisticsdown::cond_cmpl("ɛ")`/, /`r linguisticsdown::cond_cmpl("ø")`/, /`r linguisticsdown::cond_cmpl("ɑ")`/, /`r linguisticsdown::cond_cmpl("o")`/, /`r linguisticsdown::cond_cmpl("u")`/. The long allophones are [`r linguisticsdown::cond_cmpl("iː")`], [`r linguisticsdown::cond_cmpl("yː")`], [`r linguisticsdown::cond_cmpl("ʉː")`], [`r linguisticsdown::cond_cmpl("eː")`], [`r linguisticsdown::cond_cmpl("ɛː")`], [`r linguisticsdown::cond_cmpl("øː")`], [`r linguisticsdown::cond_cmpl("ɑː")`], [`r linguisticsdown::cond_cmpl("oː")`], [`r linguisticsdown::cond_cmpl("uː")`], and the short allophones are [`r linguisticsdown::cond_cmpl("ɪ")`], [`r linguisticsdown::cond_cmpl("ʏ")`], [`r linguisticsdown::cond_cmpl("ɵ")`], [`r linguisticsdown::cond_cmpl("ɛ")`], [`r linguisticsdown::cond_cmpl("ø")`], [`r linguisticsdown::cond_cmpl("a")`], [`r linguisticsdown::cond_cmpl("ɔ")`], [`r linguisticsdown::cond_cmpl("ʊ")`]. The short allophones of /`r linguisticsdown::cond_cmpl("e")`/ and /`r linguisticsdown::cond_cmpl("ɛ")`/ neutralize as [`r linguisticsdown::cond_cmpl("ɛ")`] in Central Swedish, resulting in 17 vowels, rather than 18. Some talkers of Central Swedish furthermore neutralize the short /`r linguisticsdown::cond_cmpl("ø")`/ and /`r linguisticsdown::cond_cmpl("ʉ")`/ as [`r linguisticsdown::cond_cmpl("ɵ")`], primarily in position before a retroflex [see e.g., @stahle1965; @wenner2010]. In addition to these 17 vowels, there are 4 additional long and short allophones---[`r linguisticsdown::cond_cmpl("æː")`], [`r linguisticsdown::cond_cmpl("æ")`], [`r linguisticsdown::cond_cmpl("œː")`], and [`r linguisticsdown::cond_cmpl("œ")`], as /`r linguisticsdown::cond_cmpl("ɛ")`/ and /`r linguisticsdown::cond_cmpl("ø")`/ lower in position before /r/ or any retroflex segment [e.g., @kuronen2000; @riad2014]. Traditionally, Central Swedish has been described using four height levels and three backness levels. The long high vowels [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] are the most fronted vowels, [`r linguisticsdown::cond_cmpl("ʉː")`] being more central, and [`r linguisticsdown::cond_cmpl("uː")`] back. Mid-high vowels are the front [`r linguisticsdown::cond_cmpl("eː")`], and the back [`r linguisticsdown::cond_cmpl("oː")`]. Mid vowels are [`r linguisticsdown::cond_cmpl("ɛː")`], and [`r linguisticsdown::cond_cmpl("øː")`], both front. Finally, the contextually conditioned allophones [`r linguisticsdown::cond_cmpl("æː")`], and [`r linguisticsdown::cond_cmpl("œː")`], both front, form the low part of the vowel space together with the back [`r linguisticsdown::cond_cmpl("ɑː")`]. The short vowels overall occupy the same locations along the height-backness dimensions as their long counterparts, with the exceptions of [`r linguisticsdown::cond_cmpl("ɵ")`] being high-central, [`r linguisticsdown::cond_cmpl("œ")`] mid-central, [`r linguisticsdown::cond_cmpl("ɔ")`] mid-back, and [`r linguisticsdown::cond_cmpl("a")`] low-central. The vowel space can be considered quite crowded along the high part of the front-back dimension (Figure \@ref(fig:swe-vowels)). 

Central Swedish is furthermore defined by three levels of lip-rounding, where the rounded vowels are traditionally referred to as either inrounded, with an extreme narrowing of the lips---[`r linguisticsdown::cond_cmpl("ʉː")`] and [`r linguisticsdown::cond_cmpl("uː")`], or outrounded, with less degree of lip narrowing and more protruded lips---[`r linguisticsdown::cond_cmpl("yː")`], [`r linguisticsdown::cond_cmpl("øː")`], [`r linguisticsdown::cond_cmpl("oː")`] [e.g., @fant1971]. Previous work has stressed the importance of lip-rounding for some category distinctions, especially among closely located high vowels. For instance, [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] have been described as overlapping in F1-F2 space, but exhibiting increased separability when F3 is considered [e.g., @fant1959; @fant1969; @fujimura1967; @kuronen2000].

The 9 vowel pairs differ systematically in duration, with short-long vowel to vowel ratios reported to be on average .65-.67 for Central Swedish [e.g., @elert1964; @kuronen2000; @strangert2001]. However, spectral cues also convey information for vowel pair distinctions, with previous studies finding F2 to carry most of the durational variation [e.g., @kuronen2000; @lindblom1963]. The largest spectral differences between long and short vowel pairs have been found for [`r linguisticsdown::cond_cmpl("ʉː")`] - [`r linguisticsdown::cond_cmpl("ɵ")`], and [`r linguisticsdown::cond_cmpl("ɑː")`] - [`r linguisticsdown::cond_cmpl("a")`], and the smallest differences for [`r linguisticsdown::cond_cmpl("ɛː")`] - [`r linguisticsdown::cond_cmpl("ɛ")`], [`r linguisticsdown::cond_cmpl("øː")`] - [`r linguisticsdown::cond_cmpl("ø")`] [e.g., @kuronen2000]. For pairs with small spectral differences, duration is presumably of importance for category distinction. Perceptual studies have indeed confirmed that duration is primary cue for [`r linguisticsdown::cond_cmpl("ɛː")`] - [`r linguisticsdown::cond_cmpl("ɛ")`], [`r linguisticsdown::cond_cmpl("øː")`] - [`r linguisticsdown::cond_cmpl("ø")`], [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("ɪ")`], and [`r linguisticsdown::cond_cmpl("oː")`] - [`r linguisticsdown::cond_cmpl("ɔ")`] [e.g., @behne-kirk1997; @haddingkoch1964]. The extent to which *all* vowel pairs are more or less dependent on duration is less known.

Several of the long vowels in Central Swedish tend to diphthongize in their phonetic realization. Diphthongization is prosodically conditioned and is the strongest in stressed vowels [e.g., @bleckert1987; @leinonen2010].^[In general, true phonological diphthongs are not considered part of the phonological inventory of Central Swedish [@eliasson2022].] Previous studies have characterized the diphthongal glide in the later part of the long vowels as either a centralization of the vowel segment towards [`r linguisticsdown::cond_cmpl("ə")`] or a more open quality, or as a consonantal offglide [e.g., @elert1981; @elert2000; @fant1969; @fant1971; @kuronen2000; @mcallister1974; @pelzer-boersma2019; @riad2014]. Results are inconclusive as to how widespread diphthongization is across the vowel space, and what direction it takes. Most work has however found substantial diphthongization towards a more open quality for the mid and mid-high vowels [`r linguisticsdown::cond_cmpl("eː")`], [`r linguisticsdown::cond_cmpl("øː")`] and [`r linguisticsdown::cond_cmpl("oː")`] [e.g., @eklund1997; @elert2000; @fant1969; @pelzer-boersma2019]. In addition, results have shown that diphthongization cues category distinctions for several of the high vowels ([`r linguisticsdown::cond_cmpl("iː")`], [`r linguisticsdown::cond_cmpl("yː")`], [`r linguisticsdown::cond_cmpl("ʉː")`], and [`r linguisticsdown::cond_cmpl("uː")`]) [e.g., @fant1971]. For instance, @kuronen2000 concluded that [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] - [`r linguisticsdown::cond_cmpl("eː")`], and [`r linguisticsdown::cond_cmpl("uː")`] - [`r linguisticsdown::cond_cmpl("oː")`], differed in formant patterns only at later time-points of the vowel for some talkers in his study. He also found that the contrast between [`r linguisticsdown::cond_cmpl("ɛː")`] and [`r linguisticsdown::cond_cmpl("æː")`] was maintained solely by trajectory movements. Of importance for the present study, less is known about the formant dynamics in the short vowels, given the almost exclusive focus on the long vowels in studies on diphthongization.

Some talkers of Central Swedish realize [`r linguisticsdown::cond_cmpl("iː")`], [`r linguisticsdown::cond_cmpl("yː")`], [`r linguisticsdown::cond_cmpl("ʉː")`] and [`r linguisticsdown::cond_cmpl("uː")`] with a consonantal offglide, where the end-point of [`r linguisticsdown::cond_cmpl("iː")`] is described as a palatal approximant [`r linguisticsdown::cond_cmpl("j")`], the end-point of [`r linguisticsdown::cond_cmpl("yː")`] a voiced labio-palatal approximant [`r linguisticsdown::cond_cmpl("ɥ")`], the end-point of [`r linguisticsdown::cond_cmpl("ʉː")`] and [`r linguisticsdown::cond_cmpl("uː")`] a voiced bilabial fricative [`r linguisticsdown::cond_cmpl("β")`] [@hammarstrom-norman1957; @mcallister1974]. Furthermore, both [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] can be damped and produced with a buzzing sound, phonetically realized as [`r linguisticsdown::cond_cmpl("ɨː")`]. The damped [`r linguisticsdown::cond_cmpl("ɨː")`] is found in several dialects across Sweden, both in rural areas and in the cities of Gothenburg and Stockholm [@bjorsten-engstrand1999; @engstrand2000; @gross2020; @riad2014]. In work on Swedish dialectology, it is often referred to as the Viby-i, and in the Stockholm area, as the Lidingö-i. Acoustically, it manifests primarily as a lowering of F2, thus a more centralized position in the F1-F2 space.

From previous research, we know the Central Swedish vowel space as a dense space of 21 categories. Static analyses have mapped out the space primarily along F1 and F2, but also concluded the importance of F3 for lip-rounding and duration for long-short vowel pair distinctions. Formant dynamics has furthermore been found to cue several long vowels, but little is known about its importance for the short vowels. In addition, recent changes such as the lowering of [`r linguisticsdown::cond_cmpl("ɛː")`], and the centralization of [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] have been reported in previous work [e.g., @leinonen2010; @pelzer-boersma2019; @schotz2011], however, they have often been studied using a limited number of talkers [but see @pelzer-boersma2019] and/or without comparing against their short counterparts or against the entire vowel space. The present study submits the entire vowel space to the same analyses using a fairly large corpus of hVd words for a comprehensive and potentially more complete picture of modern-day Central Swedish vowels, while assessing the scope and spread of some previously reported changes in the space.

Next, the general methods shared by both studies are presented.

```{r}
d.SweDia2000 <- read_csv('../../data/phonetic vowel statistics/Swedish/Published databases/SweDia2000/outlier-corrected-vowel-data/SweDia2000_vowels_wDistrOutliers.csv') %>%
  mutate(
    Database = "SweDia",
    category = factor(
      plyr::mapvalues(
        category,
        levels.vowel.SweFA,
        levels.vowel.IPA.swe),
      levels = levels.vowel.IPA.swe),
    Quantity = ifelse(category %in% levels.vowel.IPA.swe.long, "long", "short")) %>%
  arrange(Quantity) %>%
  ungroup()

# Before fixing all outliers
d.SweDia2000 %<>% 
  obtain_densities() %>%
  filter(!is_outlier(cumulative_probability_F1F2)) %>%
  select(-c(cumulative_probability_F1F2, mean_F1F2, cov_F1F2))

d.SweDia2000.forStudy <- d.SweDia2000 %>%
  # Remove ÄRR as this category is only recorded by 2 talkers
  filter(!is.na(F0), category != "[æ]") %>%
  rename(timepoint = Location) %>%
  group_by(Database, Talker, Age, Gender, category, Word, Token, Trial, timepoint, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ geometric.mean(.x))) %>%
  ungroup()
  
# Normalize all data based on the respective training of each cross-validation group
d.SweDia2000.forStudy %<>%
  #Add fold_type in order for norm function to work
  mutate(fold_type = "training") %>%
  group_map(
    .f = ~ apply_all_transformations_and_normalization(
      data = .x,
      normalize_based_on_fold_types = "training"),
    .keep = T) %>%
  reduce(bind_rows) %>%
  ungroup()

# Pivot data to long format for plotting and calculation of separability index
d.SweDia2000.forStudy.long <-
  d.SweDia2000.forStudy %>%
  select(-starts_with("formants")) %>%
  pivot_longer(
    cols = starts_with("F", ignore.case = FALSE),
    names_to = c("Cue", "Normalization.Scale", "Normalization.Type"),
    names_sep = "_",
    values_to = "Cue.Value") %>%
  pivot_wider(
    names_from = "Cue",
    values_from = "Cue.Value") %>%
  # Mutate both
  mutate(
    Normalization.Type = paste(Normalization.Type, Normalization.Scale, sep = "_")) %>%
  mutate(
    Normalization.Type = factor(plyr::mapvalues(Normalization.Type, levels.normalization, labels.normalization), levels = labels.normalization, ordered = T),
    Duration = ifelse(Normalization.Type %in% c("C-CuRE (Hz)", "C-CuRE (Mel)", "C-CuRE (ERB)", "C-CuRE (Bark)", "C-CuRE (semitones)"), Duration_CCuRE, Duration)) %>%
   select(-c(Duration_CCuRE, Normalization.Scale)) %>%
   relocate(Talker, Age, Gender, Normalization.Type, fold_type, category, Quantity, Token)

d.Swehvd.Swedia <- d.SweDia2000.forStudy.long %>%
  filter(Normalization.Type %in% c("no normalization (Hz)", "Nearey's uniform scaling (log)"), timepoint %in% c(35, 50, 65)) %>%
  group_by(Database, Normalization.Type, Talker, category, Token, Word, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3), ~mean(.x))) %>%
     rbind(d.SwehVd.forStudy.long.static %>%
             # Remove the ÄRR category as this is not present in all talkers in SweDia2000
             filter(category != "[æ]") %>%
             mutate(
               Database = "SwehVd"))
```

# General methods {#sec:methods}
## Measuring acoustic cues to vowel identity
Following most vowel descriptions, this study reports acoustic measurements and visualizations of the first three formants (F1, F2, F3). While fundamental frequency (F0) is not considered an important cue to vowel identity in itself, it is known to vary between languages, dialects and speech styles [e.g., @henton2005; @jacewicz-fox2018; @mennen2012] and is therefore also reported. Measurements of vowel duration are included to capture the systematic quantity distinction between long-short vowel pairs in Central Swedish.

The Swedish version of the Montreal Forced Aligner developed by @young2021 was used to obtain estimates of word and segment boundaries. The boundaries were then manually corrected by the author (an L1-talker of Swedish). The formant analysis was carried out in Praat [@boersma-weenink2022], using the Burg algorithm to extract estimates of the first three formants (F1-F3) at five time-points of the vowel (20, 35, 50, 65, 80% into the vowel), while vowel duration and F0 were extracted across the entire vowel segment. The Burg algorithm was parameterized with a time step of 0.01 seconds, a window length of 0.025 seconds, and pre-emphasis was applied from 50 Hz. The maximum number of formants was set to 5, with a formant ceiling of 5500 Hz for the female talkers, and 5000 Hz for the male talkers.

In order to correct for measurement errors in the automatic extraction of cues, the joint multivariate distribution of the five extracted cues (F0, F1, F2, F3 and duration) was estimated for each distinct combination of talker and vowel. The identification of outliers was thus relative to the joint distribution of the five cues for that vowel and talker. Points that fell outside of the $0.5th$ to the $99.5th$ quantile of the multivariate Gaussian distribution for each vowel were identified, examined for measurement errors, and subsequently corrected. This followed the approach employed in @persson-jaeger2023, and strikes a middle-ground between the ideal (manual correction of all tokens) and feasibility.

### Vowel normalization
The raw formant values were transformed into a vowel normalized space using Nearey's uniform scaling account [@nearey1978]. Vowel normalization is used in vowel production and perception studies as a means to account for phonetically irrelevant talker-specific variation, such as adjusting for differences in anatomical structure (e.g., vocal tract size) [for reviews see e.g., @barreda-nearey2018; @johnson-sjerps2021; @stilp2020]. While most vowel normalization accounts were proposed as cognitive models of speech perception, in vowel production studies such as the present, normalization is primarily used as a methodological tool. The effect of normalization on the SwehVd data is visualized in Figure \@ref(fig:swe-vowels). Transforming the formant data into a normalized space reduces differences in F1 and F2 due to physiology, which can result in increased category separability, as visualized in Figure \@ref(fig:swe-vowels) (compare left and right panel). 

Previous work on Swedish has analyzed vowel data using primarily Bark-transformed data [@fant1983; @kuronen2000; @schotz2011; @wenner2010], Mel-transformed data [@lindblom1963], Lobanov-normalized data [@gross2020], or unnormalized raw Hertz [@bjorsten-engstrand1999; @fant1969; @pelzer-boersma2019]. The choice of Nearey's uniform scaling in the present study was motivated by its previous use in socio-phonetic research to describe and compare languages and varieties [e.g., @labov2001; @labov-boberg2005], and by its plausibility as perceptual model of how we come to achieve robust cross-talker perception, as it has provided a good fit against both production [e.g., @persson-jaeger2023; @syrdal1985] and perception data [e.g., @barreda2021; @persson2024]. Furthermore, its implementation closely resembles Vocal Tract Length normalization used in automatic speech recognition applications [e.g., @gouvea-stern1997].
Henceforth, all analyses presented were applied on the normalized vowel data. Formant measurements in Hertz are reported in the SI, Section \@ref(sec:SI-raw).

## Datasets {#sec:datasets}
### The SwehVd database {#sec:swehVd}
The SwehVd database is a corpus of Swedish hVd word recordings, collected by Anna Persson and Maryann Tan (Stockholm University) in 2020-2024, and previously described in @persson-jaeger2023. All recordings, annotations, and acoustic measurements are available on an OSF separate from the paper, at <https://osf.io/ruxnb/>. For this paper, an updated version of the database has been released, including an additional 20 male talkers. 
SwehVd was collected with the goal to characterize the contemporary Central Swedish spoken in the Greater Stockholm area or surroundings (provinces of Södermanland, Uppland), within and across talkers. It covers the entire monophthong inventory of Central Swedish, including all nine long vowels (*hid*, *hyd*, *hud*, *hed*, *häd*, *höd*, *had*, *håd*, *hod*), eight short vowels (*hidd*, *hydd*, *hudd*, *hedd*, *hädd*, *hödd*, *hadd*, *hådd*, *hodd*), and four additional allophones in the position before retroflex segments (*härd*, *härr*, *hörd*, *hörr*). SwehVd focuses on a single regional variety, providing high resolution within and across talkers for this variety with N = 10 recordings of each hVd word (for a total of 220 recordings for the 22 different hVd words) from each of the N = `r n_distinct(d.SwehVd$Talker)` talkers (N = `r nrow(d.SwehVd %>% filter(Gender == "female") %>% distinct(Talker))` female), a total N of tokens = `r nrow(d.SwehVd %>% distinct(Talker, category, Token))`. All talkers in the database were L1 talkers of Swedish, born and raised in the Greater Stockholm area, of 20-40 years of age (mean age = `r mean(d.SwehVd$Age) %>% round(digits = 0)`; SD = `r sd(d.SwehVd$Age) %>% round(digits = 2)`).

For the vast majority of talkers in the SwehVd, *hädd* productions elicited the same vowel as *hedd* (see SI, Figure \@ref(fig:SI-hedd)), which confirms the commonly held assumption that the short allophone of /e/ neutralizes with the short allophone of /`r linguisticsdown::cond_cmpl("ɛ")`/ in Central Swedish. In order to have a balanced number of tokens for each category, all *hädd* words were excluded from the subsetted SwehVd materials used in this study. Recordings of which the talker did not produce the targeted vowel were also excluded.^[The SwehVd database contains information on both targeted category and what category was actually produced.] This left data from `r n_distinct(d.SwehVd$Talker)` L1 talkers (N=`r (d.SwehVd %>% filter(Gender == "female") %>% summarise(n_distinct(Talker)) %>% pull())` female talkers), with on average `r d.SwehVd.forStudy %>% filter(timepoint == 50) %>% group_by(category) %>% tally() %>% summarise(x = mean(n)) %>% pull() %>% round(digits = 0)` (SD = `r d.SwehVd.forStudy %>% filter(timepoint == 50) %>% group_by(category) %>% tally() %>% summarise(x = sd(n)) %>% pull() %>% round(digits = 1)`) tokens per vowel (range = `r d.SwehVd.forStudy %>% filter(timepoint == 50) %>% group_by(category) %>% tally() %>% summarise(x = min(n)) %>% pull() %>% round(digits = 0)` to `r d.SwehVd.forStudy %>% filter(timepoint == 50) %>% group_by(category) %>% tally() %>% summarise(x = max(n)) %>% pull() %>% round(digits = 0)`), for a total of `r nrow(d.SwehVd.forStudy %>% filter(timepoint == 50))` observations. For more details on the recruitment, recording, pre-processing, segmentation and annotation procedure, see @persson-jaeger2023.

### Reference materials - SweDia
The materials used for comparison is a database of N = `r n_distinct(d.SweDia2000.forStudy$Talker)` L1 talkers (N = `r nrow(d.SweDia2000.forStudy %>% filter(Gender == "female") %>% distinct(Talker))` female) of Central Swedish that were recorded as reference talkers of Standard Swedish for the SweDia dialect database [e.g., @eriksson2004]. All talkers were L1 talkers of Swedish, born and raised in the Greater Stockholm area or surroundings (provinces of Södermanland, Uppland), and of 20-40 years of age (mean age = `r mean(d.SweDia2000.forStudy$Age, na.rm = T) %>% round(digits = 0)`; SD = `r sd(d.SweDia2000.forStudy$Age, na.rm = T) %>% round(digits = 2)`). The talkers were recorded reading a list of 41 words containing all 21 vowels of Central Swedish (for the complete wordlist used for recording, see Table \@ref(tab:word-list) in SI). Thirty-nine words on the list were monosyllabic, and two were bi-syllabic (*leta*, *flytta*). The target vowels extracted ([`r linguisticsdown::cond_cmpl("eː")`], [`r linguisticsdown::cond_cmpl("ʏ")`]) from the bisyllabic words had primary stress and were thus included in this study.

For the current purpose of serving as a reference point for the SwehVd materials, a subset of the SweDia database was used by excluding all [`r linguisticsdown::cond_cmpl("æ")`] vowels, as this vowel was recorded by only two of the talkers (leaving a total N of tokens = `r nrow(d.SweDia2000.forStudy %>% distinct(Talker, category, Token))`). All repetitions recorded by a talker were included, even though the variability in the number of repetitions for each vowel across the 8 talkers was substantial.^[Three to 8 repetitions of [`r linguisticsdown::cond_cmpl("iː")`], [`r linguisticsdown::cond_cmpl("yː")`], [`r linguisticsdown::cond_cmpl("ʉː")`], [`r linguisticsdown::cond_cmpl("ɵ")`], [`r linguisticsdown::cond_cmpl("eː")`], [`r linguisticsdown::cond_cmpl("æː")`], [`r linguisticsdown::cond_cmpl("ø")`], [`r linguisticsdown::cond_cmpl("œː")`], [`r linguisticsdown::cond_cmpl("œ")`], [`r linguisticsdown::cond_cmpl("ɑː")`], [`r linguisticsdown::cond_cmpl("a")`], [`r linguisticsdown::cond_cmpl("oː")`], [`r linguisticsdown::cond_cmpl("uː")`], [`r linguisticsdown::cond_cmpl("ʊ")`], 6 repetitions of [`r linguisticsdown::cond_cmpl("ɪ")`], [`r linguisticsdown::cond_cmpl("ʏ")`], [`r linguisticsdown::cond_cmpl("ʏ")`], [`r linguisticsdown::cond_cmpl("ɛ")`], 3 to 14 repetitions of [`r linguisticsdown::cond_cmpl("øː")`], [`r linguisticsdown::cond_cmpl("ɔ")`], [`r linguisticsdown::cond_cmpl("ɛː")`] per talker.] The phonological contexts used were the same across talkers (with two exceptions, see Table \@ref(tab:word-list)) but differed across categories. Importantly, none of the contexts were hVd. In order to (begin to) account for the difference in phonological context between datasets, a conservative approach was adopted in the statistical analyses by including word as random effect in the models in Study 2.

For details on the recruitment, recording, pre-processing and annotation procedure in the SweDia materials, see @eriksson2004.

# Study 1: Static and dynamic analyses of the spectral and temporal properties of Central Swedish {#sec:swehVd-static}
## Methods
The spectral acoustics of SwehVd were described in static and dynamic analyses of F0, F1, F2 and F3. The static analyses presented formant measurements at the steady state of the vowel, by averaging across the three mid-points, while the dynamic analyses reported formant measurements at all five time-points.^[The choice of time-point for extracting formants, or whether to average across several time-points, affects the acoustic characterizations given how formants move across the vowel segment. The SI presents evaluations of the effect of different measurement points. For increased comparability across studies in this paper, the steady state of the vowel was selected for both studies.] Formant dynamics was furthermore modeled by generalized additive mixed models (GAMMs; see below). The materials used in Study 1 is presented in Section \@ref(sec:swehVd).

### Generalized additive mixed models
The objective of modelling formant dynamics was to investigate patterns across formant trajectories and assess the un/certainty by which we can predict acoustic cues from categorical or continuous predictors, such as word contexts and talkers. We wanted to know to what extent vowels can be distinguished by certain acoustic cues, while accounting for by-talker variability and fluctuations in formants across the vowel segment. Modelling formant dynamics thus implies that the analysis had not only a temporal structure (formant measurements at different time-points), but also a hierarchical structure, as individual measurements were grouped into contours that in turn were grouped by words and talkers. 

Both the complex dependencies generated by this structure as well as the non-monotonic behavior of formants, cannot easily be captured by traditional parametric methods, such as linear regression [e.g., @soskuthy2021]. GAMMs allow for complex models that can accommodate nonlinear relationships between predictors and response variables, and their use of smooth functions can provide more nuanced representations of these relationships. Much like other mixed models, GAMMs allow for the inclusion of random effects, which permits the model to account for dependencies in the grouped data. GAMMs have previously been used in studies on vowels in different English varieties, e.g., on /u/-fronting in Derby English [@soskuthy2018], on the front vowel system of Southern American English [@renwick-stanley2020], but to the best of my knowledge, they have yet not be implemented in studies of Swedish vowels.

In the present study, GAMMs were used for assessing the formant dynamics across all 21 categories, as well as contrasting long and short vowels in each pair against each other. The GAMMs were fit separately for each of the formants, which necessarily meant committing to the simplifying assumption of cue independence. Previous work has shown that acoustic cues tend to co-vary [for a review, see @schertz-clare2019]. For vowels, this is the case for F1 and F2, as shown by the shape and orientation of ellipses in Figure \@ref(fig:swe-vowels-by-quantity).

The \texttt{bam()} function from the \texttt{R} package \texttt{mgcv} was implemented to fit the GAMMs.

Next, the results of the static and dynamic analyses are presented.

(ref:swe-vowels-by-quantity) The SwehVd vowel data separated by quantity. Ellipses show bivariate Gaussian 95% confidence interval of category means. Vowel labels indicate category means across female and male talkers.

```{r swe-vowels-by-quantity, fig.width=base.width*4.5, fig.height=base.height*2.5, fig.align='center', fig.cap="(ref:swe-vowels-by-quantity)"}
#Generate a subset of d-measurements manually
d.F2d <- read_csv2('../../data/final-d.csv')

# Vowel plot by quantity
p.normalized <- d.SwehVd.forStudy.long.static %>%
  # Filter down to test folds
  filter(Normalization.Type == "Nearey's uniform scaling (log)") %>%
  ggplot(
    aes(
      x = F2, 
      y = F1)) +
  stat_ellipse(
    mapping = aes(
      colour = category),
    alpha = .5) +
  # Add between-talker mean
  geom_label(
    data =
      . %>%
      group_by(category, Quantity, Normalization.Type) %>%
      summarise(across(c(F1, F2), ~ mean(.x))),
    mapping = aes(
      colour = category,
      label = category,
      #fontface = ifelse(category %in% levels.vowel.IPA.swe.long, 2, 1)),
      fontface = ifelse(Quantity == "long", 2, 1)),
    alpha = .8,
    size = 4,
    label.size = NA,
    label.padding = unit(.06, "cm")) +
  scale_colour_manual(name = "category", values = colors.vowel.swe) +
  #scale_fill_manual(name = "category", values = colors.vowel.swe) +
  # scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 3), limits = c(2500, 400)) +
  # scale_y_reverse("F1", position = "right", scales::pretty_breaks(n = 8), limits = c(950, 200)) +
  scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 4)) +
  scale_y_reverse("F1", position = "right", scales::pretty_breaks(n = 4)) +
  guides(alpha = "none", color = "none", shape = "none") +
  facet_grid(factor(Normalization.Type, levels = labels.normalization) ~ Quantity)
p.normalized
```
```{r}
# ## Add code to write csv.file for gamm in munich
# write.csv(
#   d.SwehVd %>%
#     filter(Transcribed_vowel == "targeted", Word != "hädd") %>%
#     mutate(
#       rounding = case_when(category %in% c("[yː]", "[ʏ]", "[øː]", "[ø]", "[œː]", "[æ]") ~ "protruded",
#                            category %in% c("[oː]", "[ɔ]", "[uː]", "[ʊ]", "[ʉː]") ~ "inrounded",
#                            TRUE ~ "unrounded")) %>%
#     mutate(across(starts_with("F"), ~ log(.x), .names = "{.col}_log")),
#   file = "../../../data/phonetic vowel statistics/Swedish/Published databases/Persson-2021/outlier-corrected-vowel-data/Persson_2023_SwehVd_vowels.csv",
#   row.names = FALSE, quote = FALSE)

d.cue.values <- d.SwehVd %>%
  filter(Transcribed_vowel == "targeted", Word != "hädd") %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ geometric.mean(.x))) %>%
  group_by(Quantity, Gender, category) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), list(geometric.mean, sd), .names = "{.col}.fn{.fn}")) %>%
  mutate(across(where(is.numeric), ~ round(.x, digits = 2))) %>%
  rename_with(~ sub(".fn1$", ".mean", .x), ends_with(".fn1")) %>%
  rename_with(~ sub(".fn2$", ".sd", .x), ends_with(".fn2"))
```

## Results
### Static spectral and temporal cues to vowel identity
Figure \@ref(fig:swe-vowels-by-quantity), left panel, visualizes the long vowels along the two primary cues to vowel identity, F1-F2.^[The SI presents the mean cue values for the male and female talkers, Tables \@ref(tab:SI-cue-values-long) and \@ref(tab:SI-cue-values-short). As expected, the male talkers have lower formant values and lower F0s than the female talkers (average F0 across long and short categories for female talkers = `r round(d.cue.values %>% filter(Gender == "female") %>% ungroup() %>% summarise(mean = mean(F0.mean)) %>% pull(mean), digits = 0)`, for male talkers = `r round(d.cue.values %>% filter(Gender == "male") %>% ungroup() %>% summarise(mean = mean(F0.mean)) %>% pull(mean), digits = 0)`), which reflect the average differences in anatomical structure.] Four vowels cluster in the high front part of the space. The mid-high [`r linguisticsdown::cond_cmpl("eː")`] occupies a substantially higher position than in many previous descriptions, and is also the most fronted vowel [c.f., @fant1969; @kuronen2000; but see @engstrand2000; @pelzer-boersma2019]. The high [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] are more mid-central than front, and almost completely overlapping. There is a rather long gap between [`r linguisticsdown::cond_cmpl("ʉː")`] and the back vowel [`r linguisticsdown::cond_cmpl("uː")`], which is at about the same height as [`r linguisticsdown::cond_cmpl("eː")`]. Located slightly below [`r linguisticsdown::cond_cmpl("uː")`] is the mid-high back vowel [`r linguisticsdown::cond_cmpl("oː")`], partly overlapping [`r linguisticsdown::cond_cmpl("uː")`]. The mid [`r linguisticsdown::cond_cmpl("øː")`] occupies the center part of the space, together with its contextually conditioned allophone [`r linguisticsdown::cond_cmpl("œː")`], placed slightly higher and further back. The low [`r linguisticsdown::cond_cmpl("ɑː")`] occupies the mid back location. The front [`r linguisticsdown::cond_cmpl("æː")`] is the lowest vowel in the space, placed lower and slightly further back from the main allophone [`r linguisticsdown::cond_cmpl("ɛː")`]. The long allophones of /`r linguisticsdown::cond_cmpl("ø")`/ are relatively separated from each other, whereas the long allophones of /`r linguisticsdown::cond_cmpl("ɛ")`/ display more overlap.

The short vowels, Figure \@ref(fig:swe-vowels-by-quantity) right panel, present a slightly more compact space, however with increased category separability and less category overlap [c.f., @riad2014]. The high [`r linguisticsdown::cond_cmpl("ɪ")`] is the most fronted vowel, followed closely by the partly overlapping [`r linguisticsdown::cond_cmpl("ʏ")`]. These vowels are more fronted than their long counterparts, which does not replicate previous studies on Central Swedish including both the long and short vowels [e.g., @fant1971; @kuronen2000].<!--check leinonen--> Between [`r linguisticsdown::cond_cmpl("ʏ")`] and [`r linguisticsdown::cond_cmpl("ɵ")`], there is a substantial gap---[`r linguisticsdown::cond_cmpl("ɵ")`] is thus clearly more central than its long counterpart [`r linguisticsdown::cond_cmpl("ʉː")`]. The top-right corner vowel is [`r linguisticsdown::cond_cmpl("ʊ")`]. The mid-high part of the space is quite crowded, with three closely located front and center vowels, [`r linguisticsdown::cond_cmpl("ɛ")`], [`r linguisticsdown::cond_cmpl("ø")`] and [`r linguisticsdown::cond_cmpl("œ")`], and one mid back vowel [`r linguisticsdown::cond_cmpl("ɔ")`]. The lower-mid vowels [`r linguisticsdown::cond_cmpl("æ")`] and [`r linguisticsdown::cond_cmpl("a")`] are front and center, respectively. The contextually conditioned allophones of /`r linguisticsdown::cond_cmpl("ɛ")`/ and /`r linguisticsdown::cond_cmpl("ø")`/ occupy similar positions as their long counterparts. That is, the short [`r linguisticsdown::cond_cmpl("æ")`] is located below and further back in relation to [`r linguisticsdown::cond_cmpl("ɛ")`], and [`r linguisticsdown::cond_cmpl("œ")`] slightly above and further back in relation to [`r linguisticsdown::cond_cmpl("ø")`]. However, the distance between the short [`r linguisticsdown::cond_cmpl("ɛ")`] and [`r linguisticsdown::cond_cmpl("æ")`] is larger than for the long [`r linguisticsdown::cond_cmpl("ɛː")`] and [`r linguisticsdown::cond_cmpl("æː")`], [`r linguisticsdown::cond_cmpl("ɛ")`] being mid while [`r linguisticsdown::cond_cmpl("æ")`] being clearly low.

(ref:swe-vowels-all-cues) The SwehVd vowel data shown for all pairwise combinations of five cues: F0, F1, F2, F3 and duration. Panels on the diagonal show marginal cue densities of all five cues. The off-diagonal panels show category means across talkers, represented by points and with bivariate Gaussian 95% probability mass ellipses in the upper panels, and represented by vowel labels and with points for each recording in the lower panels. Note that, unlike in Figure \@ref(fig:swe-vowels), axis directions are not reversed.

```{r swe-vowels-all-cues, fig.width=base.width*4.5, fig.height=base.height*4.5, out.width = '100%', fig.align='center', fig.cap="(ref:swe-vowels-all-cues)"}
p.matrix.cues <-
  d.SwehVd.forStudy.long.static %>%
  filter(F3 > .4) %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)") %>%
  ggplot(
    aes(
      x = .panel_x,
      y = .panel_y)) +
  geom_point(
    aes(
      colour = category,
      shape = Quantity),
    alpha = .1) +
  geom_label(
    data =
      ~ .x %>%
      group_by(category, Quantity) %>%
      summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x))),
    mapping = aes(
      colour = category,
      label = category,
      fontface = ifelse(category %in% levels.vowel.IPA.swe.long, 2, 1),
      fontface = 2),
    alpha = .4, size = 4, label.size = NA) +
  geom_autodensity(
    mapping = aes(fill = category, linetype = Quantity),
    alpha = .3, position = "identity") +
  stat_ellipse(
    mapping = aes(
      colour = category,
      linetype = Quantity),
    alpha = .4) +
  geom_point(
    data =
      ~ .x %>%
      group_by(category, Quantity) %>%
      summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x))),
    mapping = aes(
      colour = category,
      shape = Quantity),
    alpha = .8, size = 1) +
  scale_colour_manual(name = "category", values = colors.vowel.swe, aesthetics = c("colour", "fill")) +
  guides(color = "none", fill = "none", shape = "none", linetype = "none") +
  #guides(color = "none", fill = "none", shape = guide_legend(override.aes = list(alpha = .4))) +
  facet_matrix(
    vars(F0, F1, F2, F3, Duration),
    layer.lower = c(1,2),
    layer.diag = 3,
    layer.upper = c(4, 5)) +
  myGplot.defaults(base_size+2)
p.matrix.cues
```

For the pairwise combinations of the five spectral and temporal cues---F0, F1, F2, F3 and duration, see Figure \@ref(fig:swe-vowels-all-cues). The densities along the diagonal suggest that F0 carries the least information about vowel identity, exhibiting less between-category separation than all other cues. This is not surprising given that F0 is not considered an important cue to vowel identity.

As is to be expected, vowels differing in quality are most separated in the F1-F2 panels. The F1-F3 and F3-F2 panels both display increased separation between the neighbouring [`r linguisticsdown::cond_cmpl("yː")`] - [`r linguisticsdown::cond_cmpl("ʉː")`] and [`r linguisticsdown::cond_cmpl("ɪ")`] - [`r linguisticsdown::cond_cmpl("ʏ")`], compared to when plotted along F1-F2, which points to the importance of F3 for rounded vowels. Interestingly, the almost complete overlap between [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] in F1-F2 space overall remains when F3 is considered, even if some individual differences in the amount of overlap exist. Most talkers produce these two vowels very close in F1-F2 space, and only slightly separated in F2-F3 space, while others display a continued overlap when considering F3. This would seem to suggest that F3 might carry less importance as distinctive feature for [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] than previously established, indicating either a need for additional cues for a stable category distinction, or a possible merger in process [c.f., @fant1959; @fant1969]. For reference, one talker of each type are displayed in Figure \@ref(fig:swe-merge).

(ref:swe-merge) The effect of including F3 for the [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] distinction. **Panel A** shows two example talkers with overlapping [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] in F1-F2 space. **Panel B** shows how considering F3 affects the space: for talker SW_005, there is very little effect of F3, thus maintained [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] overlap, while for talker SW_015, separability increases.

```{r swe-merge, fig.width=base.width*2.5, fig.height=base.height*2.5, out.width='80%', fig.align='center', fig.cap="(ref:swe-merge)", message=FALSE}
p.merge <- d.SwehVd.forStudy.long.static %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", Quantity == "long", Talker %in% c("SW_005", "SW_015")) %>%
  ggplot(
    aes(
      x = F2, 
      y = F1)) +
  geom_point(
    aes(
      colour = category,
      shape = Quantity), alpha = 0.5, size = 1.5) + 
  # Add within-talker mean
  geom_label(
    data =
      . %>%
      group_by(category, Quantity, Normalization.Type, Talker) %>%
      summarise(across(c(F1, F2), ~ mean(.x))),
    mapping = aes(
      colour = category,
      label = category,
      fontface = ifelse(category %in% c("[iː]", "[yː]"), 2, 1)),
    size = 4,
    alpha = .4,
    label.size = NA,
    label.padding = unit(.06, "cm")) +
  scale_colour_manual(name = "category", values = colors.vowel.swe) +
  scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 4)) +
  scale_y_reverse("F1", position = "right", scales::pretty_breaks(n = 4)) +
  guides(alpha = "none", color = "none", shape = "none", size = "none") +
  facet_wrap(~ Talker, ncol=2) 

p.merge.F3 <-  d.SwehVd.forStudy.long.static %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", Quantity == "long", Talker %in% c("SW_005", "SW_015")) %>%
  ggplot(
    aes(
      x = F2, 
      y = F3)) +
  geom_point(
    aes(
      colour = category,
      shape = Quantity), alpha = 0.5, size = 1.5) + 
  # Add within-talker mean
  geom_label(
    data =
      . %>%
      group_by(category, Quantity, Normalization.Type, Talker) %>%
      summarise(across(c(F3, F2), ~ mean(.x))),
    mapping = aes(
      colour = category,
      label = category,
      fontface = ifelse(category %in% c("[iː]", "[yː]"), 2, 1)),
    size = 4,
    alpha = .4,
    label.size = NA, 
    label.padding = unit(.06, "cm")) +
  scale_colour_manual(name = "category", values = colors.vowel.swe) +
  scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 4)) +
  scale_y_reverse("F3", position = "right", scales::pretty_breaks(n = 4)) +
  guides(alpha = "none", color = "none", shape = "none", size = "none") +
  facet_wrap(~ Talker, ncol=2)
  
p.merge / p.merge.F3 + theme(axis.title.x = element_blank()) +
  plot_annotation(tag_levels = 'A') &
  theme(legend.margin = margin(0, 0, 0, 0, unit = "pt"),
        plot.tag = element_text(face = "bold"))
```

As expected, long-short vowel pairs differ systematically in duration (Figure \@ref(fig:duration-densities)). For each vowel pair, the duration densities are overlapping but with two clearly separable peaks (mean duration for the long vowels = `r round(d.SwehVd.forStudy.long %>% filter(Quantity == "long") %>% summarise(mean(Duration)) %>% pull(), digits = 2)` ms, SD = `r round(d.SwehVd.forStudy.long %>% filter(Quantity == "long") %>% summarise(sd(Duration)) %>% pull(), digits = 2)`; mean duration for the short vowels = `r round(d.SwehVd.forStudy.long %>% filter(Quantity == "short") %>% summarise(mean(Duration)) %>% pull(), digits = 2)` ms, SD = `r round(d.SwehVd.forStudy.long %>% filter(Quantity == "short") %>% summarise(sd(Duration)) %>% pull(), digits = 2)`). Overall, the short vowels display less variability in duration than the long vowels, a common pattern for measures with a lower bound.

(ref:duration-densities) Category densities along duration illustrates systematic differences in duration between the long and short vowel pairs.

```{r duration-densities, fig.width= base.width * 4, fig.height = base.height * 2.5, fig.align='center', out.width='95%', fig.cap="(ref:duration-densities)"}
# Generate dataframe with vowel pairs
d.SwehVd.vowel_pairs <- d.SwehVd.forStudy.long.dynamic %>%
  select(-starts_with("formant")) %>%
  #filter(F3 > .25) %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)")  %>%
  mutate(
    vowel_pair = case_when(category %in% c("[ɪ]", "[iː]") ~ "ɪ - iː",
                           category %in% c("[ʏ]", "[yː]") ~ "ʏ - yː",
                           category %in% c("[ɵ]", "[ʉː]") ~ "ɵ - ʉː",
                           category %in% c("[ɛ]", "[eː]", "[ɛː]") ~ "ɛ - eː, ɛː",
                           category %in% c("[æ]", "[æː]") ~ "æ - æː",
                           category %in% c("[ø]", "[øː]") ~ "ø - øː",
                           category %in% c("[œ]", "[œː]") ~ "œ - œː",
                           category %in% c("[a]", "[ɑː]") ~ "a - ɑː",
                           category %in% c("[ɔ]", "[oː]") ~ "ɔ - oː",
                           category %in% c("[ʊ]", "[uː]") ~ "ʊ - uː")) %>%
  # Do additional sets for the three-way contrast
  rbind(
    d.SwehVd.forStudy.long.dynamic %>%
      select(-starts_with("formant")) %>%
      filter(Normalization.Type == "Nearey's uniform scaling (log)", category %in% c("[ɛ]", "[eː]")) %>%
      mutate(
        vowel_pair = "ɛ - eː"),
    d.SwehVd.forStudy.long.dynamic %>%
      select(-starts_with("formant")) %>%
      filter(Normalization.Type == "Nearey's uniform scaling (log)", category %in% c("[ɛ]", "[ɛː]")) %>%
      mutate(
        vowel_pair = "ɛ - ɛː"))

d.SwehVd.vowel_pairs %>%
  filter(!vowel_pair %in% c("ɛ - eː", "ɛ - ɛː")) %>%
  ggplot(
    aes(Duration, fill = factor(category))) +
  geom_density(alpha = .5) +
  theme(legend.position = "none") +
  scale_colour_manual(name = "category", values = colors.vowel.swe, aesthetics = c("colour", "fill")) +
  guides(linetype = "none", color = "none", fill = "none") +
  facet_wrap(~ factor(vowel_pair, levels = c("ɪ - iː", "ʏ - yː", "ɵ - ʉː", "ɛ - eː, ɛː", "æ - æː", "ø - øː", "œ - œː", "a - ɑː", "ɔ - oː", "ʊ - uː")), nrow = 2)
```

All long-short vowel pairs display spectral differences in F1-F2. In fact, as indicated in Figure \@ref(fig:swe-vowels), formant differences are apparent for *all* vowel pairs, even for some category distinctions for which duration has been found to be the primary cue---[`r linguisticsdown::cond_cmpl("ɛː")`] - [`r linguisticsdown::cond_cmpl("ɛ")`], [`r linguisticsdown::cond_cmpl("øː")`] - [`r linguisticsdown::cond_cmpl("ø")`], [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("ɪ")`], and [`r linguisticsdown::cond_cmpl("oː")`] - [`r linguisticsdown::cond_cmpl("ɔ")`] [e.g., @behne-kirk1997; @haddingkoch1964; @kuronen2000]. The vowel pairs that display larger spectral differences along F1-F2 seem to be [`r linguisticsdown::cond_cmpl("ʉː")`] - [`r linguisticsdown::cond_cmpl("ɵ")`] and [`r linguisticsdown::cond_cmpl("ɑː")`] - [`r linguisticsdown::cond_cmpl("a")`] [in line with e.g., @fant1983; @kuronen2000], but also [`r linguisticsdown::cond_cmpl("ɛː")`] - [`r linguisticsdown::cond_cmpl("ɛ")`], which contrasts with previous studies. The large spectral differences in [`r linguisticsdown::cond_cmpl("ɛː")`] - [`r linguisticsdown::cond_cmpl("ɛ")`] are presumably due to [`r linguisticsdown::cond_cmpl("ɛː")`] being produced very low in the SwehVd materials, which leads to a gap along the top-left to bottom-left diagonal between [`r linguisticsdown::cond_cmpl("eː")`] and [`r linguisticsdown::cond_cmpl("ɛ")`]. Overall, F2 appears to carry more of the spectral variation between the long and short vowel phonemes, as categories display increased separability in the pairwise combination of F2 and duration (Figure \@ref(fig:swe-vowels-all-cues), rightmost column, third row).

### Dynamic spectral analysis {#sec:swehVd-dynamics}
The dynamic analysis of the spectral acoustics of the 21 vowels begins with visualizations of the empirical data in formant trajectory plots. The spectral cues are then modeled with GAMMs, first focusing on the movements in all 21 categories, then on spectral differences between the long and short vowel pairs.

#### Formant movements across the space
Figure \@ref(fig:trajectories) displays the formant trajectories across all 5 time-points for the long and the short vowels. There is movement across time-points for the majority of vowels---importantly, both long *and* short. The scope and direction however vary. Considering the scope of movements across vowels, they appear to be larger moving from vowel mid-point to 80% into the vowel, as indicated by the length of the line from category label to end of arrow. Most of the formant dynamics thus take place *after* vowel mid-point. The long high front vowels are important exceptions---the dynamics in [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] mostly occur at the beginning of the vowel segment (between 20 and 50% into the vowel), whereas [`r linguisticsdown::cond_cmpl("ʉː")`] displays movements of almost equal magnitude across the first four time-points.

(ref:trajectories) The trajectory of all vowels across the five time points, along F1-F2. The arrow indicates the direction of the trajectory and ends at the final time-point, at 80% into the vowel. The vowel label is placed at the third time-point, at vowel mid-point (50%). The first (20%), second (35%) and forth (65%) time-points are represented by points.

```{r trajectories, fig.width=base.width*3.5, fig.height=base.height*2, out.width='90%', fig.align='center', fig.cap="(ref:trajectories)"}
p.trajectories <- d.SwehVd.forStudy.long.dynamic %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)") %>%
  mutate(timepoint = as.character(timepoint)) %>%
  group_by(Talker, category, timepoint, Quantity) %>%
  summarise(across(c(F1, F2), ~mean(.x))) %>%
  group_by(category, timepoint, Quantity) %>%
  summarise(across(c(F1, F2), ~mean(.x))) %>%
  ggplot(
    aes(
      x = F2,
      y = F1,
      color = category,
      group = NULL)) +
  geom_point(
    data = . %>%
      filter(!timepoint %in% c("50", "80")),
    alpha = .3) +
  geom_path(
    alpha = .5,
    arrow = arrow(ends = "last", type = "closed", angle = "22.5", length = unit(.2, "cm"))) +
  geom_label(
    data = . %>%
      filter(timepoint == "50"),
    mapping = aes(
      colour = category,
      label = category,
      fontface = ifelse(category %in% levels.vowel.IPA.swe.long, 2, 1)),
    alpha = .6,
    size = 4,
    label.size = NA,
    label.padding = unit(.01, "cm")) +
  scale_color_manual(name = "category", values = colors.vowel.swe) +
  scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 6)) +
  scale_y_reverse("F1", position = "right", scales::pretty_breaks(n = 6)) +
  guides(color = "none") +
  facet_wrap(~Quantity, labeller = label_wrap_gen(multi_line=FALSE))
p.trajectories
# d.SwehVd %>%
#   filter(Transcribed_vowel == "targeted", Word != "hädd") %>%
#   mutate(timepoint = as.character(timepoint), 
#          Token = as.factor(Token),
#          red_speaker = Talker == "SW_015") %>%
#   group_by(Talker, category, timepoint, red_speaker) %>%
#   summarise(across(c(F1, F2, F3), ~geometric.mean(.x))) %>%
#   ggplot(
#     aes(x = timepoint, y = F3, group = Talker, color = red_speaker)) +
#   geom_line() +
#   facet_wrap(~category)
#ggsave(file = "../../figures/p.png", width = 9, height = 5, dpi=300)
```

In terms of directionality, there was a general tendency of movement towards a more central vowel quality in most of the vowels, both long and short, as well as a clear diphthongization in [`r linguisticsdown::cond_cmpl("eː")`] and [`r linguisticsdown::cond_cmpl("oː")`], which replicates previous work [e.g., @eklund1997; @elert2000; @pelzer-boersma2019]. According to previous studies [e.g., @bleckert1987; @elert2000], the high vowels [`r linguisticsdown::cond_cmpl("iː")`], [`r linguisticsdown::cond_cmpl("yː")`], [`r linguisticsdown::cond_cmpl("ʉː")`] and [`r linguisticsdown::cond_cmpl("uː")`] tend to be realized with an offglide, which would generate a falling F1 for all four vowels, a rising F2 for [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] and a falling F2 for [`r linguisticsdown::cond_cmpl("ʉː")`] and [`r linguisticsdown::cond_cmpl("uː")`]. These predictions were borne out for F1 in all cases, but for F2, only for [`r linguisticsdown::cond_cmpl("ʉː")`]. Both [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] display very little movement along F2, whereas [`r linguisticsdown::cond_cmpl("uː")`] moves towards a more central quality, possibly indicating diphthongization ending in [`r linguisticsdown::cond_cmpl("ə")`] rather than a consonantal offglide. 

Figure \@ref(fig:trajectories) further demonstrates that some neighbouring categories either start at different locations but end up in the same, or start at similar locations but travel in different directions. For instance, [`r linguisticsdown::cond_cmpl("uː")`] - [`r linguisticsdown::cond_cmpl("oː")`] are fairly closely located at earlier time-points, but differ substantially towards the end of the vowel segment, while the [`r linguisticsdown::cond_cmpl("ɛː")`]-[`r linguisticsdown::cond_cmpl("æː")`], [`r linguisticsdown::cond_cmpl("øː")`]-[`r linguisticsdown::cond_cmpl("œː")`] and [`r linguisticsdown::cond_cmpl("ø")`]-[`r linguisticsdown::cond_cmpl("œ")`] pairs start at different locations but end up in approximately the same. Finally, the formant trajectories suggest that the empty spots identified in the vowel space under a static analysis (Figure \@ref(fig:swe-vowels)), may indeed be occupied when vowel dynamics is considered. This is especially true for [`r linguisticsdown::cond_cmpl("eː")`] that travels from the mid-high front to the mid center of the space as the signal unfolds---down to a position closer to its short counterpart, [`r linguisticsdown::cond_cmpl("ɛ")`]. Given the amount of overlap when static spectral cues are considered (Figure \@ref(fig:swe-vowels)), formant dynamics is likely highly informative for several of these distinctions.

#### Models of formant dynamics
The first set of GAMMs were modeled separately for each cue (F1, F2, F3) and each set of long and short vowels, with gender, category and their interaction as independent variables, and time-points as smoothing function. Random slopes and intercepts for talker were included. Category was backwards difference coded with [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("ɪ")`] as reference categories in each set, using the \texttt{contr.sdif()} function from the \texttt{R} package \texttt{MASS}. Example of model code: $F1 \sim category + Gender + s(timepoint, by = category, k = 5) + s(Talker, bs = "re") + s(Talker, category, bs = "re")$. Summary tables of all 6 models are included in the SI, Section \@ref(sec:SI-GAMMs). 

The GAMMs suggested a main effect of category on F1 for all vowels in both datasets (all $ps > .008$), except for [`r linguisticsdown::cond_cmpl("ʏ")`] ($\hat\beta = .012, se = .014, p > .37$) which indicates overlap with [`r linguisticsdown::cond_cmpl("ɪ")`] in height (Figure \@ref(fig:m-F1-short)). The difference in means between the high front vowels [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] - [`r linguisticsdown::cond_cmpl("ʉː")`] were not as substantial as for the other categories, as indicated by the overlapping confidence intervals in Figure \@ref(fig:m-F1-long) (Figures \@ref(fig:SI-m-F1-long) and \@ref(fig:SI-m-F1-short) in SI visualize all categories included in the GAMMs for F1).

```{r}
# Run GAMMs predicting F1, F2, F3 separately from the long and short categories
# Dataframe for running GAMMs on long vowels
d.model.long <- d.SwehVd.forStudy.long.dynamic %>%
  select(-starts_with("formant")) %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", category %in% levels.vowel.IPA.swe.long) %>%
  mutate(
    category = factor(category, levels = levels.vowel.IPA.swe.long)) %>%
  mutate(across(c(Talker, Quantity, Gender), ~ factor(.x)))

# Difference code the vowel variable
contrasts(d.model.long$category) <- MASS::contr.sdif(11)

# Rename contrast labels
d.model.long$category <- factor(d.model.long$category, labels = c("[iː]", "[yː]", "[ʉː]", "[eː]", "[ɛː]", "[æː]", "[øː]", "[œː]", "[ɑː]", "[oː]", "[uː]"))

# Dataframe for running GAMMs on short vowels
d.model.short <- d.SwehVd.forStudy.long.dynamic %>%
  select(-starts_with("formant")) %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", category %in% levels.vowel.IPA.swe.short) %>%
  mutate(
    category = factor(category, levels = levels.vowel.IPA.swe.short)) %>%
  mutate(across(c(Talker, Quantity, Gender), ~ factor(.x)))

# Difference code the vowel variable
contrasts(d.model.short$category) <- MASS::contr.sdif(10)

# Rename contrast labels
d.model.short$category <- factor(d.model.short$category, labels = c("[ɪ]", "[ʏ]", "[ɵ]", "[ɛ]", "[æ]", "[ø]", "[œ]", "[a]", "[ɔ]", "[ʊ]"))

m.F1.long <- mgcv::bam(F1 ~ category + Gender + s(timepoint, by = category, k = 5) 
                       + s(Talker, bs = "re")  
                       + s(Talker, category, bs = "re"),
                       # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                       data = d.model.long)

m.F2.long <- mgcv::bam(F2 ~ category + Gender + s(timepoint, by = category, k = 5) 
                       + s(Talker, bs = "re")  
                       + s(Talker, category, bs = "re"),
                       # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                       data = d.model.long)

m.F3.long <- mgcv::bam(F3 ~ category + Gender + s(timepoint, by = category, k = 5) 
                       + s(Talker, bs = "re")  
                       + s(Talker, category, bs = "re"),
                       # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                       data = d.model.long)

m.F1.short <- mgcv::bam(F1 ~ category + Gender + s(timepoint, by = category, k = 5) 
                        + s(Talker, bs = "re")  
                        + s(Talker, category, bs = "re"),
                        # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                        data = d.model.short)

m.F2.short <- mgcv::bam(F2 ~ category + Gender + s(timepoint, by = category, k = 5) 
                        + s(Talker, bs = "re")  
                        + s(Talker, category, bs = "re"),
                        # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                        data = d.model.short)

m.F3.short <- mgcv::bam(F3 ~ category + Gender + s(timepoint, by = category, k = 5) 
                        + s(Talker, bs = "re")  
                        + s(Talker, category, bs = "re"),
                        # + s(timepoint, Talker,  by = category, bs = "fs", k = 5, m=1)
                        data = d.model.short)
```

(ref:m-F1-long) Fitted smooths of GAMM for predicting F1 and 95% confidence intervals for the long vowels [`r linguisticsdown::cond_cmpl("iː")`], [`r linguisticsdown::cond_cmpl("yː")`], [`r linguisticsdown::cond_cmpl("ʉː")`], [`r linguisticsdown::cond_cmpl("eː")`] and [`r linguisticsdown::cond_cmpl("oː")`].

```{r m-F1-long, message=FALSE, fig.width=base.width*3, fig.height=base.height*2.5, out.width='60%', fig.align='center', fig.cap="(ref:m-F1-long)"}
itsadug::plot_smooth(m.F1.long, view = "timepoint", plot_all = "category", col = c("#66C2A5", "#FC8D62", "#8DA0CB", "#FB9A99", "#A6CEE3"), print.summary = FALSE, cond = list(category = c("[iː]", "[yː]", "[ʉː]", "[eː]", "[oː]")))
#plot_diff(m.F1.long, view = "timepoint", comp = list(category = c("[iː]", "[yː]")))
```

(ref:m-F1-short) Fitted smooths of GAMM for predicting F1 and 95% confidence intervals for the short vowels [`r linguisticsdown::cond_cmpl("ɪ")`] and [`r linguisticsdown::cond_cmpl("ʏ")`].

```{r m-F1-short, message=FALSE, fig.width=base.width*3, fig.height=base.height*2.5, out.width='60%', fig.align='center', fig.cap="(ref:m-F1-short)"}
itsadug::plot_smooth(m.F1.short, view = "timepoint", plot_all = "category", col = c("#1B9E77", "#D95F02"), print.summary = FALSE, cond = list(category = c("[ɪ]", "[ʏ]")))
```
For F2, there was a main effect of category for all long and short vowels (all $p > .037$; see Figures \@ref(fig:m-F2-long), \@ref(fig:m-F2-short). Figures \@ref(fig:m-F1-long), \@ref(fig:m-F1-short), \@ref(fig:m-F2-long), and \@ref(fig:m-F2-short) further highlights the clear diphthongization in [`r linguisticsdown::cond_cmpl("eː")`] and [`r linguisticsdown::cond_cmpl("oː")`], and how some neighbouring categories overlap in F2 at the end of the vowel segment (e.g., [`r linguisticsdown::cond_cmpl("ɛː")`]-[`r linguisticsdown::cond_cmpl("æː")`]). Again, [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] have overlapping confidence intervals, even if their estimated means differ (Figures \@ref(fig:SI-m-F2-long) and \@ref(fig:SI-m-F2-short) in SI visualize all categories included in the GAMMs for F2).

(ref:m-F2-long) Fitted smooths of GAMM for predicting F2 and 95% confidence intervals for the long vowels [`r linguisticsdown::cond_cmpl("iː")`], [`r linguisticsdown::cond_cmpl("yː")`], [`r linguisticsdown::cond_cmpl("eː")`], [`r linguisticsdown::cond_cmpl("ɛː")`], [`r linguisticsdown::cond_cmpl("æː")`] and [`r linguisticsdown::cond_cmpl("oː")`].

```{r m-F2-long, message=FALSE, fig.width=base.width*3, fig.height=base.height*2.5, out.width='60%', fig.align='center', fig.cap="(ref:m-F2-long)"}
itsadug::plot_smooth(m.F2.long, view = "timepoint", plot_all = "category", col = c("#66C2A5", "#FC8D62", "#FB9A99", "#E7298A", "#A6D854","#A6CEE3"), print.summary = FALSE, cond = list(category = c("[iː]", "[yː]", "[eː]", "[ɛː]", "[æː]", "[oː]")))
```

(ref:m-F2-short) Fitted smooths of GAMM for predicting F2 and 95% confidence intervals for the short vowels [`r linguisticsdown::cond_cmpl("ɪ")`] and [`r linguisticsdown::cond_cmpl("ʏ")`].

```{r m-F2-short, message=FALSE, fig.width=base.width*3, fig.height=base.height*2.5, out.width='60%', fig.align='center', fig.cap="(ref:m-F1-short)"}
itsadug::plot_smooth(m.F2.short, view = "timepoint", plot_all = "category", col = c("#1B9E77", "#D95F02"), print.summary = FALSE, cond = list(category = c("[ɪ]", "[ʏ]")))
```

A main effect of category was also found for F3 for all vowels besides [`r linguisticsdown::cond_cmpl("yː")`] ($\hat\beta = -.003, se = .011, p > .80$; all other $ps > .046$). The fact that a main effect of category on F3 was only found for the [`r linguisticsdown::cond_cmpl("ɪ")`] - [`r linguisticsdown::cond_cmpl("ʏ")`] and not the [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] distinction, suggests that F3 does not reliably distinguish between the long high front rounded-unrounded vowels, in line with the qualitative analysis (see Figures \@ref(fig:m-F3-long) and \@ref(fig:m-F3-short)) (Figures \@ref(fig:SI-m-F3-long) and \@ref(fig:SI-m-F3-short) in SI visualize all categories included in the GAMMs for F3).

<!--A main effect of Gender was only found for F2 and the short vowels ($\hat\beta = -.028, se = .01, p > .006$).-->

(ref:m-F3-long) Fitted smooths of GAMM for predicting F3 and 95% confidence intervals for the long vowels [`r linguisticsdown::cond_cmpl("iː")`], [`r linguisticsdown::cond_cmpl("yː")`], [`r linguisticsdown::cond_cmpl("ʉː")`] and [`r linguisticsdown::cond_cmpl("eː")`].

```{r m-F3-long, message=FALSE, fig.width=base.width*3, fig.height=base.height*2.5, out.width='60%', fig.align='center', fig.cap="(ref:m-F3-long)"}
itsadug::plot_smooth(m.F3.long, view = "timepoint", plot_all = "category", col = c("#66C2A5", "#FC8D62", "#8DA0CB", "#FB9A99"), print.summary = FALSE, cond = list(category = c("[iː]", "[yː]", "[ʉː]", "[eː]")))
```

(ref:m-F3-short) Fitted smooths of GAMM for predicting F3 and 95% confidence intervals for the short vowels [`r linguisticsdown::cond_cmpl("ɪ")`] and [`r linguisticsdown::cond_cmpl("ʏ")`].

```{r m-F3-short, message=FALSE, fig.width=base.width*3, fig.height=base.height*2.5, out.width='60%', fig.align='center', fig.cap="(ref:m-F3-short)"}
itsadug::plot_smooth(m.F3.short, view = "timepoint", plot_all = "category", col = c("#1B9E77", "#D95F02"), print.summary = FALSE, cond = list(category = c("[ɪ]", "[ʏ]")))
```

```{r}
# CODE KEPT FROM DISCUSSIONS W M GUBIAN??
# d.model <- d.SwehVd.forStudy.long.static %>%
#   filter(Normalization.Type %in% c("no normalization (Hz)", "Nearey2 (log)")) %>%
#   mutate(across(c(Quantity, category), factor))

# model.F3 <- glmer(category ~ F3 + (1 | Talker), family = binomial, data = d.model %>% filter(Normalization.Type == "no normalization (Hz)"))
# summary(model.F3)

# mgcv
# F3 ~ category + s(timepoint, by = category). ----corresponds to a simple lmr without random factors.
# s = smooth
# bam(F3 ~ category  + s(timepoint, by = category, k = 5))
# k = max degrees of freedom, base curves
# plot_smooth(m1, view = "timepoint", plot_all = "category")
# plot_diff(m1, view = "timepoint", comp = list(category = c())

# Model all cues
# d.model.all.cues <- d.SwehVd %>%
#   filter(Transcribed_vowel == "targeted") %>%
#   mutate(across(c(Talker, category), ~ factor(.x)),
#          across(starts_with("F"), ~ log(.x), .names = "{.col}_log"))
# 
# contrasts(d.model.all.cues$category) <- MASS::contr.sdif(21)
# m <- lmerTest::lmer(F2~ timepoint + (1 + timepoint | Talker), data = d.SwehVd %>% filter(category == "[iː]", Transcribed_vowel == "targeted"))
# summary(m)

# It needs to be a multinomial log regression if predicting vowel from cues
# m.allcues <- mgcv::bam(category ~ s(F0_log) + s(F1_log) + s(F2_log) + s(F3_log) + s(timepoint, by = category, k = 5) + s(1 | Talker), 
#                    data = d.model.all.cues, family = "multinomial")
#summary(m.allcues)
#ggeffects::ggpredict(model.cues, "F0 [all]") %>% plot()

# m.allcues.2 <- mgcv::bam(category ~ F0_log + F1_log + F2_log + F3_log + Duration + s(timepoint, by = category, k = 5), data = d.model.all.cues)
# summary(m1)

# plot_smooth(m.F1, view = "timepoint")
# plot_diff(m.F1, view = "timepoint", comp=list(category=c("[ʉː]", "[ʊ]")))

# m.all <- mgcv::gam(list(category ~ F1 + F2 + F3 + Duration + s(timepoint, by = category, k = 5)
#                         + s(Talker, bs = "re"),
#                         ~ F1 + F2 + F3 + Duration + s(timepoint, by = category, k = 5)
#                         + s(Talker, bs = "re")), data = d.model.1, family = multinom(K = 2))
# m.all <- mgcv::gam(list(category ~ s(F1) + s(F2) + s(F3),
#                         ~ s(F1) + s(F2) + s(F3)), family = multinom(K=2), data = d.model)

# d.SwehVd.forStudy.long.static %>%
#   select(-starts_with("formant")) %>%
#   #filter(F3 > .25) %>%
#   filter(Normalization.Type == "Nearey's uniform scaling (log)") %>%
#   mutate(
#     rounding = case_when(category %in% c("[ɵ]", "[ʉː]", "[ʊ]", "[uː]") ~ "inrounded",
#                          category %in% c("[ʏ]", "[yː]", "[ø]", "[øː]", "[ɔ]", "[oː]") ~ "outrounded",
#                          TRUE ~ "no rounding")) %>%
#   ggplot(
#     aes(F3, fill = factor(category))) +
#   geom_density(alpha = .5) +
#   theme(legend.position = "none") +
#   scale_colour_manual(name = "category", values = colors.vowel.swe, aesthetics = c("colour", "fill")) +
#   guides(linetype = "none", color = "none", fill = "none") +
#   facet_wrap(factor(rounding, levels = c("inrounded", "outrounded", "no rounding")) ~ Quantity, ncol = 2)
```

The effect of F1, F2, F3 and duration on all long and short vowel pairs was modeled in four separate GAMMs, with gender, category and their interaction as independent variables, and time-points as smoothing function. Random slopes and intercepts for talker were included. Category was backwards difference coded with [`r linguisticsdown::cond_cmpl("iː")`] as reference category. The category variable was ordered alternating long and short vowels by the pair, in order for the short vowel to be compared against its preceding long counterpart. Summary tables of these four GAMMs are included in the SI (Section \@ref(sec:SI-GAMMs)). 

The GAMMs suggested a main effect of category on all spectral and temporal cues for all vowel pairs (for F1, all $ps > .027$; for F2, F3 and duration, all $ps < .0001$). With the exception of the [`r linguisticsdown::cond_cmpl("yː")`] - [`r linguisticsdown::cond_cmpl("ʏ")`] vowel pair (see also previous section), all rounded vowels displayed lower F3 values, hence more lip-rounding, in their long allophones [e.g., @hadding1976; @stalhammar1973]. The results thus seem to suggest that *all* long-short Central Swedish vowel pairs differ in spectral and temporal cues, including pairs for which previous research have found primarily duration dependencies [e.g., @behne-kirk1997; @haddingkoch1964; @kuronen2000]. Some pairs did, however, display smaller differences in spectral cues than others. For instance, the [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("ɪ")`], [`r linguisticsdown::cond_cmpl("uː")`] - [`r linguisticsdown::cond_cmpl("ʊ")`] and [`r linguisticsdown::cond_cmpl("oː")`] - [`r linguisticsdown::cond_cmpl("ɔ")`] pairs predicting F1 and F3 (see Figures \@ref(fig:m-F1-paired) and \@ref(fig:m-F3-paired)).

(ref:m-F1-paired) Fitted smooths of GAMM for predicting F1 and 95% confidence intervals for the vowel pairs [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("ɪ")`] and [`r linguisticsdown::cond_cmpl("uː")`] - [`r linguisticsdown::cond_cmpl("ʊ")`].

```{r m-F1-paired, message=FALSE, fig.width=base.width*3, fig.height=base.height*2.5, out.width='60%', fig.align='center', fig.cap="(ref:m-F1-paired)"}
# Dataframe for running GAMMs on all 21 vowels grouped by pair
d.model.paired <- d.SwehVd.vowel_pairs %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", vowel_pair != "ɛ - eː, ɛː") %>%
  mutate(
    #Reorder vowel_pairs to match the levels in category
    vowel_pair = factor(vowel_pair, 
                        levels = c("ɪ - iː", "ʏ - yː", "ɵ - ʉː", "ɛ - eː", "ɛ - ɛː", "æ - æː", "ø - øː", "œ - œː", "a - ɑː", "ɔ - oː", "ʊ - uː")),
    category = factor(category, levels = levels.vowel.IPA.swe)) %>%
  mutate(across(c(Talker, Quantity, Gender), ~ factor(.x)))

# Difference code the vowel pair variable
contrasts(d.model.paired$category) <- MASS::contr.sdif(21)

# Rename contrast labels
d.model.paired$category <- factor(d.model.paired$category, labels = c("[iː]", "[ɪ]", "[yː]", "[ʏ]", "[ʉː]", "[ɵ]", "[eː]", "[ɛ]", "[ɛː]", "[æː]", "[æ]", "[øː]", "[ø]", "[œː]", "[œ]", "[ɑː]", "[a]", "[oː]", "[ɔ]", "[uː]", "[ʊ]"))

m.F1.paired <- mgcv::bam(F1 ~ category + Gender + 
                           s(timepoint, by = category, k = 5) + 
                           s(Talker, bs = "re") +
                           s(Talker, category, bs = "re"),
                         data = d.model.paired)

m.F2.paired <- mgcv::bam(F2 ~ category + Gender + 
                           s(timepoint, by = category, k = 5) + 
                           s(Talker, bs = "re") +
                           s(Talker, category, bs = "re"),
                         data = d.model.paired)

m.F3.paired <- mgcv::bam(F3 ~ category + Gender + 
                           s(timepoint, by = category, k = 5) + 
                           s(Talker, bs = "re") +
                           s(Talker, category, bs = "re"),
                         data = d.model.paired)

m.duration.paired <- mgcv::bam(Duration ~ category + Gender + 
                           s(timepoint, by = category, k = 5) + 
                           s(Talker, bs = "re") +
                           s(Talker, category, bs = "re"),
                         data = d.model.paired)

itsadug::plot_smooth(m.F1.paired, view = "timepoint", plot_all = "category", col = c("#66C2A5", "#1B9E77", "#B2DF8A", "#33A02C"), print.summary = FALSE, cond = list(category = c("[iː]", "[ɪ]", "[uː]", "[ʊ]")))
```

(ref:m-F3-paired) Fitted smooths of GAMM for predicting F3 and 95% confidence intervals for the vowel pairs [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("ɪ")`] and [`r linguisticsdown::cond_cmpl("oː")`] - [`r linguisticsdown::cond_cmpl("ɔ")`].

```{r m-F3-paired, message=FALSE, fig.width=base.width*3, fig.height=base.height*2.5, out.width='60%', fig.align='center', fig.cap="(ref:m-F3-paired)"}
itsadug::plot_smooth(m.F3.paired, view = "timepoint", plot_all = "category", col = c("#66C2A5", "#1B9E77", "#A6CEE3", "#1F78B4"), print.summary = FALSE, cond = list(category = c("[iː]", "[ɪ]", "[oː]", "[ɔ]")))
```

### Results summary
The static and dynamic analyses suggest that all cues investigated in this study are of importance for vowel identity in Central Swedish. A relatively indirect estimate of how much each cue contributed to vowel distinctions in the entire 21-category space was obtained by comparing the proportion of explained deviance in the GAMMs predicting F1, F2, F3 and duration from all vowel categories. These models confirmed the general pattern, suggesting that F1 (93.7% of explained deviance), F2 (96.2% of explained deviance) and duration (92% of explained deviance) can more easily predict all 21 categories than F3 (68.5% of explained deviance), as indicated by their relatively higher proportion of deviance explained. It is reasonable to assume formant dynamics as auxiliary cue to vowel identity, more so for some distinctions than others. For instance, the [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] , [`r linguisticsdown::cond_cmpl("oː")`] - [`r linguisticsdown::cond_cmpl("uː")`] and [`r linguisticsdown::cond_cmpl("ɛː")`] - [`r linguisticsdown::cond_cmpl("æː")`] categories display substantial overlap in static analyses but increased separability when analysed dynamically. The dynamic analyses furthermore highlighted that the short vowels also display formant movements, and that for most of both long and short categories, a larger portion of the dynamics resides in the later part of the segment.

\begin{table}
\caption{\label{tab:swehvd-vowels}The phonetic characterization of long (left) and short (right) Central Swedish vowels (as represented in the SwehVd database). Rounded vowels are shaded.}
\centering
\begin{tabular}[t]{c|c|ccc|c}
\hline
&front&\multicolumn{3}{|c|}{central}&back\\
\hline
high&&`r linguisticsdown::cond_cmpl("[iː]")`&\cellcolor[HTML]{E5E4E2}`r linguisticsdown::cond_cmpl("[yː]")`&\cellcolor[HTML]{DCDCDC}`r linguisticsdown::cond_cmpl("[ʉː]")`&\cellcolor[HTML]{DCDCDC}`r linguisticsdown::cond_cmpl("[uː]")`\\
\hline
mid-high&`r linguisticsdown::cond_cmpl("[eː]")`&&&&\cellcolor[HTML]{DCDCDC}`r linguisticsdown::cond_cmpl("[oː]")`\\
\hline
lower-mid&`r linguisticsdown::cond_cmpl("[æː]")`&&\cellcolor[HTML]{DCDCDC}`r linguisticsdown::cond_cmpl("[øː]")`&\cellcolor[HTML]{DCDCDC}`r linguisticsdown::cond_cmpl("[œː]")`&`r linguisticsdown::cond_cmpl("[ɑː]")`\\
\end{tabular}
\hspace{2em}
\begin{tabular}[t]{c|cc|cc|c}
\hline
&\multicolumn{2}{|c|}{front}&\multicolumn{2}{|c|}{central}&back\\
\hline
high&`r linguisticsdown::cond_cmpl("[ɪ]")`&\cellcolor[HTML]{DCDCDC}`r linguisticsdown::cond_cmpl("[ʏ]")`&\cellcolor[HTML]{DCDCDC}`r linguisticsdown::cond_cmpl("[ɵ]")`&&\cellcolor[HTML]{DCDCDC}`r linguisticsdown::cond_cmpl("[ʊ]")`\\
\hline
mid-high&`r linguisticsdown::cond_cmpl("[ɛ]")`&&\cellcolor[HTML]{DCDCDC}`r linguisticsdown::cond_cmpl("[ø]")`&\cellcolor[HTML]{DCDCDC}`r linguisticsdown::cond_cmpl("[œ]")`&\cellcolor[HTML]{DCDCDC}`r linguisticsdown::cond_cmpl("[ɔ]")`\\
\hline
lower-mid&`r linguisticsdown::cond_cmpl("[æ]")`&&`r linguisticsdown::cond_cmpl("[a]")`&&\\
\end{tabular}
\end{table}

The resulting phonetic characteristics of the long and short Central Swedish vowels is summarized in Table \@ref(tab:swehvd-vowels). Beginning with the long vowels, there are 4 high vowels. The current acoustic description suggests that none of these are front, rather [`r linguisticsdown::cond_cmpl("i:")`] and [`r linguisticsdown::cond_cmpl("yː")`] group with [`r linguisticsdown::cond_cmpl("ʉː")`] as central vowels, and [`r linguisticsdown::cond_cmpl("uː")`] is back [c.f., @pelzer-boersma2019; @schotz2011].^[Whether [`r linguisticsdown::cond_cmpl("yː")`] can still be considered rounded given the high F3 values, is a question for future research.] There are 2 mid-high vowels---[`r linguisticsdown::cond_cmpl("eː")`] (front) and [`r linguisticsdown::cond_cmpl("oː")`] (back)---and 4 lower-mid vowels, [`r linguisticsdown::cond_cmpl("æː")`] (front), [`r linguisticsdown::cond_cmpl("øː")`], [`r linguisticsdown::cond_cmpl("œː")`] (both central) and [`r linguisticsdown::cond_cmpl("ɑː")`] (back). Given the substantial lowering of [`r linguisticsdown::cond_cmpl("ɛː")`] and its overlap with [`r linguisticsdown::cond_cmpl("æ")`], it is reasonable to assume one long allophone for /`r linguisticsdown::cond_cmpl("ɛ")`/, which is [`r linguisticsdown::cond_cmpl("æː")`] [c.f., @pelzer-boersma2019].

The short vowel space contains 4 high vowels, two of which are front, [`r linguisticsdown::cond_cmpl("ɪ")`] and [`r linguisticsdown::cond_cmpl("ʏ")`], one is central [`r linguisticsdown::cond_cmpl("ɵ")`] and one back [`r linguisticsdown::cond_cmpl("ʊ")`]. There are 4 mid-high vowels, [`r linguisticsdown::cond_cmpl("ɛ")`] (front), [`r linguisticsdown::cond_cmpl("ø")`], [`r linguisticsdown::cond_cmpl("œ")`] (both central), and [`r linguisticsdown::cond_cmpl("ɔ")`] (back), and 2 low vowels, [`r linguisticsdown::cond_cmpl("æ")`] (front), and [`r linguisticsdown::cond_cmpl("a")`] (central). The analysis would thus seem to support what @riad2014 anticipated and @pelzer-boersma2019 suggested, more precisely, a vowel system consisting of three height levels only, in contrast to the current four height levels system [e.g., @engstrand1999; @riad2014; @engstrand2004]. 

Although suggestions for an updated vowel phonology for Central Swedish are left for future research, the motivation of the summarized acoustics presented in Table \@ref(tab:swehvd-vowels) rests on a pairwise grouping of the long and short vowels, similar to phonological analyses of Central Swedish [@riad2014]. For instance, despite its high position, [`r linguisticsdown::cond_cmpl("eː")`] is defined as mid-high, on par with [`r linguisticsdown::cond_cmpl("oː")`], as both vowels share diphthongizational patterns, and their short versions are both lower than their long counterparts. Furthermore, [`r linguisticsdown::cond_cmpl("æː")`] is front rather than central as its short version is clearly more front than [`r linguisticsdown::cond_cmpl("ø")`]. Because of their overall centralized positions in SwehVd, [`r linguisticsdown::cond_cmpl("i:")`] and [`r linguisticsdown::cond_cmpl("yː")`] groups with [`r linguisticsdown::cond_cmpl("ʉː")`], while their short versions are still clearly front. One could thus argue that both /`r linguisticsdown::cond_cmpl("i")`/ and /`r linguisticsdown::cond_cmpl("y")`/ are under-specified for the front-back dimension, which would also be the case for /`r linguisticsdown::cond_cmpl("ɑ")`/, as [`r linguisticsdown::cond_cmpl("ɑː")`] is back and [`r linguisticsdown::cond_cmpl("ɑ")`] is central.

```{r}
rm(d.model.long, d.model.short, d.model.paired)
```
```{r}
# d.SwehVd %>%
#   filter(Transcribed_vowel == "targeted", Word != "hädd") %>%
#   # Get the geometric mean F1-F2 across the five time points
#   group_by(Talker, category, Token, Quantity) %>%
#   summarise(across(c(F1, F2, F3, Duration), ~ geometric.mean(.x))) %>%
#   plot_ly(
#     x = ~ F2, 
#     y = ~ F3, 
#     z = ~ F1, 
#     color = ~ category) %>%
#     #colors = c("black", "gray"),
#     #size = ~ -Response.Entropy) %>% 
#   add_markers() %>%
#   layout(
#     autoresize = TRUE,
#     showlegend = TRUE,
#     scene = list(
#       xaxis = list(
#         title = "F2 (Hz)",
#         autorange = "reversed"),
#       yaxis = list(
#         title = "F3 (Hz)",
#         autorange = "reversed"),
#       zaxis = list(
#         title = "F1 (Hz)",
#         autorange = "reversed")))
```

# Study 2: Vowel category movements over time - from SweDia to SwehVd
The aim of Study 2 was to investigate possible vowel shifts over the last generation by comparing the acoustic characteristics of SwehVd against that of 8 reference talkers in the SweDia materials, along the two primary cues to vowel identity, F1 and F2. The materials used in Study 2 is presented in Section \@ref(sec:datasets).

## Methods
In order to assess whether the relative placement of the categories in the acoustic space was different between datasets, formant data was visualized in F1-F2 space and distances assessed with the *orthogonal projection ratio*, henceforth $op$ [@stevens-schiel2019].

### Orthogonal projection ratio {#sec:investigating-change}
The $op$, formalized in \@ref(eq:orthogonal-projection-ratio), compares the relative position of a category $V_{vowel}$ to two other anchor categories, $U_{anchor}$ and $W_{anchor}$. Typically, the anchor categories are selected based on their position in the space relative to the category of interest, $V_{vowel}$. $V_{vowel}$ is placed somewhere in between the anchors, and the relative proximity of the vowel $V_{vowel}$ to the anchors is then measured, see Figure \@ref(fig:op).

\begin{equation}
 op(V_{vowel}) = ( -2 \cdot \frac{((V_{vowel} - U_{anchor}) \texttt{\%*\%} (U_{anchor} - W_{anchor}))}{((U_{anchor} - W_{anchor}) \texttt{\%*\%} (U_{anchor} - W_{anchor}))}) - 1 (\#eq:orthogonal-projection-ratio)
\end{equation}

The anchors $U_{anchor}$ and $W_{anchor}$ are first established by calculating the by-talker centroids for each anchor. The $op$ for each observation of $V_{vowel}$ relative to the anchors is then calculated, and summarized across all tokens. This procedure is repeated for each of the $V_{vowel}$ and their respective anchors $U_{anchor}$ and $W_{anchor}$ in each dataset and then compared across datasets, in order to assess vowel shifts over time.

An $op = 0$ indicates that the category $V_{vowel}$ is positioned exactly in the middle between $U_{anchor}$ and $W_{anchor}$, an $op = \pm 1$ indicates that $V_{vowel}$ is positioned exactly on the mean of one of the two anchors, whereas a $V_{vowel}$ positioned somewhere between $U_{anchor}$ and $W_{anchor}$ has an $op$ that varies between $\pm 1$. The $op$ can thus be used as an index of e.g., lowering, fronting, or centralizing.

Table \@ref(tab:opr-table) lists all the vowels $V_{vowel}$ tested along with their corresponding anchor points, $U_{anchor}$ and $W_{anchor}$. These were all vowels for which the overall distance in F1 and F2 category means in SwehVd to SweDia, $\geq .85$ standard deviations from the category means in SweDia.^[The front `r linguisticsdown::cond_cmpl("[æː]")` was also included in this subset. Given that it occupies the lowest position in the space, the calculation of $op$ was not possible.]

\begin{table}
\caption{\label{tab:opr-table}The anchors used for calculating orthogonal projection ratio across the two datasets}
\begin{tabular}[t]{c|c|c}
$V_{vowel}$ & $U_{anchor}$ & $W_{anchor}$\\
\hline
`r linguisticsdown::cond_cmpl("[ʏ]")` & `r linguisticsdown::cond_cmpl("[ɑː]")` & `r linguisticsdown::cond_cmpl("[ɪ]")`\\
\hline
`r linguisticsdown::cond_cmpl("[ʉː]")` & `r linguisticsdown::cond_cmpl("[ɪ]")` & `r linguisticsdown::cond_cmpl("[oː]")`\\
\hline
`r linguisticsdown::cond_cmpl("[ɛː]")` & `r linguisticsdown::cond_cmpl("[eː]")` & `r linguisticsdown::cond_cmpl("[æː]")`\\
\hline
`r linguisticsdown::cond_cmpl("[øː]")` & `r linguisticsdown::cond_cmpl("[eː]")` & `r linguisticsdown::cond_cmpl("[ɑː]")`\\
\hline
`r linguisticsdown::cond_cmpl("[œ]")` & `r linguisticsdown::cond_cmpl("[ɵ]")` & `r linguisticsdown::cond_cmpl("[œː]")`\\
\hline
`r linguisticsdown::cond_cmpl("[ɔ]")` & `r linguisticsdown::cond_cmpl("[ɑː]")` & `r linguisticsdown::cond_cmpl("[oː]")`\\
\hline
\end{tabular}
\end{table}

(ref:op) Visualization of orthogonal projection ratio. Examples of target categories $V_{vowel}$ (in cyan), and their respective anchor categories, $U_{anchor}$ and $W_{anchor}$ (in red). The $op$ measures the relative distance of $V_{vowel}$ to its anchors, as represented by the lines.

```{r op, fig.width=base.width*3, fig.height=base.height*3, out.width='60%', fig.align='center', fig.cap="(ref:op)"}
d.SwehVd.forStudy.long.static %>%
  # Filter down to test folds
  filter(Normalization.Type == "Nearey's uniform scaling (log)", category %in% c("[ʉː]", "[oː]", "[ɪ]", "[eː]", "[æː]", "[ɑː]", "[ɛː]", "[øː]")) %>%
  mutate(anchor = ifelse(category %in% c("[ʉː]", "[ɛː]", "[øː]"), "vowel", "anchor"),
         op_pair = case_when(category %in% c("[ʉː]", "[oː]", "[ɪ]") ~ "A",
                             category %in% c("[eː]", "[ɑː]", "[øː]") ~ "B",
                             category %in% c("[eː]", "[æː]", "[ɛː]") ~ "C")) %>%
  ggplot(
    aes(
      x = F2, 
      y = F1)) +
  stat_ellipse(
    mapping = aes(
      colour = anchor,
      fill = anchor,
      group = interaction(anchor, category)),
    geom = "polygon",
    alpha = .4) +
  geom_path(data =
              . %>%
              filter(category %in% c("[ɪ]", "[ʉː]", "[oː]")) %>%
              mutate(
                category = factor(category, levels = c("[ɪ]", "[ʉː]", "[oː]")),
                group = "A") %>%
              group_by(category, group, Normalization.Type) %>%
              summarise(across(c(F1, F2), ~ mean(.x))),
            linetype = "dotted") +
  geom_path(data =
              . %>%
              filter(category %in% c("[eː]", "[øː]", "[ɑː]")) %>%
              mutate(
                category = factor(category, levels = c("[eː]", "[øː]", "[ɑː]")),
                group = "B") %>%
              group_by(category, group, Normalization.Type) %>%
              summarise(across(c(F1, F2), ~ mean(.x))),
            linetype = "dashed") +
  geom_path(data =
              . %>%
              filter(category %in% c("[eː]", "[ɛː]", "[æː]")) %>%
              mutate(
                category = factor(category, levels = c("[eː]", "[ɛː]", "[æː]")),
                group = "C") %>%
              group_by(category, group, Normalization.Type) %>%
              summarise(across(c(F1, F2), ~ mean(.x)))) +
  # Add between-talker mean
  geom_label(
    data =
      . %>%  
      mutate(
        anchor = ifelse(category %in% c("[ʉː]", "[ɛː]", "[øː]"), "vowel", "anchor")) %>%
      group_by(category, anchor, Normalization.Type) %>%
      summarise(across(c(F1, F2), ~ mean(.x))),
    mapping = aes(
      colour = anchor,
      label = category),
    alpha = .8,
    size = 5,
    label.size = NA,
    label.padding = unit(.06, "cm")) +
  scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 4)) +
  scale_y_reverse("F1", position = "right", scales::pretty_breaks(n = 4)) +
  guides(alpha = "none", color = "none", shape = "none", fill = "none")
```

### Sampling-with-replacement
The two datasets were of unequal sample sizes: N = `r nrow(d.SwehVd %>% distinct(Talker, category, Token))` datapoints for SwehVd, and N = `r nrow(d.SweDia2000.forStudy %>% distinct(Talker, category, Token))` datapoints for SweDia. In order to address this imbalance, the SwehVd materials was down-sampled to approximate the size of the SweDia for the calculation of the $op$. The sampling-with-replacement procedure included 1000 bootstrapping rounds. At each round, a sample of 840 tokens from the SwehVd was extracted by first randomly selecting 8 talkers (4 female) from the database, and then randomly selecting 5 tokens per vowel from each talker, with replacement. The $op$ was calculated at each round, and finally averaged across the 1000 rounds for a comparison against the $op$ calculated for the SweDia materials.

## Results
Figure \@ref(fig:db-comparison) displays the category movements in F1-F2 space across talkers between the two datasets. The figure suggests that a few vowels display considerable stability over time, for instance, the back vowel [`r linguisticsdown::cond_cmpl("ɑː")`], and to a certain extent also [`r linguisticsdown::cond_cmpl("oː")`], the center vowels [`r linguisticsdown::cond_cmpl("œː")`] and [`r linguisticsdown::cond_cmpl("a")`], and the front vowel [`r linguisticsdown::cond_cmpl("ɛ")`]. The remaining vowels have all moved to a lower or higher degree. Two general tendencies can be noted. Firstly, several of the vowels that form the edges of the space slightly shift further outwards, causing the overall space to widen. This is the case for [`r linguisticsdown::cond_cmpl("ɪ")`], [`r linguisticsdown::cond_cmpl("eː")`], [`r linguisticsdown::cond_cmpl("æː")`], [`r linguisticsdown::cond_cmpl("a")`], [`r linguisticsdown::cond_cmpl("oː")`], and [`r linguisticsdown::cond_cmpl("uː")`]. Most of the inner vowels either also shift outwards---[`r linguisticsdown::cond_cmpl("ʏ")`], [`r linguisticsdown::cond_cmpl("ɵ")`], [`r linguisticsdown::cond_cmpl("ʉː")`], [`r linguisticsdown::cond_cmpl("ʊ")`], [`r linguisticsdown::cond_cmpl("ɔ")`]---or centralize---[`r linguisticsdown::cond_cmpl("ø")`], [`r linguisticsdown::cond_cmpl("øː")`], [`r linguisticsdown::cond_cmpl("œ")`]. Second, these movements lead to neighbouring categories closing in on each other in several cases, e.g., [`r linguisticsdown::cond_cmpl("ɪ")`] - [`r linguisticsdown::cond_cmpl("ʏ")`],  [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("ʉː")`], [`r linguisticsdown::cond_cmpl("ø")`] - [`r linguisticsdown::cond_cmpl("œ")`], [`r linguisticsdown::cond_cmpl("oː")`] - [`r linguisticsdown::cond_cmpl("ɔ")`] - [`r linguisticsdown::cond_cmpl("uː")`] - [`r linguisticsdown::cond_cmpl("ʊ")`], or increasing their separability---[`r linguisticsdown::cond_cmpl("ɵ")`] - [`r linguisticsdown::cond_cmpl("œ")`], [`r linguisticsdown::cond_cmpl("ø")`] - [`r linguisticsdown::cond_cmpl("øː")`], [`r linguisticsdown::cond_cmpl("ʏ")`] - [`r linguisticsdown::cond_cmpl("ʉː")`].

Figure \@ref(fig:db-comparison) further demonstrates changes in the front part of the space that seem to suggest a series of consecutive movements, where the front vowels, [`r linguisticsdown::cond_cmpl("eː")`], [`r linguisticsdown::cond_cmpl("ɪ")`], [`r linguisticsdown::cond_cmpl("ʏ")`], all have traveled further up and front in SwehVd, while [`r linguisticsdown::cond_cmpl("iː")`], [`r linguisticsdown::cond_cmpl("yː")`] and [`r linguisticsdown::cond_cmpl("ʉː")`] have centralized. Finally, Figure \@ref(fig:db-comparison) shows that [`r linguisticsdown::cond_cmpl("ɛː")`] have lowered substantially compared to SweDia, down to a position close to [`r linguisticsdown::cond_cmpl("æː")`] (and to [`r linguisticsdown::cond_cmpl("æ")`]; recall Figure \@ref(fig:swe-vowels)). The already low [`r linguisticsdown::cond_cmpl("æː")`] has in turn moved further down, presumably constrained by the lower bound and cannot move further. This lowering of [`r linguisticsdown::cond_cmpl("ɛː")`] was anticipated by @riad2014, and was furthermore noted by @leinonen2010 and @pelzer-boersma2019.

(ref:db-comparison) Vowel category movements in the Central Swedish vowel space from 1999 to 2023, in F1-F2. The arrow indicates the direction of movement over time, with base of arrow representing SweDia (vowel label) and tip of arrow representing SwehVd. Vowel data is taken at the steady state of the vowel (averaged across the three mid-points) and averaged across talkers in the two datasets.

```{r db-comparison, fig.width=base.width*3, fig.height=base.height*3, out.width='70%', fig.align='center', fig.cap="(ref:db-comparison)"}
p.vowel.movements <- 
  d.Swehvd.Swedia %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)") %>%
  group_by(Database, category, Quantity) %>%
  summarise(across(c(F2, F1), ~mean(.x))) %>%
  ggplot(
    aes(
      x = F2,
      y = F1, 
      color = category,
      group = NULL)) +
  geom_path(
    alpha = .9,
    arrow = arrow(ends = "last", type = "closed", angle = "22.5", length = unit(.25, "cm"))) +
  # Add between-talker mean
  geom_label(
    data =
      ~ .x %>%
      filter(Database == "SweDia") %>%
      group_by(category, Quantity, Database) %>%
      summarise(across(c(F1, F2), ~ mean(.x))),
    mapping = aes(
      colour = category,
      label = category),
    fontface = 2,
    alpha = .4,
    size = 4,
    label.size = NA,
    label.padding = unit(.06, "cm")) +
  scale_color_manual(name = "category", values = colors.vowel.swe) +
  scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 6)) +
  scale_y_reverse("F1", position = "right", scales::pretty_breaks(n = 6)) +
  guides(color = "none")
p.vowel.movements
#ggsave(file = "../../figures/p.png", width = 5, height = 5, dpi=700)
```
```{r}
#Select categories for orthogonal projection ratio calc
d.vowels.for.comparison <- d.Swehvd.Swedia %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)") %>%
  group_by(Database, Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F2, F1), ~mean(.x))) %>%
  group_by(category, Database) %>%
  summarise(across(starts_with("F"), .fns = list("mean" = mean, "sd" = sd), .names = "{.col}_{.fn}")) %>%
  group_by(category) %>%
  mutate(
    within_SD_F1 = abs(F1_mean-F1_mean[Database == "SweDia"]) <= .85 * F1_sd[Database == "SweDia"],
    within_SD_F2 = abs(F2_mean-F2_mean[Database == "SweDia"]) <= .85 * F2_sd[Database == "SweDia"]) %>%
  filter(within_SD_F1 == FALSE | within_SD_F2 == FALSE)
```

```{r}
# Set seed for reproducibility
set.seed(15497221)

# Number of bootstrap rounds
n_bootstrap_rounds <- 1000

# Number of talkers to select in each round
n_talkers_per_gender <- 4

# Number of tokens to select in each round per category
n_tokens_per_category <- 5

# Create a function to perform downsampling and store values
perform_bootstrap_round <- function(data) {
  # Randomly select 4 female and 4 male talkers
  selected_talkers <- data %>%
    group_by(Gender) %>%
    slice_sample(n = n_talkers_per_gender, replace = TRUE) %>%
    ungroup()

  # Randomly select 5 tokens per category
  selected_tokens <- selected_talkers %>%
    group_by(category) %>%
    slice_sample(n = n_tokens_per_category, replace = TRUE) %>%
    ungroup()

 # Extract and store the F1 and F2 values, category, and bootstrap round
  values <- selected_tokens %>%
    select(Talker, category, Word, F1, F2) %>%
    mutate(bootstrap_round = cur_group_id()) %>%
    as.data.frame()

  return(values)
}

d.SwehVd.forStudy.opr <- d.SwehVd.forStudy.long.static %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)")

# Perform downsampling and bootstrap
bootstrap_results <- map(
  1:n_bootstrap_rounds,
  ~ perform_bootstrap_round(d.SwehVd.forStudy.opr)
)

# Create a dataframe with bootstrap results in long format
bootstrap_SwehVd <- bind_rows(bootstrap_results, .id = "bootstrap_round")
```

```{r}
# Function for calculating the orthogonal projection ratio.
orthogonal_projection <- function(V, U, W) {
  # V, U, W are vectors of the same length
  # which must be greater than 1
  # U, W are the anchors. Calculate the
  # orthogonal projection ratio of V
  # to the two anchors
  ( -2 * ((V-U) %*% (U-W)) / ((U-W) %*% (U-W)) ) - 1
}

get_centroids <- . %>%
  group_by(Talker, category) %>% 
  # compute mean F1 and F2
  summarise(cenF1 = mean(F1),
            cenF2 = mean(F2)) %>%
  # rearrange in columns
  pivot_wider(names_from = category, 
              values_from = c(cenF1, cenF2)) %>%
  ungroup() %>%
  left_join(bootstrap_SwehVd, by = "Talker") %>%
  group_by(category) %>%
  # for each row i.e. for each observation
  rowwise() %>%
  # calculate orth.proj.ratio
  mutate(orth_proj = case_when(category == "[ʏ]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɑː]`, `cenF2_[ɑː]`), c(`cenF1_[ɪ]`, `cenF2_[ɪ]`)),
                               category == "[ʉː]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɪ]`, `cenF2_[ɪ]`), c(`cenF1_[oː]`, `cenF2_[oː]`)),
                               category == "[ɛː]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[eː]`, `cenF2_[eː]`), c(`cenF1_[æː]`, `cenF2_[æː]`)),
                               category == "[øː]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[eː]`, `cenF2_[eː]`), c(`cenF1_[ɑː]`, `cenF2_[ɑː]`)),
                               category == "[œ]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɵ]`, `cenF2_[ɵ]`), c(`cenF1_[œː]`, `cenF2_[œː]`)),
                               category == "[ɔ]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɑː]`, `cenF2_[ɑː]`), c(`cenF1_[oː]`, `cenF2_[oː]`)),
                               TRUE ~ NA))

d.SwehVd.opr <-
  bootstrap_SwehVd %>% 
  # choose the anchor vowels
  filter(category %in% c("[ɪ]", "[ʏ]", "[yː]", "[ɵ]", "[eː]", "[ɛː]", "[æː]", "[œː]", "[œ]", "[ɑː]", "[oː]", "[øː]", "[ʉː]", "[ɔ]")) %>% 
  group_by(bootstrap_round) %>%
  get_centroids() %>%  
  # Then summarise across all bootstrap rounds
  group_by(category, F1, F2, Talker) %>%
  summarise(orth_proj = mean(orth_proj, na.rm = T))
```
```{r}
centroids.SweDia2000 <-
  d.SweDia2000.forStudy.long %>%
  # choose the anchor vowels
  filter(category %in% c("[ɪ]", "[ʏ]", "[yː]", "[ɵ]", "[eː]", "[ɛː]", "[æː]", "[œː]", "[œ]", "[ɑː]", "[oː]", "[øː]", "[ʉː]", "[ɔ]"), 
         Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(35, 50, 65)) %>%
  group_by(Talker, Gender, category, Token, Quantity, Word) %>%
  summarise(across(c(F2, F1), ~mean(.x))) %>%
  # for each speaker and vowel
  group_by(Talker, category) %>% 
  # compute mean F1 and F2
  summarise(cenF1 = mean(F1),
            cenF2 = mean(F2)) %>%
  # rearrange in columns
  pivot_wider(names_from = category, 
              values_from = c(cenF1, cenF2)) %>%
  ungroup()
  
d.SweDia2000.opr <-  d.SweDia2000.forStudy.long %>%
  # choose the anchor vowels
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(35, 50, 65)) %>%
  group_by(Database, Talker, Gender, category, Token, Quantity, Word) %>%
  summarise(across(c(F2, F1), ~mean(.x))) %>%
  # add the centroids %>%
  left_join(centroids.SweDia2000, by = "Talker") 

d.orth.SweDia2000 <- d.SweDia2000.opr %>% 
  group_by(category, Database) %>%
  # # for each row i.e. for each observation
  rowwise() %>%
  # calculate orth.proj.ratio
  mutate(orth_proj = case_when(category == "[ʏ]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɑː]`, `cenF2_[ɑː]`), c(`cenF1_[ɪ]`, `cenF2_[ɪ]`)),
                               category == "[ʉː]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɪ]`, `cenF2_[ɪ]`), c(`cenF1_[oː]`, `cenF2_[oː]`)),
                               category == "[ɛː]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[eː]`, `cenF2_[eː]`), c(`cenF1_[æː]`, `cenF2_[æː]`)),
                               category == "[øː]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[eː]`, `cenF2_[eː]`), c(`cenF1_[ɑː]`, `cenF2_[ɑː]`)),
                               category == "[œ]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɵ]`, `cenF2_[ɵ]`), c(`cenF1_[œː]`, `cenF2_[œː]`)),
                               category == "[ɔ]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɑː]`, `cenF2_[ɑː]`), c(`cenF1_[oː]`, `cenF2_[oː]`)),
                               TRUE ~ NA)) %>%
  ungroup()

d.orth <- rbind(
  d.orth.SweDia2000 %>%
    select(category, F1, F2, orth_proj, Database, Word, Talker),
  d.SwehVd.opr %>%
    mutate(
      Database = "SwehVd",
      # mutate back word column
      Word = category,
      Word = factor(plyr::mapvalues(
        Word,
        levels.vowel.IPA.swe,
        levels.vowel.Swe.word),
      levels = levels.vowel.Swe.word))) %>%
  filter(category %in% c("[ʏ]", "[ɛː]", "[œ]", "[ʉː]", "[øː]", "[ɔ]"))
```

To quantify these potential vowel shifts, the distances between the category locations in the two datasets were assessed with the $op$ [@stevens-schiel2019]. Figure \@ref(fig:orth-dist) visualizes the $op$ of all vowels in Table \@ref(tab:opr-table) in both datasets. By this way of assessing the relative movement between anchor points across datasets, Figure \@ref(fig:orth-dist) indicates that all selected categories have moved over time in the directions suggested by the qualitative analysis in Figure \@ref(fig:db-comparison). The largest differences appear to concern [`r linguisticsdown::cond_cmpl("ʏ")`], [`r linguisticsdown::cond_cmpl("ɛː")`], [`r linguisticsdown::cond_cmpl("œ")`], and [`r linguisticsdown::cond_cmpl("ɔ")`]. The front [`r linguisticsdown::cond_cmpl("ʏ")`] is relatively closer to [`r linguisticsdown::cond_cmpl("ɪ")`] in SwehVd than in SweDia, which suggest fronting. The [`r linguisticsdown::cond_cmpl("ɛː")`] is relatively closer to [`r linguisticsdown::cond_cmpl("æː")`] in SwehVd than in SweDia, as is [`r linguisticsdown::cond_cmpl("œ")`] to [`r linguisticsdown::cond_cmpl("œː")`], indicating the lowering of both categories over the last generation. The back [`r linguisticsdown::cond_cmpl("ɔ")`] is relatively closer to [`r linguisticsdown::cond_cmpl("oː")`] in SwehVd, indicating back raising over time.

An estimate of the magnitude of the effect was obtained by fitting separate generalized linear mixed-effects models (GLMMs) for each of the categories, predicting the orthogonal projection ratio from database, with talkers and word as random effects. Database was contrast coded with SweDia as reference category. The GLMMs confirmed the directionality of the vowel shifts, and found a main effect of Database on $op$ for [`r linguisticsdown::cond_cmpl("ʏ")`] ($\bar\beta$ = 0.162, SE = 0.076, $p >$ .043), and [`r linguisticsdown::cond_cmpl("øː")`] ($\bar\beta$ = 0.23, SE = 0.075, $p > .005$). The word context presumably carried a substantial portion of the variability found in the data, as fitting separate GLMM excluding word as random effect resulted in a main effect of database on all categories but [`r linguisticsdown::cond_cmpl("ʉ")`] (all $ps < 0.019$). This result overall highlights the importance of complementing qualitative analyses with models that can account for factors such as difference in phonological context and talker variability when comparing across datasets.

The findings from both studies are next discussed in the General discussion.

(ref:orth-dist) The relative distance between categories in SweDia and SwehVd. $op = 0$ indicates that the category is equidistant between the two anchor categories in Table \@ref(tab:opr-table), positive numbers indicate a position closer to the second anchor vowel, negative numbers indicate a position closer to the first anchor. Pointranges indicate mean and 95% bootstrapped CIs of the $op$ summarized across talkers in each database. The large CIs for the [`r linguisticsdown::cond_cmpl("œ")`] category in the SweDia materials likely reflect the tendency of some talkers to neutralize the short /`r linguisticsdown::cond_cmpl("ø")`/ and /`r linguisticsdown::cond_cmpl("ʉ")`/ as [`r linguisticsdown::cond_cmpl("ɵ")`].

```{r orth-dist, fig.width=base.width*2.5, fig.height=base.height*2.5, out.width='70%', fig.align='center', fig.cap="(ref:orth-dist)"}
# colors.anchors.swe <- c("#D95F02", "#8DA0CB", "#E7298A", "#FFD92F", "#E6AB02", "#A6761D", "#1F78B4")
# names(colors.anchors.swe) <- c("[ʏ] ~ [ɑː]-[ɪ]", "[ʉː] ~ [ɪ]-[oː]", "[ɛː] ~ [eː]-[æː]", "[øː] ~ [eː]-[ɑː]", "[ø] ~ [yː]-[a]", "[œ] ~ [ɵ]-[œː]", "[ɔ] ~ [ɑː]-[oː]")
p.orth <- d.orth %>% 
  group_by(Database, category) %>%
  mutate(
    # category = factor(category, levels = c("[ʏ]", "[ʉː]", "[ɛː]", "[øː]", "[ø]", "[œ]", "[ɔ]"), labels =  c("[ʏ] ~ [ɑː]-[ɪ]", "[ʉː] ~ [ɪ]-[oː]", "[ɛː] ~ [eː]-[æː]", "[øː] ~ [eː]-[ɑː]", "[ø] ~ [yː]-[a]", "[œ] ~ [ɵ]-[œː]", "[ɔ] ~ [ɑː]-[oː]")),
    #median = median(orth_proj),
    ci = list(enframe(Hmisc::smean.cl.boot(orth_proj)))) %>%
    unnest(ci) %>%
    spread(name, value) %>%
  group_by(Database, category) %>%
    summarise(
      mean_CI_lower = mean(Lower),
      mean_CI_upper = mean(Upper),
      mean = mean(Mean)) %>%
  ggplot(
    aes(x = Database,
        y = mean,
        ymin = mean_CI_lower,
        ymax = mean_CI_upper,
        color = category)) +
  geom_pointrange(
    position = position_dodge2(width = 0.5),
    fatten = 2.5) +
  geom_line(aes(group = category), position = position_dodge(width = 0.5), linetype = "dashed") +
  scale_color_manual(name = "category", values = colors.vowel.swe) + 
  scale_y_continuous("Orthogonal projection ratio", limits = c(-1,1)) +
  myGplot.defaults() #+
  #theme(legend.title = element_blank()) +
  # geom_text(
  #   data = d.orth %>%
  #     group_by(Database, category) %>%
  #     summarise(max = max(orth_proj)),
  #   aes(
  #     x = Database,
  #     y = max),
  #   label = "*",
  #   size = 4, color = "red", inherit.aes = F)
p.orth
```

```{r message=FALSE}
# d.orth %>%
#   filter(category == "[ʏ]") %>%
#   lmer(orth_proj ~ Database + (1|Word) + (1|Talker), .) %>%
#   anova()

# d.orth.for.model <- d.orth %>%
#   group_by(Database, category, Talker) %>%
#   mutate(
#     orth_proj = ifelse(Database == "SwehVd", mean(orth_proj), orth_proj))

m.Y <- lmerTest::lmer(orth_proj ~ Database + (1|Word) + (1|Talker),
                      data = d.orth %>%
                        filter(category == "[ʏ]", !is.na(orth_proj)) %>% 
                        mutate(across(c(Talker, Database, Word), ~ factor(.x))))
m.Ä <- lmerTest::lmer(orth_proj ~ Database + (1|Word) + (1|Talker), 
                      data = d.orth %>%
                        filter(category == "[ɛː]", !is.na(orth_proj)) %>% 
                        mutate(across(c(Talker, Database, Word), ~ factor(.x))))
m.ÖR <- lmerTest::lmer(orth_proj ~ Database + (1|Word) + (1|Talker), 
                       data = d.orth %>%
                         filter(category == "[œ]", !is.na(orth_proj)) %>% 
                         mutate(across(c(Talker, Database, Word), ~ factor(.x))))
m.U <- lmerTest::lmer(orth_proj ~ Database + (1|Word) + (1|Talker),
                      data = d.orth %>%
                        filter(category == "[ʉː]", !is.na(orth_proj)) %>% 
                        mutate(across(c(Talker, Database, Word), ~ factor(.x))))
m.Ö <- lmerTest::lmer(orth_proj ~ Database + (1|Word) + (1|Talker), 
                      data = d.orth %>%
                        filter(category == "[øː]", !is.na(orth_proj)) %>% 
                        mutate(across(c(Talker, Database, Word), ~ factor(.x))))
m.Å <- lmerTest::lmer(orth_proj ~ Database + (1|Word) + (1|Talker), 
                      data = d.orth %>%
                        filter(category == "[ɔ]", !is.na(orth_proj)) %>% 
                        mutate(across(c(Talker, Database, Word), ~ factor(.x))))
#modelsummary::modelsummary(m.Y, output = "latex")
#summary(m.ÖR)
```
```{r}
rm(d.orth, d.orth.SweDia2000, d.orth.for.model, centroids.SweDia2000, bootstrap_SwehVd)
```

```{r}
# make_ellipse <- function(mu, Sigma, cues) {
#   ellipse = rgl::ellipse3d(Sigma, subdivide = 5)
#   
#   x = as.numeric(ellipse$vb[1,] + mu[1])
#   y = as.numeric(ellipse$vb[2,] + mu[2])
#   z = as.numeric(ellipse$vb[3,] + mu[3])
#   
#   t = tibble(x, y, z)
#   names(t) = cues
#   return(t)
# }
# 
# plot_vowel3D <- function(
#     .data = d.SwehVd,
#     quantity = "long",
#     cues = c("F1", "F2", "F3"),
#     axis.expansion = 1.1
# ) {
#   require(rlang)
#   
#   .data %<>%
#     filter(Quantity %in% quantity) %>%
#     filter(Transcribed_vowel == "targeted", Word != "hädd") %>%
#     select(category, Quantity, !!! syms(cues), timepoint) %>%
#     pivot_longer(
#       cols = timepoint,
#       values_to = "timepoint") %>%
#     select(-name)
#   
#   
#   axx <- list(
#     title = cues[1],
#     nticks = 4,
#     range = range(.data[, cues[1]]) * axis.expansion,
#     gridwidth = 2,
#     autorange="reversed"
#   )
#   axy <- list(
#     title = cues[2],
#     nticks = 4,
#     range = range(.data[, cues[2]]) * axis.expansion,
#     gridwidth = 2,
#     autorange="reversed"
#   )
#   axz <- list(
#     title = cues[3],
#     nticks = 4,
#     range =  range(.data[, cues[3]]) * axis.expansion,
#     gridwidth = 2
#   )
#   
#   axes <- list(axx, axy, axz)
#   
#   m <-
#     make_MVG_ideal_observer_from_data(.data, group = "timepoint", cues = cues) %>%
#     mutate(e = map2(mu, Sigma, ~ make_ellipse(.x, .y, cues = cues))) %>%
#     unnest(e)
#   
#   .data %>%
#     plot_ly(
#       x= ~F1, y= ~F2, z= ~F3,
#       color= ~category, # here we could add *what* colors to use (in addition to the color argument): colors = colors.Category[.data.IO$Model.Sound],
#       frame = ~timepoint,
#       opacity = .2,
#       size = 1,
#       showlegend = T,
#       type="scatter3d", mode="markers") %>%
#     add_trace(
#       data = m,
#       type = 'mesh3d',
#       x= ~F1, y= ~F2, z= ~F3,
#       alphahull = 0,
#       frame = ~timepoint,
#       opacity = .1,
#       color = ~category, # here we could add *what* colors to use (in addition to the color argument): vertexcolor = colors.Category[.data.IO$Model.Sound],
#       hoverinfo = 'text',
#       text = ~paste("Category:", category,
#                     "<br>F1:", signif(F1, 4),
#                     "<br>F2:", signif(F2, 4),
#                     "<br>F3:", signif(F3, 4)),
#       inherit = F,
#       showlegend = F) %>%
#     layout(scene = list(xaxis = axes[[1]], yaxis = axes[[2]], zaxis = axes[[3]], aspectmode='cube'))
#   
#   # Talk to Xin to set perspective on 3d plot that puts x = F2 (reverted)  and y = F1 (reverted)
# }

#plot_vowel3D(quantity = "long")
#plot_vowel3D(quantity = "short")
#plot_vowel3D(quantity = c("short", "long"))
```

# General discussion {#sec:discussion}
The objective of the present study was to provide an extensive analysis of the acoustic cues to vowel identity in contemporary Central Swedish and to evaluate the extent of vowel shifts over the recent generation. The main findings from both studies are next discussed, alongside methodological considerations and future directions.

## Static and dynamic analyses of the 21 vowels of modern-day Central Swedish
The purpose of the first study was to present an up-to-date acoustic analysis of Central Swedish vowels that included both empirical formant data and models of formant dynamics. The first study thus aimed to expand on and complement previous work by 1) the scope of the analysis, performing the same type of analyses on all 21 categories, 2) the materials chosen, using a recently collected hVd corpus with high resolution within and across talkers for a single variety, and 3) the methodological approach employed, with the use of traditional formant analysis, trajectory visualizations and models of formant dynamics (GAMMs). 

The results of the static analyses indicated that the most important cues to vowel identity were F1, F2 and duration, which replicated previous work [e.g., @kuronen2000; @lindblom1963]. The results furthermore suggested that static measurements, while informative, were insufficient to accurately capture the spectral acoustics of some vowels. A more exhaustive description can be achieved by including dynamic analyses, as shown in this paper. This was especially true for the [`r linguisticsdown::cond_cmpl("ɛː")`] - [`r linguisticsdown::cond_cmpl("æː")`],  [`r linguisticsdown::cond_cmpl("uː")`] - [`r linguisticsdown::cond_cmpl("oː")`] and [`r linguisticsdown::cond_cmpl("eː")`] - [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`] distinctions, given the direction and scope of formant trajectory movements across the vowel segment. The analysis of the empirical data furthermore suggested that most of the movement took place at the later part of the segment for the majority of vowels, irrespective of the magnitude of change in F1 and F2. However, [`r linguisticsdown::cond_cmpl("iː")`], [`r linguisticsdown::cond_cmpl("yː")`] and [`r linguisticsdown::cond_cmpl("ʉː")`] constituted important exceptions as they displayed more movement in the first three to four time-points. Of note, the short vowels also displayed substantial formant movements, which seems to signal the importance of vowel dynamics for long and short vowels alike. This has been shown in work on other languages [@refs], but has largely been lacking in studies on Swedish. 

GAMMs fit to the data contributed with further insights into the formant dynamics of the vowel space in its entirety as well as for individual contrasts. For instance, GAMMs fitted to the long-short vowel pairs indicated that all pairs were reliably distinguished by at least one spectral cue, highlighting the informativity carried by formants in quantity distinctions. The GAMMs fitted to the long and short high front vowels separately furthermore suggested that F3 could not reliably distinguish between [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`]. 

What consequences these results might have for the *perception* of Central Swedish vowels should be investigated in future studies. For instance, the role of spectral cues in long-short vowel pair distinctions could be assessed by exposing listeners to synthesized versions where long and short vowel duration is crossed with the allophones' spectral information for any given phoneme. More insight into the effect of formant dynamics for vowel perception could be gained from having listeners categorize tokens extracted from different segments of the long vowels, e.g., the first three time-points vs. the three final time-points.

## Vowel shifts
In Study 2, the acoustic characteristics of SwehVd were subsequently compared to that of 8 reference talkers of Central Swedish from the SweDia materials recorded approximately one generation ago, in order to investigate the scope and spread of previously reported changes [e.g., @pelzer-boersma2019; @schotz2011]. The aim was to present a broad-scale comparison by evaluating changes in all but one of the 21 categories in the same analysis, using empirical formant analysis and a distance metric, the $op$.

Several vowels had shifted quite considerably compared to earlier work on Central Swedish [c.f., @fant1969; @kuronen2000]. Two of the most important shifts concerned the fronting of [`r linguisticsdown::cond_cmpl("ɪ")`] and [`r linguisticsdown::cond_cmpl("ʏ")`] and the centralization of [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`], where [`r linguisticsdown::cond_cmpl("ɪ")`] and [`r linguisticsdown::cond_cmpl("ʏ")`] appeared to maintain their positions as high front vowels, while [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] had centralized to the mid-center part of the space. Compared to the SweDia reference materials, the shifts in [`r linguisticsdown::cond_cmpl("ɪ")`], [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] were not as substantial as for other vowels, which suggests that these changes had begun already in 1999, and then continued throughout the first two decades of the 2000s [see also, @pelzer-boersma2019].^[Unfortunately, @pelzer-boersma2019 did not analyse the short vowels, it is therefore difficult to know whether the fronting of the short vowels was as pronounced in 2019.]

The high front vowels form a particularly interesting part of the space given that several possibly related processes might be ongoing. Besides the centralization of [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`], the visualizations and GAMMs suggested that F3 could not reliably distinguish between the two vowels, and that the differences in F1 and F2 were small. These findings conflicted with previous work on Central Swedish [e.g., @fant1969; @fujimura1967; @kuronen2000], and would seem to indicate three possibilities: 1) the contrast might be supported by additional acoustic cues not investigated in this study, 2) listeners might disambiguate the two categories using visual cues or linguistic information, or 3) perhaps these two categories are not distinguished at all, which would suggest a merger in process among these talkers. A merger might be driven by relaxation of lip-rounding, as supported by the higher F3 values found for [`r linguisticsdown::cond_cmpl("yː")`]. 

Both of these processes---centralization and sustained overlap despite all cues considered (F1, F2, F3, formant trajectories)---could however be explained by [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] being produced as damped versions. The presence of a damped [`r linguisticsdown::cond_cmpl("ɨː")`] would be supported by the lower F2 values, as the consonantal offglide in [`r linguisticsdown::cond_cmpl("ɨː")`] lowers F2 [@engstrand2000]. A merger of [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] into [`r linguisticsdown::cond_cmpl("ɨː")`] has been observed among younger talkers in other regions in Sweden. @gross2020 concluded that several of the talkers of Gothenburg Swedish in their study whose [`r linguisticsdown::cond_cmpl("iː")`] had a horizontally centralized quality closer to [`r linguisticsdown::cond_cmpl("yː")`], merged the two categories into [`r linguisticsdown::cond_cmpl("ɨː")`]. If centralization of [`r linguisticsdown::cond_cmpl("iː")`] is a prerequisite for the merger, as suggested by @gross2020, what we see here might be the beginning of a merger in process. A future scenario might involve the loss of the defining feature lip rounding in [`r linguisticsdown::cond_cmpl("yː")`] if a merger continues, given its weaker perceptual salience [as hypothesized by @gross2020].^[A parallel exist in iotacism in Norwegian and Swedish dialects [@eliasson2000], characterized by the loss of lip rounding resulting in vowel shifts such as [`r linguisticsdown::cond_cmpl("yː")`] to [`r linguisticsdown::cond_cmpl("iː")`] for monophthongs, and [`r linguisticsdown::cond_cmpl("øy")`] to [`r linguisticsdown::cond_cmpl("ei")`] for diphthongs. Iotacism describes a feature loss for a fairly unusual phonological characteristic---rounded front vowels. Talkers that merge because of iotacism would not necessarily produce [`r linguisticsdown::cond_cmpl("iː")`] as the damped [`r linguisticsdown::cond_cmpl("ɨː")`], which could potentially explain the vowel patterns of two talkers that merge the two vowels but have a rather fronted [`r linguisticsdown::cond_cmpl("iː")`] without an offglide consonant element (see \@ref(fig:swe-merge), for Talkers SW_005 and SW_015).] 

Impressionistic listening by the author did support the presence of a buzzing sound, similar to [`r linguisticsdown::cond_cmpl("ɨː")`], among the majority of talkers in SwehVd, both male and female. The strength and scope of [`r linguisticsdown::cond_cmpl("ɨː")`] varied across talkers, from a relatively strong [`r linguisticsdown::cond_cmpl("ɨː")`] (17 talkers), to a weaker buzzing sound (12 talkers), or more of a consonant element similar to [`r linguisticsdown::cond_cmpl("j")`] (3 talkers). Interestingly, several of the talkers that did not produce any apparent final consonant glide or buzz (11 talkers), seemed to have overall less retracted [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`], hence supporting the hypothesized link between centralization and consonantal offglide. Further insights into these individual differences can be gained by studying the SwehVd materials on a talker-specific level.

The changed placement of [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] might have affected the surrounding front vowels. For instance, [`r linguisticsdown::cond_cmpl("eː")`] was *not* the most fronted vowel for some of the talkers who do not centralize [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`]<!--(e.g., SW_005, SW_015, SW_020)-->. This would seem to suggest that the fronted placement of [`r linguisticsdown::cond_cmpl("eː")`] might be conditioned on a more centralized position of [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`]. The order and cause of these events is of course unclear. For instance, the raising of [`r linguisticsdown::cond_cmpl("eː")`] could be pushing back [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`]---a push chain shift [for reviews of chain shifts, see e.g., @king1969; @martinet1952]. However, what might be causing [`r linguisticsdown::cond_cmpl("eː")`] to move remains unclear, given that the neighbouring category [`r linguisticsdown::cond_cmpl("ɛː")`] substantially lowers. A future scenario might involve [`r linguisticsdown::cond_cmpl("eː")`] once more descending to a position closer to its shorter counterpart, and closer to the position it occupies in the latter part of the segment (see Figure \@ref(fig:trajectories)). Such a change would potentially result in a sparser front space, possibly setting the stage for a more compact system overall. Alternatively, the backing of [`r linguisticsdown::cond_cmpl("iː")`] and [`r linguisticsdown::cond_cmpl("yː")`] could be pulling [`r linguisticsdown::cond_cmpl("eː")`] upwards---a pull chain shift. A pull chain shift could hypothetically involve further movements downstream, for instance, a centralization of [`r linguisticsdown::cond_cmpl("ʉː")`], a lowering of [`r linguisticsdown::cond_cmpl("oː")`] and [`r linguisticsdown::cond_cmpl("uː")`], or a general backing of the entire system. Future studies should investigate the support for either hypothesis in more detail for further insight into the possible causes of the movements.

Other noticeable changes outside of the high front area concern [`r linguisticsdown::cond_cmpl("ɛː")`] and [`r linguisticsdown::cond_cmpl("œ")`], that have both lowered since 1999. The lowering of [`r linguisticsdown::cond_cmpl("ɛː")`] was anticipated by e.g., @riad2014, and confirmed in @leinonen2010 and @pelzer-boersma2019. Even if [`r linguisticsdown::cond_cmpl("œ")`] has lowered compared to SweDia, [`r linguisticsdown::cond_cmpl("øː")`] and [`r linguisticsdown::cond_cmpl("ø")`] are still located beneath [`r linguisticsdown::cond_cmpl("œ")`], which suggest a backing rather than lowering of /`r linguisticsdown::cond_cmpl("ø")`/ in position before retroflex segments [c.f., @riad2014]. Furthermore, the difference in placement of [`r linguisticsdown::cond_cmpl("œ")`] might result from a presence (SweDia) or lack of (SwehVd) neutralization of [`r linguisticsdown::cond_cmpl("ø")`] into [`r linguisticsdown::cond_cmpl("ɵ")`]. In both materials, [`r linguisticsdown::cond_cmpl("ø")`] appears in a context before a retroflex (SwehVd - *hörr*, SweDia - *dörr*), and in SweDia, [`r linguisticsdown::cond_cmpl("ø")`] is considerably closer to [`r linguisticsdown::cond_cmpl("ɵ")`] than in the SwehVd. If this result holds, it would suggest that the previously documented merger of short [`r linguisticsdown::cond_cmpl("ø")`] and [`r linguisticsdown::cond_cmpl("ɵ")`] might have lost its importance in Central Swedish over the last generation [e.g., @stahle1965; @wenner2010; but see @kotsinas1995 for younger Stockholm talkers].

## Methodological considerations
Methodological considerations for the present study concern 1) methods chosen to support claims about ongoing changes in the vowel space, 2) the choice of how to measure some of the acoustic cues, and 3) materials used for comparing vowel changes over time. Firstly, many acoustic-phonetic studies often include systematic listening by trained phoneticians, validated through measures of inter-rater reliability [for a review, see @cucchiarini1995; @gross2020; @kuronen2000; @pelzer-boersma2019]. In this study, proposals of the presence of a possible ongoing merger and/or centralization of [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`], and the absence of merger of [`r linguisticsdown::cond_cmpl("œ")`] - [`r linguisticsdown::cond_cmpl("ɵ")`], would all receive stronger support if supplemented by systematic (and not impressionistic) listening. A second consideration concerns two of the cue measurements, F0 and duration, and how they were measured and evaluated. Here, the average F0 across the vowel segment was reported. However, pitch contours have been known to influence perceived duration and prominence of vowels [e.g., @gussenhoven-zhou2013]. One could therefore claim that a more accurate acoustic representation of vowels should have included pitch contours. Similarly, the temporal analysis of the effect of duration on long-short vowel pairs could be supplemented with measures of consonant ratios [e.g., @pelzer-boersma2019; @schaeffler2005]. Both of these methodological considerations can be addressed in future studies as the SwehVd database is publicly available in an online repository (<https://osf.io/ruxnb/>).

Third and finally, there were differences in the materials compared in Study 2 that could have affected the results. For instance, the difference in size of the datasets was considerable. The sampling-with-replacement approach for the $op$ tried to address this imbalance. Furthermore, different phonological contexts were employed for recording of the two databases, which can constrain claims made about vowel changes given differences in potential effects of coarticulation. This limitation was partly mitigated by including word as random effect in the regression models. As confirmed by the GLMMs, word context *did* carry a substantial amount of the variability between datasets. Future research could address these limitations with more scrutiny, e.g., by providing models with more detailed information about the specific phonological context employed in each word. Across previous studies, there are substantial differences in word context, number of talkers recorded, number of repetitions per word, and the demographics of talkers (Table \@ref(tab:previous-work)). This variability implies that, irrespective of what reference materials chosen for comparison, there would inevitably be some loss in comparability. The SwehVd sought to address this issue by recording hVd words, with the intention of enhancing comparability in studies on Central Swedish, and with the welcome side-effect of facilitating cross-linguistic investigations.

# Conclusions
The present study has reported on the acoustic properties of Central Swedish vowels and how they have changed over the last generation. The spectral and temporal cues investigated all contributed to distinguishing between the 21 vowels in the Central Swedish vowel space, with varying weight. More insight into formant dynamics within and between quantities have been gained by the dynamic analysis presented, which is also of value for cross-linguistic research. 
What has been gained with the broad-scale approach of characterizing the *entire* vowel space adopted here, is of course lost in terms of detailed investigations of individual vowel contrasts. There is certainly a lot more to say about the centralization of [`r linguisticsdown::cond_cmpl("iː")`] - [`r linguisticsdown::cond_cmpl("yː")`], the potential loss of lip-rounding in [`r linguisticsdown::cond_cmpl("yː")`], the direction of chain shifts, and lowering of [`r linguisticsdown::cond_cmpl("ɛː")`], among other things. The acoustic descriptions outlined in this paper, together with the publicly available SwehVd database, can provide a reference point for future investigations into these acoustic events and beyond.

# Data availability statement {-}
The SwehVd dataset presented in this study can be found in an online repository (SwehVd: <https://osf.io/ruxnb/>). All analyses and visualization code can be found in a separate online repository (<https://osf.io/>).

# Funding {-}
The work presented in this study was partially funded by a grant from the Kinander's foundation (2021), a grant from Kungliga Vetenskapsakademien (2023), and by the Department of Swedish Language and Multilingualism at Stockholm University.

# Acknowledgments {-}
I thank audiences at the 2023 ASA meeting, at the Ludwig-Maximilians-Universität Institute for Phonetics and Speech Processing lab meeting October 2023, and at the Stockholm University Department of Swedish language and multilingualism for feedback on earlier presentations of this work. I am particularly grateful to Tomas Riad for discussion of Swedish vowel systems, Maryann Tan for collaboration in preparation of the SwehVd database, Jonathan Harrington for advice and support in implementing the orthogonal projection ratio, Michele Gubian for advice and support in implementing GAMMs, Anders Eriksson for providing the recordings of the SweDia reference talkers and to Nathan Young for feedback on segmentation procedures, as well as for access to, and support for, the [SweFA software](https://github.com/mcgarrah/LG-FAVE) for the automatic annotation of Swedish vowels. Finally, I thank Niclas Abrahamsson, Catrin Norrby, Tomas Riad, Maryann Tan and Carla Wikse Barrow for feedback on earlier versions of the manuscript.

# Conflict of Interest {-}
The author declares that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.

\newpage

# References

\newpage

# Supplementary information for *Persson (2024). The acoustic characteristics of Central Swedish vowels* {-}
Both the main text and these supplementary information (SI) are derived from the same R markdown document available via OSF at  [https://osf.io/](https://osf.io/)<!--enter html here-->.

## Required software {#sec:SI-software}

The document was compiled using \texttt{knitr} [@xie2023] in RStudio with R:

```{r}
version
```

The following R packages were used to create this document: `r cite_r("latex-stuff/r-referencesSweAcoust.bib")`.

```{r}

```

## Summary of previous studies on Central Swedish vowels

\begin{landscape}

```{r SI-previous-work}
# Add table of previous work on Central Swedish vowels
d.previous.work <- tibble(
  Article = c("Bruce, Elert, Engstrand & Wretling, 1999",
              "Eklund & Traunmüller, 1997",
              "Fant, Henningsson & Stålhammar, 1969",
              "Kuronen, 2000",
              "Leinonen, 2010",
              "Lindblom, 1963",
              "McAllister, Lubker & Carlson, 1974",
              "Pelzer & Boersma, 2019",
              "Schötz, Frid & Löfqvist, 2011",
              "Wenner, 2010"),
  "Speech materials" = c("3-5 repetitions of all 21 vowels, different phonological contexts for each vowel",
                         "3 repetitions of all 9 long vowels in isolation",
                         "1 repetition of all 9 long vowels in isolation",
                         "5-7 repetitions of all 21 vowels, different phonological contexts for each vowel and each repetition",
                         "a subset of 19 vowels from the SweDia database (see Bruce et al., 1999 above)",
                         "5 repetitions of all 8 short vowels in bVb, dVd, gVg contexts, with four different stress patterns",
                         "10 repetitions of all 6 long rounded vowels in itV context",
                         "8 repetitions of all 9 long vowels, different phonological contexts for each vowel and each repetition",
                         "6 repetitions of the long /i:/ in two different contexts (bibel, papipa)",
                         "3 repetitions of short /œ/ and /ɵ/ in different phonological contexts"),
  Participants = c("12 talkers (6 female; 6 young) each from 107 locations across Sweden and Finland; 8 reference talkers of standard Central Swedish from the Greater Stockholm area (4 female)",
                   "12 talkers (6 female) from the Greater Stockholm area, 20-58 years of age",
                   "24 male talkers, students at KTH in Stockholm",
                   "4 male talkers from Nyköping, 17-18 years of age (4 female reference talkers), recorded by Leinonen, Pitkänen & Vihanta, 1982",
                   "see Bruce et al., 1999",
                   "1 male talker from Stockholm, 19 years of age",
                   "6 talkers of standard Swedish",
                   "8 talkers (4 female) from Stockholm",
                   "1 male talker from Stockholm",
                   "78 talkers (40 female) from 4 locations in Uppland, 12-85 years of age"),
  Approach = c("",
               "formant trajectories at 10 measurement points; F1-F2 means and sd; linear regression",
               "F1 to F4 and duration means",
               "formant trajectories at XX measurement points (long vowels); F1 to FXX means",
               "principal component analysis of Bark-filtered vowel spectra",
               "",
               "EMG; F1 to F3 means",
               "F1-F2 median values at 20, 50, 80% into the vowel; linear mixed effects model",
               "articulography; F1 to F4 means; Bark-circles",
               "F1 to F3 means; linear regression; correlation analysis"),
  Focus = c("SweDia dialect database development",
            "Comparing the acoustics and perception of whispered and phonated vowels",
            "Formant frequencies of the long vowels",
            "Spectral and temporal acoustics of all Central Swedish vowels",
            "Acoustic characteristics of Swedish rural dialects",
            "The effect of vowel duration on formant frequencies",
            "The articulation and acoustics of rounded vowels",
            "Diphthongization in the long vowels",
            "The articulation and acoustics of the damped /i:/",
            "The merger of short /œ/ and /ɵ/"))

kable(d.previous.work,
      format = "latex",
      booktabs = TRUE,
      caption = "A selection of previous studies on Central Swedish vowels") %>%
  kable_styling(
    font_size = 7,
    latex_options = "striped") %>%
  column_spec(1, width = "2.5cm") %>%
  column_spec(2:5, width = "4.5cm")
```

\end{landscape}

## Additional information on Study 1

### Cue matrix plot in unnormalized Hertz {#sec:SI-raw}
Figure \@ref(fig:SI-swe-vowels-all-cues-hz) displays the pairwise combinations of the five spectral and temporal cues---F0, F1, F2, F3 and duration---in raw Hertz. For F0, there are two overlapping but separated peaks, reflecting the overall difference in vocal tract size for the two groups of female and male talkers in the database.

(ref:SI-swe-vowels-all-cues-hz) The SwehVd vowels in unnormalized Hertz space. Points show recordings of each of the 21 Central Swedish vowels by the talkers in the database in all five cue spaces, averaged across the three midpoints. Vowel labels indicate category means across talkers. Ellipses show bivariate Gaussian 95% confidence interval of category means. Note that, unlike in Figure \@ref(fig:swe-vowels), axis directions are not reversed. Vowels that mismatched the intended label are excluded (`r signif((d.SwehVd %>% filter(Transcribed_vowel != "targeted") %>% nrow() / d.SwehVd %>% nrow()) * 100)`% of all recordings).

```{r SI-swe-vowels-all-cues-hz, fig.width=base.width*4+.5, fig.height=base.height*4+.5, out.width = '100%', fig.align='center', fig.cap="(ref:SI-swe-vowels-all-cues-hz)"}
p.matrix.cues %+%
  (d.SwehVd.forStudy.long.static %>%
     filter(Normalization.Type == "no normalization (Hz)", F0 < 320))
```

### Neutralization of short /e/ and /`r linguisticsdown::cond_cmpl("ɛ")`/ {#sec:SI-hedd-hädd}
Figure \@ref(fig:SI-hedd) demonstrates that the short allophones of /`r linguisticsdown::cond_cmpl("e")`/ and /`r linguisticsdown::cond_cmpl("ɛ")`/ neutralize as [`r linguisticsdown::cond_cmpl("ɛ")`] in Central Swedish.

(ref:SI-hedd) The *hedd* and *hädd* words in the SwehVd vowel data in F1-F2 space. Points show recordings of the *hedd* and *hädd* words ([`r linguisticsdown::cond_cmpl("ɛ")`]) by 44 native talkers in the database, averaged across the five measurement points within each vowel segment. Word labels indicate word means across talkers. Since *hädd* and *hedd* resulted in the same allophone, all *hädd* words are excluded from the subset used in this study. This facilitates comparison of, for example, densities across vowels (see diagonal of Figure \@ref(fig:swe-vowels-all-cues)).

```{r SI-hedd, fig.width=base.width*1.5, fig.height=base.height*1.5, out.width='50%', fig.align='center', fig.cap="(ref:SI-hedd)"}
d.SwehVd %>%
  filter(Transcribed_vowel == "targeted", Word %in% c("hedd", "hädd")) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Talker, Word, Token) %>%
  summarise(across(c(F1, F2), ~ geometric.mean(.x))) %>%
  ggplot(
    aes(
      x = F2,
      y = F1,
      color = Word)) +
  geom_point(alpha = .4) +
  stat_ellipse() +
  # Add between-talker mean
  geom_label(
    data =
      ~ .x %>%
      group_by(Word) %>%
      summarise(across(c(F1, F2), ~ geometric.mean(.x))),
    mapping = aes(
      colour = Word,
      label = Word),
    alpha = .8,
    size = 4,
    label.size = NA) +
  scale_x_reverse("F2 (in Hz)", position = "top", breaks = scales::breaks_pretty(6)) +
  scale_y_reverse("F1 (in Hz)", position = "right", breaks = scales::breaks_pretty(6)) +
  guides(color = "none") +
  theme(legend.position = "top")
```

### The effect of different measurement points {#sec:SI-measurement-points}
Figure \@ref(fig:SI-time-points) visualizes the SwehVd vowel space under four different assumptions about measurement points. The figure suggests that the use of different measurement points primarily affects vowels for which formant dynamics are particularly important, e.g., [`r linguisticsdown::cond_cmpl("eː")`] and [`r linguisticsdown::cond_cmpl("oː")`]. 

(ref:SI-time-points) The SwehVd vowel data in F1-F2 space, faceted by the time points used for measuring formants: averaging across the first three time-points (at 20, 35, and 50% into the vowel; **panel A**), averaging across the mid three time-points (at 35, 50, and 65% into the vowel; **panel B**), taking the mid time-point (at 50%; **panel C**), and averaging across the final three mid-points (at 50, 65, and 80% into the vowel; **panel D**). Points show recordings of each of the 21 Central Swedish vowels by the 44 L1 talkers in the database. Vowel labels indicate category means across talkers. Long vowels are boldfaced. Vowels that mismatched intended label are excluded (`r signif((d.SwehVd %>% filter(Transcribed_vowel != "targeted") %>% nrow() / d.SwehVd %>% nrow()) * 100)`% of all recordings).

```{r SI-time-points, fig.width=base.width*4.5, fig.height=base.height*4.5+.5, out.width = '100%', fig.align='center', fig.cap="(ref:SI-time-points)"}
p.20_50 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(20,35,50)) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x)))) +
  guides(shape = "none")

p.35_65 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(35,50,65)) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x)))) +
  guides(shape = "none")

p.50 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint == 50)) +
  guides(shape = "none")

p.50_80 <- p %+%
  (d.SwehVd.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(50,65,80)) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x)))) +
  guides(shape = "none")

(p.20_50 + p.35_65) / (p.50 + p.50_80) +
  plot_annotation(tag_levels = list(c("A", "B", "C", "D"))) +
  plot_layout(
    guides = "collect") & 
  theme(legend.position = "right",
        legend.justification = "right",
        legend.margin = margin(0, 0, 0, 0, unit = "pt"),
        plot.tag = element_text(face = "bold"))
```

### Mean cue values across talkers {#sec:SI-mean-cues}
The overall mean cue values for the long and short vowels are presented in Tables \@ref(tab:SI-cue-values-long) and \@ref(tab:SI-cue-values-short), respectively.

\begin{landscape}

\begin{table}

\caption{\label{tab:SI-cue-values-long}The spectral and temporal properties of the long vowels (averaged across talkers, by gender)}
\centering
\fontsize{6}{8}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}}
\toprule
Cue & Gender & `r linguisticsdown::cond_cmpl("[iː]")`& `r linguisticsdown::cond_cmpl("[yː]")`& `r linguisticsdown::cond_cmpl("[ʉː]")` & `r linguisticsdown::cond_cmpl("[eː]")` & `r linguisticsdown::cond_cmpl("[ɛː]")` & `r linguisticsdown::cond_cmpl("[æː]")` & `r linguisticsdown::cond_cmpl("[øː]")` & `r linguisticsdown::cond_cmpl("[œː]")` & `r linguisticsdown::cond_cmpl("[ɑː]")` & `r linguisticsdown::cond_cmpl("[oː]")` & `r linguisticsdown::cond_cmpl("[uː]")`\\
\midrule
Duration & F & 0.23 (SD=0.05) & 0.24 (SD=0.05) & 0.23 (SD=0.05) & 0.26 (SD=0.05) & 0.25 (SD=0.04) & 0.25 (SD=0.05) & 0.25 (SD=0.04) & 0.26 (SD=0.05) & 0.23 (SD=0.05) & 0.25 (SD=0.05) & 0.23 (SD=0.05)\\
 & M & 0.21 (SD=0.05) & 0.22 (SD=0.06) & 0.22 (SD=0.06) & 0.23 (SD=0.06) & 0.24 (SD=0.06) & 0.24 (SD=0.06) & 0.23 (SD=0.05) & 0.23 (SD=0.06) & 0.23 (SD=0.06) & 0.24 (SD=0.05) & 0.22 (SD=0.05)\\
F0 & F & 200.53 (SD=27.45) & 196.29 (SD=30.09) & 196.1 (SD=26.89) & 195.04 (SD=25.38) & 189.86 (SD=24.03) & 187.74 (SD=25.95) & 189.77 (SD=25) & 190.65 (SD=27.04) & 192.84 (SD=25.61) & 195.51 (SD=29.33) & 199.89 (SD=26.49)\\
 & M & 118.09 (SD=22.59) & 114.94 (SD=21.79) & 116.85 (SD=22.18) & 116.12 (SD=19.12) & 109.89 (SD=22.61) & 111.56 (SD=23.07) & 112.17 (SD=28.07) & 114.12 (SD=29.46) & 113.37 (SD=NA) & 122.23 (SD=26.64) & 116.84 (SD=20.36)\\
F1 & F & 374.2 (SD=40.95) & 361.09 (SD=34.22) & 391.34 (SD=30.39) & 424.37 (SD=44.27) & 791.24 (SD=101.72) & 848.7 (SD=115.25) & 638.55 (SD=76.26) & 614.75 (SD=75.62) & 649.15 (SD=83.44) & 440.1 (SD=46.87) & 396.75 (SD=37.13)\\
\addlinespace
 & M & 315.64 (SD=37.31) & 302.43 (SD=36.81) & 333.44 (SD=41.18) & 393.6 (SD=41.58) & 682.61 (SD=69.57) & 726.06 (SD=68.57) & 553.09 (SD=41.36) & 548.04 (SD=42.07) & 582.84 (SD=54.5) & 424.82 (SD=30.21) & 354.55 (SD=32.95)\\
F2 & F & 2093.73 (SD=217.16) & 2009 (SD=186.87) & 1783.62 (SD=113.99) & 2430.25 (SD=300.79) & 1794.7 (SD=129.12) & 1648.51 (SD=86.13) & 1525.93 (SD=99.29) & 1299.7 (SD=110.97) & 1050.33 (SD=105.69) & 760.92 (SD=94.01) & 789.15 (SD=108.23)\\
 & M & 1887.17 (SD=171.28) & 1843.46 (SD=155.66) & 1590.83 (SD=88.26) & 2132.97 (SD=174.9) & 1587.19 (SD=112.9) & 1463.63 (SD=114.66) & 1358.95 (SD=106.46) & 1134.96 (SD=92.03) & 893.01 (SD=88.48) & 703.9 (SD=80.61) & 762.44 (SD=76.79)\\
F3 & F & 3173.35 (SD=351.41) & 3270.07 (SD=136.69) & 2655.81 (SD=165.76) & 3077.56 (SD=168.89) & 2785.39 (SD=244.12) & 2754.6 (SD=265.82) & 2690.74 (SD=142.9) & 2743.12 (SD=140.38) & 2864.53 (SD=195.42) & 2910.92 (SD=180.43) & 2835.14 (SD=181.03)\\
 & M & 2946.54 (SD=152.43) & 2837.89 (SD=137.97) & 2364.8 (SD=131.58) & 2697.06 (SD=166.57) & 2491.89 (SD=185.15) & 2495.84 (SD=172.42) & 2370.48 (SD=172.29) & 2417.6 (SD=164.1) & 2638.12 (SD=215.24) & 2645.54 (SD=166.4) & 2567.77 (SD=151.27)\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

```{r SI-cue-values-long}
# Keep code for reproducibility
# d.cue.values.forTable <- d.cue.values %>%
#   group_by(Quantity, Gender, category) %>%
#   mutate(
#     F0 = paste0(F0.mean, " (SD=", F0.sd, ")"),
#     F1 = paste0(F1.mean, " (SD=", F1.sd, ")"),
#     F2 = paste0(F2.mean, " (SD=", F2.sd, ")"),
#     F3 = paste0(F3.mean, " (SD=", F3.sd, ")"),
#     Duration = paste0(Duration.mean, " (SD=", Duration.sd, ")")) %>%
#   ungroup() %>%
#   select(-c(contains(c(".mean", ".sd")), Quantity)) %>%
#   arrange(factor(category, levels = levels.vowel.IPA.swe))
# 
# d.cue.values.forTable %<>%
#   pivot_longer(
#     cols = c(starts_with("F"), "Duration"),
#     names_to = "Cue",
#     values_to = "Cue.Value") %>%
#   pivot_wider(
#     names_from = "category",
#     values_from = "Cue.Value") %>%
#   relocate("Gender", .after = Cue) %>%
#   arrange(Cue) %>%
#   mutate(
#     Gender = ifelse(Gender == "female", "F", "M"),
#     Cue = ifelse(row_number() %% 2 == 0, "", Cue))
# 
# kable(d.cue.values.forTable %>%
#         select(Cue, Gender, contains(levels.vowel.IPA.swe.long)),
#       format = "latex",
#       booktabs = TRUE,
#       caption = "The spectral and temporal properties of the long vowels (averaged across talkers, by gender)") %>%
#   kable_styling(font_size = 6) %>%
#   column_spec(1:2, width = ".7cm") %>%
#   column_spec(3:13, width = "1.4cm")
```

\begin{landscape}

\begin{table}

\caption{\label{tab:SI-cue-values-short}The spectral and temporal properties of the short vowels (averaged across the female and male talkers)}
\centering
\fontsize{6}{8}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{.7cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}>{\raggedright\arraybackslash}p{1.4cm}}
\toprule
Cue & Gender & `r linguisticsdown::cond_cmpl("[ɪ]")` & `r linguisticsdown::cond_cmpl("[ʏ]")` & `r linguisticsdown::cond_cmpl("[ɵ]")` & `r linguisticsdown::cond_cmpl("[ɛ]")` & `r linguisticsdown::cond_cmpl("[æ]")` & `r linguisticsdown::cond_cmpl("[ø]")` & `r linguisticsdown::cond_cmpl("[œ]")` & `r linguisticsdown::cond_cmpl("[a]")` & `r linguisticsdown::cond_cmpl("[ɔ]")` & `r linguisticsdown::cond_cmpl("[ʊ]")`\\
\midrule
Duration & F & 0.12 (SD=0.02) & 0.12 (SD=0.02) & 0.13 (SD=0.03) & 0.12 (SD=0.02) & 0.16 (SD=0.03) & 0.13 (SD=0.03) & 0.17 (SD=0.03) & 0.13 (SD=0.03) & 0.13 (SD=0.03) & 0.13 (SD=0.03)\\
 & M & 0.11 (SD=0.02) & 0.11 (SD=0.03) & 0.13 (SD=0.02) & 0.12 (SD=0.02) & 0.14 (SD=0.03) & 0.12 (SD=0.03) & 0.14 (SD=0.03) & 0.13 (SD=0.03) & 0.12 (SD=0.03) & 0.12 (SD=0.03)\\
F0 & F & 216.61 (SD=NA) & 219.19 (SD=29.02) & 216.44 (SD=31.66) & 210.79 (SD=27.89) & 202.46 (SD=28.44) & 211.95 (SD=28.48) & 204.68 (SD=29.17) & 205.66 (SD=28.99) & 211.46 (SD=30.03) & 224.54 (SD=28.62)\\
 & M & 131.12 (SD=20.59) & 132.02 (SD=22.6) & 131.06 (SD=20.07) & 121.65 (SD=NA) & 122.6 (SD=37.4) & 128.09 (SD=33.82) & 126.12 (SD=29.65) & 117.64 (SD=NA) & 126.47 (SD=NA) & 131.03 (SD=19.12)\\
F1 & F & 380.91 (SD=38.71) & 383.73 (SD=37.34) & 422.14 (SD=37.68) & 627.89 (SD=70.76) & 801.77 (SD=97.71) & 569.1 (SD=63.26) & 533.26 (SD=55.52) & 796.35 (SD=101.31) & 500.55 (SD=53.01) & 418.77 (SD=37.08)\\
\addlinespace
 & M & 332.8 (SD=28.49) & 337.95 (SD=29.57) & 387.37 (SD=31.06) & 555.36 (SD=47.99) & 722.32 (SD=59.41) & 504.93 (SD=39.4) & 505.32 (SD=38.18) & 690.47 (SD=54.28) & 471.87 (SD=31.94) & 375.94 (SD=26.94)\\
F2 & F & 2526.8 (SD=283.27) & 2372.74 (SD=152.56) & 1322.19 (SD=112.94) & 1961.78 (SD=179.47) & 1762.92 (SD=141.77) & 1666.89 (SD=110.52) & 1399.73 (SD=151.46) & 1367.24 (SD=98.14) & 932.73 (SD=95.5) & 854.43 (SD=109.58)\\
 & M & 2257.6 (SD=149.07) & 2030.22 (SD=124.31) & 1162.53 (SD=95.09) & 1714 (SD=123.35) & 1510.17 (SD=116.33) & 1432.82 (SD=144.94) & 1190.35 (SD=82.51) & 1139.67 (SD=92.26) & 782.81 (SD=80.04) & 774.96 (SD=81.75)\\
F3 & F & 3085.16 (SD=168.28) & 2798.4 (SD=160.74) & 2781.55 (SD=188.02) & 2760.26 (SD=248.05) & 2810.87 (SD=260.86) & 2706.12 (SD=155.43) & 2781.02 (SD=171.5) & 2850.9 (SD=259.87) & 3022.98 (SD=190.97) & 2939.17 (SD=182.25)\\
 & M & 2772.45 (SD=180.21) & 2479.21 (SD=128.84) & 2459.57 (SD=148.01) & 2522.15 (SD=154.8) & 2545.13 (SD=165.44) & 2424.64 (SD=134.98) & 2470.4 (SD=165.12) & 2659.06 (SD=204.29) & 2716.22 (SD=182.89) & 2596.2 (SD=158)\\
\bottomrule
\end{tabular}
\end{table}

\end{landscape}
```{r SI-cue-values-short}
# Keep code for reproducibility
#For each cue, the highest and lowest mean and SD should be boldfaced.
# kable(d.cue.values.forTable %>%
#         select(Cue, Gender, contains(levels.vowel.IPA.swe.short)),
#       format = "latex",
#       booktabs = TRUE,
#       caption = "The spectral and temporal properties of the short vowels (averaged across the female and male talkers)") %>%
#   kable_styling(font_size = 6) %>%
#   column_spec(1:2, width = ".7cm") %>%
#   column_spec(3:12, width = "1.4cm")
```
```{r}
# itsadug::plot_smooth(m.F3.short, view = "timepoint", plot_all = "category")
# itsadug::plot_smooth(m.F3.long, view = "timepoint", plot_all = "category")
#itsadug::plot_diff(m.F3.long, view = "timepoint", comp = list(category = c("[yː]", "[iː]")))
```

### Summary tables of GAMMs {#sec:SI-GAMMs}
<!--add tables here after finished GAMMs-->
\begin{table}[ht]
\caption{\label{tab:m-F1-long}Model summary, coefficients for parametric and smooth terms for the long vowels predicting F1} 
\centering
\begin{tabular}{lrrrr}
   \hline
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
  (Intercept) & -1.2711 & 0.0149 & -85.1342 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[yː]")` & -0.0409 & 0.0157 & -2.6140 & 0.0090 \\ 
  category `r linguisticsdown::cond_cmpl("[ʉː]")` & 0.0474 & 0.0157 & 3.0261 & 0.0025 \\ 
  category `r linguisticsdown::cond_cmpl("[eː]")` & 0.1806 & 0.0157 & 11.4948 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɛː]")` & 0.7606 & 0.0157 & 48.5246 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[æː]")` & 0.8303 & 0.0157 & 52.9201 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[øː]")` & 0.5467 & 0.0157 & 34.8945 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[œː]")` & 0.5258 & 0.0157 & 33.5283 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[ɑː]")` & 0.5824 & 0.0157 & 37.1839 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[oː]")` & 0.2163 & 0.0157 & 13.8148 & $<$ 0.0001 \\ 
  category `r linguisticsdown::cond_cmpl("[uː]")` & 0.0872 & 0.0157 & 5.5664 & $<$ 0.0001 \\ 
  Gendermale & -0.0052 & 0.0156 & -0.3311 & 0.7406 \\ 
   \hline
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[iː]")` & 2.8761 & 3.3895 & 56.5201 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[yː]")` & 3.0049 & 3.5055 & 64.6035 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ʉː]")` & 3.5357 & 3.8769 & 187.3821 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[eː]")` & 3.9046 & 3.9945 & 955.4497 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɛː]")` & 2.4942 & 3.0111 & 8.0729 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[æː]")` & 3.3737 & 3.7852 & 142.0491 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[øː]")` & 1.9113 & 2.3535 & 1.3956 & 0.2221 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[œː]")` & 2.3195 & 2.8110 & 34.9363 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[ɑː]")` & 2.3980 & 2.8964 & 60.7066 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[oː]")` & 3.8313 & 3.9825 & 600.4801 & $<$ 0.0001 \\ 
  s(timepoint):category `r linguisticsdown::cond_cmpl("[uː]")` & 3.1676 & 3.6365 & 90.2477 & $<$ 0.0001 \\ 
  s(Talker) & 34.2739 & 42.0000 & 3313.6581 & $<$ 0.0001 \\ 
  s(Talker,category) & 421.7301 & 472.0000 & 99.0617 & 0.0003 \\ 
   \hline
\end{tabular}
\caption{Model summary, coefficients for parametric and smooth terms} 
\label{tab:m.F1.long}
\end{table}


```{r}
# Code kept for reproducibility
#gamtabs(m.F1.long, caption='Model summary, coefficients for parametric and smooth terms', label = "tab:m.F1.long")
#gamtabs(m.F3.short, caption='Model summary, coefficients for parametric and smooth terms', label = "tab:m.F3.short")
```

### Visualizations of GAMMs

(ref:SI-m-F1-long) Fitted smooths of GAMM for predicting F1 and 95% confidence intervals for all long vowels.

```{r SI-m-F1-long, message=FALSE, fig.width=base.width*3, fig.height=base.height*2.5, out.width='60%', fig.align='center', fig.cap="(ref:SI-m-F1-long)"}
itsadug::plot_smooth(m.F1.long, view = "timepoint", plot_all = "category", col = c("#66C2A5", "#FC8D62", "#8DA0CB", "#FB9A99", "#E7298A", "#A6D854", "#FFD92F", "#E5C494", "#B3B3B3", "#A6CEE3", "#B2DF8A"), print.summary = FALSE)
```

(ref:SI-m-F1-short) Fitted smooths of GAMM for predicting F1 and 95% confidence intervals for all short vowels.

```{r SI-m-F1-short, message=FALSE, fig.width=base.width*3, fig.height=base.height*2.5, out.width='60%', fig.align='center', fig.cap="(ref:SI-m-F1-short)"}
itsadug::plot_smooth(m.F1.short, view = "timepoint", plot_all = "category", col = c("#1B9E77", "#D95F02", "#7570B3", "#E78AC3", "#66A61E", "#E6AB02", "#A6761D", "#666666", "#1F78B4", "#33A02C"), print.summary = FALSE)
```

(ref:SI-m-F2-long) Fitted smooths of GAMM for predicting F2 and 95% confidence intervals for all long vowels.

```{r SI-m-F2-long, message=FALSE, fig.width=base.width*3, fig.height=base.height*2.5, out.width='60%', fig.align='center', fig.cap="(ref:SI-m-F2-long)"}
itsadug::plot_smooth(m.F2.long, view = "timepoint", plot_all = "category", col = c("#66C2A5", "#FC8D62", "#8DA0CB", "#FB9A99", "#E7298A", "#A6D854", "#FFD92F", "#E5C494", "#B3B3B3", "#A6CEE3", "#B2DF8A"), print.summary = FALSE)
```

(ref:SI-m-F2-short) Fitted smooths of GAMM for predicting F2 and 95% confidence intervals for all short vowels.

```{r SI-m-F2-short, message=FALSE, fig.width=base.width*3, fig.height=base.height*2.5, out.width='60%', fig.align='center', fig.cap="(ref:SI-m-F1-short)"}
itsadug::plot_smooth(m.F2.short, view = "timepoint", plot_all = "category", col = c("#1B9E77", "#D95F02", "#7570B3", "#E78AC3", "#66A61E", "#E6AB02", "#A6761D", "#666666", "#1F78B4", "#33A02C"), print.summary = FALSE)
```

(ref:SI-m-F3-long) Fitted smooths of GAMM for predicting F3 and 95% confidence intervals for all long vowels.

```{r SI-m-F3-long, message=FALSE, fig.width=base.width*3, fig.height=base.height*2.5, out.width='60%', fig.align='center', fig.cap="(ref:m-F3-long)"}
itsadug::plot_smooth(m.F3.long, view = "timepoint", plot_all = "category", col = c("#66C2A5", "#FC8D62", "#8DA0CB", "#FB9A99", "#E7298A", "#A6D854", "#FFD92F", "#E5C494", "#B3B3B3", "#A6CEE3", "#B2DF8A"), print.summary = FALSE)
```

(ref:SI-m-F3-short) Fitted smooths of GAMM for predicting F3 and 95% confidence intervals for all short vowels.

```{r SI-m-F3-short, message=FALSE, fig.width=base.width*3, fig.height=base.height*2.5, out.width='60%', fig.align='center', fig.cap="(ref:m-F3-short)"}
itsadug::plot_smooth(m.F3.short, view = "timepoint", plot_all = "category", col = c("#1B9E77", "#D95F02", "#7570B3", "#E78AC3", "#66A61E", "#E6AB02", "#A6761D", "#666666", "#1F78B4", "#33A02C"), print.summary = FALSE)
```

## Additional information on Study 2 {#sec:SI-SweDia}

### Wordlist used for recording of SweDia {#sec:SI-SweDia-wordlist}

\begin{table}
\caption{\label{tab:word-list}Words recorded by talkers of Central Swedish for the SweDia database}
\centering
\begin{tabular}[t]{c|c|c}
\hline
Vowel IPA&Target words&Comment\\
\hline
`r linguisticsdown::cond_cmpl("[æː]")`&lär&\\
\hline
`r linguisticsdown::cond_cmpl("[øː]")`&blöt&\\
\hline
&söt&\\
\hline
&lös&\\
\hline
`r linguisticsdown::cond_cmpl("[ɛː]")`&läs&missing for ym4 and yw4\\
\hline
&nät&\\
\hline
&väg&recorded by ym4 only\\
\hline
`r linguisticsdown::cond_cmpl("[uː]")`&sot&\\
\hline
`r linguisticsdown::cond_cmpl("[ʉː]")`&lus&\\
\hline
`r linguisticsdown::cond_cmpl("[yː]")`&typ&\\
\hline
`r linguisticsdown::cond_cmpl("[oː]")`&lås&\\
\hline
&låt&recorded by ym4 only\\
\hline
`r linguisticsdown::cond_cmpl("[eː]")`&leta&\\
\hline
&gles&recorded by ym4 and yw4 only\\
\hline
`r linguisticsdown::cond_cmpl("[œː]")`&dör&\\
\hline
`r linguisticsdown::cond_cmpl("[iː]")`&dis&\\
\hline
`r linguisticsdown::cond_cmpl("[ɑː]")`&lat&\\
\hline
&tak&recorded by ym4 and yw4 only\\
\hline
`r linguisticsdown::cond_cmpl("[ɪ]")`&sitt&\\
\hline
&disk&recorded by ym4 and yw4 only\\
\hline
&fisk&\\
\hline
`r linguisticsdown::cond_cmpl("[ɔ]")`&lott&\\
\hline
&såll&\\
\hline
`r linguisticsdown::cond_cmpl("[ʏ]")`&flytta&\\
\hline
&tysk&\\
\hline
&sytt&\\
\hline
`r linguisticsdown::cond_cmpl("[a]")`&lass&\\
\hline
&tack&recorded by ym4 only\\
\hline
`r linguisticsdown::cond_cmpl("[ɛ]")`&lätt&\\
\hline
&lett&\\
\hline
&vägg&recorded by ym4 only\\
\hline
`r linguisticsdown::cond_cmpl("[æ]")`&särk&recorded by ym4 and yw4 only\\
\hline
`r linguisticsdown::cond_cmpl("[ɵ]")`&ludd&\\
\hline
`r linguisticsdown::cond_cmpl("[œ]")`&dörr&\\
\hline
`r linguisticsdown::cond_cmpl("[ø]")`&blött&\\
\hline
`r linguisticsdown::cond_cmpl("[ʊ]")`&snott&\\
\hline
\end{tabular}
\end{table}

### Additional visualizations of the SweDia materials {#sec:SI-raw-Swedia}
(ref:SI-raw-Swedia) The SweDia reference talkers' vowel data along F1-F2, in unnormalized Hertz (*left*) and Nearey's uniform scaling space (*right*). Points show the productions of each of the 21 Central Swedish vowels by the 8 (4 female) L1 reference talkers in the database. Vowel labels indicate category means across talkers. Vowels that mismatched intended label are excluded (`r signif((d.SwehVd %>% filter(Transcribed_vowel != "targeted") %>% nrow() / d.SwehVd %>% nrow()) * 100)`% of all recordings).

```{r SI-raw-Swedia, fig.width=base.width*4.5, fig.height=base.height*3.5, out.width='90%', fig.align='center', fig.cap="(ref:SI-raw-Swedia)"}
p %+%
  (d.SweDia2000.forStudy.long %>%
     filter(Normalization.Type %in% c("no normalization (Hz)", "Nearey's uniform scaling (log)"), F1 < 1000, timepoint %in% c(35, 50, 65)) %>%
     # Get the geometric mean F1-F2 across the five time points
     group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
     summarise(across(c(F1, F2, F3, Duration), ~ mean(.x))))
```

(ref:SI-Swedia-normalized) The SweDia reference talkers' vowel data in F1-F2 Nearey's uniform scaling space, separated by quantity. Ellipses show bivariate Gaussian 95% confidence interval of category means. Vowel labels indicate category means across female and male talkers.

```{r SI-Swedia-normalized, fig.width=base.width*4.5, fig.height=base.height*2.5, fig.align='center', fig.cap="(ref:SI-Swedia-normalized)"}
p.normalized %+%
  (d.SweDia2000.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(35, 50, 65)) %>%
  # Get the geometric mean F1-F2 across the five time points
  group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
  summarise(across(c(F1, F2, F3, Duration), ~ mean(.x))))
```

(ref:SI-swe-vowels-all-cues-nearey-swedia) The SweDia vowels in Nearey's uniform scaling space. Points show recordings of each of the 21 Central Swedish vowels by the talkers in the database in all five cue spaces, averaged across the three midpoints. Vowel labels indicate category means across talkers. Ellipses show bivariate Gaussian 95% confidence interval of category means. Note that, unlike in Figure \@ref(fig:swe-vowels), axis directions are not reversed. Vowels that mismatched the intended label are excluded (`r signif((d.SwehVd %>% filter(Transcribed_vowel != "targeted") %>% nrow() / d.SwehVd %>% nrow()) * 100)`% of all recordings).

```{r SI-swe-vowels-all-cues-nearey-swedia, fig.width=base.width*4+.5, fig.height=base.height*4+.5, out.width = '100%', fig.align='center', fig.cap="(ref:SI-swe-vowels-all-cues-nearey-swedia)"}
set.seed(63720472)

p.matrix.cues %+%
  (d.SweDia2000.forStudy.long %>%
     filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint %in% c(35, 50, 65)) %>%
     # Get the geometric mean F1-F2 across the five time points
     group_by(Normalization.Type, Talker, Gender, category, Token, Quantity) %>%
     summarise(across(c(F0, F1, F2, F3, Duration), ~ mean(.x))) %>%
     # For plotting reasons, randomly select 3 tokens of each category
     group_by(Talker, category) %>%
     slice_sample(n = 3))
```

(ref:trajectories-swedia) The trajectory of all vowels in the SweDia reference materials across the five time points, along F1-F2. The arrow indicates the direction of the trajectory and ends at 80% into the vowel. The vowel label is placed at the third time-point, at vowel mid-point (50%).

```{r trajectories-swedia, fig.width=base.width*3.5, fig.height=base.height*2, out.width='90%', fig.align='center', fig.cap="(ref:trajectories-swedia)"}
d.SweDia2000.forStudy.long %>%
  filter(Normalization.Type == "Nearey's uniform scaling (log)") %>%
  mutate(timepoint = as.character(timepoint)) %>%
  group_by(Talker, category, timepoint, Quantity) %>%
  summarise(across(c(F2, F1), ~mean(.x))) %>%
  group_by(category, timepoint, Quantity) %>%
  summarise(across(c(F2, F1), ~mean(.x))) %>%
  ggplot(
    aes(
      x = F2,
      y = F1,
      color = category,
      group = NULL)) +
  geom_path(
    alpha = .5,
    arrow = arrow(ends = "last", type = "closed", angle = "22.5", length = unit(.1, "cm"))) +
  geom_label(
    data = . %>%
      filter(timepoint == "50"),
    mapping = aes(
      colour = category,
      label = category,
      fontface = ifelse(category %in% levels.vowel.IPA.swe.long, 2, 1)),
    alpha = .6,
    size = 4,
    label.size = NA,
    label.padding = unit(.01, "cm")) +
  # geom_point(
  #   data = d.F2d %>%
  #     #get the mean F1 F2 at 10% into d-closure across talkers
  #     summarise(
  #       F1 = mean(F1_closure),
  #       F2 = mean(F2_closure)),
  #   color = "black",
  #   alpha = .6) +
  scale_color_manual(name = "category", values = colors.vowel.swe) +
  scale_x_reverse("F2", position = "top", scales::pretty_breaks(n = 6)) +
  scale_y_reverse("F1", position = "right", scales::pretty_breaks(n = 6)) +
  guides(color = "none") +
  facet_wrap( ~ Quantity, labeller = label_wrap_gen(multi_line=FALSE))
```

## Summary tables for GLMs fit to orthogonal projection ratio {#sec:SI-GLMs}

\begin{table}
\centering
\begin{tabular}[t]{lc}
\toprule
  & (1)\\
\midrule
(Intercept) & \num{0.695}\\
 & (\num{0.071})\\
DatabaseSwehVd & \num{0.169}\\
 & (\num{0.077})\\
SD (Intercept Talker) & \num{0.196}\\
SD (Intercept Word) & \num{0.000}\\
SD (Observations) & \num{0.095}\\
\midrule
Num.Obs. & \num{300}\\
R2 Marg. & \num{0.292}\\
AIC & \num{-378.2}\\
BIC & \num{-359.7}\\
RMSE & \num{0.09}\\
\bottomrule
\end{tabular}
\end{table}

```{r subset-data-for-opr}
#Alternative downsampling approach
# set.seed(15497221)
# 
# d.SwehVd.forStudy.opr <- d.SwehVd.forStudy.long %>%
#   filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint == 50)
# 
# d.SweDia2000.forStudy.opr <- d.SweDia2000.forStudy.long %>%
#   filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint == 50)
# 
# selected_talkers <- d.SwehVd.forStudy.opr %>%
#   select(Gender, Talker) %>%
#   distinct() %>%
#   group_by(Gender) %>%
#   slice_sample(n = 4, replace = F) %>%
#   pull(Talker)
# 
# d.SwehVd.forStudy.opr %<>%
#   filter(Talker %in% selected_talkers) %>%
#   group_by(category, Talker) %>%
#   slice_sample(n = 5, replace = F)
# 
# d.SwehVd.SweDia2000.opr <- rbind(d.SwehVd.forStudy.opr, d.SweDia2000.forStudy.opr)
# 
# # Instead sample 36 tokens per category from the entire database of all talkers
# d.SwehVd.SweDia2000.opr_by_category <- d.SwehVd.forStudy.long %>%
#   filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint == 50) %>%
#   group_by(category, Gender) %>%
#   slice_sample(n = 18, replace = F) %>%
#   rbind(d.SweDia2000.forStudy.opr)
# 

# d.SwehVd.SweDia2000.opr.noBootstrap <- rbind(d.SwehVd.forStudy.long %>%
#                                                filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint == 50), 
#                                              d.SweDia2000.forStudy.long %>%
#                                                filter(Normalization.Type == "Nearey's uniform scaling (log)", timepoint == 50))
# 
# centroids.noBootstrap <- d.SwehVd.SweDia2000.opr.noBootstrap %>%
#   # choose the anchor vowels
#   filter(category %in% c("[ɪ]", "[ʏ]", "[yː]", "[ɵ]", "[eː]", "[ɛː]", "[æː]", "[œː]", "[œ]", "[ɑː]", "[a]", "[oː]", "[øː]", "[ø]", "[ʉː]", "[ɔ]")) %>%
#   # for each speaker and vowel
#   group_by(Talker, category) %>%
#   # compute mean F1 and F2
#   summarise(cenF1 = mean(F1),
#             cenF2 = mean(F2)) %>%
#   # rearrange in columns
#   pivot_wider(names_from = category,
#               values_from = c(cenF1, cenF2)) %>%
#   ungroup()
# 
# d.SwehVd.SweDia2000.opr.noBootstrap %>%
#   # add the centroids %>%
#   left_join(centroids.noBootstrap, by = "Talker")
# 
# d.orth.subset.noBootstrap <- d.SwehVd.SweDia2000.opr.noBootstrap %>%
#   group_by(category, Database) %>%
#  # for each row i.e. for each observation
#   rowwise() %>%
#   # calculate orth.proj.ratio
#   mutate(orth_proj = case_when(category == "[ʏ]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɑː]`, `cenF2_[ɑː]`), c(`cenF1_[ɪ]`, `cenF2_[ɪ]`)),
#                                category == "[ʉː]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɪ]`, `cenF2_[ɪ]`), c(`cenF1_[oː]`, `cenF2_[oː]`)),
#                                category == "[ɛː]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[eː]`, `cenF2_[eː]`), c(`cenF1_[æː]`, `cenF2_[æː]`)),
#                                category == "[øː]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[eː]`, `cenF2_[eː]`), c(`cenF1_[ɑː]`, `cenF2_[ɑː]`)),
#                                category == "[ø]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[yː]`, `cenF2_[yː]`), c(`cenF1_[a]`, `cenF2_[a]`)),
#                                category == "[œ]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɵ]`, `cenF2_[ɵ]`), c(`cenF1_[œː]`, `cenF2_[œː]`)),
#                                category == "[ɔ]" ~ orthogonal_projection(c(F1, F2), c(`cenF1_[ɑː]`, `cenF2_[ɑː]`), c(`cenF1_[oː]`, `cenF2_[oː]`)),
#                                TRUE ~ NA)) %>%
#   ungroup() %>%
#   filter(category %in% c("[ʏ]", "[ɛː]", "[œ]", "[ʉː]", "[øː]", "[ø]", "[ɔ]"))
# 
# p.orth %+%
#   (d.orth.subset.noBootstrap %>%
#      group_by(Database, category) %>%
#      mutate(
#        # category = factor(category, levels = c("[ʏ]", "[ɛː]", "[œ]", "[ɔ]", "[ʊ]", "[ɵ]"), labels =  c("[ʏ] [ɑː]-[ɪ]", "[ɛː] [eː]-[æː]", "[œ] [ɵ]-[œː]", "[ɔ] [ɑː]-[oː]", "[ʊ] [œː]-[oː]", "[ɵ] [ɛ]-[uː]")),
#        #median = median(orth_proj),
#        ci = list(enframe(Hmisc::smean.cl.boot(orth_proj)))) %>%
#      unnest(ci) %>%
#      spread(name, value) %>%
#      group_by(Database, category) %>%
#      summarise(
#        mean_CI_lower = mean(Lower),
#        mean_CI_upper = mean(Upper),
#        mean = mean(Mean)))
```
```{r}
# d.orth.subset.noBootstrap %>%
#   ggplot +
#   aes(y = orth_proj, x = Database) +
#   geom_boxplot() +
#   facet_wrap(~category, scales = "free")
```

```{r}
# d.Kuronen2000 <- read_delim('../../data/phonetic vowel statistics/Swedish/Published databases/Kuronen2000/kuronen_L1_male_talker_formants.csv') %>%
#   mutate(
#     category = replace_na(category, "9:"),
#     Gender = "male",
#     Age = 18,
#     category = factor(
#       plyr::mapvalues(
#         category,
#         levels.vowel.SweFA,
#         levels.vowel.IPA.swe),
#       levels = levels.vowel.IPA.swe),
#     Quantity = ifelse(category %in% levels.vowel.IPA.swe.long, "long", "short")) %>%
#   group_by(Talker, category) %>%
#   mutate(
#     Token = row_number()) %>%
#   rename(Word = word) %>%
#   arrange(Quantity) %>%
#   ungroup()

# Add code to write csv.file for harrington
# write.csv(
#   d.Swedish,
#   file = "../../../data/phonetic vowel statistics/Swedish/Published databases/Swedish_vowels_Persson2023_Kuronen2000.csv",
#   row.names = FALSE, quote = FALSE)
```
```{r}
# d.Swedish <- d.Kuronen2000 %>%
#   rbind(
#     d.SwehVd %>%
#       # get only mid-point for comparability to kuronen
#       filter(timepoint == 50, Transcribed_vowel == "targeted") %>%
#       select(c(Talker, Age, Gender, category, Quantity, Word, Token, F1, F2)) %>%
#       mutate(
#         Database = "Persson2021"),
#     d.SweDia2000 %>%
#       # get only mid-point for comparability to kuronen
#       filter(timepoint == 50) %>%
#       select(c(Database, Talker, Age, Gender, category, Quantity, Word, Token, F1, F2))) %>%
#     select(c(Database, Talker, Age, Gender, category, Quantity, Word, Token, F1, F2))
# 
# p.db.comparison <- 
#   d.Swedish %>%
#   filter(Gender == "male", Database != "SweDia") %>%
#   ggplot(
#     aes(
#       x = F2,
#       y = F1, 
#       color = category,
#   #shape = Gender,
#   linetype = Database)) +
#   stat_ellipse() +
#   # Add between-talker mean
#   geom_label(
#     data =
#       ~ .x %>%
#       group_by(category, Quantity, Database) %>%
#       summarise(across(c(F1, F2), ~ mean(.x))),
#     mapping = aes(
#       colour = category,
#       label = category),
#       #fontface = ifelse(Database == "SweDia", 2, 1)),
#     alpha = .8,
#     size = 4,
#     label.size = NA,
#     label.padding = unit(.06, "cm")) +
#   #geom_point(alpha = .4) +
#   scale_color_manual(name = "category", values = colors.vowel.swe) +
#   scale_x_reverse() +
#   scale_y_reverse() +
#   guides(color = "none") +
#   facet_grid(Database ~ Quantity)
# p.db.comparison
```

```{r}
# p.vowel.movements %+%
#   (d.Swedish %>%
#      filter(Gender == "male", Database != "SweDia") %>%
#      group_by(Database, Talker, category, Quantity) %>%
#      summarise(across(c(F2, F1), ~mean(.x))) %>%
#      group_by(Database, category, Quantity) %>%
#      summarise(across(c(F2, F1), ~mean(.x)))) +
#   geom_label(
#     data =
#       ~ .x %>%
#       filter(Database == "Kuronen2000") %>%
#       group_by(category, Quantity, Database) %>%
#       summarise(across(c(F1, F2), ~ mean(.x))),
#     mapping = aes(
#       colour = category,
#       label = category),
#     fontface = 2,
#     alpha = .4,
#     size = 4,
#     label.size = NA,
#     label.padding = unit(.06, "cm"))
```
